### 2021-02-09 final_report/include/CAMT.tex
moved relevant stuff to my own public repo

\chapter{Concept Alignment in Machine Translation} \label{ch5}
We conclude this work by documenting a first attempt to put our CA component to use in a MT system. 
The chapter starts with a discussion of how GF translation can benefit from using concepts extracted with our method as translation units.
After that, we give a step-by-step description of how this is achieved, also indicating the changes made to the CA module to adjust it for this specific use case. 
Finally, Section \ref{eval5} focuses on the actual evaluation of the resulting system and presents its results.

\section{Method and implementation}
As discussed in Section \ref{gf}, GF translation consists in parsing SL strings to obtain the corresponding ASTs to then linearize them according to the concrete syntax of the TL.
As well as describing the morphological and syntactical features of the languages involved, a multilingual grammar suited for this task defines a \textit{translation lexicon}, i.e. a set of word senses and the corresponding expressions in the various languages. \smallskip

Automating CA can be seen, then, as a first step for automatically generating such lexicon. 
An important point is that our approach, as opposed to word alignment techniques, identifies concepts at different levels of abstraction.
This allows us to construct lexical entries not only for the grammatical categories that typically correspond to single words (such as nouns, adjectives and adverbs), but also for phrases (with the term ``phrase'' now intended grammatically, as opposed to its meaning in SMT).
Doing so enables GF translation, designed to preserve grammatically, to also render idiomatic expressions found in the data used for generating the lexicon correctly. \smallskip

Obtaining a multilingual GF grammar from a set of aligned UD trees, however, is not immediate. In the following three sections, we illustrate the different steps of this process. We then introduce the simple GF-based MT module used for the experiments discussed in Section \ref{eval5} and describe the adjustments made to the CA component in order to optimize it for the task at hand.

\subsection{UD-to-GF conversion}
The output of both CE and CP is, unless when linearized for evaluation purposes, a set of aligned UD trees in CoNNL-U format. In order to be used for GF-based translation, such trees need to be first converted into GF ASTs, which are then used to generate a set of grammar rules constiuting the above mentioned multilingual GF lexicon. \smallskip

As discussed in Section \ref{udrelgf}, the UD-to-GF conversion, while being more complex than its reverse, can be performed by \texttt{gf-ud}, a program offering several functionalities useful to work with GF and UD simultaneously.
Such program, however, cannot be fed the alignments obtained via CE and CP directly. This is due to the fact that the members of most of the relevant alignments are subtrees (or heads of subtrees) and not complete UD trees. 
While this makes no difference in terms of the Haskell data type we represent them with (cf. Figure \ref{types}), trying to print them to files in the corresponding CoNNL-U notation with no postprocessing leads to malformed UD sentences. \smallskip

The problem, which is due to the lack of a root labelled as such and illegal \texttt{ID} and \texttt{HEAD} values, was solved by implementing a simple normalization procedure, now part of the CA library. 
Called every time a UD subtree has to be converted to CoNNL-U, it first replaces the UD label (\texttt{udDEPREL}) of its root node with \texttt{root} and assigns the special value 0 to its \texttt{udHEAD} field. After that, it proceeds to ``rescale'' the \texttt{udID} and \texttt{udHEAD} fields of all other nodes so that they end up into the range $[1,n]$, where $n$ is the number of nodes of the extracted subtrees\footnote{see Section \ref{conll} and Figure \ref{types} for definitions of all the fields of UD trees and their CoNNL-U representations.}. \smallskip

\begin{figure}[H]
\small
\begin{Verbatim}[commandchars=\\\{\}]
\textcolor{gray}{   
    1  sadly  sadly  ADV  _  _  5  advmod  _  _}
    2  this  this  DET  _  _  4  det  _  _
    3  malformed  malformed  ADJ  _  _  4  amod  _  _
    4  subtree  subtree  NOUN  _  5  nsubj  _  _
    \textcolor{gray}{5  needs  need  VERB  _  _  0  root  _  _}
    \textcolor{gray}{6  postprocessing  postprocess  VERB  _  _  5  xcomp}

\end{Verbatim}
\centering
\large
$\Downarrow$
\small
\begin{verbatim}

    1	this	this	DET	_	_	3	det	_	_
    2	malformed	malformed	ADJ	_	_	3	amod	_	_
    3	subtree	subtree	NOUN	_	_	0	root	_	_
\end{verbatim}
\caption[A UD subtree and its normalization]{A vanilla noun phrase subtree (highlighted in black) in the context it was extracted from (gray) and the same subtree after normalization.}
\label{normalization}
\end{figure}

Another practical problem that arises when trying to transform a CoNNL-U tree into a GF AST is that \texttt{gf-ud} easily becomes remarkably resource-intensive when dealing with ambiguous, large trees. 
This is due to the fact that conversion in this direction is a nondeterministic search problem \cite{udgf}. 
For this reason, it can be useful to limit the size\footnote{in this context, the size of an alignment is defined as the number of nodes of its largest member.} of the alignments that are actually kept after the extraction (or propagation) procedure. Since our CA component works at all levels of abstraction, thus also aligning full clauses, this turned out to be essential and became one of the command-line parameters of the final executables, as mentioned in Appendix \ref{b}. \smallskip

\subsection{Grammar generation} \label{grgn}
The result of this UD-to-GF conversion is still a series of tree alignments, but their resulting format makes a substantial difference, since GF ASTs can be used to obtain the rules of a generative grammar. \texttt{gf-ud} also provides a way to do this, given an \textit{extraction grammar} and a \textit{morphological dictionaries} of the languages of the alignments. \smallskip

What we refer to as an \textit{extraction grammar} is of course a GF grammar, whose objective is to define the set of basic categories and syntactic rules the entries of the automatically generated GF lexicon is going to be based on. \smallskip

A \textit{morphological dictionary}, on the other hand, contains correspondences between a large number of lemmas and various word forms. In this case, the morphological dictionaries themselves are implemented in GF. \smallskip

The program responsible for generating the GF lexicon given these components has been written in parallel with - but not as part of - this thesis project and is still under development. As a consequence, some programming work was required to automate the few parts of the process that still had to be performed manually and some (minimal) postprocessing is sometimes required before the automatically generated grammars can be compiled. This happens for instance when Italian words are incorrectly POS-tagged as verbs and, not having a word ending typical of such category, produce pattern matching errors when the corresponding strings are passed to the constructor generating all the verb forms. \smallskip

\subsubsection{Extending the extracted grammar} \label{extend}
The extraction grammar is designed for the grammar generation module to produce grammar rules corresponding to small concepts: basic categories, noun phrases, verb phrases etc. 
As a consequence, if used directly, grammars resulting from the above described generation step can only deal with very simple predicative sentences, with verbs in the present tense of the indicative mood, and their constituents, such as\smallskip

\begin{example}
    ``this sentence is simple''
\end{example}

Of course, some limited variation is still possible. For instance, the determiner can change or disappear, number can vary, the complement of the copula can become a noun and an adjectival modifiers and prepositional phrases can be added to each noun phrase. For instance:

\begin{example} \ 
    \begin{itemize}
        \item ``the sentence is simple''
        \item ``a sentence is simple''
        \item ``sentences are simple''
        \item ``these sentences are simple''
        \item ``this sentence is an example''
        \item ``this short sentence is simple''
        \item ``this sentence of the text is simple''
    \end{itemize}
\end{example}

For languages that are covered by the RGL, however, it is also extremely easy to extend such grammars with preexisting syntax rules so to make it possible to generate more complex clauses and sentences and add useful function words. The grammars used for the experiments that will be described in Section \ref{eval5}, for example, have been extended also to cover negative and interrogative forms, the past and future tense of verbs in the indicative mood and comparative adjectives. This allows for plenty of variations, such as:

\begin{example} \
    \begin{itemize}
        \item ``this sentence isn't simple''
        \item ``is this sentence simple?''
        \item ``this sentence was simple''
        \item ``this sentence will be simple''
        \item ``this sentence is simpler than that sentence''
    \end{itemize}
\end{example}

Of course, variations can be combined, so that the grammar can produce sentences like the following:

\begin{example}
    ``Won't these short sentences be simpler than that long sentence?''
\end{example}

This is accomplished by making the automatically generated grammars extend not only the extraction grammar itself, but also a module that simply imports the necessary standard RGL categories and functions.

\subsection{Translation} \label{mt}
Maybe counterintuitively, the MT module itself is by far the simplest component of the entire system. Built using PGF, the API meant to be used for embedding GF grammars in Haskell programs, it simply makes use of GF parsing and linearization.
The only difference between translating inside the GF shell and using this ad-hoc program, apart from the increased ease of use of the latter, are that: \smallskip

\begin{itemize}
    \item while GF can parse the SL string in a variety of ways and, as a consequence, return multiple TL linearizations, our program only outputs the first candidate translation, preliminarily discarding trees containing terminal nodes that have no linearization in the TL\footnote{a linearization can be missing, for instance, if a concrete syntax is postprocessed by simply getting rid of an ``illegal'' verb causing compilation problems (cf. Section \ref{grgn})} 
    \item to facilitate the experiments, our standalone program can translate several newline-separated sentences at once.
\end{itemize}

\subsection{Adjustments to the CA component} \label{tuning}

While the domain of application of CA we focus on in this work is automatic translation, the system described so far is not in any way optimized for it and can also be made use of in other contexts. 
For instance, it is easy to imagine using it to improve the user interface of online translation memories or to facilitate reading parallel texts in the context of language learning. \smallskip

In our particular setting, however, some practical problems arise and some adjustments can be done in order to make the CA software more well suited to the generation of a translation lexicon. \smallskip

In the following section, we discuss two ways to mitigate the impact of the annotation errors that inevitably derive from applying automatic dependency parsing on raw, possibly noisy data. Both techniques aim to increase the number of extracted alignments without significantly affecting precision.
Section \ref{select}, going in a sense in the opposite direction, describes the strategies we apply to filter out irrelevant alignments.

\subsubsection{Dealing with parse errors}
As discussed in Sections \ref{eval2} and \ref{eval3}, both our CE and CP modules give promising results on manually annotated data. However, when run on UDPipe-generated dependency trees, performance, and recall in particular, is significantly affected by errors happening at the parsing stage. Because parsing itself is a process consisting of several stages, errors differ based on when they happen. In particular, UDPipe goes through the following steps, each potentially causing one ore more particular kinds of errors: \smallskip

\begin{enumerate} 
    \item \textit{\textbf{tokenization}}, which can, although it is unlikely for the languages considered in this work, lead to errors in identifying word boundaries
    \item \textit{\textbf{morphological analysis}}, which can cause lemmatization and, consequently, POS-tagging errors
    \item actual \textit{\textbf{dependency parsing}}, the potential cause of attachment and labelling errors
\end{enumerate} \smallskip

In this context, we focus on the syntax-level errors: attachment and labelling. Of course, lemmatization and POS-tagging errors can be the first cause of one such error. In particular, the fact that a word is lemmatized incorrectly can make it so that it is also attributed the wrong POS tag and, as a consequence, assigned the wrong dependency label and/or attached to the wrong node. \smallskip

\begin{example}
    Consider the Italian sentence ``quel fa è diesis'' (``that F is sharp'').
    Notoriously, in Italian, the word ``fa'' can stand for the musical note F. Since ``fa'' is also the third person singular of the present indicative of the verb ``fare'' (``to do'', ``to make''), it is likely to be lemmatized so (instead of as ``fa''). If so, it will also probably POS-tagged as a \texttt{VERB} and assigned a dependency label typical of verbs (in this case, with no other lexical verbs in the sentence, most likely \texttt{root}, thus producing an attachment error as well), while the correct annotation is

    \begin{figure}[H]
        \begin{subfigure}{.5\textwidth}
            \centering
            \footnotesize
            \begin{verbatim}
    1	quel	quello	DET	_	_	2	det	_	_
    2	fa	fa	NOUN	_	_	4	nsubj	_	_
    3	è	essere	AUX	_	_	4	cop	_	_
    4	diesis	diesis	ADJ	_	_	0	root	_	_
            \end{verbatim}
        \end{subfigure}
        \begin{subfigure}{.5\textwidth}
            \centering
            \footnotesize
            \begin{picture}(223.0,90.0)
                \put(0.0,0.0){quel}
                \put(46.0,0.0){fa}
                \put(92.0,0.0){è}
                \put(129.0,0.0){diesis}
                \put(0.0,15.0){{\tiny DET}}
                \put(46.0,15.0){{\tiny NOUN}}
                \put(92.0,15.0){{\tiny AUX}}
                \put(129.0,15.0){{\tiny ADJ}}
                \put(33.0,30.0){\oval(39.47826086956522,33.333333333333336)[t]}
                \put(13.26086956521739,35.0){\vector(0,-1){5.0}}
                \put(18.0,49.66666666666667){{\tiny det}}
                \put(97.5,30.0){\oval(79.3855421686747,66.66666666666667)[t]}
                \put(57.80722891566265,35.0){\vector(0,-1){5.0}}
                \put(82.5,66.33333333333334){{\tiny nsubj}}
                \put(120.5,30.0){\oval(28.89189189189189,33.333333333333336)[t]}
                \put(106.05405405405405,35.0){\vector(0,-1){5.0}}
                \put(105.5,49.66666666666667){{\tiny cop}}
                \put(144.0,90.0){\vector(0,-1){60.0}}
                \put(149.0,80.0){{\tiny root}}
              \end{picture}
        \end{subfigure}
    \end{figure}
\end{example}
\smallskip

Nontheless, because we do not want to intervene on the parser directly (nor to modify its results), we ignore the causes and focus on the syntax-level errors directly. \smallskip

\paragraph{Working at the clause level} \label{clauses}
At the highest level of abstraction, the parser often fails to correctly identify the clauses a sentence is composed of. This happens in a variety of ways: at times, clauses are simply not identified as such or, on the contrary, non-clauses are labelled as clauses. 
Most often, however, clauses are simply not attached to the right node and/or given a label that, while still identifying them as clauses, does not give correct information about their type. \smallskip

To address this problem, the CE module has been modified so that it can try to align individual clauses instead of sentences. The idea is to obtain, from each pair of sentence trees to align, a pair of lists of subtrees whose heads are labelled \texttt{root}, \texttt{csubj}, \texttt{ccomp}, \texttt{xcomp}, \texttt{advcl} or  \texttt{acl}\footnote{all these clausal labels are defined in Appendix \ref{a_lab}}.

\begin{figure}[h]
\footnotesize
\setlength{\unitlength}{0.222mm}
\begin{picture}(817.0,150.0)
  \put(0.0,0.0){this}
  \put(46.0,0.0){sentence}
  \put(128.0,0.0){\textbf{shows}}
  \put(183.0,0.0){how}
  \put(220.0,0.0){to}
  \put(266.0,0.0){\textbf{work}}
  \put(312.0,0.0){at}
  \put(349.0,0.0){the}
  \put(386.0,0.0){clause}
  \put(450.0,0.0){level}
  \put(505.0,0.0){to}
  \put(551.0,0.0){\textbf{increase}}
  \put(633.0,0.0){recall}
  \put(0.0,15.0){{\tiny DET}}
  \put(46.0,15.0){{\tiny NOUN}}
  \put(128.0,15.0){{\tiny \textbf{VERB}}}
  \put(183.0,15.0){{\tiny ADV}}
  \put(220.0,15.0){{\tiny PART}}
  \put(266.0,15.0){{\tiny \textbf{VERB}}}
  \put(312.0,15.0){{\tiny ADP}}
  \put(349.0,15.0){{\tiny DET}}
  \put(386.0,15.0){{\tiny NOUN}}
  \put(450.0,15.0){{\tiny NOUN}}
  \put(505.0,15.0){{\tiny PART}}
  \put(551.0,15.0){{\tiny \textbf{VERB}}}
  \put(633.0,15.0){{\tiny NOUN}}
  \put(33.0,30.0){\oval(39.47826086956522,33.333333333333336)[t]}
  \put(13.26086956521739,35.0){\vector(0,-1){5.0}}
  \put(18.0,49.66666666666667){{\tiny det}}
  \put(97.0,30.0){\oval(78.34146341463415,33.333333333333336)[t]}
  \put(57.829268292682926,35.0){\vector(0,-1){5.0}}
  \put(82.0,49.66666666666667){{\tiny nsubj}}
  \put(234.5,30.0){\oval(79.3855421686747,66.66666666666667)[t]}
  \put(194.80722891566265,35.0){\vector(0,-1){5.0}}
  \put(219.5,66.33333333333334){{\tiny advmod}}
  \put(253.0,30.0){\oval(39.47826086956522,33.333333333333336)[t]}
  \put(233.26086956521738,35.0){\vector(0,-1){5.0}}
  \put(238.0,49.66666666666667){{\tiny mark}}
  \put(217.0,30.0){\oval(135.82608695652175,100.0)[t]}
  \put(284.9130434782609,35.0){\vector(0,-1){5.0}}
  \put(202.0,83.0){{\tiny \textbf{xcomp}}}
  \put(391.0,30.0){\oval(135.82608695652175,100.0)[t]}
  \put(323.0869565217391,35.0){\vector(0,-1){5.0}}
  \put(376.0,83.0){{\tiny case}}
  \put(409.5,30.0){\oval(98.02970297029702,66.66666666666667)[t]}
  \put(360.4851485148515,35.0){\vector(0,-1){5.0}}
  \put(394.5,66.33333333333334){{\tiny det}}
  \put(428.0,30.0){\oval(59.3125,33.333333333333336)[t]}
  \put(398.34375,35.0){\vector(0,-1){5.0}}
  \put(413.0,49.66666666666667){{\tiny nmod}}
  \put(378.0,30.0){\oval(182.3695652173913,133.33333333333334)[t]}
  \put(469.1847826086956,35.0){\vector(0,-1){5.0}}
  \put(363.0,99.66666666666667){{\tiny obl}}
  \put(538.0,30.0){\oval(39.47826086956522,33.333333333333336)[t]}
  \put(518.2608695652174,35.0){\vector(0,-1){5.0}}
  \put(523.0,49.66666666666667){{\tiny mark}}
  \put(428.5,30.0){\oval(283.94736842105266,166.66666666666666)[t]}
  \put(570.4736842105264,35.0){\vector(0,-1){5.0}}
  \put(413.5,116.33333333333333){{\tiny \textbf{advcl}}}
  \put(612.0,30.0){\oval(78.34146341463415,33.333333333333336)[t]}
  \put(651.170731707317,35.0){\vector(0,-1){5.0}}
  \put(597.0,49.66666666666667){{\tiny obj}}
  \put(143.0,150.0){\vector(0,-1){120.0}}
  \put(148.0,140.0){{\tiny \textbf{root}}}
\end{picture}
\begin{itemize}
    \item ``\textit{this sentence \textbf{shows} how to work at the clause level to increase recall}''
    \item ``\textit{how to \textbf{work} at the clause level to increase recall}''
    \item ``\textit{to \textbf{increase} recall}''
\end{itemize}
\caption[A sentence and its decomposition into clauses]{A sentence and its decomposition into clauses. Note how the sentence is not actually segmented, but rather used to obtain a set of subsentences, each having their main verb, marked in bold, as root.}
\end{figure}

These lists of ``clauses''\footnote{``clauses'' is in quotes since what we are referring to as a clause includes its subclauses too. This means that the sentence is not segmented into clauses in the literal sense of the term, but rather that one tree per verb is extracted from them.} are then sorted by dependency label and distance from the root, so that in the absence of annotation errors, when there is in fact an evident correspondence, they appear in the same position. Finally, the program tries to align every possible pair of clauses and the pruning procedure described in \ref{pruning} is applied to only keep the ones that are more likely to be correct.

\paragraph{Dealing with unaligned subtrees} \label{exclusion}
Clearly, however, parse errors do not happen exclusively at the clause level: smaller units, e.g. verb arguments, can also be misclassified or attached to the wrong node. To partially work around the problem, the CE module supports what one could refer to as ``alignment by exclusion'': after the extraction procedure described in Section \ref{improvements} - or its modified version outlined in the above paragraph - is carried out, the program can try to find correspondences between the remaining unaligned subtrees. 
The idea is the same as above: subtrees are sorted, aligned and lastly alignments are pruned.
This is done ignoring, if necessary, the context in which the subtrees occur and potentially making only use of a subset of the alignments criteria, for instance the strictest and safest ones. \smallskip

This is useful in cases such as that illustrated in Figure \ref{rest}, as well as in the event of actual attachment ambiguity or when the two versions of the text present syntactical differences that cannot be handled with the simple divergence patterns introduced in Section \ref{divs}.

\begin{figure}[h]
    \centering
    \footnotesize
    \begin{subfigure}{.5\textwidth}
      \centering
      \setlength{\unitlength}{0.2mm}
      \begin{picture}(315.0,110.0)
        \put(0.0,0.0){parse}
        \put(55.0,0.0){this}
        \put(101.0,0.0){sentence}
        \put(183.0,0.0){with}
        \put(229.0,0.0){care}
        \put(0.0,15.0){{\tiny VERB}}
        \put(55.0,15.0){{\tiny DET}}
        \put(101.0,15.0){{\tiny NOUN}}
        \put(183.0,15.0){{\tiny ADP}}
        \put(229.0,15.0){{\tiny NOUN}}
        \put(88.0,30.0){\oval(39.47826086956522,33.333333333333336)[t]}
        \put(68.26086956521739,35.0){\vector(0,-1){5.0}}
        \put(73.0,49.66666666666667){{\tiny det}}
        \put(70.5,30.0){\oval(98.02970297029702,66.66666666666667)[t]}
        \put(119.51485148514851,35.0){\vector(0,-1){5.0}}
        \put(55.5,66.33333333333334){{\tiny obj}}
        \put(216.0,30.0){\oval(39.47826086956522,33.333333333333336)[t]}
        \put(196.26086956521738,35.0){\vector(0,-1){5.0}}
        \put(201.0,49.66666666666667){{\tiny case}}
        \put(134.5,30.0){\oval(227.68995633187774,100.0)[t]}
        \put(248.34497816593887,35.0){\vector(0,-1){5.0}}
        \put(119.5,83.0){{\tiny obl}}
        \put(15.0,110.0){\vector(0,-1){80.0}}
        \put(20.0,100.0){{\tiny root}}
      \end{picture}
    \end{subfigure}%
    \begin{subfigure}{.5\textwidth}
        \centering
        \setlength{\unitlength}{0.25mm}
\begin{picture}(324.0,90.0)
  \put(0.0,0.0){analizza}
  \put(82.0,0.0){questa}
  \put(146.0,0.0){frase}
  \put(201.0,0.0){con}
  \put(238.0,0.0){cura}
  \put(0.0,15.0){{\tiny VERB}}
  \put(82.0,15.0){{\tiny DET}}
  \put(146.0,15.0){{\tiny NOUN}}
  \put(201.0,15.0){{\tiny ADP}}
  \put(238.0,15.0){{\tiny NOUN}}
  \put(124.0,30.0){\oval(59.3125,33.333333333333336)[t]}
  \put(94.34375,35.0){\vector(0,-1){5.0}}
  \put(109.0,49.66666666666667){{\tiny det}}
  \put(93.0,30.0){\oval(143.94520547945206,66.66666666666667)[t]}
  \put(164.97260273972603,35.0){\vector(0,-1){5.0}}
  \put(78.0,66.33333333333334){{\tiny obj}}
  \put(229.5,30.0){\oval(28.89189189189189,33.333333333333336)[t]}
  \put(215.05405405405406,35.0){\vector(0,-1){5.0}}
  \put(214.5,49.66666666666667){{\tiny case}}
  \put(212.0,30.0){\oval(88.73913043478261,66.66666666666667)[t]}
  \put(256.3695652173913,35.0){\vector(0,-1){5.0}}
  \put(197.0,66.33333333333334){{\tiny nmod}}
  \put(15.0,90.0){\vector(0,-1){60.0}}
  \put(20.0,80.0){{\tiny root}}
\end{picture}
    \end{subfigure}
    \caption[A case where alignment ``by exclusion'' is beneficial]{A case where alignment ``by exclusion'' is beneficial. In this case, while the two sentences should appear to be structured identically to a human reader, the Italian parse tree contains a (reasonable) attachment error. As a consequence, vanilla CE would not be able to align the prepositional phase ``\textit{with care}'', attached to the main verb ``\textit{parse}'', to its Italian exact counterpart ``\textit{con cura}'', incorrectly linked to the noun ``\textit{frase}'' as an \texttt{nmod}. A second CE pass trying to put unaligned subtrees in relation with each other, however, is able to detect the POS-equivalence between the two phrases, thus producing three new correct alignments: $\langle$\textit{with care, con cura}$\rangle$, $\langle$\textit{care, cura}$\rangle$ and $\langle$\textit{with, con}$\rangle$.}
    \label{rest}
\end{figure}

The results of this kind of optimization, however, as well as those of working at the clause level, vary widely depending on the quality of the parse trees and can be detrimental, in terms of precision, if applied to high-quality datasets.

\subsubsection{Selecting relevant alignments} \label{select}
The pruning procedure described in Section \ref{pruning} is aimed at filtering out alignments that, because of their number of occurrences and of the set of criteria thanks to which they have been extracted, are less likely to be correct. However, not all the alignments extracted in this way are relevant for the purposes of MT. \smallskip

On the one hand, most alignments that do not contain any content word are not useful for the generation of a domain-specific grammar, as many common function words are already covered by the GF RGL\footnote{small alignments containing both function and content words, such as prepositional phrases, are on the other hand extremely useful, as they show which specific function word should be used in a particular context.}. \smallskip

On the other hand, the concepts represented in the GF grammar should intuitively, in most cases, be ``as small as possible''. If we reconsider the example in Figure \ref{cake} (sentences \textit{``finding useful correspondences is not exactly a piece of cake''} and \textit{``trovare corrispondenze utili non è proprio scontato''}), for instance, while the multi-word correspondence $\langle$\textit{a piece of cake, scontato}$\rangle$ is indeed useful there is usually no point in keeping correspondences such as $\langle$\textit{useful correspondences, corrispondenze utili}$\rangle$, even if they are correct, when it it also possible to extract the alignments $\langle$\textit{useful, utili}$\rangle$ and $\langle$\textit{correspondences, corrispondenze}$\rangle$: one-word and, in general, shorter alignments are more easily reusable in different contexts, especially since there are language-specific RGL rules capable of handling word order properly. 
As a consequence, for the purposes of grammar generation, the CA modules should in this case only return the following set of alignments: \smallskip

\begin{enumerate}
    \item $\langle$\textit{finding, trovare}$\rangle$
    \item $\langle$\textit{useful, utili}$\rangle$
    \item $\langle$\textit{correspondences, corrispondenze}$\rangle$
    \item $\langle$\textit{is, è}$\rangle$
    \item $\langle$\textit{not, non}$\rangle$
    \item $\langle$\textit{exactly, proprio}$\rangle$
    \item $\langle$\textit{a piece of cake, scontato}$\rangle$.
\end{enumerate} \smallskip

If we consider that the RGL can already handle most function words correctly, we can even discard all alignments that do not contain any content word (in this case, alignments 4 and 5, which leaves us with only 5 alignment useful to GF). \smallskip

In some situations, however, larger alignments (such as $\langle$\textit{useful correspondences, corrispondenze utili}$\rangle$) are considered to be relevant since they help understand which words can be safely used in conjunction with each other. For instance, while the Italian adjective ``\textit{brutto}'' (most often translated as ``\textit{ugly}'') is often used in conjunction with the word ``\textit{notizia}'' (``\textit{news}'') in expressions such as ``\textit{brutte notizie}'', in English the adjective ``\textit{bad}'' (as in ``\textit{bad news}'') is definitely more appropriate. Discarding a potential $\langle\textit{bad news}, \textit{brutte notizie}\rangle$ alignment is then not always the right choice, since in a context in which there are many alternative translation equivalents for the word ``\textit{brutto}'' it provides useful information about what adjective to use to refer to the noun ``\textit{news}''. \smallskip

Because of this, and since CE could have applications other than MT, selection in this sense is performed - optionally - after the extraction phase itself and implemented as an independent function. It works as follows: first, alignments that do not present any content word are filtered out. Then, among the remaining ones, \textit{superficially} perfect alignments \textit{containing} other extracted alignments are also discarded.\smallskip

To clarify such algorithm, it is worth defining perfect alignment - which was already mentioned in Section \ref{baseline} - more rigorously. \smallskip

\begin{definition}
    An alignment $A = \langle e_1,...,e_n \rangle$ is \textit{perfect} if its member dependency trees $e_1,...,e_n$ are identical both in their topology and in the dependency labels assigned to their nodes. 
\end{definition} \smallskip

The ``shallow'' version of perfect alignment the algorithm described above refers to, which we call \textit{superficially perfect alignment}, occurs when the definition applies at least to the roots of the two member trees and the roots of their immediate subtrees.\smallskip

\begin{example}
    Consider the following pairs of English-Italian sentences:
    \begin{enumerate}
        \item $\langle$this algorithm selects \underline{the} relevant alignments, questo algoritmo seleziona gli allineamenti rilevanti$\rangle$
        \item $\langle$this algorithm selects relevant alignments, questo algoritmo seleziona gli allineamenti rilevanti$\rangle$
    \end{enumerate}

    While the two English sentences are translated to Italian in the same way, alignment is perfect only in the first case, where the determiner ``\textit{the}'' is present. The two trees are in fact identical if not for the words themselves and their order.

\begin{figure}[H]
    \centering
    \scriptsize
    \begin{subfigure}{.5\textwidth}
      \centering
      \setlength{\unitlength}{0.145mm}
\begin{picture}(479.0,110.0)
  \put(0.0,0.0){this}
  \put(46.0,0.0){algorithm}
  \put(137.0,0.0){selects}
  \put(210.0,0.0){the}
  \put(247.0,0.0){relevant}
  \put(329.0,0.0){alignments}
  \put(0.0,15.0){{\tiny DET}}
  \put(46.0,15.0){{\tiny NOUN}}
  \put(137.0,15.0){{\tiny VERB}}
  \put(210.0,15.0){{\tiny DET}}
  \put(247.0,15.0){{\tiny ADJ}}
  \put(329.0,15.0){{\tiny NOUN}}
  \put(33.0,30.0){\oval(39.47826086956522,33.333333333333336)[t]}
  \put(13.26086956521739,35.0){\vector(0,-1){5.0}}
  \put(18.0,49.66666666666667){{\tiny det}}
  \put(101.5,30.0){\oval(87.7032967032967,33.333333333333336)[t]}
  \put(57.64835164835165,35.0){\vector(0,-1){5.0}}
  \put(86.5,49.66666666666667){{\tiny nsubj}}
  \put(279.5,30.0){\oval(116.47899159663865,66.66666666666667)[t]}
  \put(221.26050420168067,35.0){\vector(0,-1){5.0}}
  \put(264.5,66.33333333333334){{\tiny det}}
  \put(298.0,30.0){\oval(78.34146341463415,33.333333333333336)[t]}
  \put(258.8292682926829,35.0){\vector(0,-1){5.0}}
  \put(283.0,49.66666666666667){{\tiny amod}}
  \put(253.0,30.0){\oval(190.4375,100.0)[t]}
  \put(348.21875,35.0){\vector(0,-1){5.0}}
  \put(238.0,83.0){{\tiny obj}}
  \put(152.0,110.0){\vector(0,-1){80.0}}
  \put(157.0,100.0){{\tiny root}}
\end{picture}
    \end{subfigure}%
    \begin{subfigure}{.5\textwidth}
        \centering
        \setlength{\unitlength}{0.16mm}
\begin{picture}(542.0,90.0)
  \put(0.0,0.0){questo}
  \put(64.0,0.0){algoritmo}
  \put(155.0,0.0){seleziona}
  \put(246.0,0.0){gli}
  \put(283.0,0.0){allineamenti}
  \put(401.0,0.0){rilevanti}
  \put(0.0,15.0){{\tiny DET}}
  \put(64.0,15.0){{\tiny NOUN}}
  \put(155.0,15.0){{\tiny VERB}}
  \put(246.0,15.0){{\tiny DET}}
  \put(283.0,15.0){{\tiny NOUN}}
  \put(401.0,15.0){{\tiny ADJ}}
  \put(42.0,30.0){\oval(59.3125,33.333333333333336)[t]}
  \put(12.34375,35.0){\vector(0,-1){5.0}}
  \put(27.0,49.66666666666667){{\tiny det}}
  \put(119.5,30.0){\oval(87.7032967032967,33.333333333333336)[t]}
  \put(75.64835164835165,35.0){\vector(0,-1){5.0}}
  \put(104.5,49.66666666666667){{\tiny nsubj}}
  \put(274.5,30.0){\oval(28.89189189189189,33.333333333333336)[t]}
  \put(260.05405405405406,35.0){\vector(0,-1){5.0}}
  \put(259.5,49.66666666666667){{\tiny det}}
  \put(239.0,30.0){\oval(125.65625,66.66666666666667)[t]}
  \put(301.828125,35.0){\vector(0,-1){5.0}}
  \put(224.0,66.33333333333334){{\tiny obj}}
  \put(362.0,30.0){\oval(115.45762711864407,33.333333333333336)[t]}
  \put(419.728813559322,35.0){\vector(0,-1){5.0}}
  \put(347.0,49.66666666666667){{\tiny amod}}
  \put(170.0,90.0){\vector(0,-1){60.0}}
  \put(175.0,80.0){{\tiny root}}
\end{picture}
    \end{subfigure}
\end{figure}

    In the second case, however, alignment is still ``perfect'' if we only compare the root node and its immediate dependents, thus ignoring the determiner ``\textit{gli}'', which is absents in English, and the adjectival modifier ``\textit{relevant/rilevanti}'':

    \begin{figure}[H]
        \centering
        \scriptsize
        \begin{subfigure}{.5\textwidth}
          \centering
          \setlength{\unitlength}{0.16mm}
          \begin{picture}(432.0,90.0)
            \put(0.0,0.0){this}
            \put(46.0,0.0){algorithm}
            \put(137.0,0.0){selects}
            \put(210.0,0.0){\textcolor{gray}{relevant}}
            \put(292.0,0.0){alignments}
            \put(0.0,15.0){{\tiny DET}}
            \put(46.0,15.0){{\tiny NOUN}}
            \put(137.0,15.0){{\tiny VERB}}
            \put(210.0,15.0){{\tiny \textcolor{gray}{ADJ}}}
            \put(292.0,15.0){{\tiny NOUN}}
            \put(33.0,30.0){\oval(39.47826086956522,33.333333333333336)[t]}
            \put(13.26086956521739,35.0){\vector(0,-1){5.0}}
            \put(18.0,49.66666666666667){{\tiny det}}
            \put(101.5,30.0){\oval(87.7032967032967,33.333333333333336)[t]}
            \put(57.64835164835165,35.0){\vector(0,-1){5.0}}
            \put(86.5,49.66666666666667){{\tiny nsubj}}
            \textcolor{gray}{\put(261.0,30.0){\oval(69,33.333333333333336)[t]}}
            \textcolor{gray}{\put(221.8292682926829,35.0){\vector(0,-1){5.0}}}
            \put(246.0,49.66666666666667){{\tiny \textcolor{gray}{amod}}}
            \put(234.5,30.0){\oval(153.06451612903226,66.66666666666667)[t]}
            \put(311.0322580645161,35.0){\vector(0,-1){5.0}}
            \put(219.5,66.33333333333334){{\tiny obj}}
            \put(152.0,90.0){\vector(0,-1){60.0}}
            \put(157.0,80.0){{\tiny root}}
          \end{picture}
        \end{subfigure}%
        \begin{subfigure}{.5\textwidth}
            \centering
            \setlength{\unitlength}{0.16mm}
    \begin{picture}(542.0,90.0)
      \put(0.0,0.0){questo}
      \put(64.0,0.0){algoritmo}
      \put(155.0,0.0){seleziona}
      \put(246.0,0.0){\textcolor{gray}{gli}}
      \put(283.0,0.0){allineamenti}
      \put(401.0,0.0){\textcolor{gray}{rilevanti}}
      \put(0.0,15.0){{\tiny DET}}
      \put(64.0,15.0){{\tiny NOUN}}
      \put(155.0,15.0){{\tiny VERB}}
      \put(246.0,15.0){{\tiny \textcolor{gray}{DET}}}
      \put(283.0,15.0){{\tiny NOUN}}
      \put(401.0,15.0){{\tiny \textcolor{gray}{ADJ}}}
      \put(42.0,30.0){\oval(59.3125,33.333333333333336)[t]}
      \put(12.34375,35.0){\vector(0,-1){5.0}}
      \put(27.0,49.66666666666667){{\tiny det}}
      \put(119.5,30.0){\oval(87.7032967032967,33.333333333333336)[t]}
      \put(75.64835164835165,35.0){\vector(0,-1){5.0}}
      \put(104.5,49.66666666666667){{\tiny nsubj}}
      \textcolor{gray}{\put(274.5,30.0){\oval(16.89189189189189,33.333333333333336)[t]}}
      \textcolor{gray}{\put(260.05405405405406,35.0){\vector(0,-1){5.0}}}
      \put(259.5,49.66666666666667){{\tiny \textcolor{gray}{det}}}
      \put(239.0,30.0){\oval(125.65625,66.66666666666667)[t]}
      \put(301.828125,35.0){\vector(0,-1){5.0}}
      \put(224.0,66.33333333333334){{\tiny obj}}
      \textcolor{gray}{\put(362.0,30.0){\oval(130,33.333333333333336)[t]}}
      \textcolor{gray}{\put(419.728813559322,35.0){\vector(0,-1){5.0}}}
      \put(347.0,49.66666666666667){{\tiny \textcolor{gray}{amod}}}
      \put(170.0,90.0){\vector(0,-1){60.0}}
      \put(175.0,80.0){{\tiny root}}
    \end{picture}
        \end{subfigure}
    \end{figure}
\end{example}

Another concept introduced by this algorithm is that of alignments \textit{containing} each other. \smallskip

\begin{definition}
    An alignment $A = \langle e_1,...,e_n \rangle$ contains an alignment $B = \langle f_1,...,f_n \rangle$ if, for each of $B$'s members, $f_i$ is a subtree of $e_i$. 
\end{definition}\smallskip

Thus, the algorithm is safe - in the sense that it does not get rid of any potentially meaningful alignment - in that it gets rid of superficially perfect alignments whose subtrees are also members of smaller alignments. This means that when sentences vary widely syntax-wise, as it happens in cases where translation is not literal or in the presence of parse errors, the resulting extracted concepts can still contain a large number of nodes.

\section{Evaluation} \label{eval5}
To evaluate the effectiveness of the translation method described in the previous section, a first, small-scale experiment was designed. The main idea is to automatically translate a set of English sentences to Italian and Swedish, make native speakers of the TLs correct the output sentences, so to obtain a set of reference, both grammatically and semantically correct translations, and compare them to the original machine-generated ones. \smallskip

The remaining part of this chapter describes the experiment in detail. Section \ref{gr} deals with the automatically generated GF lexica in play, while Sections \ref{eng} and \ref{ref} describe the way the corpus of English sentences and their reference translations were obtained. After that, Section \ref{bleu} introduces the evaluation metrics used in the experiments. Finally, the results of such experiments are presented in Section \ref{res5}.

\subsection{Lexica} \label{gr}
For this first experiment, two distinct GF lexica, built starting from the alignments extracted from the DMI and CSE course plans corpora respectively, were generated as described in Section \ref{grgn}. Both were then enriched with the syntactic constructs listed in Section \ref{extend}. \smallskip

While building a trilingual grammar would have been optimal, the idea was abandoned because of the scarce amount of domain-specific terminology obtained by CE + CP, even making use of the adjustments described in Section \ref{tuning}, for all three languages, as we commented on in Section \ref{eval4r}. \smallskip

This means that only alignments obtained via CE are put to the text in this first experiment, while CP was not made use of at all. 
Using PUD treebanks instead of the two course plans corpora could have been a partial solution to the problem, but it would have resulted in a lexicon of terms and expressions belonging to the most disparate contexts.
Most importantly, using manually annotated data would also have meant not testing all the key stages of the pipeline described in Section \ref{ourapproach}, where CP, unlike UD parsing, is seen as something optional.
\smallskip

\subsection{The English corpus} \label{eng}
Even avoiding CP, the small size of these corpora, the limitations of the language models used for parsing and the issues encountered in the UD-to-GF conversion make both the lexica resulting from this process relatively small.
This, together with the issues GF parsing suffers from, discussed in Chapter \ref{ch2}, makes working on arbitrary sentences taken from documents similar to those the course plans corpora are composed of extremely hard. \smallskip

As a consequence, our choice for this first MT experiment was to generate the sentences to translate directly in the GF shell, making use of its random AST generation functionality. As discussed in Section \ref{extend}, in fact, the (extended) CSE and DMI grammars allow for arbitrary variations - both grammatical and in terms of content words - over a small set of basic sentences. \smallskip

Grammatical variation (e.g. turning indefinite forms of nouns into definite forms, altering the tense of verbs etc.) is meant to highlight how the output of GF translation typically consists of well-formed sentences. Changes of content words, on the other hand, allow to test a larger variety of extracted concepts. \smallskip

While it is indeed possible to generate both an initial subset of sentences and the corresponding variations entirely randomly, care has been put in selecting only sentences that are semantically plausible. This is meant to facilitate the task of the human translators and to avoid having them insisting in trying to find some meaning in the sentences and correcting them using excessively rare word senses. \smallskip

The results of this process are two small testing corpora, each consisting of 50 sentences in English, generated using the abstract syntax and the English concrete syntax of the DMI and the CSE grammar respectively. 
The TL concrete syntaxes of the two grammars (Italian in the former case and Swedish in the latter) were only used at a later stage, to make sure that all sentences did have a linearization in both the source and the target language. 
A linearization can in fact be missing, even though rarely, because the grammar generation stage omits linearization rules derived from pairs of trees that cannot share the same root category.

\subsection{Reference translations} \label{ref}
Reference translations were obtained by asking two native speakers of Italian and Swedish also proficient in English to compare the original English sentences to their automatically translated counterparts and to correct the latter.
The two participants were instructed to only make the minimal changes necessary to obtain, starting from the output of the translation module described in \ref{mt} a set of grammatically and semantically correct translations. \smallskip

This makes the reference sentences as similar to the automatically obtained ones to evaluate as possible, allowing a more meaningful automatic evaluation, as we will discuss in the following Section.

\subsection{Evaluation metrics} \label{bleu}
The evaluation metric used for this experiment is the BLEU score \cite{bleu}, a widely adopted, simple, language-independent and inexpensive measure that claims good correlation with human judgments, as implemented in \cite{thesisbleu}. \smallskip

Following conventions, we report the cumulative $n$-gram scores for values of $n$ from 1 to 4 (BLEU-1 to BLEU-4). However, being a significant portion of the sentences of length 4 or less, we also report BLEU-1 to BLEU-3 scores, BLEU-1 to BLEU-2 scores and scores obtained considering unigrams only. \smallskip

The way in which the reference translations were obtained is strongly connected to the choice of this specific evaluation metric.
Without going into all the details, it is in fact important to know that the BLEU score is computed by counting matching $n$-grams in the candidate translation to $n$-grams in the reference text. 
This means that, if the reference translations are obtained independently from the automatic ones, BLEU scores can easily become misleading. \smallskip

In a preliminary experiment conducted in this way, for instance, the Italian translation ``\textit{il docente discute la Didattica}'' of the English sentence ``\textit{the teacher discusses the didactics}'', was given the minimum score despite being completely correct and even idiomatic\footnote{typically, the adjective ``didattica'' is not use in its plural form ``didattiche'' when nominalized. The capital D is often used, for this particular term, in official university documents.}. 
Looking at the two reference translations used at the time, it appears clear that such a low score is due to the capitalization of the word ``Didattica'' and the arbitrary (but equally correct) lexical choices made by the human translators, who rendered the sentence as ``il professore discute i metodi didattici'' and ``l' insegnante discute la didattica''\footnote{The terms ``docente'', ``professore'' and ``insegnante'' are almost perfect synonyms, even though ``docente'' is mostly used in formal documents to refer to university lecturers, ``professore'' is more common in secondary school and colloquial speech, and ``insegnante'' is the most generic. As such, it can also refer to elementary school teachers and in some cases even sport coaches. The expression ``metodi didattici'' is a perfectly sensible replacement for ``didattica''.}. 
Coincidentally (but not unlikely for a language like Italian), this is a case in which these lexical differences affect the choice of the determiners used in the various translations (cf. ``\underline{l'}insegnante'' vs. ``\underline{il} professore'' and ``\underline{la} didattica'' vs. ``\underline{i} metodi didattici''), producing a BLEU-1 to 4 score of 0. \smallskip

\subsection{Experimental results} \label{res5}
Corpus-level BLEU scores for the automatic translations of the 50+50 sentences of the testing corpora generated as described in Section \ref{eng} are summarized in Table \ref{tbleu}.

\begin{table}[h]
    \centering
    \begin{tabular}{|l|l|l|}
    \hline
    \textbf{}            & \textbf{DMI (en-it)} & \textbf{CSE (en-sv)} \\ \hline
    \textbf{BLEU-1 to 4} & 55.4          & 61.27         \\ \hline
    \textbf{BLEU-1 to 3} & 62.75         & 67.77         \\ \hline
    \textbf{BLEU-1 to 2} & 70.6          & 74.3          \\ \hline
    \textbf{BLEU-1}      & 79.33         & 80.99         \\ \hline
    \end{tabular}
    \caption[BLEU scores for automatic translations based on the course plans grammars]{BLEU scores obtained by comparing one candidate automatic translation per sentence to a reference translation obtained by making the necessary corrections to the automatically generated ones. Both the initial English sentences and their automatic translations were generated using the two course plans grammars.}
    \label{tbleu}
    \end{table}

These synthetic figures are useful to give an idea of the general quality of the translations: overall, although with relatively low scores, English-to-Swedish translation works significantly better than English-to-Italian. 
Looking back at the results obtained for CE (cf. Section \ref{eval2}), this is not excessively surprising, since both precision and recall for English-Swedish are consistently higher than they are for English-Italian. 
When it comes to the results obtained for CE on the course plans corpora in particular (cf. Table \ref{tcereal}), however, the difference in precision between the two language pairs is negligible.
This makes the reason for the substantial difference observed in the BLEU scores obtained for the two different corpora unclear, at least at a first glance.\smallskip

Looking at sentence-level scores is, however, sometimes more interesting. 
Regardless the corpus, scores assigned to individual segments range from the minimum possible value of 0 to the perfect score of 100, which indicates a perfect correspondence between the automatic and the reference translation. \smallskip

Examples of sentences that were assigned a perfect BLEU-1 to 4 score are ``\textit{the library provides useful textbooks}'' (translated to Italian as ``\textit{la biblioteca fornisce libri utili}'') in the DMI corpus and ``\textit{this lab is more difficult than the exam}'' (whose Swedish translation is ``\textit{den här laborationen är svårare än tentamen}'') in the CSE corpus. \smallskip

On the other hand, it is easy for shorter sentences to be assigned the minimum BLEU-1 to 4 score even when they only contain a single grammatical or semantic error. 
This is the case, for instance, of the sentence ``\textit{the test is oral}'', whose last word, ``\textit{oral}'' is translated as ``\textit{dura}'' (``\textit{hard}'') instead of ``\textit{orale}'' due to an alignment error. 
Nonetheless, it is worth noticing that this is one of the many cases in which, even though the fact that only the first translation candidate is taken into account hides it, the correct alignment $\langle$``\textit{oral}'',``\textit{orale}''$\rangle$ is also found by the CE module. 
Moreover, the such valid correspondence has a higher number of occurrences and matches stronger alignment criteria than the wrong one $\langle$``\textit{oral}'',``\textit{duro}''$\rangle$. 
This suggest that, in contexts where higher precision is necessary, there are obvious improvements that can be done either in the choice of the alignment criteria or at a later stage, when relevant alignments are selected. \smallskip

\subsubsection{Error analysis}
A problem with using the BLEU score as the only evaluation metric is the fact that it makes no distinction between content and function words, thus not allowing an evaluation focused specifically on the extracted concepts. 
The small size of the corpus, however, allows for some error analysis. 
While postprocessing the automatic translations, the participants were asked to indicate what kind of errors they encountered in each sentence (grammatical, semantical or both). Their observations are summarized in Table \ref{thuman}. \smallskip

\begin{table}[h]
    \centering
    \begin{tabular}{|l|l|l|}
    \hline
    \textbf{}                                                                   \textbf{type of errors}                 & \textbf{DMI (en-it)} & \textbf{CSE (en-sv)} \\ \hline
    \begin{tabular}[c]{@{}l@{}}semantical\end{tabular}               & 23 (46\%)    & 23 (46\%)    \\ \hline
    \begin{tabular}[c]{@{}l@{}}grammatical\end{tabular}             & 10 (20\%)      & 3 (6\%)      \\ \hline
    \begin{tabular}[c]{@{}l@{}}semantical and \\ grammatical\end{tabular} & 3 (6\%)      & 4 (8\%)      \\ \hline
    \end{tabular}
    \caption[Types of errors encountered in the automatically translated sentences]{Number of automatically translated sentences containing only semantical, only grammatical and both semantical and grammatical errors in the two synthetic course plans corpora.}
    \label{thuman}
\end{table}

Interestingly, while most errors are due to wrong alignments, the main difference between two corpora lies in the number of translations that only contain grammatical errors. 
This explains the significant difference observed in the BLEU scores.

In Italian, grammar errors often involve contractions such as ``\textit{del}'' (``\textit{di}'' + ``\textit{il}'', in English ``\textit{of the}''), some of which are systematically rendered as two separate words.
Another common case is that of wrong (or at least very confusing) adjective collocation, such as in the translation ``\textit{il libro presenta una tecnica con miglioramenti vari utile}'' (``\textit{the textbook presents a useful technique with various improvements}''), where the adjective ``\textit{utile}'' (``\textit{useful}''), referred to ``\textit{technique}'' (``\textit{tecnica}'') is placed far from such noun, making the sentence hard to interpret\footnote{the manually postprocessed translation is ``\textit{il libro presenta una utile tecnica con miglioramenti vari}''.}.
Grammatical errors in Swedish are less common and, apart from long adjectives never being turned to the comparative degree periphrastically (e.g. ``\textit{relevant\underline{are}}'' instead of ``\textit{mer relevant}''), less systematic. 
Only in one case, for instance, (``\textit{programbibliotek\underline{en}}''), gender is incorrect. \smallskip

Some alignment errors are also interesting to analyze.
In English-Swedish, while several compounds, such as $\langle$\textit{computer science, datavetenskap}$\rangle$, are aligned correctly, there are cases in which a single-root noun is translated as a compound (e.g. $\langle$ \textit{theory, automatateori} $\rangle$). This is not necessarily an actual alignment error, as it might rather be that the English version of the text was being less specific than its Swedish counterpart, thus producing an alignment that, during evaluation, would have been marked as correct but not generally reusable (\texttt{=}). \smallskip

``Reasonable'' alignment errors appear in the DMI corpus sentences too. 
The English for ``\textit{attendance of lessons}'', for instance, becomes simply ``\textit{frequenza}'' in the Italian translation. 
Such word is in fact often used alone to replace longer expressions such as ``\textit{frequenza delle lezioni}'', just like in English `\textit{of lessons}'' can be omitted when what must be attended is evident from the context.
Another interesting example is the alignment $\langle$ \textit{class, classe} $\rangle$, which causes the sentence \textit{``I will attend the class''} to be (incorrectly) translated as \textit{``io seguirò la classe''} instead of \textit{``io seguirò la lezione''} even though the correspondence is in fact valid in most of the numerous contexts in which \textit{``class''} is not to be intended as a synonym of \textit{``lesson''}.

### 2021-02-09 final_report/include/CE.tex
moved relevant stuff to my own public repo

% CREATED BY DAVID FRISK, 2016
\chapter{Concept Extraction} \label{ch3} 
As mentioned in Section \ref{aligns}, extracting a set of concepts via linguistic comparison is the first and most important step of CA. This chapter describes our CE module, delineates our strategy to evaluate it independently from the other components of the system illustrated in Figure \ref{sys} and presents the results of such preliminary evaluation.

\section{Method and implementation} \label{ce}
For the reasons discussed in Section \ref{ourapproach}, our syntax-based approach to CE is based on comparing, instead on ASTs, dependency trees, and UD trees in particular.
The task of dealing with UD trees is facilitated by the existence of a series of Haskell modules written in the context of the development of the aforementioned \texttt{gf-ud} and of some preliminary unpublished CA experiments\footnote{Aarne Ranta and Prasanth Kolachina, code available at \url{https://github.com/aarneranta/concept-alignment/tree/master/old_ca_versions/v1}}. \smallskip

In this context, as shown in Figure \ref{types}, dependency trees are represented as rose trees, i.e. tree data structures with an unbounded, variable number of branches per node.

\begin{figure}[H]
 \centering
 \scriptsize
 \begin{verbatim}
1 another another DET _ _ 3 det _ _
2 dependency dependency NOUN _ _ 3 compound _ _
3 tree tree NOUN _ _ 0 root _ _

 \end{verbatim}
 \begin{verbatim}

RTree {
 root = UDWord {
  udID = 3, udFORM = "tree", udLEMMA = "tree", udUPOS = "NOUN", (...), udHEAD = 0,
  udDEPREL = "root", (...)}, 
 subtrees = [
  RTree {
  root = UDWord {
   udID = 1, udFORM = "another", udLEMMA = "another", udUPOS = "DET", (...),
   udHEAD = 3, udDEPREL = "det", (...)}, 
  subtrees = []},
  RTree {
  root = UDWord {
   udID = 2, udFORM = "dependency", udLEMMA = "dependency", udUPOS = "NOUN", (...), 
   udHEAD = 3, udDEPREL = "compound", (...)}, 
  subtrees = []}]}

 \end{verbatim}

 \begin{verbatim}

3 tree tree NOUN _ _ 0 root _ _
 1 another another DET _ _ 3 det _ _
 2 dependency dependency NOUN _ _ 3 compound _ _
 \end{verbatim}
 \caption[The CoNLL-U and rose tree representations of a dependency tree]{The CoNLL-U and rose tree representation of a dependency tree. In the rose tree in the middle, fields left black in the CoNLL-U representation are omitted for compactness. The third representation, easier to read, resembles CoNNL-U notation while graphically showing the structure of the tree. This notation will be used for the other examples of this chapter.}
 \label{deptree}
\end{figure}

As shown in Figure \ref{types}, the type of the nodes of such trees, \texttt{UDWord}, is a record type whose fields mirror those of a CoNLL-U file (cf. \ref{conll}). Alignments are, as for Definition \ref{algndef}, simply pairs of UD trees.

\begin{figure}[H] 
 \centering
 \lstinputlisting{pseudocode/BasicTypes.hs}
 \caption[Fundamental data typesfor dependency trees]{Fundamental data types used for representing dependency trees in the CA module.}
 \label{types}
\end{figure}

\subsection{Baseline} \label{baseline}
The original CE algorithm, dating back to the above mentioned CA experiments, has two stages, sketched respectively in Figure \ref{basea} and \ref{baseb}. The first consists in, given two trees corresponding to a sentence and its translation, aligning the entire trees by recursively sorting and padding their lists of subtrees. In particular: \smallskip

\begin{itemize}
 \item sorting is based on the UD label of their root 
 \item what is meant by \textit{padding} is the insertion of a dummy dependency subtree wherever a tree in the source (resp. target) language has a subtree with UD label $l$ that is missing in its target (resp. source) language counterpart. This is useful, for instance, for dealing with cases where a determiner in a sentence in the SL is omitted in its translation.\smallskip
\end{itemize} \smallskip

The result of this first step is a pair of \textit{perfectly aligned} trees, i.e. trees with identical shape that can later be used to extract the pairs of aligned subtrees.

\begin{figure}[H]
 \centering
 \lstinputlisting{pseudocode/BaselineA.hs}
 \caption[The alignment step of the basic CE algorithm]{The alignment step of the basic CE algorithm. Given two full sentence trees, it aligns them by sorting and padding.}
 \label{basea}
\end{figure}

Once two sentence trees are aligned, extracting pairs of aligned subtrees is just a matter of recursively obtaining the lists of all subtrees of each sentence and zipping them together. Every time a pair of subtrees with depth greater than 1 is extracted, in addition, a new pair of subtrees whose only nodes are their roots is added. This makes it possible to find single-word alignments, which we refer to as \textit{head alignments}, that would otherwise be ignored.

\begin{figure}[H]
 \centering
 \lstinputlisting{pseudocode/BaselineB.hs}
 \caption[The extraction step of the basic CE algorithm]{The extraction step of the basic CE algorithm. Given two perfectly aligned trees, it returns a list of pairs of aligned subtrees. To do that, it extracts all subtrees (at any depth, cf. \texttt{subst'}) of each member of the sentence alignment. For each extracted subtree, a new tree whose only node is its root is also created (cf. \texttt{headt}).}
 \label{baseb}
\end{figure}

After extracting subtree alignments, it is possible and generally useful to remove the dummy subtrees added during alignment they contain.

Obviously, the ``full-sentence'' tree pair is itself an alignment, even though a trivial one, as it represents a sentence-level correspondence in a case where, due to the approach being syntactic comparison, the inputs are assumed to be sentences that correspond to each other\footnote{contrary to what it may seem, however, sentence-level alignment is in general not at a trivial task, as it is often the case, especially in certain language pairs, that one \textit{orthographic} sentence, i.e. a sentence defined based on the presence of a full stop, maps to more than one orthographic sentences in the translated text, or vice versa.}.

\begin{figure}[H]
 \centering
 \small
 \begin{figure}[H]
  \scriptsize
  \begin{subfigure}{.46\textwidth}
  \centering
  \begin{verbatim}
4 aligned aligned ADJ _ _ 0 root _ _
 1 we we PRON _ _ 4 nsubj _ _
 2 are be AUX _ _ 4 cop _ _
 3 perfectly perfectly ADV _ _ 4 advmod _ _
  \end{verbatim}
  \end{subfigure}
  \begin{subfigure}{.5\textwidth}
  \centering
  \begin{verbatim}
3 allineati allineato ADJ _ _ 0 root _ _
 1 siamo essere AUX _ _ 3 cop _ _
 2 perfettamente perfettamente ADV _ _ 3 advmod _ _
  \end{verbatim}
  \end{subfigure}
 \end{figure}
 $\Downarrow$ sorting
 \begin{figure}[H]
  \scriptsize
  \begin{subfigure}{.46\textwidth}
  \centering
  \begin{verbatim}
4 aligned aligned ADJ _ _ 0 root _ _
 3 perfectly perfectly ADV _ _ 4 advmod _ _
 2 are be AUX _ _ 4 cop _ _
 1 we we PRON _ _ 4 nsubj _ _
\end{verbatim}
  \end{subfigure}
  \begin{subfigure}{.5\textwidth}
  \centering
  \begin{verbatim}
3 allineati allineato ADJ _ _ 0 root _ _
 2 perfettamente perfettamente ADV _ _ 3 advmod _ _
 1 siamo essere AUX _ _ 3 cop _ _
     \end{verbatim}
  \end{subfigure}
 \end{figure}
 $\Downarrow$ padding
 \begin{figure}[H]
  \scriptsize
  \begin{subfigure}{.46\textwidth}
  \centering
  \begin{verbatim}
4 aligned aligned ADJ _ _ 0 root _ _
 3 perfectly perfectly ADV _ _ 4 advmod _ _
 2 are be AUX _ _ 4 cop _ _
 1 we we PRON _ _ 4 nsubj _ _
   \end{verbatim}
  \end{subfigure}
  \begin{subfigure}{.5\textwidth}
  \centering
  \begin{Verbatim}[commandchars=\\\{\}]
3 allineati allineato ADJ _ _ 0 root _ _
 2 perfettamente perfettamente ADV _ _ 3 advmod _ _
 1 siamo essere AUX _ _ 3 cop _ _
 \textbf{D}
        \end{Verbatim}
  \end{subfigure}
 \end{figure}
 \begin{figure}[H]
 \begin{subfigure}{.5\textwidth}
 \centering
 \setlength{\unitlength}{0.26mm}
 \scriptsize
 \begin{picture}(277.0,110.0)
 \put(0.0,0.0){we}
 \put(46.0,0.0){are}
 \put(83.0,0.0){perfectly}
 \put(174.0,0.0){aligned}
 \put(0.0,15.0){{\tiny PRON}}
 \put(46.0,15.0){{\tiny AUX}}
 \put(83.0,15.0){{\tiny ADV}}
 \put(174.0,15.0){{\tiny ADJ}}
 \put(97.0,30.0){\oval(172.27586206896552,100.0)[t]}
 \put(10.862068965517238,35.0){\vector(0,-1){5.0}}
 \put(82.0,83.0){{\tiny nsubj}}
 \put(120.0,30.0){\oval(125.65625,66.66666666666667)[t]}
 \put(57.171875,35.0){\vector(0,-1){5.0}}
 \put(105.0,66.33333333333334){{\tiny cop}}
 \put(138.5,30.0){\oval(87.7032967032967,33.333333333333336)[t]}
 \put(94.64835164835165,35.0){\vector(0,-1){5.0}}
 \put(123.5,49.66666666666667){{\tiny advmod}}
 \put(189.0,110.0){\vector(0,-1){80.0}}
 \put(194.0,100.0){{\tiny root}}
 \end{picture}
 \end{subfigure}%
 \begin{subfigure}{.5\textwidth}
 \scriptsize
 \centering
 \setlength{\unitlength}{0.26mm}
\begin{picture}(349.0,110.0)
 \put(0.0,0.0){\textbf{D}}
 \put(46.0,0.0){siamo}
 \put(101.0,0.0){perfettamente}
 \put(228.0,0.0){allineati}
 \put(0.0,15.0){{}}
 \put(46.0,15.0){{\tiny AUX}}
 \put(101.0,15.0){{\tiny ADV}}
 \put(228.0,15.0){{\tiny ADJ}}
 \put(124.0,30.0){\oval(226.68421052631578,100.0)[t]}
 \put(10.65789473684211,35.0){\vector(0,-1){5.0}}
 \put(109.0,83.0){{}}
 \put(147.0,30.0){\oval(180.35164835164835,66.66666666666667)[t]}
 \put(56.824175824175825,35.0){\vector(0,-1){5.0}}
 \put(132.0,66.33333333333334){{\tiny cop}}
 \put(174.5,30.0){\oval(124.63779527559056,33.333333333333336)[t]}
 \put(112.18110236220471,35.0){\vector(0,-1){5.0}}
 \put(159.5,49.66666666666667){{\tiny advmod}}
 \put(243.0,110.0){\vector(0,-1){80.0}}
 \put(248.0,100.0){{\tiny root}}
\end{picture}
 \end{subfigure}
\end{figure}
 \scriptsize
 \smallskip
 \smallskip
 \smallskip
 \begin{itemize}
 \item $\langle$\textit{we are perfectly aligned, siamo perfettamente allineati}$\rangle$
 \item $\langle$\textit{aligned,allineati}$\rangle$ (head alignment)
 \item $\langle$\textit{are}$\rangle$
 \item $\langle$\textit{siamo}$\rangle$
 \item $\langle$\textit{perfectly,perfettamente}$\rangle$
 \item $\langle$\textit{aligned,allineati}$\rangle$
 \end{itemize}
 \caption[Steps of the baseline CE algorithm.]{Steps of the baseline CE algorithm. First, the rose tree representations of two yet-to-align dependency trees. Just below, the two rose trees after sorting and padding. Finally, the graphical representations of the resulting perfectly aligned trees and the linearizations of the aligned subtrees extracted from them. Since, in Italian, the subject is implicit, a dummy node, marked as \textbf{D}, is added to the corresponding tree during the padding step.}
 \label{prefal}
\end{figure}

As we will see in the following sections, such algorithm is useful both as a baseline and as a source of inspiration for our improved CE program.

\subsection{Proposed improvements} \label{improvements}
While one of the most important ideas in the improved version of the CE module proposed in this thesis remains aligning (sub)trees with identical root UD labels, using this criterion alone makes our baseline both unable to detect many of the existing correspondences and prone to extract incorrect ones. This happens both when syntactic similarity is only apparent and when words are aligned based exclusively on the fact that they are the roots of two aligned subtrees\footnote{see Section \ref{heads} for a more detail discussion on head alignment.}.
As a consequence, the objective of this part of the project is to improve both the precision and the recall of the algorithm (or rather, for reasons that will be discussed in Section \ref{metrics}, two approximations of such metrics).

\subsubsection{Multiple alignment criteria} \label{criteria}
One obvious way to increase the total number of alignments the algorithm detects is to, instead of considering as aligned only subtrees in \textit{matching contexts}, i.e. whose heads are attached to the same node, and with matching UD labels, make use of a set of additional, possibly more relaxed, criteria. In the following, formal definitions of all the criteria used in the current implementation, including the original one, are given.

\paragraph{Label matching}
As mentioned in section \ref{baseline}, the only alignment criterion the original version of the CE module makes use of is based on comparing the UD labels of each pair of trees candidate for alignment, i.e. of each pair of trees in matching contexts. Formally: \smallskip

\begin{criterion}[Matching UD labels] \label{udmatch}
 Two dependency trees $t$, $u$ will be aligned if their roots share the same UD label.
\end{criterion} \smallskip

When referring to UD labels we disregard, unless otherwise specified, subtypes.

\paragraph{POS-equivalence}
As mentioned in Section \ref{ud}, dependency trees provide information not only on the syntactic role of each word, but also on their grammatical category, represented as a universal Part Of Speech (POS) tag. Intuitively, if the words corresponding to the nodes of two trees in matching contexts have the same POS tags, the two trees are generally more likely to correspond to each other than if not. 
As a consequence, a useful relation to define between dependency trees is that of \textit{POS-equivalence}: \smallskip

\begin{criterion}[POS-equivalence] \label{poseq}
 Two dependency trees $t$, $u$ are POS-equivalent (and, as such, will be aligned) if $M_1 = M_2 \neq \emptyset$, where $M_i$ is defined as the multiset of POS tags of all the meaning-carrying word nodes of $t_i$. 
\end{criterion} \smallskip

The definition specifies that the two multisets should not contain the POS tags of all the words in the sentence but rather only those of the \textit{meaning-carrying} ones, referring to words belonging to a particular set of classes. 
Words that should generally be taken into account, in fact, correspond roughly to content words (cf. Section \ref{upos}), but this term was deliberately avoided in Definition \ref{poseq} due to the fact that it can be useful also to include some function words, for instance pronouns and some kinds of determiners.
In particular, the current implementation considers as meaning-carrying all words that belong to an open class - defined as in the UD documentation \cite{openpos} - and numerals, but this set of tags has been obtained empirically working with English-Italian and might not be ideal for all language pairs.
On the other hand, words belonging to other classes, such as auxiliary verbs, adpositions and conjunctions can and should be in most cases ignored as they are often omitted or rendered with words with different POS tags when the sentence is translated to another language, especially if the two languages in the pair at hand differ significantly. \smallskip

Applied alone, this criterion can be used to capture correspondences that would otherwise be missed, thus increasing recall, but a decrease in precision is also to be expected. Perhaps more interestingly, however, another way to apply this criterion is in conjunction with other ones, and in particular together with UD label matching (cf. Criterion \ref{udmatch}), in context where high precision is more important than recall. We will get back to combining criteria in Section \ref{ceres}.

\paragraph{Handling divergences} \label{divs}
While we do not want to make our CE module language pair-specific, there are many cases where parallel texts present significant and systematic cross-linguistic distinctions. Some of these distinctions, formalized in \cite{divs}, have nothing to do with idiomatic usage or aspectual, discourse, domain or word knowledge and are not specific of particular language pairs, even though they do occur more often in some than they do in others. 
Drawing inspiration from \cite{divs} and \cite{divs2}, we refer to these distinctions as \textit{divergences} and introduce a third alignment criterion based on them: \smallskip

\begin{criterion}[Known divergence]
 Two dependency trees $t$, $u$ will be aligned if they match a known divergence pattern. 
\end{criterion} \smallskip

In \cite{divs}, seven classes of divergences are identified: \textit{thematic}, \textit{promotional}, \textit{demotional}, \textit{structural}, \textit{conflational}, \textit{categorial} and \textit{lexical}. 
However, because the author's proposed way to resolve divergences assumes the availability of more than merely syntactic information\footnote{in particular, of a lexicon of Root Lexical Conceptual Structures specifying, for instance, the logical subject, arguments and modifiers of each verb.} and because, working with UD trees, we do not have access to anything but strictly syntactical and morphological annotations, the very task of identifying these distinctions at this level of granularity becomes, in our position, extremely challenging. 
Consequently, we refer to the simpler classification and less formal definitions proposed in \cite{divs2}, where promotional and demotional divergences are merged in a wider-coverage category of \textit{head-swapping} divergences and where lexical divergences, the hardest to handle in the absence of any kind of semantic information, do not appear at all.
In the following, we give definitions and examples for each of these classes of divergences, specifying to what extent and how\footnote{unless otherwise specified, each divergence can be expressed as a set of simple rules in the form of boolean functions taking the two UD trees candidate for alignment as input.} each of them is handled in the proposed language-agnostic CE module. We do not attempt to cover all possible cases, but rather provide a few examples as a proof of concept.

\subparagraph*{Categorial divergence}
A categorial divergence happens when the translation of a word with POS tag $P_1$ is a word with a different POS tag $P_2$. 
The instances of this class of divergences our CE module handles explicitly are some of those that cannot be captured by Criterion \ref{udmatch} alone due to the fact that the difference in POS tags also causes the UD labels to be different, and in particular:

\begin{itemize}
 \item cases where an adverb in the source language corresponds to an adjective in the target language (and vice versa: all rules the program is based on are symmetrical), e.g.
 \begin{example}
 ``Roberta listens \underline{distractedly}'' VS ``Roberta lyssnar \underline{distraherad}''
 \end{example}
 where the English ``\textit{distractedly}'' is an adverb and the Swedish ``\textit{distraherad}'' is an adjective and, as such, would be labelled respectively as an \texttt{advmod} of ``\textit{lyssnar}'' and as an \texttt{amod} of ``\textit{Roberta}''
 \item cases where a nominal modifier (\texttt{nmod}) in the source language is rendered as an adjectival modifier (\texttt{amod}) in the target language, like the following, where the English adjective ``\textit{doctoral}'' becomes the Italian noun ``\textit{dottorato}'', preceded by the preposition ``\textit{di}'':
 \begin{example} \label{herb}
 ``Herbert completed his \underline{doctoral} thesis'' VS ``Herbert ha completato la sua tesi \underline{di dottorato}''
 \end{example}
 \item cases where an adverb is rendered as an oblique, such as
 \begin{example}
 ``Nicola studies \underline{consistently}'' VS ``Nicola studia \underline{con costanza}''
 \end{example}
\end{itemize} \smallbreak

There are indeed other divergences of this kind that can occur in a parallel text and cannot be captured by Criterion \ref{udmatch}, such as
\begin{example}
 ``Claudio is \underline{hungry}'' VS ``Claudio ha \underline{fame}''
\end{example}
where, in the original sentence, the complement of the copula \textit{am}, \textit{hungry}, would be labelled as its \texttt{root} but, in Italian, the noun \textit{fame} is the object of the root verb \textit{ha}. 
It proved hard to detect cases like this via purely syntactic rules without causing the program to extract a lot of incorrect alignments as well, but if necessary it is easy to modify the program by removing or adding rules of this kind, potentially even language-specific.

\subparagraph*{Conflational divergence} \label{confl}
Conflational divergences involve the translation of two or more words in the source language using a single word that combines their meanings in the target language. A straightforward example is that of compounds, for instance
\begin{example} 
 ``Filippo is interested in \underline{game design}'' VS ``Filippo är intresserad av \underline{spelutvekling}''
\end{example}
but there are also other cases, like
\begin{example} 
 ``I'll come and \underline{say hello} to Bruno and Andrea'' VS ``Verrò a \underline{salutare} Bruno e Andrea''
\end{example} 
where the single word \textit{salutare} is not a compound but expresses the same meaning as ``say'' and ``hi'' together. \smallskip
Divergences like the one described in the latter example are usually captured by Criterion \ref{udmatch}, while compounds often need to be taken care of explicitly. For reasons that will become clear later, these cases are discussed in Section \ref{heads}.

\subparagraph*{Structural divergence}
Structural divergences happen when the subject, object or indirect object of a sentence in the source language are rendered as obliques in the source language. For instance, in the sentences
\begin{example}
 ``I called \underline{Francesco}'' VS ``Ho telefonato \underline{a Francesco}''
\end{example}
what is the direct object in English (\textit{Francesco}) becomes a prepositional phrase (\textit{a Francesco}) in Italian. 
Even though in many cases these divergences can be captured by POS-equivalence, in order to ensure higher precision, the CE module handles all of them explicitly via rules that combine it with UD label checking and whose priority is higher than that of the basic Criterion \ref{poseq}.

\subparagraph*{Head swapping divergences}
Head swapping divergences always involve a head verb and a logical modifier and occur when the logical modifier is placed lower (resp. higher) in the source language sentence than in its target language counterpart. For instance, in the following example
\begin{example}
 ``Anna \underline{usually} goes for walks'' VS ``Anna \underline{brukar} promenera''
\end{example} 
in Swedish, the logical modifier \textit{usually} is implicitly expressed by the verb \textit{brukar} itself, i.e. placed ``higher up''. \smallskip

These divergences are, while not uncommon, hard to handle without access to a lexicon where entries for verbs are complete with a list of arguments and modifiers\footnote{on the other hand, developing the CA module further could be a way to construct such a ``rich'' lexicon, as will be discussed in Chapter \ref{ch6}.}.

\subparagraph*{Thematic divergence}
Thematic divergences can happen when the logical subject of a sentence in the target language differs from its grammatical subject. An example of thematic divergence is the following:
\begin{example} \label{them}
 ``\underline{Yana} likes \underline{books}'' VS ``\underline{A Yana} piacciono \underline{i libri}''
\end{example}
In this case, in the Italian translation, the (logical and grammatical) subject of the English sentence, \textit{Yana}, is expressed, even though it is still the logical subject of the sentence, as an oblique; the object \textit{books} becomes the grammatical subject (\textit{i libri}) in Italian, while the head verb remains in both cases the root. \smallskip

Similar to this are the cases where a passive sentence in the source language is rendered as active in the target language and, as a consequence, the subject and complements are swapped:
\begin{example} \label{pass}
 ``\underline{The game} had only been tried \underline{by Andrea}'' VS ``Solo \underline{Andrea} aveva provato \underline{il gioco}''
\end{example} \smallskip

One could object that divergences like the above (and some other) could - and maybe \textit{should} - be avoided when translating a text. We will not dwell upon this, since what matters for the purpose of CA is not what is desirable in natural language translation but rather what divergences do occur in the human-translated parallel texts available for analysis. \smallskip

Given that UD provides a subtype \texttt{pass} for \texttt{nsubj}s and auxiliary verbs, it is easier to explicitly handle cases such that of Example \ref{pass}, while those that resemble Example \ref{them} are left, for the moment, unhandled.

\subsubsection{New extraction algorithm} \label{extralg}
The decision to use different alignment criteria simultaneously, together with the nature of the criteria described in the above themselves, implies that perfect alignment must be given up: the new algorithm is such that aligned subtrees are not extracted from a pair of sorted and padded sentence trees but rather obtained by direct comparison between the original trees. Aligning a pair of sentences means, then, checking if the corresponding UD trees $t$, $u$ match any of the criteria at hand and, if they do, considering them as the members of an alignment, adding the alignment to a collection and recursively repeating the same procedures for all possible pairs $t'$, $u'$, where $t'$ is a subtree of $t$ and $u'$ is a subtree of $u$.

\begin{figure}[H]
 \centering
 \lstinputlisting{pseudocode/Extract.hs}
 \caption[Basic version of the improved CE algorithm]{Basic version of the improved CE algorithm. Instead of aligning full sentence trees and only then extract concepts comparing the structure of the trees and their UD labels only, it recursively extracts pairs of aligned subtrees directly. To do so, for each pair of subtrees occurring in the same context, it tries to apply multiple criteria, implemented as boolean functions telling whether two trees should become the members of an alignment.}
 \label{extr}
\end{figure}

\paragraph{Pruning alignments} \label{pruning} 
In its most basic version, this algorithm does nothing to avoid subtrees in the source language to be aligned with multiple subtrees in the target language and vice versa. As a consequence, it systematically \textit{overgenerates} alignments. \smallskip

To avoid overgeneration, or rather to counter it, we need a way to select, in the very common case that the algorithm detects several alignments alternative to each other within the same sentence, the one that is more likely to be correct. A way to do this is to sort such alignments based on their ``reliability'' and only keeping the first alternative. The question becomes, then, how to sort the alignments. \smallskip

To answer this, it must be first of all said that the criteria described in Section \ref{criteria} are not to be considered equally reliable nor describe equally frequent situations. For this reason, a way to obtain a series of alignments that are, in part, implicitly sorted is to apply criteria (or particular combinations of criteria) in a specific order that can be determined empirically and is easily modifiable thanks to the fact that, in its current implementation, the algorithm outlined above takes a list of criteria assumed to be in order of priority as one of its parameters. \smallskip

It can happen, however, that two trees match \textit{more than one} of the criteria. For instance, two trees might have their roots sharing the same UD label (cf. Criterion \ref{udmatch}) \textit{and} be POS-equivalent. In such cases, it intuitively makes sense to consider them more likely to be exact, as there are literally more reasons to align them. This makes the implicit ordering that the solution just described produces ``for free'' insufficient and leads to the necessity to keep track of the set of reasons why an alignment has been extracted and only then sort the alternative alignments. From the implementative point of view, each alignment is simply associated with a \texttt{Set} of labels, whose type is in fact named \texttt{Reason} and has an ordering defined over. The final sorting is based on:\smallskip

\begin{enumerate}
 \item the number of reasons for alignment
 \item the priority level of the highest-priority reason
\end{enumerate} \smallskip

These labels can even be part of the final output of the CE module, thus adding an ulterior level of explainability to the system it belongs to.

\paragraph{Aligning heads} \label{heads}
As mentioned in Section \ref{baseline}, the original version of the algorithm creates, every time an alignment is extracted, an additional one for the for the heads of its two members, which we refer to as a \textit{head alignment}. Doing this is a crucial part of the algorithm, as the following example, concerning a straightforward to align pair of sentences, shows: \smallskip
\begin{example}
 Consider the English sentence ``Enrico eats a banana'' and its Italian equivalent ``Enrico mangia una banana''. As the figure below shows, their trees are perfectly aligned without any need for padding or sorting:
 \begin{figure}[h]
 \centering
 \setlength{\unitlength}{0.25mm}
 \begin{subfigure}{.5\textwidth}
 \centering
 \begin{picture}(241.0,90.0)
  \put(0.0,0.0){Enrico}
  \put(64.0,0.0){eats}
  \put(110.0,0.0){a}
  \put(147.0,0.0){banana}
  \put(0.0,15.0){{\tiny PROPN}}
  \put(64.0,15.0){{\tiny VERB}}
  \put(110.0,15.0){{\tiny DET}}
  \put(147.0,15.0){{\tiny NOUN}}
  \put(42.0,30.0){\oval(59.3125,33.333333333333336)[t]}
  \put(12.34375,35.0){\vector(0,-1){5.0}}
  \put(27.0,49.66666666666667){{\tiny nsubj}}
  \put(138.5,30.0){\oval(28.89189189189189,33.333333333333336)[t]}
  \put(124.05405405405405,35.0){\vector(0,-1){5.0}}
  \put(123.5,49.66666666666667){{\tiny det}}
  \put(125.5,30.0){\oval(79.3855421686747,66.66666666666667)[t]}
  \put(165.19277108433735,35.0){\vector(0,-1){5.0}}
  \put(110.5,66.33333333333334){{\tiny obj}}
  \put(79.0,90.0){\vector(0,-1){60.0}}
  \put(84.0,80.0){{\tiny root}}
 \end{picture}
 \end{subfigure}%
 \begin{subfigure}{.5\textwidth}
  \centering
  \setlength{\unitlength}{0.25mm}
  \begin{picture}(259.0,90.0)
  \put(0.0,0.0){Enrico}
  \put(64.0,0.0){mangia}
  \put(128.0,0.0){una}
  \put(165.0,0.0){banana}
  \put(0.0,15.0){{\tiny PROPN}}
  \put(64.0,15.0){{\tiny VERB}}
  \put(128.0,15.0){{\tiny DET}}
  \put(165.0,15.0){{\tiny NOUN}}
  \put(42.0,30.0){\oval(59.3125,33.333333333333336)[t]}
  \put(12.34375,35.0){\vector(0,-1){5.0}}
  \put(27.0,49.66666666666667){{\tiny nsubj}}
  \put(156.5,30.0){\oval(28.89189189189189,33.333333333333336)[t]}
  \put(142.05405405405406,35.0){\vector(0,-1){5.0}}
  \put(141.5,49.66666666666667){{\tiny det}}
  \put(134.5,30.0){\oval(98.02970297029702,66.66666666666667)[t]}
  \put(183.5148514851485,35.0){\vector(0,-1){5.0}}
  \put(119.5,66.33333333333334){{\tiny obj}}
  \put(79.0,90.0){\vector(0,-1){60.0}}
  \put(84.0,80.0){{\tiny root}}
 \end{picture}
 \end{subfigure}
 \label{enrico}
 \end{figure}

 Without head alignments, the output of the algorithm described so far would be:
 \begin{itemize}
 \item $\langle$Enrico eats a banana, Enrico mangia una banana$\rangle$
 \item $\langle$Enrico, Enrico$\rangle$
 \item $\langle$a banana, una banana$\rangle$
 \item $\langle$a, una$\rangle$
 \end{itemize} 
 while introducing head alignment leads to detecting two additional, equally relevant one-word correspondences\footnote{note how, even with head alignments, not all potentially interesting alignments are found. In this example, for instance, it could be useful to align the verb phrase ``\textit{eats a banana}'' with its Italian counterpart ``\textit{mangia una banana}'', whose members are also not, strictly speaking, subtrees of the original sentences. The approach we use for head alignment, as we will discuss in Chapter \ref{ch6}, could be generalized to cases like this.}:
 \begin{itemize}
 \item $\langle$eats, mangia$\rangle$
 \item $\langle$banana, banana$\rangle$
 \end{itemize} 
\end{example}

While it is of extreme importance not to miss alignments like the last two in the example above, aligning heads is not always appropriate, especially when, as in the present case, common translation divergence patterns are one of the criteria. For instance, when two trees are aligned because of a categorial divergence such as that of Example \ref{herb}, where \textit{doctoral} corresponds to \textit{di dottorato}, it is at least questionable to also align \textit{doctoral} with \textit{dottorato}. Consequently, in the current implementation of the algorithm, each alignment criterion is associated with a flag telling whether heads should also be aligned whenever a new alignment is extracted because of that criterion. \smallskip

In a way, even though it is handled differently from the others discussed so far, that of aligning heads could also be seen as another alignment criterion: \smallskip

\begin{criterion}[Heads of matching trees] \label{cheads}
 The roots $n_1$, $n_2$ of two trees $t$, $u$ will be aligned if $t$ and $u$ are aligned.
\end{criterion}

There are also cases where some sort of head alignment is desirable but it is not as simple as creating an additional alignment from the roots of each pair of aligned subtrees. As a consequence, an important part of the improved CE module developed within this project is a function \texttt{alignHeads} that performs head alignment with the necessary caveats. Following are the two special cases its current implementation is able to handle correctly despite being less straightforward. As for translation divergences, there may well be other cases that can be taken care of similarly: the objective of what has been implemented so far is mostly to demonstrate how this can be done.

\subparagraph*{Auxiliaries}
When the translation counterpart of a verb in the SL is composed of a lexical verb and one or more auxiliaries (or vice versa), it is desirable to align the SL verb to the entire TL subtree composed of the head verb and its auxiliaries, like in the following example:
\begin{example}
 ``Many important decisions \underline{were taken} by Tommaso'' VS ``Många viktiga beslut \underline{togs} av Tommaso''
\end{example}
Achieving this is relatively straightforward: before aligning a pair of head verbs, their dependents labelled \texttt{aux} must be looked for. In case the head verb in the TL has one or more such dependents, but its SL counterpart does not, the TL member of the new alignment will not be composed by the head exclusively, but will include the \texttt{aux} subtrees.

\subparagraph*{Compounds}
We mentioned compounds as a type of conflational divergence (cf. Section \ref{confl}). Given the variety of ways in which the translation equivalent of a compound can be expressed and the subsequent variety of related UD labels (a compound can be rendered as another compound, as a \texttt{compound} (space-separated) expression, as a noun modified by an adjective and/or another noun, as a \texttt{flat} multiword expression, as a combination of these things...), compounds are slightly harder to align than main verb + auxiliaries constructions in practice, but the solution is conceptually the same: before aligning two heads, we check whether the list of their immediate dependents contains anything labelled \texttt{compound}, \texttt{flat}, \texttt{amod} or \texttt{nmod}. If that's the case in, for instance, only the sentence in the source language, then it is likely that the head of the tree in the target language is in fact a compound. If so, it is aligned not just with the head of the tree in the source language, but with the tree composed of the head and the list of subtrees labelled as \texttt{compound}, \texttt{flat}, \texttt{amod} or \texttt{nmod}. In this way, our program is generally able to find the counterparts of a compound even when they are very complex, such as in the following case:
\begin{example} 
 ``I took a course on \underline{Machine Learning techniques}'' VS ``Jag deltog i en kurs om \underline{maskininlärningstekniker}''
\end{example}

\paragraph{Counting and reusing alignments}
We mentioned how knowing why a certain alignment was identified is important to know to what extent we can trust the program to having taken the right decision. When working on a full parallel text instead of on a single sentence, another important information is the number of occurrences of that alignment (disregarding all sentence-specific information such as the linear positions of its nodes) throughout the entire corpus: unless it is the result of a systematic error (and even in this case, having it presented as a first-class alignment will help finding the mistake), an alignment occurring multiple times can be considered ``more reliable'' as it is in a way ``corroborated''. As a consequence each alignment produced by our CE module is associated with a pair whose first element is the set of reasons for that specific alignment and whose second element is its total number of occurrences. \smallskip

Furthermore, the fact that an alignment has already occurred can be used as an additional, backup criterion for when none of those described in section \ref{criteria} apply: \smallskip

\begin{criterion}[Known alignment] \label{known}
 Two dependency trees $t$, $u$ will be aligned if such alignment belongs to a set $K$ of known ones.
\end{criterion} \smallskip

In the case at hand, $K$ is initialized as empty and iteratively augmented with the results of aligning each pair of sentences, but it is not hard to imagine cases in which starting with a nonempty set would be useful.

\section{Evaluation} \label{eval2}
We conclude this chapter by describing the data and the approaches used to evaluate the CE module described in Section \ref{ce} independently from the subsequent stages of the proposed pipeline and by presenting the results of such early evaluation.

\subsection{Data} \label{pud}
\subsubsection{PUD treebanks}
Because we want part of our evaluation to be independent from the quality of the dependency trees obtained by automatically parsing sentence-aligned text, a portion of the data used for this purpose consist in manually annotated data. \smallskip

In particular, we use a subset of the Parallel UD (PUD) corpus, a set of handcrafted\footnote{for some languages, dependency labels were actually automatically converted to UD format from other standards. Furthermore, manual annotation often only concerns some of fields of the CoNLL-U files, but some manual annotation is generally involved in assigning POS tags and dependency relations.} multilingual treebanks in CoNLL-U format created for the CoNLL 2017 shared task on Multilingual Parsing from Raw Text to Universal Dependencies \cite{mprtud}. This prevents the CE module from failing because of parse errors, even though a small number of annotation inconsistencies and imprecisions is still present. \smallskip

PUD treebanks are available in 20+ languages, of which we selected Italian, English and Swedish, and are composed of 1000 sentences taken from the news domain and from Wikipedia. Due to the lack of a gold standard to refer to in terms of CE and to the consequent need to manually assess the correctness of each alignment obtained, for most of the evaluation we only use the first 100 of these sentences. \smallskip

\subsubsection{Course plans} \label{plans}
When it comes to testing the program on raw text to be parsed automatically, we use two bilingual sentence-aligned corpora consisting of course plans from the Department of Mathematics and Computer Science of the University of Perugia (for English-Italian) and from the Department of Computer Science and Engineering (CSE) shared between the University of Gothenburg and the Chalmers University of Technology (for English-Swedish). For brevity, we will refer to these two datasets as to the DMI and CSE corpora throughout the text.\smallskip

Part of this data were collected and sentence-aligned specifically for this work, but a core of CSE plans had already been gathered as part of another thesis project \cite{thesis}. \smallskip

Our parser of choice is UDPipe \cite{udpipe1}. For parsing the two corpora, we use language models trained on two distinct parallel treebanks: for English-Italian, we selected ParTUT, an originally Italian-only treebank including texts of various genres, while for English-Swedish our choice was LinES, a larger bilingual treebank \cite{uddocs}.

\subsection{Evaluation metrics} \label{metrics}
While precision and recall are two well-known performance metrics that would be very well suited to the evaluation of our CE module, the lack of a CE gold standard forces us to approximate them with respectively: \smallskip

\begin{itemize}
 \item the number of correct alignments the program is able to extract
 \item the ratio between such number and the total number of alignments extracted
\end{itemize} \smallskip

Determining whether an alignment is correct is not, however, a completely trivial task, since there are correspondences which are correct but not easily reusable in contexts other than that in which they were found. For instance, in the following case: \smallskip

\begin{example} \label{extrain}
 ``He missed the \underline{boat}'' VS ``Ha perso il \underline{treno}''
\end{example} \smallskip

it is hard to deny that \textit{boat} has been ``translated'' as \textit{treno}, ``train'', so our CE module would - and should - be able to detect it, but in most situations it is by no means desirable that such correspondence at the word level is made use of, for instance by storing it in an bilingual lexicon!
As a consequence, each alignment can be marked as:
\begin{itemize}
 \item correct and useful for translation (\texttt{+})\footnote{at least to some extent: we observe that perfect translation equivalents that can be replaced with each other in all contexts are extremely rare, just like perfect synonyms.}
 \item correct but not useful for translation (\texttt{=}). This means that the CE module is working as expected and that we are are faced with a divergence in the broader sense of the term, i.e. potentially due to an idiomatic usage of language
 \item incorrect (\texttt{-}).
\end{itemize} \smallskip

For instance, if we feed the CE module the pair of sentences of Example \ref{extrain}, as a result\footnote{this is what the actual output of the program in evaluation (or ``linearized") mode after annotation looks like. The abbreviations in square brackets refer to alignment Criteria \ref*{udmatch}, \ref{poseq} and \ref{cheads} described in the above.} we get: \smallskip \label{cann}
\begin{verbatim} 
 +he missed the boat|ha perso il treno[UD,POS]1
 +missed|ha perso[UD,POS,HEAD]1
 =the boat|il treno[UD,POS]1
 =boat|treno[UD,POS,HEAD]1
 +the|il[UD]1
\end{verbatim} 
\smallskip

Furthermore, the absence a reference solution makes it so that comparing the updated versions of the module with the baseline described in Section \ref{baseline}, thus focussing on measuring the improvements with respect to it rather than the quality of the results under absolute terms, is in practice the best way to assess the results.
While not ideal, repeatedly performing this kind of evaluation, starting already during the early stages of its development, proved to be useful to understand the impact of each of the changes made, thus guiding further modifications and helping debugging the program when necessary. \smallskip

\subsection{Implementation details} \label{evalign}
In order to be able to perform the type of evaluation described in the above Section and because manual inspection of the results was indispensable, the program was made able to write alignments in the reader-friendly output format shown above. The notation is the following: \smallskip

\begin{verbatim}
 linearized SL tree|linearized TL tree[R1,R2,...,Rn]N
\end{verbatim} \smallskip

where \texttt{[R1,R2,...,Rn]} is the list of criteria the members of the alignment match and \texttt{N} is the number of occurrences of the alignment in the corpus. \smallskip

With alignments in this format available, a simple way to evaluate a specific version of the CE module is to manually assess the correctness of each of the correspondences the program identifies and compute some statistics about them. 
A separate Haskell module, \texttt{EvAlign}, was written to parse files in the above format and compute a series of useful statistics: \smallskip
\begin{itemize}
 \item the total number of alignments found
 \item the number of \textit{distinct} alignments, i.e. the number of alignments with distinct linearizations
 \item the percentage of correct alignments
 \item the percentage of correct alignments that are also useful for translation
 \item the percentage of alignments found due to each combination of criteria, and how many of them are correct.
\end{itemize} \smallskip

While labelling all alignments by hand was necessary to evaluate the baseline, as long as the corpus does not change, it is possible to evaluate later versions of the program semi-automatically, as the correctness of many alignments is most likely already known thanks to the first round of manual annotation or to any of the successive ones. \texttt{EvAlign} can in fact also be run interactively with a labelled file and an unlabelled file as input. When this is the case, the program tries to label as many alignments as possible based on the content of the labelled file (which, of course, can even be the result of merging multiple files obtained with different versions of the program), asking the user to assess the correctness of newly found ones only. If run in this modality, the program also displays some more statistics that make comparison it with the older, labelled file easier, such as the number of new correct alignment found and that of incorrect alignments lost.

\subsection{Evaluating CE against a baseline} \label{ceres}
Table \ref{tbase} compares the results obtained with the baseline with those obtained with the proposed improved version on the 100-sentence manually annotated corpus described in Section \ref{pud}, both for English-Italian and for English-Swedish.

% Please add the following required packages to your document preamble:
% \usepackage[normalem]{ulem}
% \useunder{\uline}{\ul}{}
\begin{table}[H]
 \centering
 \small
 \begin{tabular}{|l|l|l|l|l|}
 \hline
   & \multicolumn{2}{l|}{\textbf{baseline}} & \multicolumn{2}{l|}{\textbf{improved version}} \\ \hline
 \textbf{}   & \textbf{en-it}  & \textbf{en-sv} & \textbf{en-it}  & \textbf{en-sv} \\ \hline
 \textbf{distinct alignments} & 1097  & 1257 & 1198   & 1314  \\ \hline
 \textbf{correct (+ and =)} & 830 (58.12\%) & 995 (79.15\%)  & 964 (80.46\%) & 1105 (84.03\%) \\ \hline
 \textbf{correct and useful (+)} & 776 (54.34\%) & 976 (77.64\%)  & 896 (74.79\%) & 1082 (82.28\%) \\ \hline
 \end{tabular}
 \caption[Comparison between the baseline and the improved version of the CE on manually annotated data]{Comparison between the baseline and the improved version of the CE module using manually annotated sentences from the PUD treebank in two different language pairs, English-Italian and English-Swedish.}
 \label{tbase}
 \end{table}

As the table shows, for both language pairs, the raw number of distinct alignments extracted increases, which suggests an increase in recall. Most importantly, the percentage of correct alignments, our approximation for precision, increases significantly, and in particular by more than 20\% for English-Italian, the language pair most used during development. 
There are also significant, but easily explainable differences between the two language pairs considered: in general, results are better for English-Swedish due to the syntactical similarity between the two. \smallskip

\subsection{Evaluating specific criteria}
When it comes to the improved version, it is also interesting to look at criterion-specific statistics, summarized in Table \ref{treas}. \smallskip

One of the main things the results suggests is that, even though most alignments are still found by applying the original criterion (cf. Criterion \ref{udmatch} in Section \ref{improvements}), a combination of criteria \ref{udmatch} and \ref{poseq} is still able to detect many correspondences (more than 30\% of the total for both language pairs, if we combine regular and head alignments extracted in this way) while having higher (>90\%) precision. \smallskip

This data also give an empirical confirmation of the observation made in Section \ref{heads} about the problem that the baseline has when it comes to head alignment, indicating that POS-equivalence is especially important when extracting head alignments. 
The small size of the corpus cause Criterion \ref{known} to remain extremely marginal.
Cross-language pair differences are, in this case, minimal, confirming that the criteria used and their priority order are not specifically tailored for the language pair used for under-development experiments. \smallskip

\begin{table}[H]
 \small
 \centering
 \begin{tabular}{|l|l|l|l|l|}
 \hline
 \textbf{criteria}       & \multicolumn{2}{l|}{\textbf{alignments extracted}} & \multicolumn{2}{l|}{\textbf{correct (+ and =)}} \\ \hline
 \textbf{}       & \textbf{en-it}  & \textbf{en-sv} & \textbf{en-it}  & \textbf{en-sv}  \\ \hline
 \textbf{matching UD labels}      & 39.39\%  & 37.71\%  & 78.60\%  & 80.24\%  \\ \hline
 \textbf{POS-equivalence}       & 3\%  & 1.82\%  & 83.33\%  & 83.33\%  \\ \hline
 \textbf{known divergence}       & 2.08\%   & 1.36\%  & 52\%   & 55.55\%  \\ \hline
 \textbf{known alignment}       & 0\%  & 0\%  & -  & -  \\ \hline
 \textbf{\begin{tabular}[c]{@{}l@{}}matching UD labels\\ + POS-equivalence\end{tabular}}  & 20.95\%  & 24.71\%  & 93.03\%  & 94.76\%  \\ \hline
 \textbf{\begin{tabular}[c]{@{}l@{}}matching UD labels\\ + head alignment\end{tabular}}  & 21.78\%  & 19.61\%  & 67.81\%  & 76.74\%  \\ \hline
 \textbf{\begin{tabular}[c]{@{}l@{}}POS-equivalence\\ + head alignment\end{tabular}}  & 1.33\%   & 1.29\%  & 75\%   & 64.70\%  \\ \hline
 \textbf{\begin{tabular}[c]{@{}l@{}}matching UD labels\\ + known alignment\end{tabular}}  & 0.25\%   & 0.45\%  & 66.66\%  & 83.33\%  \\ \hline
 \textbf{\begin{tabular}[c]{@{}l@{}}matching UD labels\\ + POS-equivalence\\ + head alignment\end{tabular}} & 10.85\%  & 12.69\%  & 95.38\%  & 90.41\%  \\ \hline
 \end{tabular}
 \caption[Criterion-specific statistics]{Criterion-specific statistics. The leftmost column specifies a criterion or combination of criteria. The central column indicates the percentage of alignments extracted because of each set of criteria, and the rightmost one specifies how many of them were marked as \texttt{+} or \texttt{-}.}
 \label{treas}
 \end{table}

\subsection{Working with raw text}
Of course, it is also interesting to compare the performance of the CE module on manually annotated data with the results obtained on the automatically parsed course plans corpora, summarized in Table \ref{tcereal}.
% Please add the following required packages to your document preamble:
% \usepackage[normalem]{ulem}
% \useunder{\uline}{\ul}{}
\begin{table}[H]
 \small
 \centering 
 \begin{tabular}{|l|l|l|}
 \hline
 \textbf{}   & \textbf{DMI (en-it, 798 sentences)} & \textbf{CSE (en-sv, 498 sentences)} \\ \hline
 \textbf{distinct alignments} & 352  & 529  \\ \hline
 \textbf{correct (+ and =)} & 243 (69.03\%) & 368 (69.56\%) \\ \hline
 \textbf{correct and useful (+)} & 229 (65.05\%) & 351 (66.35\%) \\ \hline
 \end{tabular}
 \caption[Performance of CE on barely sentence-aligned data]{Performance of CE on barely sentence-aligned data, parsed automatically. Two distinct corpora of different size have been used for the two different language pairs, so the two columns are not to be compared with each other, but rather with Table \ref{tbase}.}
 \label{tcereal}
\end{table}

As it is to be expected, results are not as good for automatically parsed sentences as they are for manually annotated data. In particular, the percentage of correct alignments drops - but interestingly it is virtually the same for both language pairs. 
The difference between the two pairs seems instead to be in the total number of alignments found: despite English-Italian corpus being significantly larger, more concepts are extracted for the Swedish-Italian one. \smallskip

\subsection{Comparison with methods from Statistical MT}
Another kind of evaluation that can be performed before the aligned dependency subtrees are converted to GF ASTs to be fed to the last stage of the pipeline is to compare them with those obtained by means of application of existing statistical algorithms. Among the well known solutions mentioned in Section \ref{statmt}, we have chosen to focus on \texttt{fast\_align} \cite{fast} because of its ease of use and installation. \smallskip

A Haskell script converts the CoNNL-U files that are fed to \texttt{AlignTrees} into the input format required by \texttt{fast\_align}, so that the exact same tokenization is used. \smallskip

A second script converting its output into that used by \texttt{EvAlign} allows to reuse all of the evaluation infrastructure described in Section \ref{metrics}.
Finally, Table \ref{tfast} summarizes the results of a comparison between the improved CE module and \texttt{fast\_align} run with 10 iterations in EM training and symmetrizing the alignments found by running the program in both directions. \smallskip

Because \texttt{fast\_align} is a purely statistical approach, we trained it not only on the same 100-sentence subset of PUD used for evaluating \texttt{AlignTrees} against the baseline, but also on the full 1000-sentences dataset. In the latter experiment, the alignments obtained for the last 900 sentences were discarded before evaluating the results.
Due to the time required to manually annotate the numerous alignments extracted by \texttt{fast\_align}, this evaluation has been conducted only for the most challenging language pair, English-Italian. 
For a fairer comparison, since \texttt{fast\_align} works at the word level, only the one-to-one, one-to-many and many-to-one word alignments produced by the CE module were kept.

\begin{table}[H]
 \centering
 \small
 \begin{tabular}{|l|l|l|l|}
 \hline
  & \textbf{\texttt{AlignTrees}} & \textbf{\texttt{fast\_align} 100} & \textbf{\texttt{fast\_align} 1000}\\ \hline
  \textbf{distinct alignments} & 716 & 1440 & 1435 \\ \hline
  \textbf{correct} & 536 (74.86\%) & 410 (28.47\%) & 656 (45.71\%)\\ \hline
  \textbf{correct and useful} & 491 (68.57\%) & 371 (25.76\%) & 590 (41.11\%)\\ \hline
 \end{tabular}
 \caption[Comparison between the improved CE module and \texttt{fast\_align}]{Comparison between the improved CE module and \texttt{fast\_align} trained on the first 100 only (column 2) and on the whole PUD corpus (column 3).}
 \label{tfast}
\end{table}

As the table shows, since \texttt{fast\_align} tries to find a correspondent in the TL for each word in the SL, the number of alignments found by our CE program is roughly half of that of those found statistically. The percentage of correct alignments is, however, much higher for our system, even when the full PUD corpus is used in the training step of \texttt{fast\_align}. This shows that our approach is particularly well suited to smaller datasets and can even be used on individual sentence pairs, provided that they are analyzed correctly.

### 2021-02-09 final_report/include/CP.tex
moved relevant stuff to my own public repo

% CREATED BY DAVID FRISK, 2016
\chapter{Concept Propagation} \label{ch4}
After a set of concepts is known, it can be useful to be able to look for the corresponding concrete expressions in a new language. In this chapter, we describe our approach to this task and present an evaluation of the corresponding module.

\section{Method and implementation}
As mentioned in Section \ref{aligns}, Concept Propagation (CP) is the task of finding the concrete expressions corresponding to a set of known concepts, represented by a set of alignments and/or their abstract representations. Since our method for propagating concepts makes use of dependency tree alignments only, CP can take place before and, in general, independently from a potential conversion of the UD trees into GF trees. \smallskip

There are two scenarios in which CP can prove useful:\smallskip
\begin{enumerate}
    \item when working on a parallel text $A$ in more than two languages. In this case, CP can be used, after CE has been performed on a pair of languages $\langle L_1,L_2 \rangle$, to propagate the extracted alignments in a third version of the text in a new language $L_3$. In this case, the expectation is for the program to be able to propagate the vast majority\footnote{even though handling common translation divergences help, free translation involving radical differences in sentence structure can make it so that some concepts lack a clear counterpart in the text in $L_3$.} of the concepts in the set
    \item when working on two bilingual parallel texts $A$ and $B$ with one language in common, say $A$ in two languages $\langle L_1,L_2 \rangle$ and $B$ in two languages $\langle L_2, L_3 \rangle$. After performing CE on the first pair, CP can be used to look for the extracted concepts in the $L_2$ version of the second and, if they do appear in such text, find their counterparts in $L_3$. Of course, the number of concepts that can be propagated in this scenario can be significantly lower, especially if the two parallel texts do not belong to the same domain.
\end{enumerate}\smallskip

\begin{figure}[h]
    \centering
    \begin{subfigure}{.5\textwidth}
      \centering
      \includegraphics[width=.7\linewidth]{figure/scen1.png}
    \end{subfigure}%
    \begin{subfigure}{.5\textwidth}
        \centering
        \includegraphics[width=.7\linewidth]{figure/scen2.png}
    \end{subfigure}
    \caption[The two CP scenarios compared]{The two CP scenarios compared. On the left, Scenario 1, when CP is applied on three translations of the same text. On the right, Scenario 2, dealing with two distinct bilingual texts.}
    \label{scens}
\end{figure}

Our algorithm is meant to be as general as possible, and as such it is not optimized for the first scenario, but rather covers both without making any assumptions in terms of sentence alignment\footnote{a way to make the algorithm more efficient for the first scenario could in fact be to rely on the fact that the $n$-lingual parallel text is sentence-aligned, keep track of the sentence from which a particular concept has been extracted and only try to find a equivalent in the corresponding $L_3$ sentence.}. 
It works as follows: for each alignment obtained by performing CE on text $A$, we consider the dependency tree in language $L_2$. We then look for such tree among all subtrees the $L_2$ version of text $B$ is composed of. If a sentence in the $L_2$ version of text $B$ does contain $a_i^2$, we align such sentence to its $L_3$ counterpart using the same procedure as described in Section \ref{extrain} for CE. Finally, if multiple possible correspondences are found, the $L_3$ dependency subtree which is closest in depth to $a_i^2$ is selected. This is meant to avoid, for instance, mistakenly aligning a full sentence tree to a dependency tree extracted when aligning heads, which would otherwise be likely, as their heads are obviously both labelled \texttt{root}. In the next two sections, we discuss some important details of the algorithm, and in Section \ref{heads2} in particular will get to other issues related to head alignment.

\begin{figure}[H]
    \centering
    \lstinputlisting{pseudocode/Propagate.hs}
    \caption[The CP algorithm]{The CP algorithm. The concept to propagate is looked for among the SL subtree of each alignment obtained by comparing the sentences of the new corpus. If it is found, the alignment it belongs to is returned. If there are multiple possible alignments, the alignment whose TL side is closest in depth to the concept itself is returned.}
    \label{prop}
\end{figure}

\subsection{Ignoring details of UD trees}
One point that becomes of extreme importance when dealing with Scenario 2, i.e. when the text used for CP is not a translation of the one used for CE, is the way we look for subtrees in the $L_2$ version of text $B$. Since we work with two different $L_2$ texts, in fact, we cannot expect text $B$'s subtrees to match $L_2$ concepts extracted from $T_1$ exactly: as we saw in Section \ref{ce}, nodes of UD trees contain information, particularly about the position of each word and its head, that should be disregarded in this context.\smallskip

To make the CP algorithm more robust to minor parse errors, our decision was to consider two (sub)trees equivalent based on a comparison of just their shape and their \texttt{FORM}, \texttt{LEMMA}, \texttt{UPOS} and \texttt{DEPREL} fields. 
In this way, not only all optional fields (which might well differ or be blank for only one of the texts), but word position information is also ignored.

\subsection{Propagating head alignments} \label{heads2}
The propagation of head alignments is by far the most challenging aspect of CP. 
This is due to the fact that the head of a dependency tree $t$, as defined for the purposes of CE (cf. Section \ref{heads}), is not, strictly speaking, a subtree of $t$ (even though in a broader sense it is, of course, a part of $t$). When head alignments are extracted, in fact, their members are either: \smallskip

\begin{itemize}
    \item trees composed exclusively of the root of a dependency subtree
    \item trees composed of the root of a dependency subtree and some of its immediate subtrees. As discussed in Section \ref{heads}, this happens in cases such as when the root in the SL is a compound (written as a one-word) and its TL counterpart is a multiword.
\end{itemize} \smallskip

Not only, then, is it necessary to look for each concept to propagate among all the dependency subtrees of each SL sentence of the corpus CP is performed on, but for each subtree it is also necessary to check whether the concept could be a head of such subtree. This is implemented by means of a function \texttt{isHeadOf} specular to \texttt{alignHeads}. \smallskip

If this is the case, i.e. if the concept the program is trying to propagate is the head of a certain SL subtree, then the resulting alignment should be with the head of the TL counterpart of such subtree.

\section{Evaluation} \label{eval3}
In the following three sections, we describe both the CP experiments performed and their results. Clearly, for all the experiments, we make use of the evaluation script used for CE, which has been extended to also compute CP-specific statistics, such as the percentage of (successfully) propagated concepts.

\subsection{Preliminary testing} 
One way to check the CP module for obvious flaws before trying to assess its performance in the two scenarios described above is to try to re-obtain the alignments obtained with the CE module via propagation. This means running both programs on the exact same input treebanks as used for CE. It represents a special case of Scenario 1 in which the program is expected to be able to propagate all alignments, with no exceptions. \smallskip

The fact that many of the alignments obtained via propagation are the same as the ones extracted initially is generally a good indication that the program is working according to expectations. However, this has to be taken with a grain of salt: as discussed in the above, the program looks for one possible counterpart per concept only and is provided no indication about what sentence in particular it should look into, so that in presence of multiple valid alternative alignments the user should not expect it to identify the same correspondences as the extraction procedure, but rather just correct, potentially different correspondences.

\subsubsection{Experimental results} \label{eval4r}
This kind of pre-evaluation has been run on the first 100 sentences of the PUD treebanks in English and Italian, one of the corpora used for evaluating CE. Most importantly, the experiment confirms that the CP module is in fact able to find all the alignments extracted by CE. 
The results also show that the vast majority of the propagated alignments is identical to its extracted counterpart, and that most often, even when it is not, that does not imply that the propagated alignment is incorrect. Out of a total of 224 (18.69\%) incorrect propagated alignments, only 28 (12.5\%) introduce an alignment error which was not already present in their extracted counterparts, and there are even instances where propagated alignments are, conincidentally, correct even though their extracted counterparts are not.

\subsection{Scenario 1}
For evaluating how well CP works in the first scenario, the corpus we use is the same as for testing, but in three instead of two languages: the concept extracted by means of comparing the English to the Italian version are propagated to Swedish, using both English and Italian UD trees as the starting point. \smallskip

Considering that we are working with three versions of the same text, the expectation is for the program to be able to propagate the vast majority of the concepts with a precision similar to that obtained during the extraction stage, even though free translation, unhandled translation divergences and annotation inconsistencies can make it possible for some of them not to be found in the Swedish translation or not to be aligned correctly. \smallskip

As for when it comes for CE, comparing the results of CP for two different language pairs, English-Swedish and Italian-Swedish, can be interesting to see how much easier it is to propagate concepts in two more syntactically similar languages.

\subsubsection{Experimental results}
As Table \ref{tcp1} shows, the experimental results of this type of evaluation match the above expectations. First, the number of alignments the program is able to propagate is, for both language pairs, smaller than that obtained in the testing phase but still relatively large, while the amount of propagation errors increases marginally. Furthermore, the English-Swedish pair gives slightly better results than the Italian-Swedish pair, both in terms of the number of alignments the program is able to propagate and in terms of the amount of alignment errors. \smallskip

It is interesting to notice, however, that the amount of errors is significantly (> 5\%) inferior when concepts are propagated to Swedish starting from English, as opposed to Italian, dependency trees. This is most likely due to the syntactical features shared by the former two languages. This means that not all Italian-English extraction errors are propagated and it seems to suggest that, if we were to only use the trilingual alignments obtained by English-Italian extraction followed by English-Swedish propagation, some of the initial incorrectly extracted alignments would be automatically discarded.

\begin{table}[H]
    \centering
    \begin{tabular}{|l|l|l|}
    \hline
    \textbf{}              & \textbf{en-sv} & \textbf{it-sv} \\ \hline
    propagated             & 1019 (85.05\%) & 979 (84.64\%) \\ \hline
    tot. errors            & 133 (13.05\%)  & 187 (19.1\%)  \\ \hline
    CP-introduced          & 75 (56.39\%)   & 84 (44.91\%)   \\ \hline
    \end{tabular}
    \caption[Performance of CP Scenario 1 on manually annotated data]{Results of propagating the Italian-English alignments obtained via CE on the first 100 sentences of the PUD corpus to Swedish, using either the English or Italian subtrees as a starting point.}
    \label{tcp1}
    \end{table}

\subsection{Scenario 2}

The PUD corpus is composed of extracts of texts on a variety of subjects. As a consequence, performing an evaluation of CP in Scenario 2 using, for extraction and propagation, two different subsets of the corpus, would not help assess whether the module is well suited for the most relevant subcase of such scenario, i.e. when working with two texts belonging to the same domain. \smallskip

On top of evaluating the program on PUD data dividing the set of sentences used so far into two equally sized subsets, then, the two course plans corpora described in Section \ref{plans} have been also used to put the program to the test. \smallskip

These two small bilingual corpora allow us to perform two additional experiments: \smallskip
\begin{enumerate}
    \item to extract English-Swedish alignments from the CSE corpus and propagate them to Italian using the DMI corpus
    \item vice versa, to extract English-Italian alignments from the DMI corpus and propagate them to Swedish using the CSE corpus
\end{enumerate} \smallskip

\subsubsection{Experimental results}

\paragraph{Propagation with texts in different domains}
When it comes to the first set of experiments, where we use PUD data, being the subsets of selected sentences extremely small and having the correctness of most alignments been already assessed in previous experiments, it is easy to work with all possible pairs of languages. The results, summarized in Table \ref{tcp2pud}, show that, in general, only a small fraction of the extracted alignments are actually propagated in this case. As mentioned above, this is to be expected since different sentences belong to different domain. A closer look to the set of alignments returned by propagation, the vast majority of which has single words as members, shows that most of them are common function words, which, alone, are not always relevant for a GF-based MT pipeline. 
Common determiners such as articles, for instance, can be better handled by RGL grammars alone. 
Prepositions, on the other hand, are only useful with some information about the context in which they occur in, since it is customary for them to be used differently across different, even closely related languages. 
However, some common content words, like $\langle$\textit{always, sempre, alltid}$\rangle$, $\langle$\textit{new, nuovi, nya}$\rangle$ and $\langle$\textit{same, stesso, samma}$\rangle$ are also aligned correctly.

\begin{table}[H]
    \centering
    \scriptsize
    \begin{tabular}{|l|l|l|l|l|l|l|}
    \hline
    \textbf{}                     & \textbf{en-it-sv} & \textbf{it-en-sv} & \textbf{en-sv-it} & \textbf{sv-en-it} & \textbf{it-sv-en} & \textbf{sv-it-en} \\ \hline
    extracted   & 638               & 638               & 687               & 687               & 608               & 608               \\ \hline
    propagated                    & 92 (14.42\%)      & 92 (14.42\%)      & 98 (14.26\%)      & 84 (12.22\%)      & 101 (16.61\%)     & 87 (14.37\%)      \\ \hline
    tot. errors                   & 46 (50\%)         & 21 (22.82\%)      & 42 (42.85\%)      & 24 (28.57\%)      & 21 (20.79\%)      & 28 (32.18\%)      \\ \hline
    CP-introduced                 & 33 (71.73\%)      & 11 (52.38\%)      & 21 (50\%)         & 12 (50\%)         & 12 (57.14\%)      & 21 (75\%)         \\ \hline
    \end{tabular}
    \caption[Performance of CP Scenario 2 on manually annotated non-homogeneous data]{Results of CP for various language pairs using PUD (non-homogeneous) data. For each column, the first two languages are the ones used for extraction and the second two the ones used for propagation.}
    \label{tcp2pud}
    \end{table}

The table also shows, like the results obtained from Scenario 1 suggest too, that the order in which languages are used is not completely irrelevant: while the percentage of propagated alignments does not vary widely, the amount of errors is significantly lower when the propagation step involves the two syntactically more similar languages, English and Swedish. The fact that CP works better with English-Italian than with Italian-Swedish, on the other hand, may simply be a result of the fact that the latter language pair was never used for testing the program during its development, which might have caused some of the criteria to be not completely language pair-independent. \smallskip

Finally, the percentage of errors introduced by CP is significantly higher than that recorded for Scenario 1. One factor that contributes to this is the fact that function words, which are by far more numerous in the propagated alignments than content words, are often used differently in different languages and contexts. When it comes to the languages we are considering, this applies in particular to prepositions: because it is hard to know what the correct translation equivalent is without looking at the specific sentence, when CP is performed on a text different from the one used for CE, many preposition-to-preposition alignments have to be marked as incorrect even though they might be correct in some specific cases.

\paragraph{Propagation with texts within the same domains}

Quantitatively, the results of the last pair of experiments, performed on two different automatically parsed bilingual corpora composed of course plans of Computer Science programmes, are to some extent similar to those obtained using different texts in different domains (cf. Tables \ref{tcp2courses} and \ref{tcp2pud}). Only a tiny fraction of the extracted alignments is propagated and errors are numerous, as in the previous setting.

\begin{table}[H]
    \centering
    \begin{tabular}{|l|l|l|}
    \hline
                           & \textbf{sv-en-it} & \textbf{it-en-sv} \\ \hline
    extracted concepts     & 1950              & 1823              \\ \hline
    propagated             & 205 (10.51\%)     & 200 (10.97\%)     \\ \hline
    tot. errors            & 66 (32.19\%)      & 61 (30.5\%)       \\ \hline
    errors introduced by CP & 33 (50\%)        & 33 (54.09\%)      \\ \hline
    \end{tabular}
    \caption[Performance of CP Scenario 2 on automatically parsed homogeneous data]{Results of CP for using two different bilingual corpora belonging to the same domain of course plans. For each column, the first two languages are the ones used for extraction and the second two the ones used for propagation.}
    \label{tcp2courses}
    \end{table}

    More than the quantitative results, however, it is interesting to take a look at some of the propagated concepts. Because the text are similar both stylistically and, most importantly, in terms of content, the program is in this case also able to find trilingual correspondences between content words, sometimed specific of the domain. Examples of such alignments are $\langle$\textit{skills, capacità, färdigheter}$\rangle$, $\langle$\textit{functional, funzionale, funktionell}$\rangle$, $\langle$\textit{exam, prova, tentamen}$\rangle$, $\langle$\textit{course, corso, kurs}$\rangle$, $\langle$\textit{prerequisites, prerequisiti, behörigheter}$\rangle$, $\langle$\textit{lectures, lezioni, föreläsningar}$\rangle$ and $\langle$\textit{knowledge, conoscenza, kunskap}$\rangle$. \smallskip

    Sometimes, however, the meaning of a word is highly context-dependent and as a consequence, even if both the pairs of extracted and propagated alignments are correct, the corresponding trilingual alignment we could infer isn't (e.g. $\langle$\textit{learning, conoscere}$\rangle$ and $\langle$\textit{learning, inlärning}$\rangle$). \smallskip

    Finally, in some rare cases, even correspondences regarding longer expressions, such as $\langle$\textit{the aim of the course, l'obiettivo del corso, syftet med kursen}$\rangle$, can be found. In this case, the English-Italian correspondence is not especially useful, since the phrase could be translated almost word by word between the two languages, but the translation to Swedish is not literal, thus making the alignment useful for the purposes of MT.

### 2021-02-09 final_report/include/Conclusions.tex
moved relevant stuff to my own public repo

% CREATED BY DAVID FRISK, 2016
\chapter{Conclusions} \label{ch6}

The main objective of this thesis project was to develop a syntax-based Concept Alignment system and put it to the test in the context of domain-specific MT. The approach we propose basically consists in analyzing texts with a dependency parser, UDPipe, and comparing the resulting trees. For the MT evaluation, such dependency trees are converted to Abstract Syntax Trees, from which the rules of a multilingual generative phrase-structure grammar are derived. The framework we make use of when it comes to this grammar, GF, makes having a grammar sufficient to perform simple MT experiments.\smallskip

The tangible fruit of this work is a Haskell library, as well as a number of executables offering an easy to use and configure interface to perform both variants of CA, extraction and propagation, and a variety of kinds of evaluations. \smallskip

To assess the quality of the alignments obtained with such system, the extraction and propagation components were first tested without taking the target application context into account. \smallskip

When it comes to CE, the core part of this work, comparison against a pre-existing basic implementation shows significant improvements in terms of both the quality and the quantity of the resulting alignments. 
With respects to traditional statistical approaches, one of the most useful consequences of our syntax-based approach is that it makes it possible to identify correspondences between multiword, even discontinuous expressions. 
Another important advantage over standard word alignment techniques is that, even though it can still benefit from large amounts of data, our system works consistently well even when run on extremely small corpora. This is confirmed by one of our experiments, where we restrict it so to only extract one-to-many and many-to-one alignments and compare it with \texttt{fast\_align}. 
Furthermore, the alignment criteria our software makes use of are easy to modify, add and remove, paving the way to a variety of experiments. \smallskip

CP, a less studied variant of CA, has been explored in two different scenarios. On the one hand, it can be useful to generate large multilingual lexica if applied to more-than-bilingual parallel corpora. On the other, it can also be used to identify shared terminology between two bilingual corpora having one language in common. The first experiments show promising results, but suffer from the scarce availability of sentence-segmented large bilingual corpora. \smallskip

Finally, the CA module was adapted to the task of MT. The results of the first, small-scale experiments conducted in this sense highlight the strengths of GF in terms of NLG but also indicate the need for a more aggressive alignment selection strategy, as the numerous correct alignments are sometimes overshadowed by alternative incorrect ones.

\section*{Future work} \label{future}
These results, while encouraging, suggest that there is still much room for improvement and that work in this area can be carried on in many different directions even constaining ourselves to CA \textit{per se}, without focussing on its MT applications.

\subsection*{Further integration with statistical alignment techniques} 
With regards to the CA component itself, it is important to remember that both the quantity and the quality of the alignments are a direct consequence of those of the dependency trees they are obtained from. Apart from working on dependency parsers directly, a possibility is to experiment more with hybrid alignment approaches, integrating the results of statistical methods (which only depend on the quantity and quality of raw data rather than on that of their analyses) with those of our rule-based system. One way to do that has already been implemented, but has not been made use of in the final MT experiments due to the small size of the corpora involved.

\subsection*{Aligning verb phrases}
Early in this work we noticed how only aligning subtrees is not enough to identify all the concepts that are present in a pair of sentences. As a consequence, significant effort has been made in also aligning subtree heads correctly. While this increases the number of concepts the program is able to find, it does not cover all cases. For instance, aligning complete verb phrases (i.e. verbs together with their arguments) could prove extremely useful. Generalizing the approach used for head alignment could be a way of addressing the issue and would make it possible to use CA to construct ``rich'' lexical entries for verbs, whose lack, as discussed in Chapter \ref{ch3}, currently makes it hard to handle some types of common translations divergences.

\subsection*{Iterative CA}
The current implementations of CE and CP try to apply a set of priority-sorted criteria one after another to each pair of sentences in the corpus, allowing for later ranking of the alignments obtained based on the criteria their members match. Another possibility is to go through the entire corpus multiple times, for instance using only the strictest criteria first and only falling back to the less reliable ones at later iterations, until no more alignments are found.
The same hypothetical iterative algorithm can have the size of the alignments, and not (or not only) the criteria they have to satisfy, as a parameter.

\subsection*{Optimizing propagation for multilingual corpora}
CP was not the main focus of this work. As noted in Chapter \ref{ch4}, however, optimizing it for the simpler scenario in which it is applied to a more-than-2-lingual parallel text is straightforward and would improve both the quality of the results and, arguably, running time. Once this is done, generating multi-, as opposed to bi-, lingual domain-specific grammars will be more realistic, especially if \texttt{gf-ud}'s grammar generation functionality will also be further developed in parallel.

\subsection*{Generalizing CE to $n$ languages}
As an alternative to this optimization, CE could be generalized to an arbitrary number of languages. The expectation is for the resulting concepts to be of slightly larger size than those obtained with the current system, since comparing more trees simultaneously increases the possibility of encountering small divergences, this term being intended in its broad sense. This approach would probably help identifying longer idiomatic expression and could provide a better way to deal with multilingual corpora, even though, while intuitive, it does present some implementation challenges.

\subsection*{Stricter and language pair-specific criteria} 
When it comes to using the concepts as actual translation units, getting rid of incorrect alignments is crucial. While doing so manually is still more feasible than extracting them manually, it would be desirable to devise better alignment selection policies or stricter alignment criteria. One idea would is to tune the alignment criteria for the language pair at hand, both by removing the noisier ones for semantically related languages and by adding language-pair specific criteria. The criteria-specific statistics reported in Section \ref{eval2} can serve as a starting point for work in this direction.

### 2021-02-09 final_report/include/Introduction.tex
moved relevant stuff to my own public repo

% CREATED BY DAVID FRISK, 2016
\chapter{Introduction} \label{ch1}
% CA definition
Concept Alignment (CA) consists in finding semantical correspondences between parts of multilingual parallel texts.
% CA inuition
Such task, often preliminary to further linguistic analysis, is routinely performed by learners of classical languages when working with a translation alongside the original text. It can - and usually does happen simultaneously - at different degrees of abstraction, ranging from word to sentence level. 
While in some cases the student identifies new concepts thanks to language comparison itself (\textit{concept extraction}), there are instances where a set of concepts is already known and the objective becomes finding the corresponding expressions in a certain language (\textit{concept propagation}). \smallskip

% CA in translation 
Another task that involves CA is natural language translation: the human translator, almost subconsciously, first identifies concepts in the source text and only then looks for ways to render them in the target language. 
It is then natural to wonder whether it is possible to make use of CA in Machine Translation (MT) as well.
The hypothesis motivating this project, whose objective is to develop and test strategies for automating CA, is that it can serve as a step of a compositional MT pipeline. 
In such case, its users could easily be provided with a way to verify the correctness of the results by examining - at any level - the ways a concept is expressed in different languages instead of having to compare the entire text in the source language to its automatically translated counterparts.
From this perspective, CA can help develop a more easily interpretable, and therefore more reliable MT system. \smallskip

% motivation
While MT, and in particular Statistical Machine Translation (SMT), does make use of some alignment models, traditional solutions focus on aligning words or, at most, sequences of words, thus generally putting raw strings of text in different languages in relation with each other. This makes it hard to operate at multiple levels of abstraction simultaneously. 
Nevertheless, there are several reasons why a system able to do that is desirable. First of all, choosing the abstraction level to operate at is not trivial, to the point that one could argue that, even within the particular context of translation, correspondences at different levels of abstraction may be more or less useful according to the specific pair of sentences they occur in. 
As an example, let us take the following two English-Italian pairs of sentences: \smallskip

\begin{enumerate}
    \item ``\textit{May I have a piece of cake?}'' and ``\textit{Potrei avere un pezzo di torta?}''
    \item ``\textit{Finding useful correspondences isn't exactly a piece of cake}'' and ``\textit{Trovare corrispondenze utili non è proprio scontato}''
\end{enumerate} \smallskip

In the first case, correspondences on the word level, like ``\textit{piece}''-``\textit{pezzo}'' and ``\textit{cake}''-``\textit{torta}'' are definitely relevant, but in the second sentence pair ``\textit{piece of cake}'' is used idiomatically and it would probably be more useful to just put the whole phrase in relation with ``\textit{scontato}''. \smallskip

\begin{figure}[H]
    \centering
    \includegraphics[width=.4\linewidth]{figure/alignment.png}
    \caption[An English sentence aligned with its Italian translation]{An English sentence aligned with its Italian translation. In this case, we are looking for the smallest possible alignments, but it is also possible to find correspondences at a higher level of abstractions, such as ``\textit{useful correspondences}'' and ``\textit{corrispondenze utili}'', which puts two noun phrases in relation with each other.} \label{cake}
\end{figure}

In cases like this, a syntactical comparison between the two sentences can help, if not to extract meaningful correspondences only, which would be the case in the example above, where the complement of the copula ''\textit{is}'' (resp. ``\textit{è}'') is a multiword expression in English and a one-word in Italian, at least to obtain a series of alignments at all levels of abstraction from which to select the most relevant at a later stage\footnote{an example of a more difficult case could be the alternative translation ``Trovare corrispondenze utili non è proprio \textit{un gioco da ragazzi}'', roughly corresponding to ``\textit{a child's play}'' (literally ``\textit{a game (played) by children}''). In this case, trying to align the sentence based on a syntactical comparison would yield both ``\textit{piece of cake}''-``\textit{gioco da ragazzi}'' and the more questionable correspondences ``\textit{piece}''-``\textit{gioco}'' and ``\textit{of cake}-``\textit{da ragazzi}''.}. 
The idea is that, in rule-based MT, using aligned phrases (as opposed to aligned words) is often useful for translating idiomatic expressions correctly and, in general, beneficial in terms of fluency of the output sentences.\smallskip

Furthermore, taking syntax into account allows to easily deal with non-contiguous multiword expressions, another situation which is challenging to deal with by means of traditional statistical approaches. For instance, if we consider the sentence ``\textit{Without any linguistic knowledge, it is definitely hard}'' and its Italian translation ``\textit{Senza conoscenze linguistiche, è decisamente arduo}'', a syntactical analysis would make it possible to extract the correspondence ``\textit{is hard}''-``\textit{è arduo}'', even if in both cases the copula and its complement are separated by an adverb. \smallskip

% contribution (+ sw deliverables w technologies involved)
With this thesis, we propose a syntax-based approach to both of the above mentioned variants of CA - concept extraction and concept propagation - and put it to the test by integrating it in a prototype domain-specific MT system composed of a neural Universal Dependencies (UD) parser, a rule-based alignment module and a target language generation component based on Grammatical Framework (GF).

\section*{Structure of the thesis}
This work is structured as follows. 
Chapter \ref{ch2} gives a few basic definitions, including a more rigorous one of CA itself, as well as providing the necessary background and contextualizing our project by reviewing a few related works. 
Chapters \ref{ch3} and \ref{ch4} focus respectively on extraction and propagation, presenting both our approach to each task and the corresponding experimental results, while Chapter \ref{ch5} describes the experiment designed to produce evaluate how well our CA component performs in the context of MT. 
Finally, Chapter \ref{ch6} consists of a discussion of the overall results and proposes some ideas for future work.

### 2021-02-09 final_report/include/TT.tex
moved relevant stuff to my own public repo

% CREATED BY DAVID FRISK, 2016
\chapter{Theory and Technologies} \label{ch2} 
In this chapter, we first review the preliminary notions of Computational Linguistics relevant to this work. We then give a more rigorous definition of the problem at hand and present the basic idea behind our proposed automation approach, comparing it to the existing ones and mentioning the technologies involved.

\section{Preliminary notions}
This first section consists of an overview of the basic notions that are necessary for a full understanding of the approach we propose. First, we discuss the principle of compositionality. After that, we describe and compare the grammar formalisms involved in the project, focussing on the two playing the most prominent roles in this work and comparing them to other, related approaches.

\subsection{Semantic compositionality}
The principle of semantic compositionality states, in its most general form, that the meaning of a complex expression is determined solely by the meanings of its components and by the manner in which these components are combined \cite{frege}. \smallskip

A useful variant of this formulation, which refers to natural languages in particular, is the following: \smallskip

\begin{definition} \label{sc}
For every complex expression $e$ in a language $L$, the meaning of $e$ in $L$ is determined by the structure of $e$ and the meanings of the constituents of $e$ in $L$.
\end{definition}
\smallskip

Questions of structure and constituency are settled by the syntax of $L$, while the meanings of simple expressions are given by the lexical semantics of $L$, meaning that syntax and lexical semantics, together, are sufficient to determine the entire semantics of $L$ \cite{frege}. \smallskip

According to this definition, then, if $L$ is compositional, the meaning of an expression $e$ in $L$ cannot depend directly on the context in which $S$ is used in or on the intentions of the speaker who uses it, even though it might be the case that the meanings of the constituents of $e$ depend on the context or on the speaker's intentions, thus making the sentence indirectly depend on them \cite{semcom2}. \smallskip

While many artificial languages, such as programming languages, are compositional by construction, the general validity of this principle for natural languages is, despite many arguments in favor, under debate. In this work, we make the assumption that compositionality also holds for natural languages, or at least that it is a useful notion applying to most natural language sentences\footnote{in Section \ref{evalign}, however, we'll see that idiomatic expressions, often used as an arguments against compositionality, are also challenging to deal with in the context of grammar-based CA.}. \smallskip

In particular, we adapt the notion of semantic compositionality to translation. 
The intuitive idea is that the translation of a sentence is composed of translations of each part of the original one. Adjusting Definition \ref{sc}:

\begin{definition}
    For every complex expression $e$ in a source language $S$, the translation of $e$ in a target language $T$ is determined by the structure of $e$ and the translations of the constituents of $e$ in $T$.
\end{definition}

\subsection{Constituency grammars}
The idea of taking advantage of semantic compositionality is not new in the field of Computational Linguistics. Doing so traditionally involves using some kind of \textit{constituency grammar}. \smallskip

The notion of constituency or \textit{phrase structure} grammar, introduced by Noam Chomsky in \cite{chomsky}, is usually known to computer scientists due to the fact that a class of phrase structure grammars, namely that of \textit{Context-Free Grammars} (CFGs), that can be used to describe programming languages and plays an central role in compiler theory. 
As a consequence, we find giving a complete formal definition of the latter unnecessary and will restrict ourselves to an informal review of the aspects that are particularly relevant for the specific subject of this thesis. \smallskip

A constituency grammar consists essentially in a set of \textit{rewrite rules}, such as

\begin{lstlisting}[frame=none]
    S -> NP VP,
\end{lstlisting}

which we can read as ``$S$ can be \textit{rewritten as} (or \textit{substituted with}) a $\mathit{NP}$ followed by a $\mathit{VP}$'', or as ``a sentence $S$ is constituted by a noun phrase $\mathit{NP}$ and a verb phrase $\mathit{VP}$''. \smallskip

\begin{figure}[h]
    \centering
    \begin{subfigure}{.35\textwidth}
      \centering
      \includegraphics[width=.4\linewidth]{figure/parsetree.pdf}
    \end{subfigure}%
    \begin{subfigure}{.6\textwidth}
        \centering
        \begin{lstlisting}
            S -> NP VP
            NP -> CN
            VP -> V
            CN -> N
            N -> "grammars"
            V -> "rule"
        \end{lstlisting}
    \end{subfigure}
    \caption[Parse tree of the sentence ``\textit{grammars rule}'' and corresponding grammar fragment]{Parse tree of the simple sentence ``\textit{grammars rule}''. On the right, the fragment of the simple but linguistically informed grammar necessary to obtain such tree.}
    \label{pt}
\end{figure}

Replacing the term ``constituted'' with ``composed'' makes it easy to understand in which way phrase structure grammars are related to the aforementioned principle of compositionality. \smallskip

In the context of MT, the idea of exploiting the constituency relation and, as such, compositionality, was proposed by Curry in the early 1960s \cite{curry61} and first put in practice two decades later in the form of an experimental interlingual translation system, not coincidentally named Rosetta, which requires the definition of two distinct logically isomorphic \textit{Montague grammars}\footnote{Montague grammars are a semantics-oriented development of \textit{categorial grammars}, which are in turn a type of constituency grammars.} - one for the Source Language (SL), one for the Target Language (TL) - and constructs an intermediate representation based on such isomorphism \cite{rosetta}. \smallskip

\subsubsection{Synchronous grammars}
Among constituency grammars, other formalisms that have been employed in more recent MT systems are that of \textit{synchronous CFG}, originally developed for programming language compilation \cite{scfg} and adapted to natural language translation in several settings \cite{nlpscfg0,nlpscfg1,nlpscfg2,nlpscfg3}, and, later on, that of \textit{synchronous TAG} \cite{stag}, a variation of \textit{TAG} (Tree-Adjoining Grammar) \cite{tag}.
Both formalisms are meant to characterize correspondences between languages by having as their elements, instead of single rewrite rules, pairs of rules - one for the source and one for the target language. In a synchronous TAG, the constituents of a SL rule may be linked to their counterparts in the corresponding TL rule, and such links may be used to identify concepts in a syntax-based fashion. \smallskip

\subsubsection{Grammatical Framework} \label{gf}
% translation as compilation
As the repeated mention of programming language compilers in the above may suggest, it is possible to draw a very close parallel between compiler and MT pipelines. From this perspective, a natural language translation system can consist of: \smallskip

\begin{itemize}
    \item a \textit{frontend}, where the SL is analyzed or \textit{parsed} and whose output is an \textit{intermediate representation} most frequently in the form of an \textit{Abstract Syntax Tree} (AST)
    \item a \textit{backend}, where the AST is \textit{linearized} by means of application of a set of rules, a process usually referred to as \textit{code generation} in the context of programming languages and that we will refer to as \text{Target Language Generation} (TLG) or, when the context makes it clear that what the TL is, \textit{Natural Language Generation} (NLG). \smallskip
\end{itemize}
\begin{figure}[H]
    \centering
    \includegraphics[width=.8\linewidth]{figure/compiler.png}
    \caption[Steps of programming language compilation and, in the case of GF, natural language translation]{Steps of programming language compilation and, in the case of GF, natural language translation. In compilation, the input is usually a high-level programming language and the output is machine code (or code in a lower-level language). In the case of MT, the input and the output are strings in two different natural languages.} \label{compilerd}
\end{figure}

% intro to GF
Grammatical Framework (GF), being a grammar formalism and programming language designed with this parallel in mind, goes a step further in the direction of synchronous CFGs and TAGs by introducing a clear distinction between the \textit{abstract syntax}, whose aim is to capture the syntactic structures all natural languages taken into account have in common, and the \textit{concrete syntaxes}, specific to each individual language, consisting in their linearization rules \cite{gf2004} \cite{gfbook}. 
This makes it possible to deal with multiple languages by writing only one grammar, whose components are an abstract syntax and several concrete syntaxes, providing a solid basis for a system able to translate between any pair of for which a concrete syntax is available, in any direction. \smallskip

% GF as an interlingua-based MT system
With respects to this, a GF can be also seen as an \textit{interlingual} MT system where the intermediate representation or \textit{interlingua} is an AST. Interlingua-based systems have the advantage, in terms of efficiency, of making it unnecessary to build $n (n - 1)$ translation functions to cover all possible pairs of $n$ languages: having $2n$ is in general enough. In GF in particular, using a multilingual grammar makes it so that translation can be performed, for each pair of languages, in both directions, so that $n$ translation functions (or rahter $n$ concrete syntaxes and an abstract syntax) are already sufficient.\smallskip

% the RGL
An advantage of GF in particular, in addition, is the availability of a set of \textit{resource grammars} - the Resource Grammar Library (RGL) - for a variety of languages. A resource grammar is essentially a grammar that captures only the syntactic and morphological structures of a language, i.e. a grammar in the traditional sense of the term, and that can be easily extended to construct what is usually referred to as an \textit{application grammar}, i.e. a domain-specific grammar aiming to describe the language in a way that is not only syntactically correct, but also semantically accurate \cite{gfbook}. \smallskip

% GF and xMT
Extending the RGL in various ways, translation experiments with GF grammars have been conducted in the field of eXplainable Machine Translation (XMT), as the ASTs produced by SL analysis can serve both as to some extent automatically checkable certificates for the correctness of translation - by backlinearization to the SL - and as fully inspectable explanations aimed towards expert users \cite{rantaxmt}. \smallskip

% limitations
While the approach described above has proved successful for domain-specific translation, i.e. in cases where the natural languages in question can be reduced to \textit{Constrained Natural Languages} (CNLs), when it comes to open-domain translation, while target language generation remains effective, the results are negatively affected by the lack of robustness of the GF-based parsers available at the time of writing.

\subsection{Dependency grammars} \label{depg}
% DGs as opposed to PSGs
A class of grammar formalisms alternative to phrase structure, first proposed in \cite{dg}, is that of \textit{dependency grammars}, the main difference between the two being the relation they are based on. 
As opposed to constituency, \textit{dependency} is a word-to-word correspondence, meaning that words are simply put in relation with each other via directed links, called in fact \textit{dependencies}. Each dependency is, then, composed of two words: a \textit{head} and a \textit{dependent} that refers to it. For instance, if we try to look at syntactic dependencies in the sentence ``\textit{grammars rule}'', we could identify the verb ``\textit{rule}'' as the head and the noun ``\textit{children}'', its subject, as its dependent. \smallskip

% pros
Intuitively, this makes it so that dependency trees are simpler, which makes them an easier target for the frontend, possibly ML-based, of a MT system such as the one outlined in the above. 
Existing \textit{dependency parsers}, such as UDPipe \cite{udpipe1} and the Standford parser \cite{standford}, are often - but not always \cite{rasp} - neural pipelines trained on dependency treebanks. As a result, they are significantly more robust than their phrase structure counterparts.

% cons
On the other hand, there is currently no effective way to use these trees as the starting point for NLG: dependency grammars are an effective way to describe language, but are not \textit{generative}, unlike their phrase-structure counterparts\footnote{that does not mean, however, that dependency grammars have not been made use of in MT research, in conjunction with statistical techniques, in order to better capture grammatical generalizations \cite{treelet}.}. \smallskip

One idea is, then, to develop a hybrid system where the frontend is a dependency parser and the backend a grammaticality-preserving GF-based target language generation module. 
The connecting link between these two seemingly incompatible stages of our MT pipeline can be, as we will elaborate on in Section \ref{ourapproach}, a CA component able to identify matching dependency trees and generate the corresponding GF concrete and abstract syntax functions: concepts.

While there are several different dependency-based frameworks, the following section focuses on the one we find most well suited to this purpose, \textit{Universal Dependencies}.

\subsubsection{Universal Dependencies} \label{ud} 
Universal Dependencies (UD) is a framework for cross-linguistically consistent grammatical annotation. 
The UD project aims at developing parallel treebanks for many languages in order to support, among other things, the development of multilingual dependency parsers, such as the aforementioned UDPipe \cite{udpipe1}. 
In order to do so, it specifies an annotation scheme and a standard format for dependency trees consisting in an evolution of (universal) Stanford dependencies \cite{st1, st2}, Google universal part-of-speech tags \cite{upos}, and the Interset interlingua for morphosyntactic tagsets \cite{tagconv}. The basic idea behind such standard is to provide a set of Part-Of-Speech (POS) tags, dependency relations and annotation guidelines as language-agnostic as possible - so to facilitate its application in multilingual settings - while allowing language-specific extensions whenever necessary. \smallskip

In the following, we give a quick overview of the standard format UD uses to store dependency trees and of the aspects of the annotation scheme that are most relevant to the task at hand. Appendix \ref{a} provides a more comprehensive, but not completely exhaustive, description of all the POS tags and dependency labels mentioned in the examples appearing in this work based on that in \cite{compsyn}. The reader interested in a full specification of the UD annotation scheme may refer to the official UD documentation\footnote{\url{universaldependencies.org}}.

\paragraph{The CoNLL-U format} \label{conll}
The standard plain text format for dependency trees, CoNLL-U, is an extension of CoNNL-X \cite{conllx} which may contain \textit{comment lines}, starting with \texttt{\#}, \textit{blank lines} marking sentence boundaries and \textit{word lines} containing the annotation of a single token (generally a word, with some exceptions\footnote{examples of such exceptions are Italian contractions such as ``\textit{dello}'', which is generally divided into ``\textit{del}'' + ``\textit{lo}''.} in 10 tab-separated fields: \smallskip

\begin{enumerate}
    \item \texttt{ID}: integer\footnote{again, there are exceptions: in the case of Italian contractions and other multiword tokens, ranges may be used. Furthermore, empty nodes are characterized by decimal numbers between 0 and 1.} representing the position of the word in the sentence, starting from 1
    \item \texttt{FORM}: inflected from of the word
    \item \texttt{LEMMA}: lemmatized form of the word
    \item \texttt{UPOS}: universal POS tag
    \item \texttt{XPOS}: language-specific POS tag (optional)
    \item \texttt{FEATS}: list of (universal or language-specific) morphological features (optional)
    \item \texttt{HEAD}: head of the current word, i.e. either the value of the \texttt{ID} of another word in the same sentence or 0 in case the word at hand is a \texttt{root}
    \item \texttt{DEPREL}: universal dependency relation to the \texttt{HEAD} (\texttt{root} in case the word at hand is itself the \texttt{root})
    \item \texttt{DEPS}: enhanced dependency graph in the form of a list of \texttt{HEAD}-\texttt{DEPREL} pairs (optional)
    \item \texttt{MISC}: any other annotation (optional)
\end{enumerate} \smallskip

\begin{figure}[H]
    \centering
    \begin{subfigure}{.55\textwidth}
      \centering
      \footnotesize
        \begin{verbatim}
1	this	this	PRON	_	_	5	nsubj	_	_
2	is	be	AUX	_	_	5	cop	_	_
3	a	a	DET	_	_	5	det	_	_
4	dependency	dependency	NOUN	_	_	5	compound	_	_
5	tree	tree	NOUN	_	_	0	root	_	_
        \end{verbatim}
    \end{subfigure}%
    \begin{subfigure}{.45\textwidth}
        \centering
        \footnotesize
        \setlength{\unitlength}{0.21mm}
        \begin{picture}(371.0,130.0)
            \put(0.0,0.0){this}
            \put(46.0,0.0){is}
            \put(83.0,0.0){a}
            \put(120.0,0.0){dependency}
            \put(220.0,0.0){tree}
            \put(0.0,15.0){{\tiny PRON}}
            \put(46.0,15.0){{\tiny AUX}}
            \put(83.0,15.0){{\tiny DET}}
            \put(120.0,15.0){{\tiny NOUN}}
            \put(220.0,15.0){{\tiny NOUN}}
            \put(120.0,30.0){\oval(218.63636363636363,133.33333333333334)[t]}
            \put(10.681818181818187,35.0){\vector(0,-1){5.0}}
            \put(105.0,99.66666666666667){{\tiny nsubj}}
            \put(143.0,30.0){\oval(172.27586206896552,100.0)[t]}
            \put(56.86206896551724,35.0){\vector(0,-1){5.0}}
            \put(128.0,83.0){{\tiny cop}}
            \put(161.5,30.0){\oval(134.8102189781022,66.66666666666667)[t]}
            \put(94.0948905109489,35.0){\vector(0,-1){5.0}}
            \put(146.5,66.33333333333334){{\tiny det}}
            \put(180.0,30.0){\oval(97.0,33.333333333333336)[t]}
            \put(131.5,35.0){\vector(0,-1){5.0}}
            \put(165.0,49.66666666666667){{\tiny compound}}
            \put(235.0,130.0){\vector(0,-1){100.0}}
            \put(240.0,120.0){{\tiny root}}
        \end{picture}
    \end{subfigure}
    \caption[A dependency tree in CoNNL-U format alongside its graphical representation]{A dependency tree in CoNNL-U format alongside its graphical representation. Optional fields in the CoNLL-U file are left blank.}
    \label{deptree}
\end{figure}

\paragraph{Universal POS tags} \label{upos}
Universal POS tags mark the core part-of-speech categories, such as nouns, verbs, pronouns and determiners. An important - as we will see in Section \ref{criteria} - distinction we can make based on POS tags is that between \textit{content} or \textit{open class} words, i.e. words that contribute to the meaning of the sentences they appear in (nouns, lexical verbs, adjectives, adverbs and interjections), and \textit{function} or \textit{closed class} words, like pronouns and determiners, so called because they do not readily accept new members\footnote{one exception is the recent introduction of the gender-neutral pronoun ``\textit{hen}'' in Swedish.}. \smallskip

While the majority of Universal POS tags correspond to the grammatical categories of traditional grammars, the UD annotation scheme does have its peculiarities. Most importantly for the following discussion, verbs are divided into lexical verbs, tagged \texttt{VERB}, and auxiliaries, tagged \texttt{AUX}, thus providing an easy way to distinguish between content verbs and function verbs. A more systematic description of \texttt{UPOS} tags is given in \ref{a_pos}.

\paragraph{Universal dependency relations}
%subtypes
Universal dependency relations, largely based on \cite{st2}, represent, as mentioned in the above, syntactic dependencies between individual pairs of words occurring in the same sentence. In particular, according to the CoNLL-U standard (cf. Section \ref{conll}), each word is assigned a \textit{dependency label} indicating in which way it is liked to its \texttt{HEAD}. \smallskip

In this sense, the only exceptional case, which will be the starting point for our short overview of dependency relations, is that of the \texttt{root} label, generally assigned to the main verb of a sentence, ignoring any auxiliaries. For instance, ``\textit{smoked}'' is the root of the sentence \smallskip

\begin{example} \label{katia1}
    ``Katia has just \underline{smoked} a cigarette''
\end{example} \smallskip

In cases such as the following, where the main verb is a copula, the root is its complement and the verb is linked to it with the label \texttt{cop}. \smallskip

\begin{example} \label{katia2}
    ``Katia is a \underline{psychologist}''
\end{example} \smallskip

 If a sentence contains is no verbs at all, there is no fixed rule excepts that in order to avoid unnecessary discrepancies between languages, the root should be a content word. \smallskip

\begin{figure}[H]
   \centering
   \scriptsize
   \begin{subfigure}{.5\textwidth}
        \centering
        \setlength{\unitlength}{0.24mm}
        \begin{picture}(380.0,110.0)
            \put(0.0,0.0){Katia}
            \put(55.0,0.0){has}
            \put(92.0,0.0){just}
            \put(138.0,0.0){smoked}
            \put(202.0,0.0){a}
            \put(239.0,0.0){cigarette}
            \put(0.0,15.0){{\tiny PROPN}}
            \put(55.0,15.0){{\tiny AUX}}
            \put(92.0,15.0){{\tiny ADV}}
            \put(138.0,15.0){{\tiny VERB}}
            \put(202.0,15.0){{\tiny DET}}
            \put(239.0,15.0){{\tiny NOUN}}
            \put(79.0,30.0){\oval(135.82608695652175,100.0)[t]}
            \put(11.086956521739125,35.0){\vector(0,-1){5.0}}
            \put(64.0,83.0){{\tiny nsubj}}
            \put(106.5,30.0){\oval(79.3855421686747,66.66666666666667)[t]}
            \put(66.80722891566265,35.0){\vector(0,-1){5.0}}
            \put(91.5,66.33333333333334){{\tiny aux}}
            \put(125.0,30.0){\oval(39.47826086956522,33.333333333333336)[t]}
            \put(105.26086956521739,35.0){\vector(0,-1){5.0}}
            \put(110.0,49.66666666666667){{\tiny advmod}}
            \put(230.5,30.0){\oval(28.89189189189189,33.333333333333336)[t]}
            \put(216.05405405405406,35.0){\vector(0,-1){5.0}}
            \put(215.5,49.66666666666667){{\tiny det}}
            \put(208.5,30.0){\oval(98.02970297029702,66.66666666666667)[t]}
            \put(257.5148514851485,35.0){\vector(0,-1){5.0}}
            \put(193.5,66.33333333333334){{\tiny obj}}
            \put(153.0,110.0){\vector(0,-1){80.0}}
            \put(158.0,100.0){{\tiny root}}
          \end{picture}
   \end{subfigure}%
   \begin{subfigure}{.5\textwidth}
        \centering
        \setlength{\unitlength}{0.20mm}
        \begin{picture}(474.0,150.0)
            \put(0.0,0.0){Katia}
            \put(55.0,0.0){has}
            \put(92.0,0.0){worked}
            \put(156.0,0.0){in}
            \put(193.0,0.0){a}
            \put(230.0,0.0){big}
            \put(267.0,0.0){school}
            \put(331.0,0.0){library}
            \put(0.0,15.0){{\tiny PROPN}}
            \put(55.0,15.0){{\tiny AUX}}
            \put(92.0,15.0){{\tiny VERB}}
            \put(156.0,15.0){{\tiny ADP}}
            \put(193.0,15.0){{\tiny DET}}
            \put(230.0,15.0){{\tiny ADJ}}
            \put(267.0,15.0){{\tiny NOUN}}
            \put(331.0,15.0){{\tiny NOUN}}
            \put(56.0,30.0){\oval(88.73913043478261,66.66666666666667)[t]}
            \put(11.630434782608695,35.0){\vector(0,-1){5.0}}
            \put(41.0,66.33333333333334){{\tiny nsubj}}
            \put(83.5,30.0){\oval(28.89189189189189,33.333333333333336)[t]}
            \put(69.05405405405405,35.0){\vector(0,-1){5.0}}
            \put(68.5,49.66666666666667){{\tiny aux}}
            \put(253.5,30.0){\oval(173.28571428571428,133.33333333333334)[t]}
            \put(166.85714285714286,35.0){\vector(0,-1){5.0}}
            \put(238.5,99.66666666666667){{\tiny case}}
            \put(272.0,30.0){\oval(135.82608695652175,100.0)[t]}
            \put(204.08695652173913,35.0){\vector(0,-1){5.0}}
            \put(257.0,83.0){{\tiny det}}
            \put(290.5,30.0){\oval(98.02970297029702,66.66666666666667)[t]}
            \put(241.4851485148515,35.0){\vector(0,-1){5.0}}
            \put(275.5,66.33333333333334){{\tiny amod}}
            \put(309.0,30.0){\oval(59.3125,33.333333333333336)[t]}
            \put(279.34375,35.0){\vector(0,-1){5.0}}
            \put(294.0,49.66666666666667){{\tiny compound}}
            \put(231.5,30.0){\oval(237.744769874477,166.66666666666666)[t]}
            \put(350.3723849372385,35.0){\vector(0,-1){5.0}}
            \put(216.5,116.33333333333333){{\tiny obl}}
            \put(107.0,150.0){\vector(0,-1){120.0}}
            \put(112.0,140.0){{\tiny root}}
          \end{picture}
   \end{subfigure}
   \begin{subfigure}{.4\textwidth}
        \centering
        \setlength{\unitlength}{0.2mm}
        \begin{picture}(277.0,110.0)
            \put(0.0,0.0){Katia}
            \put(55.0,0.0){is}
            \put(92.0,0.0){a}
            \put(129.0,0.0){psychologist}
            \put(0.0,15.0){{\tiny PROPN}}
            \put(55.0,15.0){{\tiny AUX}}
            \put(92.0,15.0){{\tiny DET}}
            \put(129.0,15.0){{\tiny NOUN}}
            \put(74.5,30.0){\oval(126.67441860465117,100.0)[t]}
            \put(11.162790697674417,35.0){\vector(0,-1){5.0}}
            \put(59.5,83.0){{\tiny nsubj}}
            \put(102.0,30.0){\oval(69.94594594594595,66.66666666666667)[t]}
            \put(67.02702702702703,35.0){\vector(0,-1){5.0}}
            \put(87.0,66.33333333333334){{\tiny cop}}
            \put(120.5,30.0){\oval(28.89189189189189,33.333333333333336)[t]}
            \put(106.05405405405405,35.0){\vector(0,-1){5.0}}
            \put(105.5,49.66666666666667){{\tiny det}}
            \put(144.0,110.0){\vector(0,-1){80.0}}
            \put(149.0,100.0){{\tiny root}}
          \end{picture}
   \end{subfigure}%
   \begin{subfigure}{.6\textwidth}
        \centering
        \setlength{\unitlength}{0.2mm}
        \begin{picture}(530.0,110.0)
            \put(0.0,0.0){Katia}
            \put(55.0,0.0){showed}
            \put(119.0,0.0){me}
            \put(165.0,0.0){a}
            \put(202.0,0.0){book}
            \put(248.0,0.0){about}
            \put(303.0,0.0){the}
            \put(340.0,0.0){Stone}
            \put(395.0,0.0){Age}
            \put(0.0,15.0){{\tiny PROPN}}
            \put(55.0,15.0){{\tiny VERB}}
            \put(119.0,15.0){{\tiny PRON}}
            \put(165.0,15.0){{\tiny DET}}
            \put(202.0,15.0){{\tiny NOUN}}
            \put(248.0,15.0){{\tiny ADP}}
            \put(303.0,15.0){{\tiny DET}}
            \put(340.0,15.0){{\tiny PROPN}}
            \put(395.0,15.0){{\tiny PROPN}}
            \put(37.5,30.0){\oval(49.54545454545455,33.333333333333336)[t]}
            \put(12.727272727272727,35.0){\vector(0,-1){5.0}}
            \put(22.5,49.66666666666667){{\tiny nsubj}}
            \put(107.0,30.0){\oval(59.3125,33.333333333333336)[t]}
            \put(136.65625,35.0){\vector(0,-1){5.0}}
            \put(92.0,49.66666666666667){{\tiny iobj}}
            \put(193.5,30.0){\oval(28.89189189189189,33.333333333333336)[t]}
            \put(179.05405405405406,35.0){\vector(0,-1){5.0}}
            \put(178.5,49.66666666666667){{\tiny det}}
            \put(148.5,30.0){\oval(144.9591836734694,66.66666666666667)[t]}
            \put(220.9795918367347,35.0){\vector(0,-1){5.0}}
            \put(133.5,66.33333333333334){{\tiny obj}}
            \put(304.0,30.0){\oval(88.73913043478261,66.66666666666667)[t]}
            \put(259.6304347826087,35.0){\vector(0,-1){5.0}}
            \put(289.0,66.33333333333334){{\tiny case}}
            \put(331.5,30.0){\oval(28.89189189189189,33.333333333333336)[t]}
            \put(317.05405405405406,35.0){\vector(0,-1){5.0}}
            \put(316.5,49.66666666666667){{\tiny det}}
            \put(217.5,30.0){\oval(283.94736842105266,100.0)[t]}
            \put(359.47368421052636,35.0){\vector(0,-1){5.0}}
            \put(202.5,83.0){{\tiny obl}}
            \put(387.5,30.0){\oval(49.54545454545455,33.333333333333336)[t]}
            \put(412.27272727272725,35.0){\vector(0,-1){5.0}}
            \put(372.5,49.66666666666667){{\tiny flat}}
            \put(70.0,110.0){\vector(0,-1){80.0}}
            \put(75.0,100.0){{\tiny root}}
          \end{picture}
    \end{subfigure}

   \caption[UD trees showing the common dependency relations that are discussed in Chapter \ref{ch3}]{UD trees showing the common dependency relations that are discussed in Chapter \ref{ch3}. The three on the left correspond to examples \ref{katia1} and \ref{katia2}.}
    \label{katia}
\end{figure}

Other dependency labels commonly found in simple clauses and that will be dwelt on in Chapter \ref{ch3}, all exemplified in Figure \ref{katia}, are: \smallskip
\begin{itemize}
    \item \texttt{nsubj}, marking the link between a noun, proper noun, pronoun or numeral to the \texttt{root} of a sentence or, more in general, to the head of a clause
    \item \texttt{aux}, marking auxiliary verbs other than the copula
    \item \texttt{obj}, \texttt{iobj} and \texttt{obl} marking the link between a verb and its object, indirect object and other complements respectively
    \item \texttt{advmod}, \texttt{amod}, \texttt{nummod} and \texttt{nmod}, marking links between modifiers and the nouns or verbs they refer to
    \item \texttt{flat}, indicating a flat multiword expression, and \texttt{compound}, appearing in compound nouns whenever they are written as two or more separate words.
\end{itemize} \smallskip

A more detailed description of all the UD labels mentioned in this work is given in \ref{a_lab}. 
However, because the topic will arise in Chapter \ref{ch3}, it is also worth mentioning that CoNNL-U trees can present UD labels followed by a \textit{subtype}, separated by the label itself by a semicolon and used to indicate grammatical relations that are specific to one language or a small group of related languages. A subtype that is commonly used in English is, for instance, \texttt{pass}, added to both the (clausal or nominal) syntactic subject and \texttt{aux} of a sentence in passive voice.

\begin{figure}[H]
    \centering
    \begin{subfigure}{.5\textwidth}
        \centering
        \setlength{\unitlength}{0.25mm}
\begin{picture}(259.0,90.0)
  \put(0.0,0.0){Katia}
  \put(55.0,0.0){reads}
  \put(110.0,0.0){a}
  \put(147.0,0.0){magazine}
  \put(0.0,15.0){{\tiny PROPN}}
  \put(55.0,15.0){{\tiny VERB}}
  \put(110.0,15.0){{\tiny DET}}
  \put(147.0,15.0){{\tiny NOUN}}
  \put(37.5,30.0){\oval(49.54545454545455,33.333333333333336)[t]}
  \put(12.727272727272727,35.0){\vector(0,-1){5.0}}
  \put(22.5,49.66666666666667){{\tiny nsubj}}
  \put(138.5,30.0){\oval(28.89189189189189,33.333333333333336)[t]}
  \put(124.05405405405405,35.0){\vector(0,-1){5.0}}
  \put(123.5,49.66666666666667){{\tiny det}}
  \put(121.0,30.0){\oval(88.73913043478261,66.66666666666667)[t]}
  \put(165.3695652173913,35.0){\vector(0,-1){5.0}}
  \put(106.0,66.33333333333334){{\tiny obj}}
  \put(70.0,90.0){\vector(0,-1){60.0}}
  \put(75.0,80.0){{\tiny root}}
\end{picture}
    \end{subfigure}%
    \begin{subfigure}{.5\textwidth}
        \centering
        \setlength{\unitlength}{0.25mm}
        \begin{picture}(344.0,90.0)
            \put(0.0,0.0){A}
            \put(37.0,0.0){magazine}
            \put(119.0,0.0){is}
            \put(156.0,0.0){read}
            \put(202.0,0.0){by}
            \put(239.0,0.0){Katia}
            \put(0.0,15.0){{\tiny DET}}
            \put(37.0,15.0){{\tiny NOUN}}
            \put(119.0,15.0){{\tiny AUX}}
            \put(156.0,15.0){{\tiny VERB}}
            \put(202.0,15.0){{\tiny ADP}}
            \put(239.0,15.0){{\tiny PROPN}}
            \put(28.5,30.0){\oval(28.89189189189189,33.333333333333336)[t]}
            \put(14.054054054054054,35.0){\vector(0,-1){5.0}}
            \put(13.5,49.66666666666667){{\tiny det}}
            \put(106.5,30.0){\oval(116.47899159663865,66.66666666666667)[t]}
            \put(48.260504201680675,35.0){\vector(0,-1){5.0}}
            \put(91.5,66.33333333333334){{\tiny nsubj:pass}}
            \put(147.5,30.0){\oval(28.89189189189189,33.333333333333336)[t]}
            \put(133.05405405405406,35.0){\vector(0,-1){5.0}}
            \put(132.5,49.66666666666667){{\tiny aux:pass}}
            \put(230.5,30.0){\oval(28.89189189189189,33.333333333333336)[t]}
            \put(216.05405405405406,35.0){\vector(0,-1){5.0}}
            \put(215.5,49.66666666666667){{\tiny case}}
            \put(217.5,30.0){\oval(79.3855421686747,66.66666666666667)[t]}
            \put(257.1927710843373,35.0){\vector(0,-1){5.0}}
            \put(202.5,66.33333333333334){{\tiny obl}}
            \put(171.0,90.0){\vector(0,-1){60.0}}
            \put(176.0,80.0){{\tiny root}}
          \end{picture}
    \end{subfigure}
    \caption[The UD trees of an example active sentence and its passive counterpart]{The UD trees of an example active sentence and its passive counterpart.}
    \label{actpass}
\end{figure}

\paragraph{Relation to GF} \label{udrelgf}
The complementarity of constituency and dependency grammars and the multilingual nature of UD make it interesting to use in conjunction with GF, and experiments aimed at exploiting their similarities have already been performed \cite{gfud, udgf}. 
In the case of hybrid MT pipelines as the one sketched in Section \ref{depg}, UD trees need to be at some point converted into GF ASTs, albeit with the disadvantage that, unlike its reverse, the UD-to-GF conversion is a non-deterministic search problem.
An algorithm for conversion in this direction is presented in \cite{udgf}, where \texttt{gf-ud}, a program for converting GF ASTs into UD trees initially presented in \cite{gfud}, is extended to also be able to do the reverse.

\begin{figure}[H]
    \centering
    \begin{subfigure}{.7\textwidth}
        \centering
        \includegraphics[width=.4\linewidth]{figure/different.pdf}
    \end{subfigure}%
    \begin{subfigure}{.3\textwidth}
        \centering
        \setlength{\unitlength}{0.25mm}
        \begin{picture}(194.0,90.0)
          \put(0.0,0.0){they}
          \put(46.0,0.0){are}
          \put(83.0,0.0){different}
          \put(0.0,15.0){{\tiny PRON}}
          \put(46.0,15.0){{\tiny AUX}}
          \put(83.0,15.0){{\tiny ADJ}}
          \put(51.5,30.0){\oval(79.3855421686747,66.66666666666667)[t]}
          \put(11.807228915662648,35.0){\vector(0,-1){5.0}}
          \put(36.5,66.33333333333334){{\tiny nsubj}}
          \put(74.5,30.0){\oval(28.89189189189189,33.333333333333336)[t]}
          \put(60.054054054054056,35.0){\vector(0,-1){5.0}}
          \put(59.5,49.66666666666667){{\tiny cop}}
          \put(98.0,90.0){\vector(0,-1){60.0}}
          \put(103.0,80.0){{\tiny root}}
        \end{picture}
    \end{subfigure}
    \caption[A GF AST and its UD counterpart]{A GF AST and its UD counterpart. Note how the copula is implicit in the GF AST and subordinate to its complement in UD.}
    \label{parsetree}
\end{figure}

\section{Concept Alignment}
In this section, we aim to give an exhaustive description of CA and of the subtasks it consists of. In order to do that, we deem it necessary to start by giving a definition of what we commonly refer to as \textit{concepts}.

\subsection{Concepts} \label{concepts}
Intuitively, concepts are the components of meaning, and therefore, in a multilingual context, the units of translation. 
If we assume the principle of compositionality to be valid and apply it to translation, these meaning components are the common denominator between an expression in the source language and its translation, which can be generated using them as the starting point. \smallskip

From the compiler-like perspective described in Section \ref{gf}, this means that concepts are what the abstract syntax should represent, i.e. that they \textit{are} the ``interlingua'' of the MT system. 
This raises the question of how to obtain the abstract syntax needed for GF-based MT starting from a corpus of parallel texts, and it is at this point that language comparison and, as such, CA, come into play.

\subsection{Alignments} \label{aligns}
In terms of parallel text analysis, an \textit{alignment} consists in a pair of semantically equivalent (i.e., in this context, sharing the same abstract syntax) concrete expressions, one in the source and one in the target language. 
Even though, unless otherwise specified, we will keep referring to \textit{language pairs}, with a source and a target language, it is easy to see how this can be generalized to a more-than-bilingual, multidirectional definition: the pair just mentioned simply becomes a tuple of equivalent expressions in different languages. Formally, \smallskip

\begin{definition} \label{algndef}
    An $n$-lingual \textit{alignment} is an $n$-uple $\langle e_1,...,e_n \rangle$ of semantically equivalent expressions, where each expressions $e_i$ is in a different language $L_i$.
\end{definition} \smallskip

While this general definition does not specify it, expressions are not necessarily represented as strings, but can rather be defined as tree-like structures, and in the present case as dependency trees that can later be replaced by GF ASTs. Matching ASTs can finally be used to generate the rules of a multilingual GF grammar, as we will discuss in Chapter \ref{ch5}. \smallskip

As mentioned in the introduction, the task of aligning concepts comes in two variants:
\begin{enumerate}
    \item \textbf{\textit{Concept Extraction}} (CE), which consists in identifying new concepts via linguistic comparison and whose output is a set of bilingual alignments
    \item \textbf{\textit{Concept Propagation}} (CP), i.e. the task of finding the concrete expressions corresponding to a set of known concepts in a particular language. Here, by concepts we do not mean necessarily ASTs, as a suitable approach is to propagate UD tree alignments before obtaining the corresponding abstract representation.
\end{enumerate} \smallskip

If, as in the case of this project, the aim is to develop a multilingual MT system, these two tasks can be seen as two potentially subsequent steps. A first objective, in fact, can be to extract a set of concepts by comparing two translations of the same text. Once these concepts are in adequate number and of sufficient quality, it becomes possible to provide support for additional languages by simply looking for the concrete expressions that, in the new language, correspond to each of the previously gathered concepts. In other situations, clearly, CE alone can be enough or, if a set of concepts is already known, CP can be applied independently from CE to find their equivalents in a new language.

\subsection{Approaches to automation} 
Some form of manual, semi- or fully automated CA is in a sense at the heart of all traditional, even very early, MT systems. 
In the following section, we review some standard approaches. After discussing their limitations, we conclude the chapter with an overview of the grammar-based approach proposed in this thesis.

\subsubsection{Existing methods} \label{statmt}  
In the simplest case, \textit{word alignment} consists in finding pairs of individual words that translate to each other and store them in a dictionary. 
The standard automatic approaches to this task are statistical, and among them the five IBM Models \cite{ibm} and their numerous variations stand out. 
The IBM Models are a sequence of increasingly complex alignment models \cite{bitext}: \smallskip

\begin{itemize}
    \item Model 1 is based exclusively on lexical translation probabilities
    \item Model 2 takes word order (i.e. the absolute word positions) into account 
    \item Model 3 takes even a \textit{fertility} parameter into account (fertility being the number of target words that can be generated from a given source word)
    \item Models 4 and 5 takes the \textit{context} (i.e. the relative position of the words) in which each word occurs into account in a broader sense
\end{itemize}

The GIZA++ toolkit \cite{gizapp}, an open source implementation of the IBM models based on the initial GIZA package \cite{giza}, is widely used, for instance in SMT systems such as Moses \cite{moses}. \smallskip

As mentioned in the introduction, however, alignment can be performed at different levels of abstraction: to bring this to the extreme, there are contexts in which we might want to align full sentences, or even whole documents. While the latter two levels of abstraction are not necessarily relevant to MT, only aligning individual words is also, as discussed in the Introduction, not always the best option. For this reason, word alignment is often generalized to \textit{phrase alignment}. In systems like GIZA++, whose output is a list of pairs of word positions indicating which words in the SL string are to be aligned with which words in the TL string (commonly referred to as the \textit{pharaoh format}), extracting phrase alignments by combining the indices is realtively straightforward. \smallskip

\begin{figure}[h]
    \centering
    \begin{subfigure}{.55\textwidth}
      \centering
      \footnotesize
      \begin{itemize}
          \item[SL] ``\textit{Word alignment with standard statistical tools}''
          \item[TL] ``\textit{Allineamento parola per parola con strumenti statistici standar}''
      \end{itemize}
    \end{subfigure}%
    \begin{subfigure}{.45\textwidth}
        \centering
        \footnotesize
        \setlength{\unitlength}{0.21mm}
        \begin{verbatim}
    0-1 0-2 0-3 1-0 2-4 3-7 4-6 5-5
        \end{verbatim}
    \end{subfigure}
    \footnotesize
    \smallskip
    \smallskip
    \begin{itemize}
        \item \texttt{0-1 0-2 0-3}: $\langle$\textit{``word'', ``parola per parola''}$\rangle$
        \item \texttt{1-0}: $\langle$\textit{``alignment'', ``allineamento''}$\rangle$
        \item \texttt{2-4}: $\langle$\textit{``with'', ``con''}$\rangle$
        \item \texttt{3-7}: $\langle$\textit{``standard'', ``standard''}$\rangle$
        \item \texttt{4-6}: $\langle$\textit{``statistical'', ``statistici''}$\rangle$
        \item \texttt{5-5}: $\langle$\textit{``tools'', ``strumenti''}$\rangle$
        \item \texttt{0-1 0-2 0-3 1-0}: $\langle$\textit{``word alignment'', ``allineamento parola per parola''}$\rangle$
        \item ...
    \end{itemize}
    ...
    \caption[The pharaoh format output of optimal word alignment on a pair of Italian-English sentences and some derived phrase alignments]{The pharaoh format output of optimal word alignment on a pair of Italian-English sentences and, below, some of the phrase alignments that can be derived from it.}
    \label{wordalgn}
\end{figure}

However, it must be noted that, in the context of statistical MT, the term ``phrase'' is not to be intended in its grammatical sense, but can refer to any sequence of - typically contiguous - words and in any case, both for word and phrase alignment, the relation that is established with these statistical methods is merely between strings in different languages: there is no intermediate representation capturing the concepts themselves. \smallskip

The parallel between MT and compiler pipelines proposed in Section \ref{gf} suggests instead that a more flexible approach, applicable at any level of abstraction, could be grammar-based. 
This is particularly useful in cases, not at all uncommon, where the minimal translation units are, in one or both the source and the target language, multiword - potentially discontinuous - expressions or even more complex constructions that phrase alignment is not able to handle. \smallskip

Over time, various constituency grammar-based approaches have been proposed \cite{t2t1, t2t2}. These approaches, which make use of either parallel - manually or automatically annotated - treebanks, can be referred to as \textit{tree-to-tree alignment} methods \cite{bitext}. While theoretically appealing, however, their use in data-driven MT is limited, since they tend to suffer not only from the scarce availability of large-scale manually annotated corpora and the lack of robustness of the existing parsers, but also from the fact that, in general, the monolingual grammars used for parsing tend to be designed independently from each other, following different traditions, formalisms and linguistic theories \cite{nlphandbook}, thus making it often impossible to find a common representation describing a complete mapping from one tree to another \cite{bitext}.

\subsubsection{Our approach} \label{ourapproach}
Given that GF solves the latter problem, that it provides a good basis for NLG and that the intermediate representation it makes use of preserves all the syntactic information present in a sentence, the idea of trying to perform CA by comparing GF ASTs seems tempting. However, since ASTs need to be obtained from raw - or, at most, barely sentence-segmented - text, CA would suffer from the same problem that, as mentioned in in Section \ref{gf}, affects other tree-to-tree alignment methods and GF-based open-domain translation: the lack of a sufficiently robust analysis stage and, as a consequence, the inadequacy of the resulting parse trees. \smallskip

As anticipated throughout this chapter, we attempt to solve this problem by taking advantage of dependency parsing. In particular, UD is our formalism of choice since its focus on ``universality'' (i.e. on abstracting away from cross-lingual differences) makes it as interesting as GF itself when it comes to the possibility of establishing mappings between trees in different languages. Furthermore, when it comes to MT itself, using UD allows us to take advantage of the tools that have been developed with the intention of leveraging its similarities with GF (cf. Section \ref{ud}), thus enabling GF target language generation. \smallskip

Concretely, this means that the system we propose requires the following elements, whose reciprocal relations are shown in Figure \ref{elems}: \smallskip

\begin{itemize}
    \item a UD parser
    \item an alignment module based on dependency tree comparison, which is the tangible output of this project and whose CE and CP components are described and evaluated in detail in Chapters \ref{ch3} and \ref{ch4} respectively
    \item a program, based on \texttt{gf-ud}, that converts the alignments produced in step 2 into GF ASTs to then generate a domain-specific GF lexicon\footnote{CA is performed on UD trees, i.e. before conversion to GF, to reduce the possibility of errors. The reason is that, as mentioned in Section \ref{udrelgf}, in fact, such conversion is not obtained by means of a deterministic procedure and, as such, is not always guaranteed to be correct.}
\end{itemize} \smallskip

The CE and CP modules are implemented as part of the project, while the UD parser and \texttt{gf-ud} are independently developed open source software, even though the latter, being currently under development was adapted and extended as part of the process. In addition, we provide a simple GF-based translation script to evaluate the performance of the CA module to the test.

\begin{figure}[h]
    \centering
    \includegraphics[width=.6\linewidth]{figure/elems.png}
    \caption[Relationship between the different components of the CA system]{Relationship between the different components of the system proposed in this work. The diagram shows the minimal, trilingual use case in which both CE and CP can be made use of.} \label{elems}
    \label{sys}
\end{figure}

### 2021-02-09 final_report/include/backmatter/Appendix_1.tex
moved relevant stuff to my own public repo

% CREATED BY DAVID FRISK, 2016
\chapter{Universal POS tags and dependency labels} \label{a}
While some of them are also discussed briefly in Section \ref{ud}, this appendix provides systematic definitions for all the main UD Universal Part Of Speech tags and dependency relations used in the examples of this work. Complete lists and a good number of examples in several languages can be found in \cite{uddocs} and \cite{compsyn}.

\section{\texttt{UPOS} tags} \label{a_pos}
The official UD documentation \cite{uddocs} distinguishes between \textit{open class} or \textit{content} words and \textit{closed class} or \textit{function} words.

\subsection{Open class words}
When it comes to open class words, the categories that are relevant to this work are the following:\smallskip
\begin{itemize}
    \item \texttt{NOUN}, the class of words typically denoting a person, place, thing, animal or idea. Unlike nouns in traditional grammars, \texttt{NOUN}s are intended for \textit{common} (as opposed to \textit{proper}) nouns only
    \item \texttt{PROPN}, the class of nominals used as names (or part of names) of a specific individual, place, or object, i.e. \textit{proper nouns}
    \item \texttt{VERB}, designating events and actions. This category refers to content verbs only, while auxiliaries should be assigned the tag \texttt{AUX} (cf. \ref{cc})
\end{itemize}

\subsection{Closed class words} \label{cc}
Among closed class words, the categories used in the examples of this thesis are: \smallskip
\begin{itemize}
    \item \texttt{PRON}, defined as in traditional grammars: the class of substitutes for nouns or noun phrases, whose meaning is recoverable from the context
    \item \texttt{AUX}, the class of function words that accompany the lexical verb of a verb phrase and express grammatical distinctions not carried by such lexical verb. The most common example of verbs that fall into this category is perhaps that of \textit{tense auxiliaries} (e.g. ``\textit{has}'' in the verb phrase ``\textit{\underline{has} completed}''), but \textit{passive auxiliaries}, like ``\textit{was}'' in ``\textit{\underline{was} completed}, modal auxiliaries like ``\textit{must}'' and ``\textit{should}'' and verbal copulas should also be tagged as \texttt{AUX}
    \item \texttt{DET}, a wide-coverage class for the different types of determiners, i.e. words that modify nouns or noun phrases expressing the reference of the noun phrase in context. It includes articles, pronominal numerals and words like \textit{all} and \textit{every}.
\end{itemize} \smallskip \smallskip

Finally, UD groups together categories that apply to tokens that are not even necessarily considered words in traditional grammars, such as symbols. The only category of this kind mentioned in this work is \texttt{PUNCT}, which refers of course to punctuation.

\section{\texttt{DEPREL}s} \label{a_lab}
Dependency relation are used to label links connecting the word nodes of a dependency tree with their heads, indicating the syntactic relation occurring between them. \smallskip

The only ``exceptional'' label is \texttt{root}, which connects a fake node to the \texttt{root}, i.e. usually the main \texttt{VERB}, of a sentence. To avoid discrepancies between languages, the \texttt{root} of a sentence is always supposed to be a content word, meaning for instance that, if the main verb is a copula, it is its complement that ought to be labelled \texttt{root}.

The other dependency labels we refer to in this text can, like POS tags, be grouped according to the nature of the nature of the dependent they link to a head.

\subsection{Open class dependents}
Open class words can be assigned various kinds of labels: 
\begin{itemize}
    \item labels linking the core arguments of a verb to the verb they refer to (i.e., in case of a single-clause sentence, to its \texttt{root}): \begin{itemize}
        \item \texttt{nsubj}, indicating the nominal subject of a clause (a noun, proper noun, pronoun or numeral)
        \item \texttt{obj} and \texttt{iobj}, indicating its direct (resp. indirect) object. With \textit{direct object} we refer to the noun phrase that denotes the entity acted upon or which undergoes a change of state or motion, while the \textit{indirect object} of a verb is a nominal phrase that is a core argument of the verb but is not its subject or (direct) object 
    \end{itemize}
    \item \texttt{obl}, linking complements (i.e. non-core arguments, introduced by prepositions) to the verb they refer to
    \item labels for modifier words:    
    \begin{itemize}
        \item \texttt{amod}, linking adjectives to the nominals they modify
        \item \texttt{advmod}, linking adverbs to the words (which can belong to several categories) they modify
        \item \texttt{nmod}, used for marking the relation between a nominal and another \texttt{NOUN}, \texttt{PNOUN} or noun phrase
        \item \texttt{nummod}, linking number phrases to the \texttt{NOUN} they modify with a quantity
    \end{itemize}
    \item \texttt{flat}, used to connect the words following (in terms of position in the sentence) the head (i.e. the first word) of a flat multiword expression, and \texttt{compound}, playing a similar role in compound written as two or more separate words
    \item labels linking non-\texttt{root} verbs to other verbs, specifying what kind of clause they are heads of:
    \begin{itemize}
        \item \texttt{csubj}, for clausal subjects
        \item \texttt{ccomp}, for clausal complements, i.e. clauses that function as objects of a verb
        \item \texttt{xcomp}, for open clausal complements, i.e. clausal complements whose subject is determined by the higher clause
        \item \texttt{advcl}, for adverbial clauses modifying a predicate
        \item \texttt{acl}, for clauses modifying a nominal
    \end{itemize}
\end{itemize}

\subsection{Closed class dependents}
When it comes to closed class words, commonly used dependency labels are:

\begin{itemize}
    \item \texttt{cop}, linking copulas to their complements
    \item \texttt{aux}, linking an \texttt{AUX} to the lexical verb it refers to
    \item \texttt{det}, marking the link between a determiner and the nominal it refers to
    \item \texttt{case}, linking words (such as pre- and postpositions) marking the case of a \texttt{nmod} to the \texttt{nmod} itself in languages that do not express case morphologically
\end{itemize} \smallskip

Finally, corresponding to the POS tag \texttt{PUNCT} is \texttt{punct}, which links punctuation marks to the head of the clause they belong to. \smallskip

### 2021-02-09 final_report/include/backmatter/Appendix_2.tex
moved relevant stuff to my own public repo

\chapter{Installing, using and configuring the CA module} \label{b}
This appendix provides installation and usage instructions for the CA software developed as part of this thesis project, whose source code, as well as this report, is available at \url{https://github.com/aarneranta/concept-alignment}.

\section{Installation}
To compile the CA module, we recommend the Haskell Stack\footnote{\url{https://docs.haskellstack.org/en/stable/README/}}. 
To build the project, clone the above mentioned GitHub repository, move into the corresponding directory and run \texttt{stack build}. 
This will generate a haskell library, named \texttt{concept-alignment}, and the following five executables: \smallskip

\begin{itemize}
    \item \texttt{extract-concepts}, the CE module described in Section \ref{ch3}
    \item \texttt{propagate-concepts}, the CP module described in Section \ref{ch4}
    \item \texttt{evalign}, a script for evaluating CP and CE
    \item \texttt{generate-grammar}, for automatically generating a grammar as described in Section \ref{grgn}
    \item \texttt{translate}, the simple MT module presented in Section \ref{mt}
\end{itemize}

\section{Usage}
\subsection{\texttt{extract-concepts}}
Using the CE module with the default parameters is straightforward: \smallskip

\begin{verbatim}
    stack exec -- extract-concepts SL.conllu TL.conllu
\end{verbatim} \smallskip

The program, however, supports a number of command-line options. Most importantly:
\begin{itemize}
    \item \texttt{--file=FILE} can be used to specify where to write the resulting alignments. Unless otherwise specified (see below), the output consists in two new aligned \texttt{.conllu} files, stored at the chosen location, whose names are prefixed with \texttt{SL} and \texttt{TL} respectively
    \item \texttt{--linearize} specifies that concepts should be in the linearized format described in Section \ref{evalign}, useful for evaluation purposes
    \item \texttt{--maxsize=INT} sets the maximum size of the extracted alignments
    \item \texttt{--all} specifies that the selection step described in Section \ref{select} should be skipped
    \item \texttt{--clauses} enables clause segmentation (cf. Section \ref{clauses})
    \item \texttt{--rest} enables a second pass of alignment ``by exclusion'' (cf. Section \ref{exclusion}) 
    \item \texttt{--pharaoh=FILE} is used to specify a file in pharaoh format (cf. Figure \ref{wordalgn}) to use as backup
\end{itemize}

\subsection{\texttt{propagate-concepts}}
CP requires two, instead of three, \texttt{.conllu} files, one with the concepts to propagate and the following two containing the corpus of annotated sentences where their should be looked for. As a consequence, running it with the default parameters should be done as follows: \smallskip

\begin{verbatim}
    stack exec -- propagate-concepts SL_concepts.conllu SL.conllu 
    TL.conllu
\end{verbatim} \smallskip

All \texttt{extract-concepts} options that are also relevant for CP (i.e. all of them excepts \texttt{--maxsize} and \texttt{pharaoh}) are also valid for this second executable.

\subsection{\texttt{evalign}}
The evaluation scrip \texttt{evalign} can be run in three different ``modes'': \smallskip

\begin{enumerate}
    \item single-file mode (\texttt{stack exec -- evalign annotated.txt}): given a pre-annotated file in the linearized format described in Section \ref{evalign}, it prints out basic statistics about precision, recall and amount of reusable alignments.
    \item extraction interactive mode (\texttt{stack exec -- evalign extraction annotated.txt new.txt}): given an annotated and a new, possibly yet-to-annotate file containing linearized alignments, it allows interactive minimal annotation of the latter (if needed) and, on top of printing out the basic statitics, it compares the new alignments to the old ones, telling how many correct and incorrect alignments were lost and/or found
    \item propagation interactive mode (\texttt{stack exec -- evalign propagation annotated.ca new.ca}): similar to extraction interactive mode, excepts that the statistics are CP-specific (percentage of successfully propagated alignments, number of errors introduced by CP etc.)
\end{enumerate} \smallskip

The command line option \texttt{--reasons} can be used when criterion-wise statistics are needed.

\subsection{\texttt{generate-grammar}}
The grammar generation module can be run as follows: \smallskip

\begin{verbatim}
    stack exec -- GenerateGrammar path_to_extract_grammar 
    paths_to_morphodicts paths_to_aligned_conllu_files
\end{verbatim} \smallskip

Here, \texttt{path\_to\_extract\_grammar} should point to the extraction grammar in \texttt{.pgf}\footnote{Portable Grammar Format, generated by the GF compiler.} format. \texttt{paths\_to\_morphodicts} stands for the path of the morphological dictionaries of the languages involved, also in \texttt{.pgf} format. Similarly, \texttt{paths\_to\_aligned\_conllu\_files} is to be replaced with the files containing the concepts in CoNNL-u format. These files should be, in terms of the languages they are written in, in the same order as the morphological dictionaries.

\subsection{\texttt{translate}}
The simplicity of the translation module makes it also extremely easy to use. The string or list of newline-separate strings to translate comes from the standard input and the only argument is the path to the automatically generated GF grammar to be used, again in \texttt{.pgf} format. Here is an example of using the program to translate a single sentence: \smallskip

\begin{verbatim}
    echo "this sentence will be translated" | stack exec -- translate
    Extracted.pgf             
\end{verbatim}

\section{Configuration: modifying the alignment criteria}
As mentioned in Section \ref{criteria}, modifying the alignment criteria the program makes use of requires little effort. 
All criteria are in fact defined in a separate Haskell module \texttt{Criteria}\footnote{\url{https://github.com/aarneranta/concept-alignment/blob/master/Criteria.hs}} that exports only the list of criteria to be used. 
In the current implementation, it looks like this: \smallskip

\begin{lstlisting}
    criteria :: [Criterion]
    criteria = [udpos, ud, divs, pass, pos]
\end{lstlisting}

Removing and/or changing the priority of the criteria is then just a matter of altering such list. \smallskip

Adding new criteria is also simple, but it requires an understanding of the data type \texttt{Criterion}, defined in module \texttt{ConceptAlignment}:

\begin{lstlisting}
    data Criterion = C {
        func :: UDTree -> UDTree -> Bool, 
        reas :: S.Set Reason,
        headAlign :: Bool,
        strict :: Bool
}
\end{lstlisting}

As the above code block shows, \texttt{Criterion} is a record type whose fields are:

\begin{itemize}
    \item a boolean function \texttt{func} specifying a rule to decide whether two trees should be aligned. For instance Criterion \ref{udmatch} (UD label matching) is implemented as
    \begin{lstlisting}
        udMatch :: UDTree -> UDTree -> Bool
        (RTree n ts) `udMatch` (RTree m us) = 
            udDEPREL n == udDEPREL m
    \end{lstlisting}
    \item a set of \texttt{Reason} \texttt{reas} to be used for ranking the alignments, shown in the linearized files and used to compute criterion-specific statistics. Reasons, and not alignment rules (\texttt{func}s), are what mirrors criteria as defined in Section \ref{criteria} exactly. A \texttt{func} can define in fact a more specific alignment rule, such as ``type of categorial divergence in which an adjective is replaced by adverb'', while a \texttt{Reason} is in practice a more coarse-grained label associated to potentially many rules
    \item two boolean flags, \texttt{headAlign} and \texttt{strict}, the former specifying whether head alignment should be performed for UD tree pairs matching that criterion, the latter marking the criterion as either strict (i.e. to be also used for ``alignment by exclusion'', cf. Section \ref{exclusion}) or not.
\end{itemize}

### 2021-02-09 final_report/include/frontmatter/Abstract.tex
moved relevant stuff to my own public repo

% CREATED BY DAVID FRISK, 2016
\oneLineTitle\\
Arianna Masciolini\\
Department of Computer Science and Engineering\\
Chalmers University of Technology and University of Gothenburg\setlength{\parskip}{0.5cm}

\thispagestyle{plain}
\setlength{\parskip}{0pt plus 1.0pt}
\section*{Abstract}
This thesis presents a syntax-based approach to Concept Alignment (CA), the task of finding semantical correspondences between parts of multilingual parallel texts, with a focus on Machine Translation (MT).
Two variants of CA are taken into account: Concept Extraction (CE), whose aim is to identify new concepts my means of mere linguistic comparison, and Concept Propagation (CP), which consists in looking for the translation equivalents of a set of known concepts in a new language.
As opposed to standard statistical alignment methods, our approach allows to simultaneously align individual words and multiword expressions (even discontinuous). 
Since phrase-level alignments are useful to correctly translate idiomatic expressions, this can be beneficial for grammar-based translation pipelines, such as those based on Grammatical Framework (GF), which we use to put our system to the test. 
This is made possible by the fact that the alignments extracted by our CA model are not correspondences between strings, but rather between grammatical objects.
Another advantage of our system with respects to the solutions adopted in statistical MT is that, being essentially rule-based, it performs consistently well even on extremely small amounts of data.
Our system does, however, rely on the quality of the analyses of the parallel corpora it is applied to.
In order to mitigate the consequences of the lack of robustness of existing GF and, in general, constituency parsers, alignment is performed on the Universal Dependency (UD) trees generated by a neural dependency parser.
The resulting concepts are then used, exploiting the similarities between UD and GF, as a starting point for automatically generating a GF lexicon to be used in translation.
The tangible fruit of this work is a Haskell library, accompanied by a number of executables offering a user-friendly interface to perform both variants of CA, extraction and propagation, evaluate their results and use them in MT experiments.
% CE + CP

% KEYWORDS (MAXIMUM 10 WORDS)
\vfill
Keywords: computational linguistic, machine translation, concept alignment, syntax, dependency parsing, Universal Dependencies, Grammatical Framework

\newpage				% Create empty back of side
\thispagestyle{empty}
\mbox{}

### 2021-02-09 final_report/include/frontmatter/Acknowledgements.tex
moved relevant stuff to my own public repo

% CREATED BY DAVID FRISK, 2016
\thispagestyle{plain}
\section*{Acknowledgements}
I would like to start by thanking my supervisor Aarne Ranta, not only for giving me an opportunity to work on a topic that puts together so many of my interests, but also for the constant dialogue we had throughout the project. \smallskip

For their time and valuable feedback, I am also grateful to my Examiner Carl-Johan Seger, my opponent Tarik Ala Hadi and Marie-Philippe Gill. With her I had an informal but prolonged exchange of views on our respective thesis projects. \smallskip

I also want to express my gratitude to the friends who helped me with the translations necessary for one of this thesis' experiments: Michael, Haidar and both Claudios. \smallskip

Furthermore, I want to thank the Museo Nazionale Etrusco di Villa Giulia for letting me use a reproduction of the Pyrgi Tablets as the cover image for this report. That of the Tablets is an early example of parallel text that, for a series of coincidences, kept coming back to my mind during the six months I dedicated to this thesis. In particular, I am thankful to Alessia Argento for taking the time to look for the image based on my vague description. \smallskip

There is also a long list of people I would like to mention despite not being directly involved in the project. 
First of all, I have to thank Claudio once again, both for encouraging me and for putting up with living together during such a busy time (we would say ``\textit{per avermi supportato e sopportato}'').
I also want to thank my family for supporting my sudden decision to pursue my studies in Sweden.
Finally, the list would easily grow too long if I mentioned all the close and distant friends that made my life enjoyable during this time, but I cannot avoid mentioning Herbert, with whom I share a love for languages and Computer Science, and Quelli della B1, the irreplaceable group of people I went through my Bachelor's degree in Perugia with, that still continues being such an important part of my daily life and thoughts.

\vspace{1.5cm}
\hfill
Arianna Masciolini, Gothenburg, \monthname \space \the\year

\newpage				% Create empty back of side
\thispagestyle{empty}
\mbox{}

### 2021-02-09 final_report/include/frontmatter/Titlepage.tex
moved relevant stuff to my own public repo

% CREATED BY DAVID FRISK, 2016
% MODIFIED BY JAKOB JARMAR, 2016
% A few changes by Birgit Grohe, 2017 and \the\year
% Adjustments with the help of Gustav Örtenberg 2019

% COVER PAGE
\begin{titlepage}
\newgeometry{top=3cm, bottom=3cm,
			left=2.25 cm, right=2.25cm}	% Temporarily change margins

% Cover page background 
\AddToShipoutPicture*{\backgroundpic{-4}{56.7}{figure/auxiliary/frontpage_gu_eng_vec_m2.pdf}}
\addtolength{\voffset}{2cm}

% Cover picture (replace with your own or delete)		
\begin{figure}[H]
\centering
\vspace{1cm}	% Adjust vertical spacing here
\includegraphics[width=1\linewidth]{figure/cover/pyrgi.png}
\end{figure}

% Cover text
\mbox{}
\vfill
\renewcommand{\familydefault}{\sfdefault} \normalfont % Set cover page font

\textbf{\Huge \multiLineTitle{0.2cm}} 
\\[0.5cm]

%{\Large A Subtitle that can be Very Much Longer if Necessary}\\[0.5cm]

Master's thesis in Computer Science\setlength{\parskip}{1cm}

{\Large Arianna Masciolini} \setlength{\parskip}{2.9cm}

Department of Computer Science and Engineering \\
\textsc{Chalmers University of Technology} \\
\textsc{University of Gothenburg} \\
Gothenburg, Sweden \the\year

\renewcommand{\familydefault}{\rmdefault} \normalfont % Reset standard font
\end{titlepage}

% BACK OF COVER PAGE (BLANK PAGE)
\newpage
\restoregeometry
\thispagestyle{empty}
\mbox{}

% TITLE PAGE
\newpage
\thispagestyle{empty}
\begin{center}
	\textsc{\large Master's thesis \the\year}\\[4cm]		% Report number is currently not in use
	\textbf{\Large \multiLineTitle{0.2cm}} \\[1cm]
	{\large Arianna Masciolini}

	\vfill	
	% Logotype on titlepage	
	\begin{figure}[H]
	\centering
	% Remove the following line to remove the titlepage logotype
	\includegraphics[width=0.25\pdfpagewidth]{figure/auxiliary/ChGULogoHog.pdf}
	\end{figure}	\vspace{5mm}

	Department of Computer Science and Engineering\\
	%\emph{Division of Division name}\\
	%Name of research group (if applicable)\\
	\textsc{Chalmers University of Technology} \\
	\textsc{University of Gothenburg} \\
	Gothenburg, Sweden \the\year \\
\end{center}

% IMPRINT PAGE (BACK OF TITLE PAGE)
\newpage
\thispagestyle{plain}
\vspace*{4.5cm}
\oneLineTitle\\
Arianna Masciolini \setlength{\parskip}{1cm}

%\copyright ~ NAME FAMILYNAME, \the\year. \setlength{\parskip}{1cm}

Supervisor: Aarne Ranta, Department of Computer Science and Engineering\\
Examiner: Carl-Johan Seger, Department of Computer Science and Engineering\setlength{\parskip}{1cm}

Master's Thesis \the\year\\	% Report number currently not in use 
Department of Computer Science and Engineering\\
%Division of Division name\\
%Name of research group (if applicable)\\
Chalmers University of Technology and University of Gothenburg\\
SE-412 96 Gothenburg\\
Telephone +46 31 772 1000 \setlength{\parskip}{0.5cm}

\vfill
Cover: Reproduction of the Pyrgi Tablets, a bilingual Etruscan-Phoenician dedicatory text. Courtesy of the Museo Nazionale Etrusco di Villa Giulia.

Typeset in \LaTeX \\
%Printed by [Name of printing company]\\
Gothenburg, Sweden \the\year

### 2021-02-09 final_report/include/settings/Settings.tex
moved relevant stuff to my own public repo

% CREATED BY DAVID FRISK, 2016

% BASIC SETTINGS
\usepackage{fancyvrb}								% fancy verbatim
\usepackage{xcolor}									% Colors
\usepackage{moreverb}								% List settings
\usepackage{textcomp}								% Fonts, symbols etc.
\usepackage{lmodern}								% Latin modern font
\usepackage{helvet}									% Enables font switching
\usepackage[T1]{fontenc}							% Output settings
\usepackage[english]{babel}							% Language settings
\usepackage[utf8]{inputenc}							% Input settings
\usepackage{amsmath}								% Mathematical expressions (American mathematical society)
\usepackage{amssymb}								% Mathematical symbols (American mathematical society)
\usepackage{graphicx}								% Figures
\usepackage{subcaption}								% Enables subfigures
\numberwithin{equation}{chapter}					% Numbering order for equations
\numberwithin{figure}{chapter}						% Numbering order for figures
\numberwithin{table}{chapter}						% Numbering order for tables
\usepackage{minted}						    		% Enables source code listings
\usepackage{chemfig}								% Chemical structures
\usepackage[top=3cm, bottom=3cm,
			inner=3cm, outer=3cm]{geometry}			% Page margin lengths			
\usepackage{eso-pic}								% Create cover page background
\newcommand{\backgroundpic}[3]{
	\put(#1,#2){
	\parbox[b][\paperheight]{\paperwidth}{
	\centering
	\includegraphics[width=\paperwidth,height=\paperheight,keepaspectratio]{#3}}}}
\usepackage{float} 									% Enables object position enforcement using [H]
\usepackage{parskip}								% Enables vertical spaces correctly 
\usepackage{datetime} %date formatting tools
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}

\makeatletter
\def\BState{\State\hskip-\ALG@thistlm}
\makeatother

% OPTIONAL SETTINGS (DELETE OR COMMENT TO SUPRESS)

% Disable automatic indentation (equal to using \noindent)
\setlength{\parindent}{0cm}

% Caption settings (aligned left with bold name)
\usepackage[labelfont=bf, textfont=normal,
			justification=justified,
			singlelinecheck=false]{caption}

% Activate clickable links in table of contents  	
\usepackage{hyperref}								
\hypersetup{colorlinks, citecolor=black,
   		 	filecolor=black, linkcolor=black,
    		urlcolor=black}

% Define the number of section levels to be included in the t.o.c. and numbered	(3 is default)	
\setcounter{tocdepth}{5}							
\setcounter{secnumdepth}{5}

% Chapter title settings
\usepackage{titlesec}		
\titleformat{\chapter}[display]
  {\Huge\bfseries\filcenter}
  {{\fontsize{50pt}{1em}\vspace{-4.2ex}\selectfont \textnormal{\thechapter}}}{1ex}{}[]
\titleformat{\paragraph}
{\normalfont\normalsize\bfseries}{\theparagraph}{1em}{}
\titlespacing*{\paragraph}
{0pt}{3.25ex plus 1ex minus .2ex}{1.5ex plus .2ex}

% Header and footer settings (Select TWOSIDE or ONESIDE layout below)
\usepackage{fancyhdr}								
\pagestyle{fancy}  
\renewcommand{\chaptermark}[1]{\markboth{\thechapter.\space#1}{}}

% Select one-sided (1) or two-sided (2) page numbering
\def\layout{2}	% Choose 1 for one-sided or 2 for two-sided layout
% Conditional expression based on the layout choice
\ifnum\layout=2	% Two-sided
    \fancyhf{}			 						
	\fancyhead[LE,RO]{\nouppercase{ \leftmark}}
	\fancyfoot[LE,RO]{\thepage}
	\fancypagestyle{plain}{			% Redefine the plain page style
	\fancyhf{}
	\renewcommand{\headrulewidth}{0pt} 		
	\fancyfoot[LE,RO]{\thepage}}	
\else			% One-sided  	
  	\fancyhf{}					
	\fancyhead[C]{\nouppercase{ \leftmark}}
	\fancyfoot[C]{\thepage}
\fi

% Enable To-do notes
\usepackage[textsize=tiny]{todonotes}   % Include the option "disable" to hide all notes
\setlength{\marginparwidth}{2.5cm}

\usepackage{listings}
\lstset{
  xleftmargin=1em,
  literate={->}{$\rightarrow$}{2},
  language=haskell,
  frame=single,
  basicstyle=\footnotesize\ttfamily
}

% Supress warning from Texmaker about headheight
\setlength{\headheight}{15pt}

\usepackage{amsthm}
\newtheoremstyle{indented}% name of the style to be used
  {5pt}% measure of space to leave above the theorem. E.g.: 3pt
  {5pt}% measure of space to leave below the theorem. E.g.: 3pt
  {\itshape}% name of font to use in the body of the theorem
  { }% measure of space to indent
  {\bfseries}% name of head font
  { }% punctuation between head and body
  { }% space after theorem head; " " = normal interword space
  {\thmname{#1}\thmnumber{ #2}\thmnote{ (#3)}}

\theoremstyle{indented}

\newtheorem{definition}{Definition}

\newtheorem{criterion}{Criterion}

\newtheorem{example}{Example}

### 2021-02-09 ArgvParse.hs

@@ -1,0 +1,29 @@
+module ArgvParse where
+
+import System.Console.GetOpt
+import System.Exit
+
+-- | All flags used in the three executables. Which ones is used in which is
+-- specified in each module's options constant
+data Flag = Pharaoh String
+          | Path String
+          | MaxSize String
+          | All
+          | Clauses 
+          | Rest
+          | Linearize
+          | Help
+          | Reasons
+  deriving (Show,Eq)
+
+-- | Type synonym for command line args (non-options)
+type Arg = String
+
+-- | Given a list of command line args, a help message and a list of
+-- option descriptions, return flags and args separately or quit the program
+-- showing usage if the input is invalid
+parseArgv :: [String] -> String -> [OptDescr Flag] -> IO ([Flag], [Arg])
+parseArgv argv help options = 
+  case getOpt Permute options argv of
+    (flags,args,[]) -> return (flags,args)
+    (_,_,errs) -> putStrLn (concat errs ++ help) >> exitWith (ExitFailure 1)

### 2021-02-09 ConceptAlignment.hs

@@ -1,0 +1,483 @@
+module ConceptAlignment where
+
+import Data.Maybe
+import Data.List
+import Data.List.Split
+import qualified Data.MultiSet as MS
+import qualified Data.Map as M
+import qualified Data.Set as S
+import RTree
+import UDConcepts
+
+{- Basic (mostly data types) definitions and instances for alignments etc. -}
+
+-- | An alignment is a pair of corresponding UD subtrees (not a type synonym
+-- because of custom Eq and Ord instances)
+newtype Alignment = A (UDTree,UDTree)
+
+-- | Left (SL) tree of an alignment
+sl :: Alignment -> UDTree
+sl (A (t,u)) = t
+
+-- | Correct (TL) tree of an alignment
+tl :: Alignment -> UDTree
+tl (A (t,u)) = u
+
+-- | Alignments are shown as pairs
+instance Show Alignment where
+  show (A a) = show a
+
+-- | prAlignment is used instead of show to "visually" inspect the alignments
+prAlignment :: Alignment -> String
+prAlignment a = t' ++ "|" ++ u'
+  where (t',u') = linearize a
+
+-- | Two alignment are equal whenever their "linearizations" are the same
+instance Eq Alignment where
+  a == b = linearize a == linearize b
+
+-- | Linearize the alignment to a pair of strings, ignoring initial whitespace
+linearize :: Alignment -> (String,String)
+linearize (A (t,u)) = (prUDTreeString' t,prUDTreeString' u)
+  where prUDTreeString' = dropWhile (== ' ') . prUDTreeString
+
+-- | Alignments can be ordered based on:
+-- 1. size of their left tree
+-- 2. size of their Correct tree
+-- 3. "linearization" of their left tree (alphabetically)
+-- 4. "linearization" of their Correct tree (alphabetically)
+instance Ord Alignment where
+  (A (t1,u1)) <= (A (t2,u2)) = k t2 u2 <= k t1 u1
+    where 
+      k t u = (depthRTree t,depthRTree u,prUDTreeString t,prUDTreeString u)
+
+-- | Check if an alignment contains another;
+-- used both in pruning and selection of alignments for MT
+contains :: Alignment -> Alignment -> Bool
+(A (t1,u1)) `contains` (A (t2,u2)) = t2 `isSubRTree` t1 && u2 `isSubRTree` u1
+
+-- | Alignments is a mapping where keys are of type Alignment and values are 
+-- tuples with other relevant info 
[…]

### 2021-02-09 CoursePlansPrepoc.hs

@@ -1,0 +1,20 @@
+module CoursePlansPreproc where
+    import System.Environment (getArgs)
+    import Data.List.Split
+
+    main = do
+        args <- getArgs
+        file' <- case args of
+            [file] -> do
+                readFile file >>= return . map unwhitespace . concatMap sentences . lines
+            _ -> do 
+                putStrLn "Usage: runghc CoursePlansPreproc file"
+                return []
+        putStrLn $ unlines file'
+
+    sentences :: String -> [String]
+    sentences = splitOn ". "
+
+    unwhitespace :: String -> String
+    unwhitespace = dropWhile isWhitespace
+        where isWhitespace c = c == ' ' || c == '\t'

### 2021-02-09 Criteria.hs

@@ -1,0 +1,97 @@
+module Criteria (criteria) where
+
+import qualified Data.Set as S
+import RTree
+import UDConcepts
+import ConceptAlignment
+
+-- Modify this modules to add/remove/change alignments criteria.
+-- NOTE: this module will only export the list of criteria below! 
+
+-- | List of criteria used by align, sorted by priority
+criteria :: [Criterion]
+criteria = [udpos, ud, divs, pass, pos]
+
+{- Functions used in criteria -}
+
+-- | Original basic criterion: matching root UD labels 
+udMatch :: UDTree -> UDTree -> Bool
+(RTree n ts) `udMatch` (RTree m us) = udSimpleDEPREL n == udSimpleDEPREL m
+
+-- | POS-equivalence
+posEquiv :: UDTree -> UDTree -> Bool
+t1 `posEquiv` t2 = (not . null) ct1 && (ct1 == ct2)
+  where (ct1, ct2) = (contentTags t1, contentTags t2)
+  
+-- | A an adverb is translated as a PP (structural divergence)
+advmodObl :: UDTree -> UDTree -> Bool
+advmodObl t u = t `isLabelled` "advmod" && u `isLabelled` "obl"
+                -- POS-equiv would be too strict here
+                && length (contentTags t) == length (contentTags u)
+
+-- | An adjective is translated as a nmod (categorial divergence)
+amodNmod :: UDTree -> UDTree -> Bool
+amodNmod t u = t `isLabelled` "amod" && u `isLabelled` "nmod"
+                && subtreesTags t == subtreesTags u 
+
+-- | An adjective is translated as an adverb (categorial divergence)
+amodAdvmod :: UDTree -> UDTree -> Bool
+amodAdvmod t u = t `isLabelled` "amod" && u `isLabelled` "advmod"
+                && subtreesTags t == subtreesTags u
+
+-- | A verb is transitive in the SL but not in the TL
+-- (structural divergences regarding object)
+objObl :: UDTree -> UDTree -> Bool
+objObl t u = t `isLabelled` "obj" && u `isLabelled` "obl"
+             && t `posEquiv` u
+
+-- | The indirect object of a verb in the SL is rendered as a PP in the TL
+-- (structural divergences regarding indirect object)
+iobjObl :: UDTree -> UDTree -> Bool
+iobjObl t u = t `isLabelled` "iobj" && u `isLabelled` "obl"
[…]

### 2021-02-09 EvAlign.hs

@@ -1,0 +1,231 @@
+module EvAlign where
+import System.Environment (getArgs)
+import System.Console.GetOpt
+import System.Exit
+import Data.Maybe
+import Data.List
+import Data.List.Split
+import qualified Data.Set as S
+import qualified Data.Map as M
+import ConceptAlignment
+import UDConcepts
+import ArgvParse
+
+-- | Horrible evaluation script.
+--   Usage: runghc EvAlign labelled_old.ca new.ca 
+--   to complete annotation and compare the new alignments with the old ones
+-- , or runghc EvAlign labelled.ca to get stats on a single file
+main = do
+  argv <- getArgs
+  (flags,args) <- parseArgv argv help options
+  if Help `elem` flags 
+    then putStrLn help >> exitSuccess 
+    else do
+      alignments <- case args of
+        [p] -> readFile p >>= return . map parseLine . nonEmptyLines 
+        (cmd:p1:p2:_) -> do
+          olds <- readFile p1 >>= return . nonEmptyLines 
+          news <- readFile p2 >>= return . nonEmptyLines
+          let olds' = map parseLine olds
+          -- weak annotation check
+          news' <- if head (head news) `elem` ['+','-','=']
+            then return (map parseLine news)
+            else annotate (reverse olds') (reverse $ map read news :: [LinAlignment])
+          case cmd of
+            "extraction" -> putStrLn $ diffStats olds' news'
+            "propagation" -> putStrLn $ propStats olds' news'
+          writeFile p2 (unlines $ map (\(x,y) -> show x ++ show y) (reverse news'))
+          return news'
+        _ -> do 
+          putStrLn "Wrong number of arguments."
+          putStrLn help
+          exitWith (ExitFailure 1)
+      let stats = basicStats alignments ++ if Reasons `elem` flags then reasonStats alignments else "" 
+      putStrLn stats
+    where nonEmptyLines = filter (not . null) . lines
+
+-- | Parse a line of a .ca annotated file 
+-- (not trivial because files coming from older versions may not have reasons 
+-- or occurrence counts)
+parseLine :: String -> AnnotatedAlignment
+parseLine s = (a, la)
+  where 
+    a = read [head s] :: Annotation
+    s' = tail s
+    [t,u] = splitOn "|" $ takeWhile (/='[') s'
[…]

### 2021-02-09 ExtractConcepts.hs

@@ -1,0 +1,91 @@
+module ExtractConcepts where
+
+import Data.List
+import qualified Data.Map as M
+import qualified Data.Set as S
+import Data.Maybe
+import System.Environment (getArgs)
+import System.Console.GetOpt
+import System.Exit
+import UDConcepts
+import ConceptAlignment
+import FastAlignUtils
+import Criteria 
+import ArgvParse
+
+main = do
+  argv <- getArgs
+  (flags,args) <- parseArgv argv help options
+  case args of
+    [src,trg] -> if Help `elem` flags
+      then putStrLn help >> exitSuccess
+      else do
+        ts <- conlluFile2UDTrees src
+        us <- conlluFile2UDTrees trg
+        let tus = zip ts us -- let tus = take 50 $ zip ts us
+        smtAs <- getSmtAlignments flags src trg
+        let segment = Clauses `elem` flags
+        let byExcl = Rest `elem` flags
+        let fp = listToMaybe [path | Path path <- flags] 
+        let fp' = fromJust fp
+        let as = M.toList $ align smtAs criteria segment byExcl tus
+        let m = listToMaybe [read mmax :: Int | MaxSize mmax <- flags]
+        let as' = if All `elem` flags then as else selectForMT m as
+        let as'' = sortByConfidence as'
+        if Linearize `elem` flags
+          then 
+            if isJust fp 
+              then writeFile fp' (unlines $ map (show . toLinAlignment) as'') 
+              else mapM_ (print . toLinAlignment) as''
+          else 
+            if isJust fp 
+              then do
+                let numberedSentPairs = [1..] `zip` map (alignment2sentencePair . fst) as''
+                writeFile ("SL" ++ fp') (unlines [prUDSentence (fst x) (fst $ snd x) | x <- numberedSentPairs])
+                writeFile ("TL" ++ fp') (unlines [prUDSentence (fst x) (snd $ snd x) | x <- numberedSentPairs]) 
+              else mapM_ (print . fst) as''
+    _ -> do
+      putStrLn "Wrong number of arguments."
+      putStrLn help
+      exitWith (ExitFailure 1)
+  where 
+    getSmtAlignments :: [Flag] -> FilePath -> FilePath -> IO [LinAlignment]
+    getSmtAlignments flags src trg = case [path | Pharaoh path <- flags] of
+      [] -> return []
+      [path] -> do
+        indices <- readFile path >>= return . parsePh
+        srcConllu <- readFile src
+        trgConllu <- readFile trg
[…]

### 2021-02-09 FastAlignUtils.hs

@@ -1,0 +1,60 @@
+module FastAlignUtils where
+
+import System.Environment (getArgs)
+import Data.List
+import Data.List.Split
+import qualified Data.Set as S
+import qualified Data.Map as M
+import UDConcepts
+import ConceptAlignment
+
+-- parse a string in pharaoh format to a list of one-to-many indices
+-- correspondences 
+parsePh :: String -> [[(Int,[Int])]]
+parsePh s = map (oneToMany . map toIdxPair . words) (lines s)
+    where 
+        toIdxPair :: String -> (Int,[Int])
+        toIdxPair p = (read i :: Int, [read j :: Int])
+            where [i,j] = splitOn "-" p
+        oneToMany = M.toList . M.fromListWith (++)
+
+-- parse a bitext in fast_align format to a list of pairs of sentences
+parseBi :: String -> [(String, String)]
+parseBi s = [(head s',last s') | s' <- map (splitOn " ||| ") (lines s)]
+
+-- given a list of (target, language) sentences and a list of alignment
+-- indices, return the corresponding pairs of (multi)words as 
+-- LinAlignments
+phToLas :: [(String,String)] -> [[(Int,[Int])]] -> [LinAlignment]
+phToLas ss is = 
+    map 
+        (\(lts,(r,o)) -> LAlignment lts (S.toList r) o) 
+        (M.toList . M.fromListWith combineVals $ phToLas' ss is)
+  where 
+    phToLas' _ [] = [] -- length is supposed to be the same though
+    phToLas' [] _ = []
+    phToLas' (s:ss) (ps:pss) = phToLa s ps ++ phToLas' ss pss
+        where
+            phToLa _ [] = []
+            phToLa (s,t) ((i,js):ps) = 
+                (lt,(S.singleton FAST,1)):phToLa (s,t) ps
+                where 
+                    lt = (filter (/= '\n') one,filter (/= '\n') many) 
+                    one = words s !! i
+                    many = unlines $ reverse $ 
+                           intersperse " " [words t !! j | j <- js]
+
+-- | CoNLL-U aligned strings to fast-align bitext conversion (useful to feed 
+--   fast_aling the same tokens)
+conllu2bi :: (String,String) -> String
+conllu2bi (c1,c2) = unlines $ zipWith (\s1 s2 -> s1 ++ " ||| " ++ s2) ts1 ts2
+    where
+        (ss1,ss2) = mapPair parseUDText (c1,c2) 
[…]

### 2021-02-09 GenerateGrammar.hs

@@ -1,0 +1,110 @@
+module GenerateGrammar where
+
+import System.Environment (getArgs)
+import System.Exit
+import System.FilePath.Posix
+import System.Timeout
+import Data.List
+import Data.Maybe
+import Data.Char
+import PGF
+import BuildGFGrammar
+import UD2GF
+import UDAnnotations
+import UDConcepts
+import GFConcepts
+
+main = do
+    argv <- getArgs
+    case argv of
+        -- args should be the extraction grammar + the morphodicts + the conllu file (in this order)
+        (pr:dcs) -> do 
+            let ds = -- paths to morphodicts
+                    takeWhile isPfg dcs 
+            let cs = -- paths to aligned (by id) .conllu files
+                    dropWhile isPfg dcs
+            let gr = pr ++ ".pgf"
+            -- read grammars and conllu files
+            gr' <- readPGF gr
+            ds' <- mapM readPGF ds
+            cs' <- mapM parseUDFile cs
+            -- discard concepts not instantiated in all languages
+            let cs'' = getAlignments cs'
+            -- TODO: language order can be a problem, is it alphabetical...?
+            -- TODO: it uses all langs of the pgf grammar...
+            let langs = map (reverse . take 3 . reverse . showLanguage) (languages gr')
+            --print langs
+            -- create UD environments (one per language)
+            envs <- mapM (flip (getEnv pr) "Utt") langs
+            -- associate envs (langs) to lists of UD sentences
+            let ecs = envs `zip` cs'' :: [(UDEnv, [UDSentence])]
+            -- transform UD sentences into (LISTS OF, cf. uds2ast) ASTs, 
+            -- translation equivalents are on the same row
+            as <- mapM (\(e,ss) -> mapM (uds2astTimed 1 e) ss) ecs
+            let langNotes = map (++ ": ") langs
+            let aas = map (zip langNotes) (transpose as)
+            -- cf. $paste out/gfts-en.tmp out/gfts-it.tmp but with additional formatting required by BuildGrammar
+            let stras = map (intercalate "\n" . map (\(a,ts) -> a ++ concatMap prAbsTree ts)) aas
+            -- cf. $grep -v Backup | sort -u 
+            let stras' = intersperse "\n" $ sort $ nub $ filter (not . isInfixOf "Backup") stras
+            -- putStrLn stras' -- just because "se non vedo non credo" 
+            -- filter out trees containing weird characters and numbers
[…]

### 2021-02-09 PreprocessUtils.hs

@@ -1,0 +1,22 @@
+module PreprocessUtils where
+    import System.Environment (getArgs)
+    import Data.List
+    import Data.List.Split
+
+
+    -- Usage: runhaskell tagged_file
+    main = do
+        args <- getArgs
+        case args of
+            [f] -> do
+                content <- readFile f
+                let ss = map (map fields . lines) (sentences content)
+                let ss' = map (map wordform . denumber) ss
+                mapM_ (putStrLn . unlines) ss'
+            _ -> putStrLn "Usage: runhaskell tagged_file"
+            where
+                sentences = splitOn "\n\n"
+                fields = splitOn "\t"
+                wordform = (!! 0)
+                denumber = dropWhile (\fs -> (fs !! 2) `elem` numberingTags)
+                    where numberingTags = ["Z", "Fz", "Fpa", "Fpt"]

### 2021-02-09 PropagateConcepts.hs

@@ -1,0 +1,74 @@
+module PropagateConcepts where
+
+import Data.List
+import Data.Maybe
+import System.Environment (getArgs)
+import System.Console.GetOpt
+import System.Exit
+import RTree
+import UDConcepts
+import ConceptAlignment
+import Criteria
+import ArgvParse
+
+main = do
+    argv <- getArgs
+    (flags,args) <- parseArgv argv help options
+    case args of
+        [old,src,trg] -> if Help `elem` flags
+            then putStrLn help >> exitSuccess
+            else do
+                cs <- conlluFile2UDTrees old
+                ts <- conlluFile2UDTrees src
+                us <- conlluFile2UDTrees trg
+                let segment = Clauses `elem` flags
+                let byExcl = Rest `elem` flags
+                let fp = listToMaybe [path | Path path <- flags] 
+                let fp' = fromJust fp
+                let cs' = map rmAddedRoot cs
+                let as = map (propagate criteria segment byExcl (ts,us)) cs'
+                if Linearize `elem` flags
+                    then 
+                        if isJust fp 
+                            then writeFile fp' (unlines $ map (maybe "" (show . toLinAlignment)) as) 
+                            else mapM_ (putStrLn . maybe "" (show . toLinAlignment)) as
+                    else 
+                        if isJust fp 
+                            then do
+                                let numberedAs = [1..] `zip` map maybeFst as
+                                writeFile fp' (unlines [prUDSentence (fst x) (snd $ alignment2sentencePair $ fromJust (snd x)) | x <- numberedAs, isJust (snd x)])
+                            else mapM_ print (filter isJust (map maybeFst as)) 
+        _ -> do
+            putStrLn "Wrong number of arguments."
+            putStrLn help
+            exitWith (ExitFailure 1)
+    where
+        maybeFst :: Maybe (a,b) -> Maybe a
+        maybeFst Nothing = Nothing 
+        maybeFst (Just (a,_)) = Just a  
+        -- remove root "supertype" added by subtree2Sentence
+        rmAddedRoot :: UDTree -> UDTree
+        rmAddedRoot (RTree n ts) = 
+            RTree n { udDEPREL = if r `isPrefixOf` label
+                then drop (length r) label
+                else label
+            } ts
+            where
+                label = udDEPREL n
+                r = "root:"
+
+{- Argument parsing -} 
+
+options :: [OptDescr Flag]
+options = [ 
[…]

### 2021-02-09 Translate.hs

@@ -1,0 +1,21 @@
+module Translate where
+  
+import PGF
+import System.Environment (getArgs)
+  
+main :: IO ()
+main = do
+    file:_ <- getArgs
+    gr <- readPGF file
+    interact (translate gr)
+
+translate :: PGF -> String -> String
+translate gr s = unlines $ map (translateSent gr) (lines s)
+
+translateSent :: PGF -> String -> String
+-- try to parse in all languages
+translateSent gr s = case parseAllLang gr (startCat gr) s of
+    -- print all translations (exclude lg, the source language)
+    (lg,ts):_ -> head $ 
+        filter (notElem '[') [linearize gr l t | l <- languages gr, l /= lg, t <- ts]
+    _ -> "no parse"

### 2021-02-09 baseline/AlignTrees.hs

@@ -1,0 +1,127 @@
+module Main where
+
+import TreeConv
+import GetConfig (getSentences)
+import Data.List
+import qualified Data.Map as M
+import System.Environment (getArgs)
+
+-- usage: runghc AlignTrees file1.conllu file2.conllu
+
+main = do
+  a:b:_ <- getArgs
+  testAlignDepTrees a b
+
+-- read conll format from two files, which contain parallel UD trees
+testAlignDepTrees :: FilePath -> FilePath -> IO ()
+testAlignDepTrees a b = do
+  ts <- readFile a >>= return . map (conll2deptree . unlines) . getSentences . lines
+  us <- readFile b >>= return . map (conll2deptree . unlines) . getSentences . lines
+  let tus = take 100 $ zip ts us -- TODO: rm take & drop!
+  -- aligned trees
+  let atus  = map (uncurry alignDepTrees) tus
+  -- subtree alignments found in each tree
+  let past (t,u) = prAlignment 2 (prDepTreeString t, prDepTreeString u)
+  let pastus = [--prDepTree (fst tu) ++ prDepTree (snd tu) 
+                  -- ++ uncurry prAlignedDepTrees atu ++
+                  unlines (map past (alignedDepSubtrees atu)) 
+                | (tu,atu) <- zip tus atus]
+  mapM_ putStrLn pastus
+
+padDepNode   :: DepNode
+isPadDepNode :: DepNode -> Bool
+padDepNode     = initDepNode{label="pad",source=666,udlabel="pad"}
+isPadDepNode n = label n == "pad"
+
+{- not used
+alignUDTrees  :: DepTree -> DepTree -> (DepTree,DepTree)
+alignUDTrees  = albl 
+  where
+   albl    = alignTrees (label . root)   (T padDepNode [])
+   audlbl  = alignTrees (udlabel . root) (T padDepNode [])
+   aclslbl = alignTrees (udlabel . root) (T padDepNode [])
+-}
+
+-- align two dep trees by sorting and padding
+alignDepTrees :: DepTree -> DepTree -> (DepTree,DepTree)
+alignDepTrees = alignTrees (udlabel . root) (\t -> (udlabel (root t), distanceDepNode (root t))) (T padDepNode [])
+---alignDepTrees = (alignTrees (udlabel . root) (T padDepNode []))
+
[…]

### 2021-02-09 baseline/GetConfig.hs

@@ -1,0 +1,269 @@
+module GetConfig where
+
+import Utils 
+import TreeConv
+import Data.List (intersperse,nub,isPrefixOf,isSuffixOf)
+import Data.Char (isSpace)
+import qualified Data.Map.Strict as M
+import qualified Data.Set as S 
+import qualified PGF
+import GF (BindType(Explicit)) ---- should not need to import this!!
+import Text.PrettyPrint (render)
+
+getConfiguration :: [FilePath] -> Maybe PGF.PGF -> IO Configuration
+getConfiguration files mpgf = do
+  ss <- mapM readFile files >>= return . concatMap lines
+  let ls  = [takeWhile (/="--") (words l) |                                    -- possible tail comment separated by --
+                   l@(c:_) <- ss, notElem c "-"]                               -- non-empty lines, not starting with -
+  let metalines = [ws | ws@(_:_) <- ls, elem "#" ws]
+  let catlines  = [ws | ws@(_:_) <- ls, not (any (flip elem ws) [";", ":", "=", "#"]) ]
+  let funlines  = [ws | ws@(w:_) <- ls, w /="*",elem ":" ws]                   -- functions with type-sigs in the config file
+  let backuplines = [ws | "*":ws <- ls]
+  let deflines  = [ws | ws@(w:_) <- ls, elem "=" ws, w /= "cat", w /= "lbl"]
+  let helpcatlines = [(k,c) | "cat":k:"=":c:_ <- ls]
+
+  let helplablines = [(k,[c]) | "lbl":k:"=":cs <- ls, c <- cs]                 -- macros used for labels
+  let labmacros = M.fromListWith (++) helplablines 
+  let extfunlines = [ews | ws <- funlines, ews <- replaceAll labmacros ws] 
+
+  let lablines  = [ws | ws@(w:_) <- ls, w /="*",elem ";" ws, notElem ":" ws]   -- labels for pgf functions without type-sigs
+  let extlablines = [ews | ws <- lablines, ews <- replaceAll labmacros ws] 
+  let labmap    = M.fromList [(f,(ls,ms)) | f:";":lms <- extlablines, let (ls,ms) = break (elem '=') lms]
+  
+  let pgfmap    = maybe M.empty (M.fromList . pgfFunctions) mpgf
+  let pgffuns   = [initFunInfo {funid = f,
+                                valtype = c,
+                                argtypes = zipCatsLabels cs ls,
+                                morphoconstraints = map pMorConstraint ms      -- morph. constraints
+                                }
+                        |
[…]

### 2021-02-09 baseline/TreeConv.hs

@@ -1,0 +1,611 @@
+module TreeConv    where
+
+import Utils
+import Data.List
+import Data.Char
+import qualified Data.Map.Strict as M
+import Data.Maybe (fromMaybe)
+
+conll2deptree :: String -> DepTree
+conll2deptree =
+  depnodes2tree .
+  map row2depnode .
+  filter validDepNode .
+  map (splitWithSep '\t') .
+  filter ((/="#") . take 1) .
+  lines
+
+validDepNode ws = length ws > 7 && all isDigit (head ws)
+
+data Tree n = T n [Tree n] deriving (Show,Eq,Ord)
+
+root     :: Tree a -> a 
+children :: Tree a -> [Tree a]
+allnodes :: Tree a -> [a]
+subtrees :: Tree a -> [Tree a]
+root     (T n _)  = n
+children (T _ ts) = ts
+-- all nodes in a tree
+allnodes t = root t : concatMap allnodes (children t)
+-- all subtrees of a tree
+subtrees t = t : concatMap subtrees (children t)
+
+mapTree :: (n -> m) -> Tree n -> Tree m
+mapTree f (T n ts) = T (f n) (map (mapTree f) ts)
+
+sizeTree :: Tree n -> Int
+sizeTree (T _ ts) = 1 + sum (map sizeTree ts)
+
+-- foldTree :: (m -> m -> m) -> (n -> m) -> Tree n -> m -- can be used to implement span
+-- elemTree     :: (n -> Int) -> Tree n -> Tree n -> Bool
+-- notElemTree  :: (n -> Int) -> Tree n -> Tree n -> Bool
+
+-- to normalize trees by sorting the subtrees based on label and distance. PK (idea from AR)
+normalizeTree :: Ord a => (Tree n -> a) -> Tree n -> Tree n
+normalizeTree key (T rt rts) = T rt (sortOn key (map (normalizeTree key) rts))
+
+type MorphoTags = [(String,String)]   -- String
+-- changed by PK on 2017-09-01
+-- also changed hasMorpho, hasLemma, hasTagandMorpho functions
+-- switched from plain string to list of tuples
+
+data DepNode = DN {
+  source   :: Int,
+  word     :: String,
+  lemma    :: String,
+  postag   :: POS,
+  xpostag  :: POS,         -- added later. helpful for use in PrepocUD.hs
+  morpho   :: MorphoTags,
+  position :: Int,
+  label    :: Label,       -- udlabel found in the tree
[…]

### 2021-02-09 baseline/Utils.hs

@@ -1,0 +1,70 @@
+module Utils where
+
+import Data.List
+import Data.Char
+import qualified Data.Map.Strict as M
+import Data.Maybe (fromMaybe)
+
+-- best to move the debugging functions here
+-- to avoid replicating this in every module
+import Debug.Trace
+-- the trace functions have been modified to accepted variable names as arguments 
+tracePrS   varname v = v -- trace ("#" ++ varname ++ "\t" ++ show v) v
+tracePrSub varname v = v -- trace ("\t" ++ "#--" ++ varname ++ "\t" ++ show v) v
+tracePrL   varname v = v -- trace ("#" ++ varname ++ "\t" ++ show v ++ "\n\n") v
+traceMsgPr mesg    v = v -- trace ("#" ++ mesg ++ " " ++ show v) v
+
+
+-- Moved from GetConfig.hs
+splitWithSep :: Eq a => a -> [a] -> [[a]]
+splitWithSep sep xs = case break (==sep) xs of
+  (x1,_:x2) -> x1 : splitWithSep sep x2
+  (x1,[]) -> x1 : []
+
+-- sequence to lattice to enumerate sequences list
+replaceAll :: Ord a => M.Map a [a] -> [a] -> [[a]] 
+replaceAll dict ws = case ws of
+  []   -> []
+  h:[] -> [[x] | x <- replh]
+  h:_  -> [x:y | x <- replh, y <- replhs]
+  where
+    replh  = fromMaybe [head ws] $ M.lookup (head ws) dict
+    replhs = replaceAll dict (tail ws) 
+
+-- Read the morph feature string and return 
+-- MorphoTags / [MorphoConstraints]
+str2fsList :: Char -> Char -> String -> String -> [(String,String)]
+str2fsList lsep psep def str = case (str == def) of
+  True  -> []                          -- default value found. return Empty list
+  False -> map (i2pair psep) $ splitWithSep lsep str
+  where
+    i2pair sep mf = case break (==sep) mf of
+     (fn, sep:fv) -> (fn, fv)
+     (fn, "")     -> (fn, "")
+
+lpairs2Str :: Char -> Char -> String -> [(String,String)] -> String
+lpairs2Str lsep psep def lst = case lst of 
+  [] -> def
+  _  -> concat . intersperse (lsep:[]) $ map (\(x,y) -> x++(psep:y)) lst
+
+-- cousin of groupSortOn in Data.List.Extra
[…]

### 2021-02-09 final_report/pseudocode/BaselineA.hs

@@ -1,0 +1,21 @@
+align :: (UDTree,UDTree) -> Alignment
+align (RTree n ts,RTree m us) = (RTree n pts, RTree m pus)
+   where 
+      (pts,pus) = unzip [align (t,u) | 
+      (t,u) <- padSubtrees (sortSubtrees ts) (sortSubtrees us)]
+         where
+            root (RTree n _) = n
+            sortSubtrees = sortOn (udDEPREL . root)
+            padSubtrees xs ys = case (xs,ys) of
+               (x:xs', y:ys')
+                  | k x == k y -> (x,y):padSubtrees xs' ys'
+                  | k x <  k y -> (x,d):padSubtrees xs' ys
+                  | k x >  k y -> (d,y):padSubtrees xs  ys'
+               (_,[]) -> [(x,d) | x <- xs]
+               ([],_) -> [(d,y) | y <- ys]
+               _ -> []
+               where
+                  d = padUDNode -- dummy node used for padding
+                  -- padding key to compare subtrees
+                  k t = (udDEPREL (root t), 
+                        distanceDepNode (root t))

### 2021-02-09 final_report/pseudocode/BaselineB.hs

@@ -1,0 +1,8 @@
+extract :: Alignment -> [Alignment]
+extract (t,u) = zip (subts' t) (subts' u) 
+    where
+        subts' t = let ts = subts t in ts ++ map headt ts
+            where 
+                subts t = t:concatMap subts (immediateSubts t) 
+                    where immediateSubts (RTree _ ts) = ts
+                headt (RTree n _) = RTree n []

### 2021-02-09 final_report/pseudocode/BasicTypes.hs

@@ -1,0 +1,18 @@
+data UDWord = UDWord {
+  udID     :: Int,
+  udFORM   :: String,
+  udLEMMA  :: String, 
+  udUPOS   :: String,
+  udXPOS   :: String,
+  udFEATS  :: [String],
+  udHEAD   :: Int,
+  udDEPREL :: String,
+  udDEPS   :: String,
+  udMISC   :: [String]
+  } 
+
+data RTree n = RTree n [RTree n] 
+
+type UDTree = RTree UDWord
+
+type Alignment = (UDTree,UDTree)

### 2021-02-09 final_report/pseudocode/Extract.hs

@@ -1,0 +1,11 @@
+type Criterion = UDTree -> UDTree -> Bool
+
+alignExtract :: [Criterion] -> (UDTree,UDTree) -> [Alignment]
+alignExtract cs (t@(RTree n ts), u@(RTree m us)) = 
+    if (not . null) applyingCriteria 
+        then (t,u):concatMap (alignExtract cs) [(t',u') | 
+            t' <- sortSubtrees ts, u's <- sortSubtrees us]
+        else []
+        where
+            applyingCriteria = filter (\c -> c t u) cs
+            sortSubtrees = sortOn (udDEPREL . root)

### 2021-02-09 final_report/pseudocode/Propagate.hs

@@ -1,0 +1,12 @@
+propagate :: [Criterion] -> ([UDTree],[UDTree]) -> UDTree -> 
+    Maybe Alignment
+propagate cs ([],_) _ = Nothing 
+propagate cs (t:ts,u:us) c = 
+    let as = sortAlignments (alignExtract cs (t,u)) 
+    in case find (\(sl,_) -> sl == c) as of
+        Nothing -> propagate cs (ts,us) c
+        a -> a
+    where 
+        sortAlignments = sortOn (depthDiff . fst)
+            where
+                depthDiff t = abs (depth t - depth c)

### 2021-02-09 final_report/pseudocode/Utils.hs

@@ -1,0 +1,45 @@
+module Utils where
+
+import Data.List
+
+-- UD words/nodes are represented as a record type whose 
+-- fields mimic those of a CoNLL-U file 
+data UDWord = UDWord {
+  udID     :: Int,
+  udFORM   :: String,
+  udLEMMA  :: String, 
+  udUPOS   :: String,
+  udXPOS   :: String,
+  udFEATS  :: [String],
+  udHEAD   :: Int,
+  udDEPREL :: String,
+  udDEPS   :: String,
+  udMISC   :: [String]
+  } deriving Eq
+
+-- a rose tree is a root node followed by a list of rose trees
+data RTree n = RTree n [RTree n] deriving Eq
+
+-- a UD tree is a rose tree whose nodes are UD words
+type UDTree = RTree UDWord
+
+-- an alignment is a pair of dependency trees
+type Alignment = (UDTree,UDTree)
+
+type Criterion = UDTree -> UDTree -> Bool
+
+alignSent :: [Criterion] -> (UDTree,UDTree) -> [Alignment]
+alignSent cs (t@(RTree n ts), u@(RTree m us)) = 
+    if (not . null) applyingCriteria 
+        then (t,u):concatMap (alignSent cs) [(t,u) | 
+            t <- sortSubtrees ts, u <- sortSubtrees us]
+        else []
+        where
+            applyingCriteria = filter (\c -> c t u) cs
+            sortSubtrees = sortOn (udDEPREL . root)
+
+root :: RTree a -> a
+root (RTree n _) = n 
+
+depth :: RTree a -> Int
+depth = undefined 

### 2021-02-09 final_report/include/CAMT.tex
added part about depconfigs

@@
 In order to be used for GF-based translation, such trees need to be first converted into GF ASTs, which are then used to generate a set of grammar rules constiuting the above mentioned multilingual GF lexicon.
+\smallskip
+
+Conversion is based on annotations defining the relation between UD trees and GF ASTs, which we refer to as \textit{dependency configurations}. 
+While going into all the notational details is beyond the scope of this project\footnote{a comprehensive description of \texttt{gf-ud}'s annotation scheme is available at \url{github.com/GrammaticalFramework/gf-ud/blob/master/doc/annotations.md}.}, an example of one such configuration is
+
+\begin{verbatim}
+    #fun PredVP	nsubj head
+\end{verbatim}
+
+which specifies that a \texttt{PredVP} (predicative Verb Phrase), defined in the RGL as a function of type
+
+\begin{verbatim}
+    PredVP : NP -> VP -> Cl
+\end{verbatim}
+
+i.e.
+a function taking two arguments, an \texttt{NP} (Noun Phrase) and a \texttt{VP} (Verb Phrase) and returning a \texttt{Cl} (clause), corresponds to a UD tree whose head, the VP, has a \texttt{nsubj} attached to it.
+Appendix \ref{c} lists the dependency configurations used for UD-to-GF conversion in the context of this project.
 \smallskip
@@
 The result of this UD-to-GF conversion is still a series of tree alignments, but their resulting format makes a substantial difference, since GF ASTs can be used to obtain the rules of a generative grammar.
-\texttt{gf-ud} also provides a way to do this, given an \textit{extraction grammar} and a \textit{morphological dictionaries} of the languages of the alignments.
+\texttt{gf-ud} also provides a way to do this, given an \textit{extraction grammar} and a \textit{morphological dictionary} of the languages of the alignments.
 \smallskip

### 2021-02-09 final_report/include/backmatter/Appendix_3.tex
added part about depconfigs

\chapter{Dependency configurations} \label{c}
This appendix provides the complete list of the abstract dependency configurations used for the UD-to-GF conversion of the alignments extracted by the CA module.

\section{Category annotation}
Category annotations describe a mapping between GF categories and UD \texttt{UPOS} tags (cf. Appendix \ref{a_pos}). \smallskip

\begin{verbatim}
    #cat A        ADJ primary
    #cat Adv      ADV primary
    #cat Det      DET primary
    #cat N        NOUN primary
    #cat PN       PROPN primary
    #cat Prep     ADP primary
    #cat Pron     PRON primary
    #cat V        VERB primary
    #cat Conj     CCONJ primary
\end{verbatim}

\section{Function annotations}
Function annotations, on the other hand, put GF functions in relation with UD \texttt{DEPREL}s (cf. Appendix \ref{a_lab}). 
The following list also serves the purpose of enumerating the types of concepts GF lexica automatically generated as described in Chapter \ref{ch5} consist of. \smallskip

\begin{verbatim}
    #fun AdjCN       amod head
    #fun AdvVP       head advmod
    #fun ComplV      head obj
    #fun DetCN       det head
    #fun AdvCN       head advmod
    #fun PredVP      nsubj head
    #fun PrepNP      case head
    #fun PrepCN      head nmod
    #fun FlatPN      head flat -- PN -> PN -> PN
    #fun CompoundN   compound head
    #fun PrepPP      head obl
\end{verbatim}

### 2021-02-09 final_report/include/TT.tex
reviewed TT chapter

@@
 While many artificial languages, such as programming languages, are compositional by construction, the general validity of this principle for natural languages is, despite many arguments in favor, under debate.
-In this work, we make the assumption that compositionality also holds for natural languages, or at least that it is a useful notion applying to most natural language sentences\footnote{in Section \ref{evalign}, however, we'll see that idiomatic expressions, often used as an arguments against compositionality, are also challenging to deal with in the context of grammar-based CA.}.
+In this work, we make the assumption that compositionality also holds for natural languages, or at least that it is a useful notion applying to most natural language sentences\footnote{in Section \ref{evalign}, however, we will see that idiomatic expressions, often used as an arguments against compositionality, are also challenging to deal with in the context of grammar-based CA.}.
 \smallskip
@@
 The idea of taking advantage of semantic compositionality is not new in the field of Computational Linguistics. 
-Doing so traditionally involves using some kind of \textit{constituency grammar}.
-\smallskip
+Doing so traditionally involves using some kind of \textit{constituency grammar}. \smallskip
 
@@
 % intro to GF
-Grammatical Framework (GF), being a grammar formalism and programming language designed with this parallel in mind, goes a step further in the direction of synchronous CFGs and TAGs by introducing a clear distinction between the \textit{abstract syntax}, whose aim is to capture the syntactic structures all natural languages taken into account have in common, and the \textit{concrete syntaxes}, specific to each individual language, consisting in their linearization rules \cite{gf2004} \cite{gfbook}.
+Grammatical Framework (GF), being a grammar formalism and programming language designed with this parallel in mind, goes a step further in the direction of synchronous CFGs and TAGs. 
+It introduces in fact a clear distinction between the \textit{abstract syntax}, whose aim is to capture the syntactic structures all natural languages taken into account have in common, and the \textit{concrete syntaxes}, specific to each individual language, consisting in their linearization rules \cite{gf2004} \cite{gfbook}.
 This makes it possible to deal with multiple languages by writing only one grammar, whose components are an abstract syntax and several concrete syntaxes, providing a solid basis for a system able to translate between any pair of for which a concrete syntax is available, in any direction.
@@
 % GF as an interlingua-based MT system
-With respects to this, a GF can be also seen as an \textit{interlingual} MT system where the intermediate representation or \textit{interlingua} is an AST.
+With respects to this, GF can be also seen as an \textit{interlingual} MT system where the intermediate representation or \textit{interlingua} is an AST.
 Interlingua-based systems have the advantage, in terms of efficiency, of making it unnecessary to build $n (n - 1)$ translation functions to cover all possible pairs of $n$ languages: having $2n$ is in general enough.
@@
 Each dependency is, then, composed of two words: a \textit{head} and a \textit{dependent} that refers to it.
-For instance, if we try to look at syntactic dependencies in the sentence ``\textit{grammars rule}'', we could identify the verb ``\textit{rule}'' as the head and the noun ``\textit{children}'', its subject, as its dependent.
+For instance, if we try to look at syntactic dependencies in the sentence ``\textit{grammars rule}'', we could identify the verb ``\textit{rule}'' as the head and the noun ``\textit{grammars}'', its subject, as its dependent.
 \smallskip
@@
 % pros
-Intuitively, this makes it so that dependency trees are simpler, which makes them an easier target for the frontend, possibly ML-based, of a MT system such as the one outlined in the above. 
-Existing \textit{dependency parsers}, such as UDPipe \cite{udpipe1} and the Standford parser \cite{standford}, are often - but not always \cite{rasp} - neural pipelines trained on dependency treebanks.
-As a result, they are significantly more robust than their phrase structure counterparts.
+Intuitively, this means that dependency trees are simpler than their phrase structure counterparts.
+This makes them an easier target for the frontend, possibly ML-based, of a MT system such as the one outlined in the above.
+Existing \textit{dependency parsers}, such as UDPipe \cite{udpipe1} and the Standford parser \cite{standford}, are often - but not always \cite{rasp} - neural pipelines trained on dependency treebanks, significantly more robust than their phrase structure counterparts.
 
 % cons
-On the other hand, there is currently no effective way to use these trees as the starting point for NLG: dependency grammars are an effective way to describe language, but are not \textit{generative}, unlike their phrase-structure counterparts\footnote{that does not mean, however, that dependency grammars have not been made use of in MT research, in conjunction with statistical techniques, in order to better capture grammatical generalizations \cite{treelet}.}.
+On the other hand, there is currently no effective way to use these trees as a starting point for NLG: dependency grammars are an effective way to describe language, but, unlike phrase-structure grammars, they are not \textit{generative},\footnote{that does not mean, however, that dependency grammars have not been made use of in MT research, in conjunction with statistical techniques, in order to better capture grammatical generalizations \cite{treelet}.}.
 \smallskip
@@
 One idea is, then, to develop a hybrid system where the frontend is a dependency parser and the backend a grammaticality-preserving GF-based target language generation module. 
-The connecting link between these two seemingly incompatible stages of our MT pipeline can be, as we will elaborate on in Section \ref{ourapproach}, a CA component able to identify matching dependency trees and generate the corresponding GF concrete and abstract syntax functions: concepts.
+This work goes in this direction, and the connecting link between these two seemingly incompatible stages of the pipeline is, as we will elaborate on in Section \ref{ourapproach}, a CA component able to identify matching dependency trees and generate the corresponding GF concrete and abstract syntax functions: concepts.
 
@@
 Appendix \ref{a} provides a more comprehensive, but not completely exhaustive, description of all the POS tags and dependency labels mentioned in the examples appearing in this work based on that in \cite{compsyn}.
-The reader interested in a full specification of the UD annotation scheme may refer to the official UD documentation\footnote{\url{universaldependencies.org}}.
+The reader interested in a full specification of the UD annotation scheme may refer to the official UD documentation\footnote{available at \url{universaldependencies.org}.}.
 
@@
 
-\begin{figure}[H]
+\begin{figure}[h]
     \centering
@@
 \paragraph{Universal POS tags} \label{upos}
-Universal POS tags mark the core part-of-speech categories, such as nouns, verbs, pronouns and determiners.
-An important - as we will see in Section \ref{criteria} - distinction we can make based on POS tags is that between \textit{content} or \textit{open class} words, i.e.
-words that contribute to the meaning of the sentences they appear in (nouns, lexical verbs, adjectives, adverbs and interjections), and \textit{function} or \textit{closed class} words, like pronouns and determiners, so called because they do not readily accept new members\footnote{one exception is the recent introduction of the gender-neutral pronoun ``\textit{hen}'' in Swedish.}.
+Universal POS tags mark the core part-of-speech categories, such as nouns, verbs, pronouns and determiners. \smallskip
+
+An important distinction we can make based on POS tags is, as we will see in Section \ref{criteria}, that between \textit{content} and \textit{function} words.
+\textit{Content} or \textit{open class} words are words with a lexical meaning (nouns, lexical verbs, adjectives, adverbs and interjections).
+\textit{Function} words, like pronouns and determiners, only have a grammatical meaning.
+Since they do not readily accept new members\footnote{one exception is the recent introduction of the gender-neutral pronoun ``\textit{hen}'' in Swedish.}, they are often referred to also as \textit{closed class} words.
 \smallskip
@@
 Universal dependency relations, largely based on \cite{st2}, represent, as mentioned in the above, syntactic dependencies between individual pairs of words occurring in the same sentence. 
-In particular, according to the CoNLL-U standard (cf.
-Section \ref{conll}), each word is assigned a \textit{dependency label} indicating in which way it is liked to its \texttt{HEAD}.
-\smallskip
+In particular, according to the CoNLL-U standard (cf. Section \ref{conll}), each word is assigned a \textit{dependency label} indicating in which way it is liked to its \texttt{HEAD}. \smallskip
 
@@
 
- If a sentence contains is no verbs at all, there is no fixed rule excepts that in order to avoid unnecessary discrepancies between languages, the root should be a content word. \smallskip
+ If a sentence contains is no verbs at all, there is no fixed rule excepts that, in order to avoid unnecessary discrepancies between languages, the root should be a content word. \smallskip
 
-\begin{figure}[H]
+\begin{figure}[h]
    \centering
@@
 In terms of parallel text analysis, an \textit{alignment} consists in a pair of semantically equivalent (i.e., in this context, sharing the same abstract syntax) concrete expressions, one in the source and one in the target language.
-Even though, unless otherwise specified, we will keep referring to \textit{language pairs}, with a source and a target language, it is easy to see how this can be generalized to a more-than-bilingual, multidirectional definition: the pair just mentioned simply becomes a tuple of equivalent expressions in different languages.
-Formally, \smallskip
+Even though, unless otherwise specified, we will keep referring to \textit{language pairs}, with a source and a target language, it is easy to see how this can be generalized to a more-than-bilingual, multidirectional definition.
+The pair just mentioned simply becomes a tuple of equivalent expressions in different languages. Formally, \smallskip
 
@@
     \item Model 3 takes even a \textit{fertility} parameter into account (fertility being the number of target words that can be generated from a given source word)
-    \item Models 4 and 5 takes the \textit{context} (i.e. the relative position of the words) in which each word occurs into account in a broader sense
+    \item Models 4 and 5 take the \textit{context} (i.e. the relative position of the words) in which each word occurs into account in a broader sense
 \end{itemize}
@@
 
-However, it must be noted that, in the context of statistical MT, the term ``phrase'' is not to be intended in its grammatical sense, but can refer to any sequence of - typically contiguous - words and in any case, both for word and phrase alignment, the relation that is established with these statistical methods is merely between strings in different languages: there is no intermediate representation capturing the concepts themselves.
+However, it must be noted that, in the context of statistical MT, the term ``phrase'' is not to be intended in its grammatical sense, but can refer to any sequence of - typically contiguous - words. 
+In any case, both for word and phrase alignment, the relation that is established with these statistical methods is merely between strings in different languages: there is no intermediate representation capturing the concepts themselves.
 \smallskip
@@
 Over time, various constituency grammar-based approaches have been proposed \cite{t2t1, t2t2}. 
-These approaches, which make use of either parallel - manually or automatically annotated - treebanks, can be referred to as \textit{tree-to-tree alignment} methods \cite{bitext}.
+These approaches, which make use of parallel treebanks, can be referred to as \textit{tree-to-tree alignment} methods \cite{bitext}. 
 While theoretically appealing, however, their use in data-driven MT is limited, since they tend to suffer not only from the scarce availability of large-scale manually annotated corpora and the lack of robustness of the existing parsers, but also from the fact that, in general, the monolingual grammars used for parsing tend to be designed independently from each other, following different traditions, formalisms and linguistic theories \cite{nlphandbook}, thus making it often impossible to find a common representation describing a complete mapping from one tree to another \cite{bitext}.
@@
 \subsubsection{Our approach} \label{ourapproach}
-Given that GF solves the latter problem, that it provides a good basis for NLG and that the intermediate representation it makes use of preserves all the syntactic information present in a sentence, the idea of trying to perform CA by comparing GF ASTs seems tempting.
+Given that GF solves the latter problem, that it provides a good basis for NLG and that the intermediate representations it makes use of preserve all the syntactic information present in a sentence, the idea of trying to perform CA by comparing GF ASTs seems tempting.
 However, since ASTs need to be obtained from raw - or, at most, barely sentence-segmented - text, CA would suffer from the same problem that, as mentioned in in Section \ref{gf}, affects other tree-to-tree alignment methods and GF-based open-domain translation: the lack of a sufficiently robust analysis stage and, as a consequence, the inadequacy of the resulting parse trees.
@@
 on abstracting away from cross-lingual differences) makes it as interesting as GF itself when it comes to the possibility of establishing mappings between trees in different languages.
-Furthermore, when it comes to MT itself, using UD allows us to take advantage of the tools that have been developed with the intention of leveraging its similarities with GF (cf.
+Furthermore, when it comes to MT itself, using UD allows us to take advantage of the tools that have been developed to leverage its similarities with GF (cf.
 Section \ref{ud}), thus enabling GF target language generation.
@@
     \item an alignment module based on dependency tree comparison, which is the tangible output of this project and whose CE and CP components are described and evaluated in detail in Chapters \ref{ch3} and \ref{ch4} respectively
-    \item a program, based on \texttt{gf-ud}, that converts the alignments produced in step 2 into GF ASTs to then generate a domain-specific GF lexicon\footnote{CA is performed on UD trees, i.e.
+    \item a program, based on \texttt{gf-ud}, that converts the alignments into GF ASTs to then generate a domain-specific GF lexicon\footnote{CA is performed on UD trees, i.e.
 before conversion to GF, to reduce the possibility of errors.
-The reason is that, as mentioned in Section \ref{udrelgf}, in fact, such conversion is not obtained by means of a deterministic procedure and, as such, is not always guaranteed to be correct.}
+The reason is that, as mentioned in Section \ref{udrelgf}, in fact, such conversion is not obtained by means of a deterministic procedure and, as such, is not always guaranteed to be correct.}.
 \end{itemize} \smallskip

### 2021-02-09 final_report/include/Introduction.tex
added par on grammatical objects to the intro

@@
 
+Last but not least, a syntax-based approach can produce correspondences between grammatical objects, as opposed to pairs of strings.
+Alignments of this kind are valuable, as they can more easily be utilized in rule-based MT systems.
+\smallskip
+
 % contribution (+ sw deliverables w technologies involved)
-With this thesis, we propose a syntax-based approach to both of the above mentioned variants of CA - concept extraction and concept propagation - and put it to the test by integrating it in a prototype domain-specific MT system composed of a neural Universal Dependencies (UD) parser, a rule-based alignment module and a target language generation component based on Grammatical Framework (GF).
+With this thesis, we propose a syntax-based approach to both of the above mentioned variants of CA - concept extraction and concept propagation - composed of a neural Universal Dependencies (UD) parser and a rule-based alignment module.
+We put our system to the test by integrating it in a prototype domain-specific MT system based on Grammatical Framework (GF).
 

### 2021-02-09 final_report/include/CE.tex
quickly reviewed CE chapter

@@
 In this context, as shown in Figure \ref{types}, dependency trees are represented as rose trees, i.e. tree data structures with an unbounded, variable number of branches per node.
+
+\begin{figure}[H] 
+  \centering
+  \lstinputlisting{pseudocode/BasicTypes.hs}
+  \caption[Fundamental data typesfor dependency trees]{Fundamental data types used for representing dependency trees in the CA module.}
+  \label{types}
+ \end{figure}
+
+As shown in Figure \ref{types}, the type of the nodes of such trees, \texttt{UDWord}, is a record type whose fields mirror those of a CoNLL-U file (cf.
+\ref{conll}).
+Alignments are, as for Definition \ref{algndef}, simply pairs of UD trees.
 
@@
    udDEPREL = "root", (...)}, 
+  udDEPREL = "root", (...)}, 
+   udDEPREL = "root", (...)}, 
   subtrees = [
@@
     udHEAD = 3, udDEPREL = "det", (...)}, 
+   udHEAD = 3, udDEPREL = "det", (...)}, 
+    udHEAD = 3, udDEPREL = "det", (...)}, 
    subtrees = []},
@@
     udID = 2, udFORM = "dependency", udLEMMA = "dependency", udUPOS = "NOUN", (...), 
+   udID = 2, udFORM = "dependency", udLEMMA = "dependency", udUPOS = "NOUN", (...), 
+    udID = 2, udFORM = "dependency", udLEMMA = "dependency", udUPOS = "NOUN", (...), 
+    udHEAD = 3, udDEPREL = "compound", (...)}, 
+   udHEAD = 3, udDEPREL = "compound", (...)}, 
     udHEAD = 3, udDEPREL = "compound", (...)}, 
@@
   \label{deptree}
-\end{figure}
-
-As shown in Figure \ref{types}, the type of the nodes of such trees, \texttt{UDWord}, is a record type whose fields mirror those of a CoNLL-U file (cf.
-\ref{conll}).
-Alignments are, as for Definition \ref{algndef}, simply pairs of UD trees.
-
-\begin{figure}[H] 
- \centering
- \lstinputlisting{pseudocode/BasicTypes.hs}
- \caption[Fundamental data typesfor dependency trees]{Fundamental data types used for representing dependency trees in the CA module.}
- \label{types}
  \end{figure}
@@
 After extracting subtree alignments, it is possible and generally useful to remove the dummy subtrees added during alignment they contain.
-
-Obviously, the ``full-sentence'' tree pair is itself an alignment, even though a trivial one, as it represents a sentence-level correspondence in a case where, due to the approach being syntactic comparison, the inputs are assumed to be sentences that correspond to each other\footnote{contrary to what it may seem, however, sentence-level alignment is in general not at a trivial task, as it is often the case, especially in certain language pairs, that one \textit{orthographic} sentence, i.e.
-a sentence defined based on the presence of a full stop, maps to more than one orthographic sentences in the translated text, or vice versa.}.
 
@@
  \begin{itemize}
- \item $\langle$\textit{we are perfectly aligned, siamo perfettamente allineati}$\rangle$
- \item $\langle$\textit{aligned,allineati}$\rangle$ (head alignment)
- \item $\langle$\textit{are}$\rangle$
- \item $\langle$\textit{siamo}$\rangle$
- \item $\langle$\textit{perfectly,perfettamente}$\rangle$
- \item $\langle$\textit{aligned,allineati}$\rangle$
+ \item $\langle$\textit{``we are perfectly aligned'', ``siamo perfettamente allineati''}$\rangle$
+ \item $\langle$\textit{``aligned'', ``allineati''}$\rangle$ (head alignment)
+ \item $\langle$\textit{``are'', ``siamo''}$\rangle$
+ \item $\langle$\textit{``perfectly'', ``perfettamente}$\rangle$
+ \item $\langle$\textit{``aligned'', ``allineati}$\rangle$
  \end{itemize}
@@
 \end{figure}
+
+Obviously, the ``full-sentence'' tree pair is itself an alignment, even though a trivial one, as it represents a sentence-level correspondence in a case where, due to the approach being syntactic comparison, the inputs are assumed to be sentences that correspond to each other\footnote{contrary to what it may seem, however, sentence-level alignment is in general not at a trivial task, as it is often the case, especially in certain language pairs, that one \textit{orthographic} sentence, i.e.
+a sentence defined based on the presence of a full stop, maps to more than one orthographic sentences in the translated text, or vice versa.}.
 
@@
 \end{example}
-where, in the original sentence, the complement of the copula \textit{am}, \textit{hungry}, would be labelled as its \texttt{root} but, in Italian, the noun \textit{fame} is the object of the root verb \textit{ha}.
+where, in the original sentence, the complement of the copula ``\textit{am}'', ``\textit{hungry}'', would be labelled as its \texttt{root} but, in Italian, the noun ``\textit{fame}'' is the object of the root verb ``\textit{ha}''.
 It proved hard to detect cases like this via purely syntactic rules without causing the program to extract a lot of incorrect alignments as well, but if necessary it is easy to modify the program by removing or adding rules of this kind, potentially even language-specific.
@@
 \end{example} 
-where the single word \textit{salutare} is not a compound but expresses the same meaning as ``say'' and ``hi'' together. \smallskip
+where the single word ``\textit{salutare}'' is not a compound but expresses the same meaning as ``\textit{say}'' and ``\textit{hi}'' together. \smallskip
 Divergences like the one described in the latter example are usually captured by Criterion \ref{udmatch}, while compounds often need to be taken care of explicitly.
@@
 \end{example}
-what is the direct object in English (\textit{Francesco}) becomes a prepositional phrase (\textit{a Francesco}) in Italian. 
+what is the direct object in English (``\textit{Francesco}'') becomes a prepositional phrase (``\textit{a Francesco}'') in Italian. 
 Even though in many cases these divergences can be captured by POS-equivalence, in order to ensure higher precision, the CE module handles all of them explicitly via rules that combine it with UD label checking and whose priority is higher than that of the basic Criterion \ref{poseq}.
@@
 \end{example} 
-in Swedish, the logical modifier \textit{usually} is implicitly expressed by the verb \textit{brukar} itself, i.e. placed ``higher up''. \smallskip
+in Swedish, the logical modifier ``\textit{usually}'' is implicitly expressed by the verb ``\textit{brukar}'' itself, i.e. placed ``higher up''. \smallskip
 
@@
 
-One could object that divergences like the above (and some other) could - and maybe \textit{should} - be avoided when translating a text.
+One could object that divergences like the above (and some other) could - and maybe should - be avoided when translating a text.
 We will not dwell upon this, since what matters for the purpose of CA is not what is desirable in natural language translation but rather what divergences do occur in the human-translated parallel texts available for analysis.
@@
 In its most basic version, this algorithm does nothing to avoid subtrees in the source language to be aligned with multiple subtrees in the target language and vice versa.
-As a consequence, it systematically \textit{overgenerates} alignments.
+As a consequence, it systematically overgenerates alignments.
 \smallskip
@@
 
-It can happen, however, that two trees match \textit{more than one} of the criteria.
+It can happen, however, that two trees match more than one of the criteria.
 For instance, two trees might have their roots sharing the same UD label (cf.
-Criterion \ref{udmatch}) \textit{and} be POS-equivalent.
+Criterion \ref{udmatch}) and be POS-equivalent.
 In such cases, it intuitively makes sense to consider them more likely to be exact, as there are literally more reasons to align them.
@@
  \begin{itemize}
- \item $\langle$Enrico eats a banana, Enrico mangia una banana$\rangle$
- \item $\langle$Enrico, Enrico$\rangle$
- \item $\langle$a banana, una banana$\rangle$
- \item $\langle$a, una$\rangle$
+ \item $\langle$``Enrico eats a banana'', ``Enrico mangia una banana''$\rangle$
+ \item $\langle$``Enrico'', ``Enrico''$\rangle$
+ \item $\langle$``a banana'', ``una banana''$\rangle$
+ \item $\langle$``a'', ``una''$\rangle$
  \end{itemize} 
@@
  \begin{itemize}
- \item $\langle$eats, mangia$\rangle$
- \item $\langle$banana, banana$\rangle$
+ \item $\langle$``eats'', ``mangia''$\rangle$
+ \item $\langle$``banana'', ``banana''$\rangle$
  \end{itemize} 
@@
 While it is of extreme importance not to miss alignments like the last two in the example above, aligning heads is not always appropriate, especially when, as in the present case, common translation divergence patterns are one of the criteria.
-For instance, when two trees are aligned because of a categorial divergence such as that of Example \ref{herb}, where \textit{doctoral} corresponds to \textit{di dottorato}, it is at least questionable to also align \textit{doctoral} with \textit{dottorato}.
+For instance, when two trees are aligned because of a categorial divergence such as that of Example \ref{herb}, where ``\textit{doctoral}'' corresponds to ``\textit{di dottorato}'', it is at least questionable to also align ``\textit{doctoral}'' with ``\textit{dottorato}''.
 Consequently, in the current implementation of the algorithm, each alignment criterion is associated with a flag telling whether heads should also be aligned whenever a new alignment is extracted because of that criterion.
@@
 \begin{example} 
- ``I took a course on \underline{Machine Learning techniques}'' VS ``Jag deltog i en kurs om \underline{maskininlärningstekniker}''
+ ``Giorgio took a course on \underline{Machine Learning techniques}'' VS ``Giorgio deltog i en kurs om \underline{maskininlärningstekniker}''
 \end{example}
@@
 
-it is hard to deny that \textit{boat} has been ``translated'' as \textit{treno}, ``train'', so our CE module would - and should - be able to detect it, but in most situations it is by no means desirable that such correspondence at the word level is made use of, for instance by storing it in an bilingual lexicon!
+it is hard to deny that ``\textit{boat}'' has been ``translated'' as ``\textit{treno}'', ``\textit{train}'', so our CE module would - and should - be able to detect it, but in most situations it is by no means desirable that such correspondence at the word level is made use of, for instance by storing it in an bilingual lexicon!
 As a consequence, each alignment can be marked as:
@@
 With alignments in this format available, a simple way to evaluate a specific version of the CE module is to manually assess the correctness of each of the correspondences the program identifies and compute some statistics about them.
-A separate Haskell module, \texttt{EvAlign}, was written to parse files in the above format and compute a series of useful statistics: \smallskip
+A separate Haskell module, whose usage is described in Appendix \ref{b}, was written to parse files in the above format and compute a series of useful statistics: \smallskip
 \begin{itemize}
  \item the total number of alignments found
- \item the number of \textit{distinct} alignments, i.e. the number of alignments with distinct linearizations
+ \item the number of distinct alignments, i.e. the number of alignments with distinct linearizations
  \item the percentage of correct alignments
@@
 % \useunder{\uline}{\ul}{}
-\begin{table}[H]
+\begin{table}[h]
  \centering
- \small
+ \footnotesize
  \begin{tabular}{|l|l|l|l|l|}
@@
 \begin{table}[H]
- \small
+ \footnotesize
  \centering
@@
 \begin{table}[H]
- \small
+ \footnotesize
  \centering 
@@
 
-A Haskell script converts the CoNNL-U files that are fed to \texttt{AlignTrees} into the input format required by \texttt{fast\_align}, so that the exact same tokenization is used. \smallskip
+A Haskell script converts the CoNNL-U files that are fed to our CE module into the input format required by \texttt{fast\_align}, so that the exact same tokenization is used. \smallskip
 
@@
 
-Because \texttt{fast\_align} is a purely statistical approach, we trained it not only on the same 100-sentence subset of PUD used for evaluating \texttt{AlignTrees} against the baseline, but also on the full 1000-sentences dataset.
+Because \texttt{fast\_align} is a purely statistical approach, we trained it not only on the same 100-sentence subset of PUD used for evaluating our proposed improved CA module against the baseline, but also on the full 1000-sentences dataset.
 In the latter experiment, the alignments obtained for the last 900 sentences were discarded before evaluating the results.
@@
  \hline
-  & \textbf{\texttt{AlignTrees}} & \textbf{\texttt{fast\_align} 100} & \textbf{\texttt{fast\_align} 1000}\\ \hline
+  & \textbf{our system} & \textbf{\texttt{fast\_align} 100} & \textbf{\texttt{fast\_align} 1000}\\ \hline
   \textbf{distinct alignments} & 716 & 1440 & 1435 \\ \hline

### 2021-02-09 final_report/pseudocode/BasicTypes.hs

@@ -8,8 +8,7 @@ data UDWord = UDWord {
   udHEAD   :: Int,
   udDEPREL :: String,
   udDEPS   :: String,
-  udMISC   :: [String]
-  } 
+  udMISC   :: [String] } 
 
 data RTree n = RTree n [RTree n] 
 

### 2021-02-09 final_report/include/CE.tex
quickly reviewed chapters about CP and CE

@@
 For the reasons discussed in Section \ref{ourapproach}, our syntax-based approach to CE is based on comparing, instead on ASTs, dependency trees, and UD trees in particular.
-The task of dealing with UD trees is facilitated by the existence of a series of Haskell modules written in the context of the development of the aforementioned \texttt{gf-ud} and of some preliminary unpublished CA experiments\footnote{Aarne Ranta and Prasanth Kolachina, code available at \url{https://github.com/aarneranta/concept-alignment/tree/master/old_ca_versions/v1}}.
+The task of dealing with UD trees is facilitated by the existence of a series of Haskell modules written in the context of the development of the aforementioned \texttt{gf-ud} and of some preliminary unpublished CA experiments\footnote{Aarne Ranta and Prasanth Kolachina, code available at \url{github.com/aarneranta/concept-alignment/tree/master/old_ca_versions/v1}}.
 \smallskip

### 2021-02-09 final_report/include/CP.tex
quickly reviewed chapters about CP and CE

@@
 \chapter{Concept Propagation} \label{ch4}
-After a set of concepts is known, it can be useful to be able to look for the corresponding concrete expressions in a new language.
+After a set of concepts is known, it can be useful to look for the corresponding concrete expressions in a new language.
 In this chapter, we describe our approach to this task and present an evaluation of the corresponding module.
@@
 Prepositions, on the other hand, are only useful with some information about the context in which they occur in, since it is customary for them to be used differently across different, even closely related languages.
-However, some common content words, like $\langle$\textit{always, sempre, alltid}$\rangle$, $\langle$\textit{new, nuovi, nya}$\rangle$ and $\langle$\textit{same, stesso, samma}$\rangle$ are also aligned correctly.
+However, some common content words, like $\langle$\textit{``always'', ``sempre'', ``alltid''}$\rangle$, $\langle$\textit{``new'', ``nuovi'', ``nya''}$\rangle$ and $\langle$\textit{``same'', ``stesso'', ``samma''}$\rangle$ are also aligned correctly.
 
@@
 Because the text are similar both stylistically and, most importantly, in terms of content, the program is in this case also able to find trilingual correspondences between content words, sometimed specific of the domain.
-Examples of such alignments are $\langle$\textit{skills, capacità, färdigheter}$\rangle$, $\langle$\textit{functional, funzionale, funktionell}$\rangle$, $\langle$\textit{exam, prova, tentamen}$\rangle$, $\langle$\textit{course, corso, kurs}$\rangle$, $\langle$\textit{prerequisites, prerequisiti, behörigheter}$\rangle$, $\langle$\textit{lectures, lezioni, föreläsningar}$\rangle$ and $\langle$\textit{knowledge, conoscenza, kunskap}$\rangle$.
+
+    Examples of such alignments are $\langle$\textit{``skills'', ``capacità'', ``färdigheter''}$\rangle$, $\langle$\textit{``functional'', ``funzionale'', ``funktionell''}$\rangle$, $\langle$\textit{``exam'', ``prova'', ``tentamen''}$\rangle$, $\langle$\textit{``course'', ``corso'', ``kurs''}$\rangle$, $\langle$\textit{``lectures'', ``lezioni'', ``föreläsningar''}$\rangle$, $\langle$\textit{``prerequisites'', ``prerequisiti'', ``behörigheter''}$\rangle$ and $\langle$\textit{``knowledge'', ``conoscenza'', ``kunskap''}$\rangle$.
 \smallskip
@@
     Sometimes, however, the meaning of a word is highly context-dependent and as a consequence, even if both the pairs of extracted and propagated alignments are correct, the corresponding trilingual alignment we could infer isn't (e.g.
-$\langle$\textit{learning, conoscere}$\rangle$ and $\langle$\textit{learning, inlärning}$\rangle$).
+$\langle$\textit{``learning'', ``conoscere''}$\rangle$ and $\langle$\textit{``learning'', ``inlärning''}$\rangle$).
 \smallskip
 
-    Finally, in some rare cases, even correspondences regarding longer expressions, such as $\langle$\textit{the aim of the course, l'obiettivo del corso, syftet med kursen}$\rangle$, can be found.
+    Finally, in some rare cases, even correspondences regarding longer expressions, such as $\langle$\textit{``the aim of the course'', ``l'obiettivo del corso'', ``syftet med kursen''}$\rangle$, can be found.
 In this case, the English-Italian correspondence is not especially useful, since the phrase could be translated almost word by word between the two languages, but the translation to Swedish is not literal, thus making the alignment useful for the purposes of MT.

### 2021-02-09 final_report/include/backmatter/Appendix_2.tex
linked to new repo

@@
 \chapter{Installing, using and configuring the CA module} \label{b}
-This appendix provides installation and usage instructions for the CA software developed as part of this thesis project, whose source code, as well as this report, is available at \url{https://github.com/aarneranta/concept-alignment}.
+This appendix provides installation and usage instructions for the CA software developed as part of this thesis project, whose source code, as well as this report, is available at \url{github.com/harisont/concept-alignment}.
 
@@
 As mentioned in Section \ref{criteria}, modifying the alignment criteria the program makes use of requires little effort. 
-All criteria are in fact defined in a separate Haskell module \texttt{Criteria}\footnote{\url{https://github.com/aarneranta/concept-alignment/blob/master/Criteria.hs}} that exports only the list of criteria to be used.
+All criteria are in fact defined in a separate Haskell module \texttt{Criteria}\footnote{\url{https://github.com/harisont/concept-alignment/blob/master/Criteria.hs}} that exports only the list of criteria to be used.
 In the current implementation, it looks like this: \smallskip

### 2021-02-09 final_report/include/CAMT.tex
reviewed MT eval chapter

@@
 This allows us to construct lexical entries not only for the grammatical categories that typically correspond to single words (such as nouns, adjectives and adverbs), but also for phrases (with the term ``phrase'' now intended grammatically, as opposed to its meaning in SMT).
-Doing so enables GF translation, designed to preserve grammatically, to also render idiomatic expressions found in the data used for generating the lexicon correctly. \smallskip
+Doing so enables GF translation, designed to preserve grammaticality, to also render idiomatic expressions found in the data used for generating the lexicon correctly. \smallskip
 
@@
 \subsection{UD-to-GF conversion}
-The output of both CE and CP is, unless when linearized for evaluation purposes, a set of aligned UD trees in CoNNL-U format.
+The output of both CE and CP is, excepts when linearized for evaluation purposes, a set of aligned UD trees in CoNNL-U format. 
 In order to be used for GF-based translation, such trees need to be first converted into GF ASTs, which are then used to generate a set of grammar rules constiuting the above mentioned multilingual GF lexicon.
@@
 
-which specifies that a \texttt{PredVP} (predicative Verb Phrase), defined in the RGL as a function of type
+which specifies that a \texttt{PredVP} (predicative Verb Phrase), defined in the RGL as of type
 
@@
 The result of this UD-to-GF conversion is still a series of tree alignments, but their resulting format makes a substantial difference, since GF ASTs can be used to obtain the rules of a generative grammar.
-\texttt{gf-ud} also provides a way to do this, given an \textit{extraction grammar} and a \textit{morphological dictionary} of the languages of the alignments.
-\smallskip
+\texttt{gf-ud} also provides a way to do this, given an \textit{extraction grammar} and a \textit{morphological dictionary} of the languages of the alignments. \smallskip
 
@@
 A \textit{morphological dictionary}, on the other hand, contains correspondences between a large number of lemmas and various word forms. 
-In this case, the morphological dictionaries themselves are implemented in GF.
-\smallskip
+In this case, the morphological dictionaries themselves are implemented in GF, as part of the RGL. \smallskip
 
@@
 For instance, the determiner can change or disappear, number can vary, the complement of the copula can become a noun and an adjectival modifiers and prepositional phrases can be added to each noun phrase.
-For instance:
+Here are some examples:
 
-\begin{example} \ 
+\begin{example} \ \smallskip
     \begin{itemize}
@@
 
-\begin{example} \
+\begin{example} \ \smallskip
     \begin{itemize}
@@
 Section \ref{grgn})}
-    \item to facilitate the experiments, our standalone program can translate several newline-separated sentences at once.
+    \item to facilitate the experiments, our standalone has been written so that it can translate several newline-separated sentences at once.
 \end{itemize}
@@
 
-In this context, we focus on the syntax-level errors: attachment and labelling.
-Of course, lemmatization and POS-tagging errors can be the first cause of one such error.
+In this context, we focus on the syntax-level errors: attachment and labelling. Of course, lemmatization and POS-tagging errors can be the first cause of one such error. 
 In particular, the fact that a word is lemmatized incorrectly can make it so that it is also attributed the wrong POS tag and, as a consequence, assigned the wrong dependency label and/or attached to the wrong node.
@@
 As a consequence, vanilla CE would not be able to align the prepositional phase ``\textit{with care}'', attached to the main verb ``\textit{parse}'', to its Italian exact counterpart ``\textit{con cura}'', incorrectly linked to the noun ``\textit{frase}'' as an \texttt{nmod}.
-A second CE pass trying to put unaligned subtrees in relation with each other, however, is able to detect the POS-equivalence between the two phrases, thus producing three new correct alignments: $\langle$\textit{with care, con cura}$\rangle$, $\langle$\textit{care, cura}$\rangle$ and $\langle$\textit{with, con}$\rangle$.}
+A second CE pass trying to put unaligned subtrees in relation with each other, however, is able to detect the POS-equivalence between the two phrases, thus producing three new correct alignments: $\langle$\textit{with care, con cura}$\rangle$, $\langle$\textit{``care'', ``cura''}$\rangle$ and $\langle$\textit{``with'', ``con''}$\rangle$.}
     \label{rest}
@@
 On the other hand, the concepts represented in the GF grammar should intuitively, in most cases, be ``as small as possible''.
-If we reconsider the example in Figure \ref{cake} (sentences \textit{``finding useful correspondences is not exactly a piece of cake''} and \textit{``trovare corrispondenze utili non è proprio scontato''}), for instance, while the multi-word correspondence $\langle$\textit{a piece of cake, scontato}$\rangle$ is indeed useful there is usually no point in keeping correspondences such as $\langle$\textit{useful correspondences, corrispondenze utili}$\rangle$, even if they are correct, when it it also possible to extract the alignments $\langle$\textit{useful, utili}$\rangle$ and $\langle$\textit{correspondences, corrispondenze}$\rangle$: one-word and, in general, shorter alignments are more easily reusable in different contexts, especially since there are language-specific RGL rules capable of handling word order properly.
+If we reconsider the example in Figure \ref{cake} (sentences \textit{``finding useful correspondences is not exactly a piece of cake''} and \textit{``trovare corrispondenze utili non è proprio scontato''}), for instance, while the multi-word correspondence $\langle$\textit{``a piece of cake'', ``scontato''}$\rangle$ is indeed useful there is usually no point in keeping correspondences such as $\langle$\textit{``useful correspondences'', ``corrispondenze utili''}$\rangle$, even if they are correct, when it it also possible to extract the alignments $\langle$\textit{``useful'', ``utili''}$\rangle$ and $\langle$\textit{``correspondences'', ``corrispondenze''}$\rangle$: one-word and, in general, shorter alignments are more easily reusable in different contexts, especially since there are language-specific RGL rules capable of handling word order properly.
 As a consequence, for the purposes of grammar generation, the CA modules should in this case only return the following set of alignments: \smallskip
@@
 \begin{enumerate}
-    \item $\langle$\textit{finding, trovare}$\rangle$
-    \item $\langle$\textit{useful, utili}$\rangle$
-    \item $\langle$\textit{correspondences, corrispondenze}$\rangle$
-    \item $\langle$\textit{is, è}$\rangle$
-    \item $\langle$\textit{not, non}$\rangle$
-    \item $\langle$\textit{exactly, proprio}$\rangle$
-    \item $\langle$\textit{a piece of cake, scontato}$\rangle$.
+    \item $\langle$\textit{``finding'', ``trovare''}$\rangle$
+    \item $\langle$\textit{``useful'', ``utili''}$\rangle$
+    \item $\langle$\textit{``correspondences'', ``corrispondenze''}$\rangle$
+    \item $\langle$\textit{``is'', ``è''}$\rangle$
+    \item $\langle$\textit{``not'', ``non''}$\rangle$
+    \item $\langle$\textit{``exactly'', ``proprio''}$\rangle$
+    \item $\langle$\textit{``a piece of cake'', ``scontato''}$\rangle$.
 \end{enumerate} \smallskip
@@
 
-In some situations, however, larger alignments (such as $\langle$\textit{useful correspondences, corrispondenze utili}$\rangle$) are considered to be relevant since they help understand which words can be safely used in conjunction with each other.
+In some situations, however, larger alignments (such as $\langle$\textit{``useful correspondences'', ``corrispondenze utili''}$\rangle$) are considered to be relevant since they help understand which words can be safely used in conjunction with each other.
 For instance, while the Italian adjective ``\textit{brutto}'' (most often translated as ``\textit{ugly}'') is often used in conjunction with the word ``\textit{notizia}'' (``\textit{news}'') in expressions such as ``\textit{brutte notizie}'', in English the adjective ``\textit{bad}'' (as in ``\textit{bad news}'') is definitely more appropriate.
-Discarding a potential $\langle\textit{bad news}, \textit{brutte notizie}\rangle$ alignment is then not always the right choice, since in a context in which there are many alternative translation equivalents for the word ``\textit{brutto}'' it provides useful information about what adjective to use to refer to the noun ``\textit{news}''.
+Discarding a potential $\langle$\textit{``bad news''}, \textit{``brutte notizie''}$\rangle$ alignment is then not always the right choice, since in a context in which there are many alternative translation equivalents for the word ``\textit{brutto}'' it provides useful information about what adjective to use to refer to the noun ``\textit{news}''.
 \smallskip
@@
 
-Thus, the algorithm is safe - in the sense that it does not get rid of any potentially meaningful alignment - in that it gets rid of superficially perfect alignments whose subtrees are also members of smaller alignments.
+Thus, the algorithm is ``safe'' in the sense that it does not get rid of any potentially meaningful alignment. 
+In fact, it only gets rid of superficially perfect alignments whose subtrees are also members of smaller alignments. 
 This means that when sentences vary widely syntax-wise, as it happens in cases where translation is not literal or in the presence of parse errors, the resulting extracted concepts can still contain a large number of nodes.
@@
 
-In a preliminary experiment conducted in this way, for instance, the Italian translation ``\textit{il docente discute la Didattica}'' of the English sentence ``\textit{the teacher discusses the didactics}'', was given the minimum score despite being completely correct and even idiomatic\footnote{typically, the adjective ``didattica'' is not use in its plural form ``didattiche'' when nominalized.
+In a preliminary experiment conducted in this way, for instance, the Italian translation ``\textit{il docente discute la Didattica}'' of the English sentence ``\textit{the teacher discusses the didactics}'', was given the minimum score despite being completely correct and even idiomatic\footnote{typically, the adjective ``\textit{didattica}'' is not used in its plural form ``\textit{didattiche}'' when nominalized.
 The capital D is often used, for this particular term, in official university documents.}.
-Looking at the two reference translations used at the time, it appears clear that such a low score is due to the capitalization of the word ``Didattica'' and the arbitrary (but equally correct) lexical choices made by the human translators, who rendered the sentence as ``il professore discute i metodi didattici'' and ``l' insegnante discute la didattica''\footnote{The terms ``docente'', ``professore'' and ``insegnante'' are almost perfect synonyms, even though ``docente'' is mostly used in formal documents to refer to university lecturers, ``professore'' is more common in secondary school and colloquial speech, and ``insegnante'' is the most generic.
+Looking at the two reference translations used at the time, it appears clear that such a low score is due to the capitalization of the word ``\textit{Didattica}'' and the arbitrary (but equally correct) lexical choices made by the human translators, who rendered the sentence as ``\textit{il professore discute i metodi didattici}'' and ``\textit{l' insegnante discute la didattica}''\footnote{The terms ``\textit{docente}'', ``\textit{professore}'' and ``\textit{insegnante}'' are almost perfect synonyms, even though ``\textit{docente}'' is mostly used in formal documents to refer to university lecturers, ``\textit{professore}'' is more common in secondary school and colloquial speech, and ``\textit{insegnante}'' is the most generic.
 As such, it can also refer to elementary school teachers and in some cases even sport coaches.
-The expression ``metodi didattici'' is a perfectly sensible replacement for ``didattica''.}.
+The expression ``\textit{metodi didattici}'' is a perfectly sensible replacement for ``\textit{didattica}''.}.
 Coincidentally (but not unlikely for a language like Italian), this is a case in which these lexical differences affect the choice of the determiners used in the various translations (cf.
-``\underline{l'}insegnante'' vs.
-``\underline{il} professore'' and ``\underline{la} didattica'' vs.
-``\underline{i} metodi didattici''), producing a BLEU-1 to 4 score of 0.
+``\textit{\underline{l'}insegnante}'' vs.
+``\textit{\underline{il} professore}'' and ``\textit{\underline{la} didattica}'' vs.
+``\textit{\underline{i} metodi didattici}''), producing a BLEU-1 to 4 score of 0.
 \smallskip
@@
 When it comes to the results obtained for CE on the course plans corpora in particular (cf. Table \ref{tcereal}), however, the difference in precision between the two language pairs is negligible.
-This makes the reason for the substantial difference observed in the BLEU scores obtained for the two different corpora unclear, at least at a first glance.\smallskip
+This makes the reason for the substantial difference in the BLEU scores obtained for the two different corpora unclear, at least at a first glance.\smallskip
 
@@
 This is the case, for instance, of the sentence ``\textit{the test is oral}'', whose last word, ``\textit{oral}'' is translated as ``\textit{dura}'' (``\textit{hard}'') instead of ``\textit{orale}'' due to an alignment error.
-Nonetheless, it is worth noticing that this is one of the many cases in which, even though the fact that only the first translation candidate is taken into account hides it, the correct alignment $\langle$``\textit{oral}'',``\textit{orale}''$\rangle$ is also found by the CE module.
-Moreover, the such valid correspondence has a higher number of occurrences and matches stronger alignment criteria than the wrong one $\langle$``\textit{oral}'',``\textit{duro}''$\rangle$. 
+Nonetheless, it is worth noticing that this is one of the many cases in which the correct alignment $\langle$``\textit{oral}'',``\textit{orale}''$\rangle$ is also found by the CE module. 
+This is hidden by the fact that, for simplicity, only the first translation candidate is taken into account.
+Moreover, such valid correspondence has a higher number of occurrences and matches stronger alignment criteria than the wrong one $\langle$``\textit{oral}'',``\textit{duro}''$\rangle$. 
 This suggest that, in contexts where higher precision is necessary, there are obvious improvements that can be done either in the choice of the alignment criteria or at a later stage, when relevant alignments are selected.
@@
 ``\textit{relevant\underline{are}}'' instead of ``\textit{mer relevant}''), less systematic.
-Only in one case, for instance, (``\textit{programbibliotek\underline{en}}''), gender is incorrect. \smallskip
+Only in one case, for instance, gender is incorrect (``\textit{programbibliotek\underline{en}}''). \smallskip
 
-Some alignment errors are also interesting to analyze.
-In English-Swedish, while several compounds, such as $\langle$\textit{computer science, datavetenskap}$\rangle$, are aligned correctly, there are cases in which a single-root noun is translated as a compound (e.g.
-$\langle$ \textit{theory, automatateori} $\rangle$).
-This is not necessarily an actual alignment error, as it might rather be that the English version of the text was being less specific than its Swedish counterpart, thus producing an alignment that, during evaluation, would have been marked as correct but not generally reusable (\texttt{=}).
+Some errorsregarding the extracted concepts are also interesting to analyze.
+In English-Swedish, while several compounds, such as $\langle$\textit{``computer science'', ``datavetenskap''}$\rangle$, are aligned correctly, there are cases in which a single-root noun is translated as a compound (e.g.
+$\langle$ \textit{``theory'', ``automatateori''} $\rangle$).
+This is not necessarily an actual alignment error, as it might rather be that the English version of the text was being less specific than its Swedish counterpart, thus producing an alignment that, during evaluation, would have been marked as ``correct but not reusable'' (\texttt{=}).
 \smallskip
@@
 ``Reasonable'' alignment errors appear in the DMI corpus sentences too. 
-The English for ``\textit{attendance of lessons}'', for instance, becomes simply ``\textit{frequenza}'' in the Italian translation. 
-Such word is in fact often used alone to replace longer expressions such as ``\textit{frequenza delle lezioni}'', just like in English `\textit{of lessons}'' can be omitted when what must be attended is evident from the context.
-Another interesting example is the alignment $\langle$ \textit{class, classe} $\rangle$, which causes the sentence \textit{``I will attend the class''} to be (incorrectly) translated as \textit{``io seguirò la classe''} instead of \textit{``io seguirò la lezione''} even though the correspondence is in fact valid in most of the numerous contexts in which \textit{``class''} is not to be intended as a synonym of \textit{``lesson''}.
+The English phrase ``\textit{attendance of lessons}'', for instance, becomes simply ``\textit{frequenza}'' in the Italian translation. 
+Such word is in fact often used alone to replace longer expressions such as ``\textit{frequenza delle lezioni}'', just like in English ``\textit{of lessons}'' can be omitted when what must be attended is evident from the context.
+Another interesting example is the alignment $\langle$\textit{``class'', ``classe''}$\rangle$, which causes the sentence \textit{``I will attend the class''} to be (incorrectly) translated as \textit{``io seguirò la classe''} instead of \textit{``io seguirò la lezione''} even though the correspondence is in fact valid in most of the numerous contexts in which \textit{``class''} is not to be intended as a synonym of \textit{``lesson''}.

### 2021-02-09 final_report/include/CP.tex
reviewed MT eval chapter

@@
     \textbf{}              & \textbf{en-sv} & \textbf{it-sv} \\ \hline
-    propagated             & 1019 (85.05\%) & 979 (84.64\%) \\ \hline
-    tot. errors            & 133 (13.05\%)  & 187 (19.1\%)  \\ \hline
-    CP-introduced          & 75 (56.39\%)   & 84 (44.91\%)   \\ \hline
+    \textbf{propagated}             & 1019 (85.05\%) & 979 (84.64\%) \\ \hline
+    \textbf{tot. errors}            & 133 (13.05\%)  & 187 (19.1\%)  \\ \hline
+    \textbf{CP-introduced}          & 75 (56.39\%)   & 84 (44.91\%)   \\ \hline
     \end{tabular}
@@
     \textbf{}                     & \textbf{en-it-sv} & \textbf{it-en-sv} & \textbf{en-sv-it} & \textbf{sv-en-it} & \textbf{it-sv-en} & \textbf{sv-it-en} \\ \hline
-    extracted   & 638               & 638               & 687               & 687               & 608               & 608               \\ \hline
-    propagated                    & 92 (14.42\%)      & 92 (14.42\%)      & 98 (14.26\%)      & 84 (12.22\%)      & 101 (16.61\%)     & 87 (14.37\%)      \\ \hline
-    tot. errors                   & 46 (50\%)         & 21 (22.82\%)      & 42 (42.85\%)      & 24 (28.57\%)      & 21 (20.79\%)      & 28 (32.18\%)      \\ \hline
-    CP-introduced                 & 33 (71.73\%)      & 11 (52.38\%)      & 21 (50\%)         & 12 (50\%)         & 12 (57.14\%)      & 21 (75\%)         \\ \hline
+    \textbf{extracted}   & 638               & 638               & 687               & 687               & 608               & 608               \\ \hline
+    \textbf{propagated}                    & 92 (14.42\%)      & 92 (14.42\%)      & 98 (14.26\%)      & 84 (12.22\%)      & 101 (16.61\%)     & 87 (14.37\%)      \\ \hline
+    \textbf{tot. errors}                   & 46 (50\%)         & 21 (22.82\%)      & 42 (42.85\%)      & 24 (28.57\%)      & 21 (20.79\%)      & 28 (32.18\%)      \\ \hline
+    \textbf{CP-introduced}                 & 33 (71.73\%)      & 11 (52.38\%)      & 21 (50\%)         & 12 (50\%)         & 12 (57.14\%)      & 21 (75\%)         \\ \hline
     \end{tabular}
@@
                            & \textbf{sv-en-it} & \textbf{it-en-sv} \\ \hline
-    extracted concepts     & 1950              & 1823              \\ \hline
-    propagated             & 205 (10.51\%)     & 200 (10.97\%)     \\ \hline
-    tot. errors            & 66 (32.19\%)      & 61 (30.5\%)       \\ \hline
-    errors introduced by CP & 33 (50\%)        & 33 (54.09\%)      \\ \hline
+    \textbf{extracted}     & 1950              & 1823              \\ \hline
+    \textbf{propagated}             & 205 (10.51\%)     & 200 (10.97\%)     \\ \hline
+    \textbf{tot. errors}            & 66 (32.19\%)      & 61 (30.5\%)       \\ \hline
+    \textbf{CP-introduced} & 33 (50\%)        & 33 (54.09\%)      \\ \hline
     \end{tabular}

### 2021-02-09 final_report/include/CAMT.tex
reviewed conclusions and minor changes in MT eval chapter

@@
 
-Because of this, and since CE could have applications other than MT, selection in this sense is performed - optionally - after the extraction phase itself and implemented as an independent function.
+Because of this, and since CE could have applications other than MT, selection in this sense is performed - optionally, as described in Appendix \ref{b} - after the extraction phase itself and implemented as an independent function.
 It works as follows: first, alignments that do not present any content word are filtered out.

### 2021-02-09 final_report/include/Conclusions.tex
reviewed conclusions and minor changes in MT eval chapter

@@
 For the MT evaluation, such dependency trees are converted to Abstract Syntax Trees, from which the rules of a multilingual generative phrase-structure grammar are derived.
-The framework we make use of when it comes to this grammar, GF, makes having a grammar sufficient to perform simple MT experiments.\smallskip
+The framework we make use of to this extent, GF, makes the grammar itself sufficient to perform simple MT experiments.\smallskip
 
-The tangible fruit of this work is a Haskell library, as well as a number of executables offering an easy to use and configure interface to perform both variants of CA, extraction and propagation, and a variety of kinds of evaluations.
+The tangible fruits of this work are a Haskell library and a number of executables, whose usage is described in Appendix \ref{b}, offering an easy to use and configure interface to perform both variants of CA, extraction and propagation, and a variety of kinds of evaluations.
 \smallskip
@@
 When it comes to CE, the core part of this work, comparison against a pre-existing basic implementation shows significant improvements in terms of both the quality and the quantity of the resulting alignments.
-With respects to traditional statistical approaches, one of the most useful consequences of our syntax-based approach is that it makes it possible to identify correspondences between multiword, even discontinuous expressions.
+With respects to traditional statistical approaches, one of the most useful consequences of adopting a syntax-based approach is that it makes it possible to identify correspondences between multiword expressions, even when discontinuous.
 Another important advantage over standard word alignment techniques is that, even though it can still benefit from large amounts of data, our system works consistently well even when run on extremely small corpora.
-This is confirmed by one of our experiments, where we restrict it so to only extract one-to-many and many-to-one alignments and compare it with \texttt{fast\_align}.
+This is confirmed by one of our experiments, where we compare it with \texttt{fast\_align}.
 Furthermore, the alignment criteria our software makes use of are easy to modify, add and remove, paving the way to a variety of experiments. \smallskip
@@
 On the other, it can also be used to identify shared terminology between two bilingual corpora having one language in common.
-The first experiments show promising results, but suffer from the scarce availability of sentence-segmented large bilingual corpora.
+The first experiments show promising results, but suffer from the scarce availability of sentence-segmented large multilingual corpora.
 \smallskip
@@
 \section*{Future work} \label{future}
-These results, while encouraging, suggest that there is still much room for improvement and that work in this area can be carried on in many different directions even constaining ourselves to CA \textit{per se}, without focussing on its MT applications.
+These results, while encouraging, suggest that there is still much room for improvement in many different directions, even if we were to restrict ourselves to CA \textit{per se}, without focussing on its MT applications.
 
@@
 With regards to the CA component itself, it is important to remember that both the quantity and the quality of the alignments are a direct consequence of those of the dependency trees they are obtained from.
-Apart from working on dependency parsers directly, a possibility is to experiment more with hybrid alignment approaches, integrating the results of statistical methods (which only depend on the quantity and quality of raw data rather than on that of their analyses) with those of our rule-based system.
-One way to do that has already been implemented, but has not been made use of in the final MT experiments due to the small size of the corpora involved.
+Apart from working on dependency parsers directly, a possibility is to experiment more with hybrid alignment approaches, integrating the results prodiced by statistical tools (which only depend on the quantity and quality of raw data rather than also on that of their analyses) with those of our rule-based system.
+A basic way to do that has already been implemented, but has not been made use of in the final MT experiments due to the small size of the corpora involved.
 
 \subsection*{Aligning verb phrases}
-Early in this work we noticed how only aligning subtrees is not enough to identify all the concepts that are present in a pair of sentences.
-As a consequence, significant effort has been made in also aligning subtree heads correctly.
+Early in this work we noticed how only aligning subtrees is not enough to identify all the concepts present in a pair of sentences.
+As a consequence, significant effort has been made to also align subtree heads correctly.
 While this increases the number of concepts the program is able to find, it does not cover all cases.
@@
 verbs together with their arguments) could prove extremely useful.
-Generalizing the approach used for head alignment could be a way of addressing the issue and would make it possible to use CA to construct ``rich'' lexical entries for verbs, whose lack, as discussed in Chapter \ref{ch3}, currently makes it hard to handle some types of common translations divergences.
+Generalizing the approach used for head alignment could be a way of addressing this problem.
+This would make it possible to use CA to construct ``rich'' lexical entries for verbs, whose lack, as discussed in Chapter \ref{ch3}, currently makes it hard to handle some types of common translations divergences.
 
 \subsection*{Iterative CA}
-The current implementations of CE and CP try to apply a set of priority-sorted criteria one after another to each pair of sentences in the corpus, allowing for later ranking of the alignments obtained based on the criteria their members match.
+The current implementations of CE and CP try to apply a set of priority-sorted criteria one after another to each pair of sentences in the corpus, allowing for later ranking of the alignments obtained based on the criteria they match.
 Another possibility is to go through the entire corpus multiple times, for instance using only the strictest criteria first and only falling back to the less reliable ones at later iterations, until no more alignments are found.
-The same hypothetical iterative algorithm can have the size of the alignments, and not (or not only) the criteria they have to satisfy, as a parameter.
+The same hypothetical iterative algorithm could also have the size of the alignments, and not (or not only) the criteria they have to satisfy, as a parameter.
 
@@
 As an alternative to this optimization, CE could be generalized to an arbitrary number of languages.
-The expectation is for the resulting concepts to be of slightly larger size than those obtained with the current system, since comparing more trees simultaneously increases the possibility of encountering small divergences, this term being intended in its broad sense.
-This approach would probably help identifying longer idiomatic expression and could provide a better way to deal with multilingual corpora, even though, while intuitive, it does present some implementation challenges.
+The expectation is for the resulting concepts to be of slightly larger size than those obtained with the current system, since comparing more trees simultaneously increases the possibility of encountering small divergences, with the term ``divergence'' intended in its broader sense.
+This approach would probably help identifying longer idiomatic expression and could provide a better way to deal with multilingual corpora.
+While intuitive, however, it does present some implementation challenges.
 
@@
 When it comes to using the concepts as actual translation units, getting rid of incorrect alignments is crucial.
-While doing so manually is still more feasible than extracting them manually, it would be desirable to devise better alignment selection policies or stricter alignment criteria.
-One idea would is to tune the alignment criteria for the language pair at hand, both by removing the noisier ones for semantically related languages and by adding language-pair specific criteria.
+While manual selection is still more feasible than manual extraction, it would be desirable to devise better alignment selection policies or stricter alignment criteria.
+One idea would is to tune the alignment criteria for the language pair(s) at hand, both by removing the noisier ones for semantically related languages and by adding language-pair specific criteria.
 The criteria-specific statistics reported in Section \ref{eval2} can serve as a starting point for work in this direction.

### 2021-02-09 final_report/include/CE.tex
reviewed all appendices

@@
 Consequently, we refer to the simpler classification and less formal definitions proposed in \cite{divs2}, where promotional and demotional divergences are merged in a wider-coverage category of \textit{head-swapping} divergences and where lexical divergences, the hardest to handle in the absence of any kind of semantic information, do not appear at all.
-In the following, we give definitions and examples for each of these classes of divergences, specifying to what extent and how\footnote{unless otherwise specified, each divergence can be expressed as a set of simple rules in the form of boolean functions taking the two UD trees candidate for alignment as input.} each of them is handled in the proposed language-agnostic CE module.
+In the following, we give definitions and examples for each of these classes of divergences, specifying to what extent and how\footnote{unless otherwise specified, each divergence can be expressed as a set of simple rules in the form of boolean functions taking the pair of UD trees candidate for alignment as input.
+An example of this can be found in Appendix \ref{cconf}.} each of them is handled in the proposed language-agnostic CE module.
 We do not attempt to cover all possible cases, but rather provide a few examples as a proof of concept.

### 2021-02-09 final_report/include/backmatter/Appendix_2.tex
reviewed all appendices

@@
 \section{Installation}
-To compile the CA module, we recommend the Haskell Stack\footnote{\url{https://docs.haskellstack.org/en/stable/README/}}. 
+To compile the CA module, we recommend the Haskell Stack\footnote{\url{docs.haskellstack.org}}. 
 To build the project, clone the above mentioned GitHub repository, move into the corresponding directory and run \texttt{stack build}. 
-This will generate a haskell library, named \texttt{concept-alignment}, and the following five executables: \smallskip
+This will generate a Haskell library, named \texttt{concept-alignment}, and the following five executables: \smallskip
 
@@
     \item \texttt{--rest} enables a second pass of alignment ``by exclusion'' (cf. Section \ref{exclusion}) 
-    \item \texttt{--pharaoh=FILE} is used to specify a file in pharaoh format (cf. Figure \ref{wordalgn}) to use as backup
+    \item \texttt{--pharaoh=FILE} is used to specify a file in pharaoh format (cf. Figure \ref{wordalgn}) to use as backup.
 \end{itemize}
@@
 \subsection{\texttt{propagate-concepts}}
-CP requires two, instead of three, \texttt{.conllu} files, one with the concepts to propagate and the following two containing the corpus of annotated sentences where their should be looked for.
-As a consequence, running it with the default parameters should be done as follows: \smallskip
+CP requires two, instead of three, \texttt{.conllu} files, one with the concepts to propagate and the following two containing the corpus of annotated sentences where they should be looked for.
+As a consequence, running it with the default parameters is as follows: \smallskip
 
@@
 \subsection{\texttt{evalign}}
-The evaluation scrip \texttt{evalign} can be run in three different ``modes'': \smallskip
+The evaluation script \texttt{evalign} can be run in three different ``modes'': \smallskip
 
 \begin{enumerate}
-    \item single-file mode (\texttt{stack exec -- evalign annotated.txt}): given a pre-annotated file in the linearized format described in Section \ref{evalign}, it prints out basic statistics about precision, recall and amount of reusable alignments.
-    \item extraction interactive mode (\texttt{stack exec -- evalign extraction annotated.txt new.txt}): given an annotated and a new, possibly yet-to-annotate file containing linearized alignments, it allows interactive minimal annotation of the latter (if needed) and, on top of printing out the basic statitics, it compares the new alignments to the old ones, telling how many correct and incorrect alignments were lost and/or found
-    \item propagation interactive mode (\texttt{stack exec -- evalign propagation annotated.ca new.ca}): similar to extraction interactive mode, excepts that the statistics are CP-specific (percentage of successfully propagated alignments, number of errors introduced by CP etc.)
+    \item single-file mode (\texttt{stack exec -- evalign linearized.txt}): \\given a pre-annotated file in the linearized format described in Section \ref{evalign}, it prints out basic statistics about precision, recall and amount of reusable alignments.
+    \item extraction mode (\texttt{stack exec -- evalign extraction old.txt new.txt}): \\given an annotated and a new, possibly yet-to-annotate file containing linearized alignments, it allows interactive minimal annotation of the latter (if needed) and, on top of printing out the basic statistics, it compares the new alignments to the old ones, telling how many correct and incorrect alignments were lost and/or found
+    \item propagation mode (\texttt{stack exec -- evalign propagation new.txt new.txt}): similar to extraction interactive mode, excepts that the statistics are CP-specific (percentage of successfully propagated alignments, number of errors introduced by CP etc.).
 \end{enumerate} \smallskip
@@
 \begin{verbatim}
-    stack exec -- GenerateGrammar path_to_extract_grammar 
+    stack exec -- generate-grammar path_to_extract_grammar 
     paths_to_morphodicts paths_to_aligned_conllu_files
@@
 \texttt{paths\_to\_morphodicts} stands for the path of the morphological dictionaries of the languages involved, also in \texttt{.pgf} format.
+
 Similarly, \texttt{paths\_to\_aligned\_conllu\_files} is to be replaced with the files containing the concepts in CoNNL-u format.
@@
 
-\section{Configuration: modifying the alignment criteria}
-As mentioned in Section \ref{criteria}, modifying the alignment criteria the program makes use of requires little effort. 
-All criteria are in fact defined in a separate Haskell module \texttt{Criteria}\footnote{\url{https://github.com/harisont/concept-alignment/blob/master/Criteria.hs}} that exports only the list of criteria to be used.
+\section{Configuration: modifying the alignment \\criteria} \label{cconf}
+As mentioned in Section \ref{criteria}, modifying the criteria CA makes use of requires little effort. 
+All criteria are in fact defined in a separate Haskell module \texttt{Criteria}\footnote{\url{https://github.com/harisont/concept-alignment/blob/master/Criteria.hs}} that exports the list of criteria to be used only.
 In the current implementation, it looks like this: \smallskip
@@
 
-Adding new criteria is also simple, but it requires an understanding of the data type \texttt{Criterion}, defined in module \texttt{ConceptAlignment}:
+Adding new criteria is simple too, but it also requires an understanding of the data type \texttt{Criterion}, defined in module \texttt{ConceptAlignment}:
 
@@
 \begin{itemize}
-    \item a boolean function \texttt{func} specifying a rule to decide whether two trees should be aligned. For instance Criterion \ref{udmatch} (UD label matching) is implemented as
+    \item a boolean function \texttt{func} specifying a rule to decide whether two trees should be aligned. For instance, Criterion \ref{udmatch} (UD label matching) is implemented as
     \begin{lstlisting}
@@
     \end{lstlisting}
-    \item a set of \texttt{Reason} \texttt{reas} to be used for ranking the alignments, shown in the linearized files and used to compute criterion-specific statistics.
-Reasons, and not alignment rules (\texttt{func}s), are what mirrors criteria as defined in Section \ref{criteria} exactly.
+    \item a set of ``reasons'' \texttt{reas} to be used for ranking the alignments, shown in the linearized files and used to compute criterion-specific statistics.
+\texttt{Reason}s, and not alignment rules (\texttt{func}s), are what mirrors alignment criteria (as they are defined in Section \ref{criteria}) exactly.
 A \texttt{func} can define in fact a more specific alignment rule, such as ``type of categorial divergence in which an adjective is replaced by adverb'', while a \texttt{Reason} is in practice a more coarse-grained label associated to potentially many rules
-    \item two boolean flags, \texttt{headAlign} and \texttt{strict}, the former specifying whether head alignment should be performed for UD tree pairs matching that criterion, the latter marking the criterion as either strict (i.e.
+    \item two boolean flags, \texttt{headAlign} and \texttt{strict}, the former specifying whether head alignment (cf.
+Section \ref{heads}) should be performed for UD tree pairs matching that criterion, the latter marking the criterion as either strict (i.e.
 to be also used for ``alignment by exclusion'', cf.

### 2021-02-09 final_report/include/CE.tex
checked bibliography entries + minor changes

@@
 This notation will be used for the other examples of this chapter.}
-  \label{deptree}
  \end{figure}
@@
 Section \ref{upos}), but this term was deliberately avoided in Definition \ref{poseq} due to the fact that it can be useful also to include some function words, for instance pronouns and some kinds of determiners.
-In particular, the current implementation considers as meaning-carrying all words that belong to an open class - defined as in the UD documentation \cite{openpos} - and numerals, but this set of tags has been obtained empirically working with English-Italian and might not be ideal for all language pairs.
+In particular, the current implementation considers as meaning-carrying all words that belong to an open class - defined as in the UD documentation \cite{uddocs} - and numerals, but this set of tags has been obtained empirically working with English-Italian and might not be ideal for all language pairs.
 On the other hand, words belonging to other classes, such as auxiliary verbs, adpositions and conjunctions can and should be in most cases ignored as they are often omitted or rendered with words with different POS tags when the sentence is translated to another language, especially if the two languages in the pair at hand differ significantly.

### 2021-02-09 final_report/include/frontmatter/Abstract.tex
of course there was a typo in the abstract

@@
 This thesis presents a syntax-based approach to Concept Alignment (CA), the task of finding semantical correspondences between parts of multilingual parallel texts, with a focus on Machine Translation (MT).
-Two variants of CA are taken into account: Concept Extraction (CE), whose aim is to identify new concepts my means of mere linguistic comparison, and Concept Propagation (CP), which consists in looking for the translation equivalents of a set of known concepts in a new language.
+Two variants of CA are taken into account: Concept Extraction (CE), whose aim is to identify new concepts by means of mere linguistic comparison, and Concept Propagation (CP), which consists in looking for the translation equivalents of a set of known concepts in a new language.
 As opposed to standard statistical alignment methods, our approach allows to simultaneously align individual words and multiword expressions (even discontinuous). 

### 2021-03-28 final_report/include/CAMT.tex
minor changes
capitalized footnotes

@@
 Conversion is based on annotations defining the relation between UD trees and GF ASTs, which we refer to as \textit{dependency configurations}. 
-While going into all the notational details is beyond the scope of this project\footnote{a comprehensive description of \texttt{gf-ud}'s annotation scheme is available at \url{github.com/GrammaticalFramework/gf-ud/blob/master/doc/annotations.md}.}, an example of one such configuration is
+While going into all the notational details is beyond the scope of this project\footnote{A comprehensive description of \texttt{gf-ud}'s annotation scheme is available at \url{github.com/GrammaticalFramework/gf-ud/blob/master/doc/annotations.md}.}, an example of one such configuration is
 
@@
 Called every time a UD subtree has to be converted to CoNNL-U, it first replaces the UD label (\texttt{udDEPREL}) of its root node with \texttt{root} and assigns the special value 0 to its \texttt{udHEAD} field.
-After that, it proceeds to ``rescale'' the \texttt{udID} and \texttt{udHEAD} fields of all other nodes so that they end up into the range $[1,n]$, where $n$ is the number of nodes of the extracted subtrees\footnote{see Section \ref{conll} and Figure \ref{types} for definitions of all the fields of UD trees and their CoNNL-U representations.}.
+After that, it proceeds to ``rescale'' the \texttt{udID} and \texttt{udHEAD} fields of all other nodes so that they end up into the range $[1,n]$, where $n$ is the number of nodes of the extracted subtrees\footnote{See Section \ref{conll} and Figure \ref{types} for definitions of all the fields of UD trees and their CoNNL-U representations.}.
 \smallskip
@@
 This is due to the fact that conversion in this direction is a nondeterministic search problem \cite{udgf}. 
-For this reason, it can be useful to limit the size\footnote{in this context, the size of an alignment is defined as the number of nodes of its largest member.} of the alignments that are actually kept after the extraction (or propagation) procedure.
+For this reason, it can be useful to limit the size\footnote{In this context, the size of an alignment is defined as the number of nodes of its largest member.} of the alignments that are actually kept after the extraction (or propagation) procedure.
 Since our CA component works at all levels of abstraction, thus also aligning full clauses, this turned out to be essential and became one of the command-line parameters of the final executables, as mentioned in Appendix \ref{b}.
@@
 \begin{example}
-    ``Won't these short sentences be simpler than that long sentence?''
+    ``won't these short sentences be simpler than that long sentence?''
 \end{example}
@@
 \begin{itemize}
-    \item while GF can parse the SL string in a variety of ways and, as a consequence, return multiple TL linearizations, our program only outputs the first candidate translation, preliminarily discarding trees containing terminal nodes that have no linearization in the TL\footnote{a linearization can be missing, for instance, if a concrete syntax is postprocessed by simply getting rid of an ``illegal'' verb causing compilation problems (cf.
+    \item while GF can parse the SL string in a variety of ways and, as a consequence, return multiple TL linearizations, our program only outputs the first candidate translation, preliminarily discarding trees containing terminal nodes that have no linearization in the TL\footnote{A linearization can be missing, for instance, if a concrete syntax is postprocessed by simply getting rid of an ``illegal'' verb causing compilation problems (cf.
 Section \ref{grgn})}
@@
 To address this problem, the CE module has been modified so that it can try to align individual clauses instead of sentences. 
-The idea is to obtain, from each pair of sentence trees to align, a pair of lists of subtrees whose heads are labelled \texttt{root}, \texttt{csubj}, \texttt{ccomp}, \texttt{xcomp}, \texttt{advcl} or  \texttt{acl}\footnote{all these clausal labels are defined in Appendix \ref{a_lab}}.
+The idea is to obtain, from each pair of sentence trees to align, a pair of lists of subtrees whose heads are labelled \texttt{root}, \texttt{csubj}, \texttt{ccomp}, \texttt{xcomp}, \texttt{advcl} or  \texttt{acl}\footnote{All these clausal labels are defined in Appendix \ref{a_lab}}.
 
@@
 
-These lists of ``clauses''\footnote{``clauses'' is in quotes since what we are referring to as a clause includes its subclauses too.
+These lists of ``clauses''\footnote{Note that ``clauses'' is in quotes since what we are referring to as a clause includes its subclauses too.
 This means that the sentence is not segmented into clauses in the literal sense of the term, but rather that one tree per verb is extracted from them.} are then sorted by dependency label and distance from the root, so that in the absence of annotation errors, when there is in fact an evident correspondence, they appear in the same position.
@@
 
-On the one hand, most alignments that do not contain any content word are not useful for the generation of a domain-specific grammar, as many common function words are already covered by the GF RGL\footnote{small alignments containing both function and content words, such as prepositional phrases, are on the other hand extremely useful, as they show which specific function word should be used in a particular context.}.
+On the one hand, most alignments that do not contain any content word are not useful for the generation of a domain-specific grammar, as many common function words are already covered by the GF RGL\footnote{Small alignments containing both function and content words, such as prepositional phrases, are on the other hand extremely useful, as they show which specific function word should be used in a particular context.}.
 \smallskip
@@
 
-In a preliminary experiment conducted in this way, for instance, the Italian translation ``\textit{il docente discute la Didattica}'' of the English sentence ``\textit{the teacher discusses the didactics}'', was given the minimum score despite being completely correct and even idiomatic\footnote{typically, the adjective ``\textit{didattica}'' is not used in its plural form ``\textit{didattiche}'' when nominalized.
+In a preliminary experiment conducted in this way, for instance, the Italian translation ``\textit{il docente discute la Didattica}'' of the English sentence ``\textit{the teacher discusses the didactics}'', was given the minimum score despite being completely correct and even idiomatic\footnote{Typically, the adjective ``\textit{didattica}'' is not used in its plural form ``\textit{didattiche}'' when nominalized.
 The capital D is often used, for this particular term, in official university documents.}.
@@
 In Italian, grammar errors often involve contractions such as ``\textit{del}'' (``\textit{di}'' + ``\textit{il}'', in English ``\textit{of the}''), some of which are systematically rendered as two separate words.
-Another common case is that of wrong (or at least very confusing) adjective collocation, such as in the translation ``\textit{il libro presenta una tecnica con miglioramenti vari utile}'' (``\textit{the textbook presents a useful technique with various improvements}''), where the adjective ``\textit{utile}'' (``\textit{useful}''), referred to ``\textit{technique}'' (``\textit{tecnica}'') is placed far from such noun, making the sentence hard to interpret\footnote{the manually postprocessed translation is ``\textit{il libro presenta una utile tecnica con miglioramenti vari}''.}.
+Another common case is that of wrong (or at least very confusing) adjective collocation, such as in the translation ``\textit{il libro presenta una tecnica con miglioramenti vari utile}'' (``\textit{the textbook presents a useful technique with various improvements}''), where the adjective ``\textit{utile}'' (``\textit{useful}''), referred to ``\textit{technique}'' (``\textit{tecnica}'') is placed far from such noun, making the sentence hard to interpret\footnote{The manually postprocessed translation is ``\textit{il libro presenta una utile tecnica con miglioramenti vari}''.}.
 Grammatical errors in Swedish are less common and, apart from long adjectives never being turned to the comparative degree periphrastically (e.g.

### 2021-03-28 final_report/include/CE.tex
minor layout changes
minor changes
capitalized footnotes

@@
 
-Obviously, the ``full-sentence'' tree pair is itself an alignment, even though a trivial one, as it represents a sentence-level correspondence in a case where, due to the approach being syntactic comparison, the inputs are assumed to be sentences that correspond to each other\footnote{contrary to what it may seem, however, sentence-level alignment is in general not at a trivial task, as it is often the case, especially in certain language pairs, that one \textit{orthographic} sentence, i.e.
+Obviously, the ``full-sentence'' tree pair is itself an alignment, even though a trivial one, as it represents a sentence-level correspondence in a case where, due to the approach being syntactic comparison, the inputs are assumed to be sentences that correspond to each other\footnote{Contrary to what it may seem, however, sentence-level alignment is in general not at a trivial task, as it is often the case, especially in certain language pairs, that one \textit{orthographic} sentence, i.e.
 a sentence defined based on the presence of a full stop, maps to more than one orthographic sentences in the translated text, or vice versa.}.
@@
 While one of the most important ideas in the improved version of the CE module proposed in this thesis remains aligning (sub)trees with identical root UD labels, using this criterion alone makes our baseline both unable to detect many of the existing correspondences and prone to extract incorrect ones.
-This happens both when syntactic similarity is only apparent and when words are aligned based exclusively on the fact that they are the roots of two aligned subtrees\footnote{see Section \ref{heads} for a more detail discussion on head alignment.}.
+This happens both when syntactic similarity is only apparent and when words are aligned based exclusively on the fact that they are the roots of two aligned subtrees\footnote{See Section \ref{heads} for a more detail discussion on head alignment.}.
 As a consequence, the objective of this part of the project is to improve both the precision and the recall of the algorithm (or rather, for reasons that will be discussed in Section \ref{metrics}, two approximations of such metrics).
@@
 In \cite{divs}, seven classes of divergences are identified: \textit{thematic}, \textit{promotional}, \textit{demotional}, \textit{structural}, \textit{conflational}, \textit{categorial} and \textit{lexical}.
-However, because the author's proposed way to resolve divergences assumes the availability of more than merely syntactic information\footnote{in particular, of a lexicon of Root Lexical Conceptual Structures specifying, for instance, the logical subject, arguments and modifiers of each verb.} and because, working with UD trees, we do not have access to anything but strictly syntactical and morphological annotations, the very task of identifying these distinctions at this level of granularity becomes, in our position, extremely challenging.
+However, because the author's proposed way to resolve divergences assumes the availability of more than merely syntactic information\footnote{In particular, of a lexicon of Root Lexical Conceptual Structures specifying, for instance, the logical subject, arguments and modifiers of each verb.} and because, working with UD trees, we do not have access to anything but strictly syntactical and morphological annotations, the very task of identifying these distinctions at this level of granularity becomes, in our position, extremely challenging.
 Consequently, we refer to the simpler classification and less formal definitions proposed in \cite{divs2}, where promotional and demotional divergences are merged in a wider-coverage category of \textit{head-swapping} divergences and where lexical divergences, the hardest to handle in the absence of any kind of semantic information, do not appear at all.
-In the following, we give definitions and examples for each of these classes of divergences, specifying to what extent and how\footnote{unless otherwise specified, each divergence can be expressed as a set of simple rules in the form of boolean functions taking the pair of UD trees candidate for alignment as input.
+In the following, we give definitions and examples for each of these classes of divergences, specifying to what extent and how\footnote{Unless otherwise specified, each divergence can be expressed as a set of simple rules in the form of boolean functions taking the pair of UD trees candidate for alignment as input.
 An example of this can be found in Appendix \ref{cconf}.} each of them is handled in the proposed language-agnostic CE module.
@@
 \begin{example} 
- ``Filippo is interested in \underline{game design}'' VS ``Filippo är intresserad av \underline{spelutvekling}''
+ ``Filippo is interested in \underline{game development}'' VS ``Filippo är intresserad av \underline{spelutveckling}''
 \end{example}
@@
 
-These divergences are, while not uncommon, hard to handle without access to a lexicon where entries for verbs are complete with a list of arguments and modifiers\footnote{on the other hand, developing the CA module further could be a way to construct such a ``rich'' lexicon, as will be discussed in Chapter \ref{ch6}.}.
+These divergences are, while not uncommon, hard to handle without access to a lexicon where entries for verbs are complete with a list of arguments and modifiers\footnote{On the other hand, developing the CA module further could be a way to construct such a ``rich'' lexicon, as will be discussed in Chapter \ref{ch6}.}.
 
@@
 
-\begin{figure}[H]
+\begin{figure}[h]
  \centering
@@
  \end{itemize} 
- while introducing head alignment leads to detecting two additional, equally relevant one-word correspondences\footnote{note how, even with head alignments, not all potentially interesting alignments are found.
+ while introducing head alignment leads to detecting two additional, equally relevant one-word correspondences\footnote{Note how, even with head alignments, not all potentially interesting alignments are found.
 In this example, for instance, it could be useful to align the verb phrase ``\textit{eats a banana}'' with its Italian counterpart ``\textit{mangia una banana}'', whose members are also not, strictly speaking, subtrees of the original sentences.
@@
 
-In particular, we use a subset of the Parallel UD (PUD) corpus, a set of handcrafted\footnote{for some languages, dependency labels were actually automatically converted to UD format from other standards.
+In particular, we use a subset of the Parallel UD (PUD) corpus, a set of handcrafted\footnote{For some languages, dependency labels were actually automatically converted to UD format from other standards.
 Furthermore, manual annotation often only concerns some of fields of the CoNLL-U files, but some manual annotation is generally involved in assigning POS tags and dependency relations.} multilingual treebanks in CoNLL-U format created for the CoNLL 2017 shared task on Multilingual Parsing from Raw Text to Universal Dependencies \cite{mprtud}.
@@
 \begin{itemize}
- \item correct and useful for translation (\texttt{+})\footnote{at least to some extent: we observe that perfect translation equivalents that can be replaced with each other in all contexts are extremely rare, just like perfect synonyms.}
+ \item correct and useful for translation (\texttt{+})\footnote{At least to some extent: we observe that perfect translation equivalents that can be replaced with each other in all contexts are extremely rare, just like perfect synonyms.}
  \item correct but not useful for translation (\texttt{=}).
@@
 
-For instance, if we feed the CE module the pair of sentences of Example \ref{extrain}, as a result\footnote{this is what the actual output of the program in evaluation (or ``linearized") mode after annotation looks like.
+For instance, if we feed the CE module the pair of sentences of Example \ref{extrain}, as a result\footnote{This is what the actual output of the program in evaluation (or ``linearized") mode after annotation looks like.
 The abbreviations in square brackets refer to alignment Criteria \ref*{udmatch}, \ref{poseq} and \ref{cheads} described in the above.} we get: \smallskip \label{cann}
@@
 
-\begin{table}[H]
+\begin{table}[h]
  \footnotesize

### 2021-03-28 final_report/include/CP.tex
capitalized footnotes

@@
 In this case, CP can be used, after CE has been performed on a pair of languages $\langle L_1,L_2 \rangle$, to propagate the extracted alignments in a third version of the text in a new language $L_3$.
-In this case, the expectation is for the program to be able to propagate the vast majority\footnote{even though handling common translation divergences help, free translation involving radical differences in sentence structure can make it so that some concepts lack a clear counterpart in the text in $L_3$.} of the concepts in the set
+In this case, the expectation is for the program to be able to propagate the vast majority\footnote{Even though handling common translation divergences help, free translation involving radical differences in sentence structure can make it so that some concepts lack a clear counterpart in the text in $L_3$.} of the concepts in the set
     \item when working on two bilingual parallel texts $A$ and $B$ with one language in common, say $A$ in two languages $\langle L_1,L_2 \rangle$ and $B$ in two languages $\langle L_2, L_3 \rangle$.
@@
 
-Our algorithm is meant to be as general as possible, and as such it is not optimized for the first scenario, but rather covers both without making any assumptions in terms of sentence alignment\footnote{a way to make the algorithm more efficient for the first scenario could in fact be to rely on the fact that the $n$-lingual parallel text is sentence-aligned, keep track of the sentence from which a particular concept has been extracted and only try to find a equivalent in the corresponding $L_3$ sentence.}.
+Our algorithm is meant to be as general as possible, and as such it is not optimized for the first scenario, but rather covers both without making any assumptions in terms of sentence alignment\footnote{A way to make the algorithm more efficient for the first scenario could in fact be to rely on the fact that the $n$-lingual parallel text is sentence-aligned, keep track of the sentence from which a particular concept has been extracted and only try to find a equivalent in the corresponding $L_3$ sentence.}.
 It works as follows: for each alignment obtained by performing CE on text $A$, we consider the dependency tree in language $L_2$.

### 2021-03-28 final_report/include/TT.tex
checked bibliography entries + minor changes
capitalized footnotes

@@
 While many artificial languages, such as programming languages, are compositional by construction, the general validity of this principle for natural languages is, despite many arguments in favor, under debate.
-In this work, we make the assumption that compositionality also holds for natural languages, or at least that it is a useful notion applying to most natural language sentences\footnote{in Section \ref{evalign}, however, we will see that idiomatic expressions, often used as an arguments against compositionality, are also challenging to deal with in the context of grammar-based CA.}.
+In this work, we make the assumption that compositionality also holds for natural languages, or at least that it is a useful notion applying to most natural language sentences\footnote{In Section \ref{evalign}, however, we will see that idiomatic expressions, often used as an arguments against compositionality, are also challenging to deal with in the context of grammar-based CA.}.
 \smallskip
@@
 % cons
-On the other hand, there is currently no effective way to use these trees as a starting point for NLG: dependency grammars are an effective way to describe language, but, unlike phrase-structure grammars, they are not \textit{generative},\footnote{that does not mean, however, that dependency grammars have not been made use of in MT research, in conjunction with statistical techniques, in order to better capture grammatical generalizations \cite{treelet}.}.
+On the other hand, there is currently no effective way to use these trees as a starting point for NLG: dependency grammars are an effective way to describe language, but, unlike phrase-structure grammars, they are not \textit{generative},\footnote{That does not mean, however, that dependency grammars have not been made use of in MT research, in conjunction with statistical techniques, in order to better capture grammatical generalizations \cite{treelet}.}.
 \smallskip
@@
 Appendix \ref{a} provides a more comprehensive, but not completely exhaustive, description of all the POS tags and dependency labels mentioned in the examples appearing in this work based on that in \cite{compsyn}.
-The reader interested in a full specification of the UD annotation scheme may refer to the official UD documentation\footnote{available at \url{universaldependencies.org}.}.
+The reader interested in a full specification of the UD annotation scheme may refer to the official UD documentation\footnote{Available at \url{universaldependencies.org}.}.
 
 \paragraph{The CoNLL-U format} \label{conll}
-The standard plain text format for dependency trees, CoNLL-U, is an extension of CoNNL-X \cite{conllx} which may contain \textit{comment lines}, starting with \texttt{\#}, \textit{blank lines} marking sentence boundaries and \textit{word lines} containing the annotation of a single token (generally a word, with some exceptions\footnote{examples of such exceptions are Italian contractions such as ``\textit{dello}'', which is generally divided into ``\textit{del}'' + ``\textit{lo}''.} in 10 tab-separated fields: \smallskip
+The standard plain text format for dependency trees, CoNLL-U, is an extension of CoNNL-X \cite{conllx} which may contain \textit{comment lines}, starting with \texttt{\#}, \textit{blank lines} marking sentence boundaries and \textit{word lines} containing the annotation of a single token (generally a word, with some exceptions\footnote{Examples of such exceptions are Italian contractions such as ``\textit{dello}'', which is generally divided into ``\textit{del}'' + ``\textit{lo}''.} in 10 tab-separated fields: \smallskip
 
 \begin{enumerate}
-    \item \texttt{ID}: integer\footnote{again, there are exceptions: in the case of Italian contractions and other multiword tokens, ranges may be used.
+    \item \texttt{ID}: integer\footnote{Again, there are exceptions: in the case of Italian contractions and other multiword tokens, ranges may be used.
 Furthermore, empty nodes are characterized by decimal numbers between 0 and 1.} representing the position of the word in the sentence, starting from 1
@@
 Optional fields in the CoNLL-U file are left blank.}
-    \label{deptree}
 \end{figure}
@@
 \textit{Function} words, like pronouns and determiners, only have a grammatical meaning.
-Since they do not readily accept new members\footnote{one exception is the recent introduction of the gender-neutral pronoun ``\textit{hen}'' in Swedish.}, they are often referred to also as \textit{closed class} words.
+Since they do not readily accept new members\footnote{One exception is the recent introduction of the gender-neutral pronoun ``\textit{hen}'' in Swedish.}, they are often referred to also as \textit{closed class} words.
 \smallskip

### 2021-03-28 final_report/include/CAMT.tex
fixed typos and other minor minor things

@@
 \subsection{The English corpus} \label{eng}
-Even avoiding CP, the small size of these corpora, the limitations of the language models used for parsing and the issues encountered in the UD-to-GF conversion make both the lexica resulting from this process relatively small.
-This, together with the issues GF parsing suffers from, discussed in Chapter \ref{ch2}, makes working on arbitrary sentences taken from documents similar to those the course plans corpora are composed of extremely hard.
+Even without CP, the small size of these corpora, the limitations of the language models used for parsing and the issues encountered in the UD-to-GF conversion make both of the lexica resulting from this process relatively small.
+This, together with the issues GF parsing suffers from (as discussed in Chapter \ref{ch2}), makes working on arbitrary sentences extremely hard.
+Even only using sentences extracted from documents that are similar, both stylistically and in terms of content, to those the course plans corpora are composed of is not especially helpful.
 \smallskip
@@
 
-Some errorsregarding the extracted concepts are also interesting to analyze.
+Some errors regarding the extracted concepts are also interesting to analyze.
 In English-Swedish, while several compounds, such as $\langle$\textit{``computer science'', ``datavetenskap''}$\rangle$, are aligned correctly, there are cases in which a single-root noun is translated as a compound (e.g.
 $\langle$ \textit{``theory'', ``automatateori''} $\rangle$).
-This is not necessarily an actual alignment error, as it might rather be that the English version of the text was being less specific than its Swedish counterpart, thus producing an alignment that, during evaluation, would have been marked as ``correct but not reusable'' (\texttt{=}).
+This is not necessarily an actual alignment error, as it might well be the case that the English version of the text was being less specific than its Swedish counterpart, thus producing an alignment that, during evaluation, would have been marked as ``correct but not reusable'' (\texttt{=}).
 \smallskip

### 2021-03-28 final_report/include/CE.tex
fixed typos and other minor minor things

@@
 
-Obviously, the ``full-sentence'' tree pair is itself an alignment, even though a trivial one, as it represents a sentence-level correspondence in a case where, due to the approach being syntactic comparison, the inputs are assumed to be sentences that correspond to each other\footnote{Contrary to what it may seem, however, sentence-level alignment is in general not at a trivial task, as it is often the case, especially in certain language pairs, that one \textit{orthographic} sentence, i.e.
+Obviously, the ``full-sentence'' tree pair is itself an alignment, even though a trivial one, as it represents a sentence-level correspondence in a case where, due to the approach being syntactic comparison, the inputs are assumed to be sentences that correspond to each other\footnote{Contrary to what it may seem, however, sentence-level alignment is in general not na trivial task, as it is often the case, especially in certain language pairs, that one \textit{orthographic} sentence, i.e.
 a sentence defined based on the presence of a full stop, maps to more than one orthographic sentences in the translated text, or vice versa.}.
@@
 \begin{example} 
- ``I'll come and \underline{say hello} to Bruno and Andrea'' VS ``Verrò a \underline{salutare} Bruno e Andrea''
+ ``I'll come and \underline{say hi} to Bruno and Andrea'' VS ``Verrò a \underline{salutare} Bruno e Andrea''
 \end{example} 

### 2021-03-28 final_report/include/Conclusions.tex
fixed typos and other minor minor things

@@
 With regards to the CA component itself, it is important to remember that both the quantity and the quality of the alignments are a direct consequence of those of the dependency trees they are obtained from.
-Apart from working on dependency parsers directly, a possibility is to experiment more with hybrid alignment approaches, integrating the results prodiced by statistical tools (which only depend on the quantity and quality of raw data rather than also on that of their analyses) with those of our rule-based system.
+Apart from working on dependency parsers directly, a possibility is to experiment more with hybrid alignment approaches, integrating the results produced by statistical tools (which only depend on the quantity and quality of raw data rather than also on that of their analyses) with those of our rule-based system.
 A basic way to do that has already been implemented, but has not been made use of in the final MT experiments due to the small size of the corpora involved.

### 2021-03-28 final_report/include/TT.tex
fixed typos and other minor minor things

@@
 
-As mentioned in the introduction, the task of aligning concepts comes in two variants:
+As mentioned in the Introduction, the task of aligning concepts comes in two variants:
 \begin{enumerate}
@@
 
-As mentioned in the introduction, however, alignment can be performed at different levels of abstraction: to bring this to the extreme, there are contexts in which we might want to align full sentences, or even whole documents.
+As mentioned in the Introduction, however, alignment can be performed at different levels of abstraction: to bring this to the extreme, there are contexts in which we might want to align full sentences, or even whole documents.
 While the latter two levels of abstraction are not necessarily relevant to MT, only aligning individual words is also, as discussed in the Introduction, not always the best option.

### 2021-03-28 final_report/include/CE.tex
no decimals in tables

@@
  \textbf{distinct alignments} & 1097  & 1257 & 1198   & 1314  \\ \hline
- \textbf{correct (+ and =)} & 830 (58.12\%) & 995 (79.15\%)  & 964 (80.46\%) & 1105 (84.03\%) \\ \hline
- \textbf{correct and useful (+)} & 776 (54.34\%) & 976 (77.64\%)  & 896 (74.79\%) & 1082 (82.28\%) \\ \hline
+ \textbf{correct (+ and =)} & 830 (58\%) & 995 (79\%)  & 964 (80\%) & 1105 (84\%) \\ \hline
+ \textbf{correct and useful (+)} & 776 (54\%) & 976 (78.64\%)  & 896 (75\%) & 1082 (82\%) \\ \hline
  \end{tabular}
@@
  \textbf{}       & \textbf{en-it}  & \textbf{en-sv} & \textbf{en-it}  & \textbf{en-sv}  \\ \hline
- \textbf{matching UD labels}      & 39.39\%  & 37.71\%  & 78.60\%  & 80.24\%  \\ \hline
- \textbf{POS-equivalence}       & 3\%  & 1.82\%  & 83.33\%  & 83.33\%  \\ \hline
- \textbf{known divergence}       & 2.08\%   & 1.36\%  & 52\%   & 55.55\%  \\ \hline
+ \textbf{matching UD labels}      & 39\%  & 38\%  & 79\%  & 80\%  \\ \hline
+ \textbf{POS-equivalence}       & 3\%  & 2\%  & 83\%  & 83\%  \\ \hline
+ \textbf{known divergence}       & 2\%   & 1\%  & 52\%   & 56\%  \\ \hline
  \textbf{known alignment}       & 0\%  & 0\%  & -  & -  \\ \hline
- \textbf{\begin{tabular}[c]{@{}l@{}}matching UD labels\\ + POS-equivalence\end{tabular}}  & 20.95\%  & 24.71\%  & 93.03\%  & 94.76\%  \\ \hline
- \textbf{\begin{tabular}[c]{@{}l@{}}matching UD labels\\ + head alignment\end{tabular}}  & 21.78\%  & 19.61\%  & 67.81\%  & 76.74\%  \\ \hline
- \textbf{\begin{tabular}[c]{@{}l@{}}POS-equivalence\\ + head alignment\end{tabular}}  & 1.33\%   & 1.29\%  & 75\%   & 64.70\%  \\ \hline
- \textbf{\begin{tabular}[c]{@{}l@{}}matching UD labels\\ + known alignment\end{tabular}}  & 0.25\%   & 0.45\%  & 66.66\%  & 83.33\%  \\ \hline
- \textbf{\begin{tabular}[c]{@{}l@{}}matching UD labels\\ + POS-equivalence\\ + head alignment\end{tabular}} & 10.85\%  & 12.69\%  & 95.38\%  & 90.41\%  \\ \hline
+ \textbf{\begin{tabular}[c]{@{}l@{}}matching UD labels\\ + POS-equivalence\end{tabular}}  & 21\%  & 25\%  & 93\%  & 95\%  \\ \hline
+ \textbf{\begin{tabular}[c]{@{}l@{}}matching UD labels\\ + head alignment\end{tabular}}  & 22\%  & 20\%  & 68\%  & 77\%  \\ \hline
+ \textbf{\begin{tabular}[c]{@{}l@{}}POS-equivalence\\ + head alignment\end{tabular}}  & 1\%   & 1\%  & 75\%   & 65\%  \\ \hline
+ \textbf{\begin{tabular}[c]{@{}l@{}}matching UD labels\\ + known alignment\end{tabular}}  & 0\%   & 0\%  & 67\%  & 83\%  \\ \hline
+ \textbf{\begin{tabular}[c]{@{}l@{}}matching UD labels\\ + POS-equivalence\\ + head alignment\end{tabular}} & 11\%  & 13\%  & 95\%  & 90\%  \\ \hline
  \end{tabular}
@@
  \textbf{distinct alignments} & 352  & 529  \\ \hline
- \textbf{correct (+ and =)} & 243 (69.03\%) & 368 (69.56\%) \\ \hline
- \textbf{correct and useful (+)} & 229 (65.05\%) & 351 (66.35\%) \\ \hline
+ \textbf{correct (+ and =)} & 243 (69\%) & 368 (70\%) \\ \hline
+ \textbf{correct and useful (+)} & 229 (65\%) & 351 (66\%) \\ \hline
  \end{tabular}
@@
   \textbf{distinct alignments} & 716 & 1440 & 1435 \\ \hline
-  \textbf{correct} & 536 (74.86\%) & 410 (28.47\%) & 656 (45.71\%)\\ \hline
-  \textbf{correct and useful} & 491 (68.57\%) & 371 (25.76\%) & 590 (41.11\%)\\ \hline
+  \textbf{correct} & 536 (75\%) & 410 (28\%) & 656 (46\%)\\ \hline
+  \textbf{correct and useful} & 491 (69\%) & 371 (26\%) & 590 (41\%)\\ \hline
  \end{tabular}

### 2021-03-28 final_report/include/CE.tex
added a couple of clarifications

@@
 \begin{criterion}[POS-equivalence] \label{poseq}
- Two dependency trees $t$, $u$ are POS-equivalent (and, as such, will be aligned) if $M_1 = M_2 \neq \emptyset$, where $M_i$ is defined as the multiset of POS tags of all the meaning-carrying word nodes of $t_i$.
+ Two dependency trees $t$, $u$ are POS-equivalent (and, as such, will be aligned) if $M_1 = M_2 \neq \emptyset$, where $M_i$ is defined as the multiset of POS tags of all the \normalfont{meaning-carrying} \textit{(see below) word nodes of $t_i$}.
 \end{criterion} \smallskip

### 2021-03-28 final_report/include/CAMT.tex
no decimals in tables

@@
 \begin{itemize}
-    \item while GF can parse the SL string in a variety of ways and, as a consequence, return multiple TL linearizations, our program only outputs the first candidate translation, preliminarily discarding trees containing terminal nodes that have no linearization in the TL\footnote{A linearization can be missing, for instance, if a concrete syntax is postprocessed by simply getting rid of an ``illegal'' verb causing compilation problems (cf.
-Section \ref{grgn})}
+    \item while GF can parse the SL string in a variety of ways and, as a consequence, return multiple TL linearizations, our program only outputs the first candidate translation, preliminarily discarding trees containing terminal nodes that have no linearization in the TL\footnote{A linearization can be missing, for instance, if a concrete syntax is postprocessed by simply getting rid of an ``illegal'' verb causing compilation problems.
+In the experiments conducted during this project, this initially happened even with some irregular (but correctly lemmatized!) Italian verbs whose infinitive does not end in ``\textit{-are}'', ``\textit{-ere}'' or ``\textit{-ire}'', such as ``\textit{porre}.'' The Italian morphological dictionary was then modified so to also handle this type of verbs.}
     \item to facilitate the experiments, our standalone has been written so that it can translate several newline-separated sentences at once.
@@
 To address this problem, the CE module has been modified so that it can try to align individual clauses instead of sentences. 
-The idea is to obtain, from each pair of sentence trees to align, a pair of lists of subtrees whose heads are labelled \texttt{root}, \texttt{csubj}, \texttt{ccomp}, \texttt{xcomp}, \texttt{advcl} or  \texttt{acl}\footnote{All these clausal labels are defined in Appendix \ref{a_lab}}.
+The idea is to obtain, from each pair of sentence trees to align, a pair of lists of subtrees whose heads are labelled \texttt{root}, \texttt{csubj}, \texttt{ccomp}, \texttt{xcomp}, \texttt{advcl} or  \texttt{acl}\footnote{All these clausal labels are defined in Appendix \ref{a_lab}.}.
 
@@
     \textbf{}            & \textbf{DMI (en-it)} & \textbf{CSE (en-sv)} \\ \hline
-    \textbf{BLEU-1 to 4} & 55.4          & 61.27         \\ \hline
-    \textbf{BLEU-1 to 3} & 62.75         & 67.77         \\ \hline
-    \textbf{BLEU-1 to 2} & 70.6          & 74.3          \\ \hline
-    \textbf{BLEU-1}      & 79.33         & 80.99         \\ \hline
+    \textbf{BLEU-1 to 4} & 55         & 61         \\ \hline
+    \textbf{BLEU-1 to 3} & 63         & 68         \\ \hline
+    \textbf{BLEU-1 to 2} & 70         & 74          \\ \hline
+    \textbf{BLEU-1}      & 79         & 81         \\ \hline
     \end{tabular}

### 2021-04-27 ConceptAlignment.hs
commented out propagation section for later adaptation

@@ -337,52 +337,52 @@ alignHeads (A (RTree n ts,RTree m us))
 {- Propagation functions -}
 
 -- | Generic (not optimized for same text in n languages) propagation function
-propagate :: [Criterion]            -- ^ a list of criteria 
-                                    -- (sorted by priority)
-          -> Bool                   -- ^ -- ^ a flag indicating whether clause 
-                                    -- segmentation should be performed
-          -> Bool                   -- ^ a flag indicating whether alignment
-                                    --"by exclusion" should also be performed 
-          -> ([UDTree],[UDTree])    -- a pair of lists of UD trees (the
-                                    -- sentences to propagate on) in L1, L2
-          -> UDTree                 -- a previously extracted L1 concept
-          -> Maybe (Alignment,Info) -- an (Alignment,Info) pair, if found
-propagate cs segment byExcl ([],_) _ = Nothing 
-propagate cs segment byExcl (t:ts,u:us) c =
-  let 
-    as = M.toList $ alignSent M.empty [] cs Nothing segment byExcl (t,u)
-    as' = case (c `isSubUDTree'` t, c `isHeadSubUDTree` t) of
-      (True,_) -> sortOnDepth as
-      (_,True) -> sortOnDepth (filter (\(_,(rs,_)) -> HEAD `elem` rs) as)
-      (False,False) -> []
-    in case find (\(a,_) -> c =~ sl a) as' of
-      Nothing -> propagate cs segment byExcl (ts,us) c
-      n -> n
-    where 
-      -- difference between the depth of the SL aligned subtree and c:
-      -- the smaller, the better (mostly used to avoid that root nodes are
-      -- aligned with full sentences, but also makes sense in general)
-      depthDiff :: RTree a -> Int
-      depthDiff t = abs (depthRTree t - depthRTree c)
-      sortOnDepth :: [(Alignment,Info)] ->[(Alignment,Info)]
-      sortOnDepth = sortOn (depthDiff . sl . fst)
-      -- check if c is the head of one of t's subtrees
-      isHeadSubUDTree :: UDTree -> UDTree -> Bool
-      isHeadSubUDTree c t = 
-        isJust $ listToMaybe $ 
-          sortOn depthDiff (filter (isHeadUDTree c) (allSubRTrees t))
[…]

### 2021-04-27 ConceptAlignment.hs
commented out MT selection section for later adaptation

@@ -459,28 +459,28 @@ alignment2sentencePair (A (t,u)) =
 -- - remove redundant alignments (i.e. alignments that can be inferred
 --   from their sub-alignments)
 -- - optionally, remove alignments where the size of both trees is > n
-selectForMT :: Maybe Int -> [(Alignment,Info)] -> [(Alignment,Info)]
-selectForMT mmax as = nubBy 
-                  (\(a,_) (b,_) -> b `contains` a && isPerfectShallow b) 
-                  (reverse $ sort as'')
-  where 
-    as' = filter (\(a,_) -> hasContent a) as -- remove function-only stuff
-      where hasContent (A (t,u)) = (not . null) (contentTags t) 
-                                || (not . null) (contentTags u)
-    as'' = case mmax of
-      (Just m) -> 
-        filter (\(A (t,u),_) -> sizeRTree t <= m || sizeRTree u <= m) as'
-      Nothing -> as'
+--selectForMT :: Maybe Int -> [(Alignment,Info)] -> [(Alignment,Info)]
+--selectForMT mmax as = nubBy 
+--                  (\(a,_) (b,_) -> b `contains` a && isPerfectShallow b) 
+--                  (reverse $ sort as'')
+--  where 
+--    as' = filter (\(a,_) -> hasContent a) as -- remove function-only stuff
+--      where hasContent (A (t,u)) = (not . null) (contentTags t) 
+--                                || (not . null) (contentTags u)
+--    as'' = case mmax of
+--      (Just m) -> 
+--        filter (\(A (t,u),_) -> sizeRTree t <= m || sizeRTree u <= m) as'
+--      Nothing -> as'
 
 -- | Check if an alignment is perfect, i.e. ig the structure of the two trees
 -- is the same
-isPerfect :: Alignment -> Bool
-isPerfect (A (t,u)) = abstractUDTree t == abstractUDTree u 
+--isPerfect :: Alignment -> Bool
+--isPerfect (A (t,u)) = abstractUDTree t == abstractUDTree u 
 
 -- | Check the top-level structure of the two trees composing an alignment
 -- is the same
-isPerfectShallow :: Alignment -> Bool
-isPerfectShallow (A (t,u)) 
-  = root t' == root u' 
-  && map root (childrenRTree t') == map root (childrenRTree u') 
-  where (t',u') = (abstractUDTree t,abstractUDTree u)
+--isPerfectShallow :: Alignment -> Bool
[…]

### 2021-04-28 ConceptAlignment.hs
moved sentId to ConceptAlignment (provisorily)

@@ -484,3 +484,9 @@ alignment2sentencePair (A (t,u)) =
 --  = root t' == root u' 
 --  && map root (childrenRTree t') == map root (childrenRTree u') 
 --  where (t',u') = (abstractUDTree t,abstractUDTree u)
+
+-- | Return the id of a sentence, taken from the comment that precedes it
+-- assumption: the CoNLL-U files are generated by ExtractConcepts and
+-- PropagateConcepts and have not been modified
+sentId :: UDSentence -> Int -- TODO: mv (to gf-ud), make more robust
+sentId = read . drop 4 . last . words . head . udCommentLines

### 2021-04-28 GenerateGrammar.hs

@@ -195,13 +195,6 @@ getAlignments ss =
         in if all isJust as then Just $ map fromJust as else Nothing
       o = 1000000 -- numbering offset (cf. prUDSentence in UDConcepts)
 
--- | Return the id of a sentence, taken from the comment that precedes it
--- assumption: the CoNLL-U files are generated by ExtractConcepts and
--- PropagateConcepts and have not been modified
-sentId :: UDSentence -> Int
-sentId = read . drop 4 . last . words . head . udCommentLines
-
-
 -- | Like UDAnnotations.getEnv, but uses pgf instead of paths
 -- (this is a bad solution but I don't necessarily want to rely on the pgf
 -- to be in a certain location)

### 2021-04-28 ConceptAlignment.hs
draft major type redesign; should close #17 and #23

@@ -4,87 +4,125 @@ import Data.Maybe
 import Data.List
 import Data.List.Split
 import qualified Data.MultiSet as MS
-import qualified Data.Map as M
 import qualified Data.Set as S
+import qualified Data.Map as M
 import RTree
 import UDConcepts
 import UDPatterns
 
 {- Basic (mostly data types) definitions and instances for alignments etc. -}
 
--- | An alignment is a pair of corresponding UD subtrees (not a type synonym
--- because of custom Eq and Ord instances)
--- TODO: type synonym instances and just a pair?
-newtype Alignment = A (UDTree,UDTree)
+-- | An alignment is a pair of corresponding UD (sub)trees associated to some 
+-- metadata 
+data Alignment = A {
+  trees :: AlignedTrees,    -- ^ pair of UD subtrees
+  meta :: Meta              -- ^ metadata
+} deriving (Show,Read,Eq,Ord)
 
--- | Left (SL) tree of an alignment
+-- | Initialize a potential alignment given just the pair of trees
+initAlignment :: AlignedTrees -> Alignment
+initAlignment tu = A {
+  trees = tu,
+  meta = initMeta
+}
+
+-- | Convert an alignment into a (key,value) pair, where the key is the tree
+-- pair and the "value" is the metadata
+alignment2keyval :: Alignment -> (AlignedTrees,Meta)
+alignment2keyval a = (trees a,meta a)
+
+-- | Convert a (key,value) pair, where the key is the tree pair and the 
+-- "value" is the metadata, into an Alignment
+keyval2alignment :: (AlignedTrees,Meta) -> Alignment
+keyval2alignment (k,v) = A {
+  trees = k,
+  meta = v
+}
+
+-- | Left (source language) tree of an alignment
 sl :: Alignment -> UDTree
-sl (A (t,u)) = t
+sl a = let (AT (t,u)) = trees a in t
 
--- | Correct (TL) tree of an alignment
+-- | Right (target language) tree of an alignment
 tl :: Alignment -> UDTree
-tl (A (t,u)) = u
+tl a = let (AT (t,u)) = trees a in u
 
--- | Alignments are shown as pairs
-instance Show Alignment where
-  show (A a) = show a
-
--- | prAlignment is used instead of show to "visually" inspect the alignments
+-- | prAlignment is used instead of show for inspecting the alignments in an
+-- easier way, e.g. in EvAlign and for debugging
[…]

### 2021-04-28 ExtractConcepts.hs
prepared for restoring basic CE functionalities, ignoring SMT stuff for the moment

@@ -11,7 +11,7 @@ import System.Exit
 import UDConcepts
 import UDPatterns
 import ConceptAlignment
-import FastAlignUtils
+--import FastAlignUtils
 import Criteria 
 import ArgvParse
 
@@ -58,12 +58,13 @@ main = do
     getSmtAlignments flags src trg = case [path | Pharaoh path <- flags] of
       [] -> return []
       [path] -> do
-        indices <- readFile path >>= return . parsePh
-        srcConllu <- readFile src
-        trgConllu <- readFile trg
-        let bitext = take 100 $ parseBi $ conllu2bi (srcConllu,trgConllu)
-        print $ ltrees $ phToLas bitext indices !! 5
-        return $ phToLas bitext indices
+        undefined
+        --indices <- readFile path >>= return . parsePh
+        --srcConllu <- readFile src
+        --trgConllu <- readFile trg
+        --let bitext = take 100 $ parseBi $ conllu2bi (srcConllu,trgConllu)
+        --print $ ltrees $ phToLas bitext indices !! 5
+        --return $ phToLas bitext indices
       _ -> undefined
     getPattern :: [Flag] -> IO (Maybe UDPattern)
     getPattern flags = do

### 2021-04-28 ExtractConcepts.hs
actually ignore SMT stuff altogether

@@ -54,18 +54,18 @@ main = do
   where 
     insertLang :: FilePath -> String -> FilePath
     insertLang fp l = dropExtension fp ++ l ++ takeExtension fp
-    getSmtAlignments :: [Flag] -> FilePath -> FilePath -> IO [LinAlignment]
-    getSmtAlignments flags src trg = case [path | Pharaoh path <- flags] of
-      [] -> return []
-      [path] -> do
-        undefined
-        --indices <- readFile path >>= return . parsePh
-        --srcConllu <- readFile src
-        --trgConllu <- readFile trg
-        --let bitext = take 100 $ parseBi $ conllu2bi (srcConllu,trgConllu)
-        --print $ ltrees $ phToLas bitext indices !! 5
-        --return $ phToLas bitext indices
-      _ -> undefined
+    getSmtAlignments = undefined
+    --getSmtAlignments :: [Flag] -> FilePath -> FilePath -> IO [LinAlignment]
+    --getSmtAlignments flags src trg = case [path | Pharaoh path <- flags] of
+    --  [] -> return []
+    --  [path] -> do
+    --    indices <- readFile path >>= return . parsePh
+    --    srcConllu <- readFile src
+    --    trgConllu <- readFile trg
+    --    let bitext = take 100 $ parseBi $ conllu2bi (srcConllu,trgConllu)
+    --    print $ ltrees $ phToLas bitext indices !! 5
+    --    return $ phToLas bitext indices
+    --  _ -> undefined
     getPattern :: [Flag] -> IO (Maybe UDPattern)
     getPattern flags = do
           let rp = listToMaybe [p | Pattern p <- flags] 

### 2021-04-28 ConceptAlignment.hs
temporarily ignore sentId

@@ -499,4 +499,4 @@ alignment2sentencePair a =
 -- assumption: the CoNLL-U files are generated by ExtractConcepts and
 -- PropagateConcepts and have not been modified
 sentId :: UDSentence -> Int -- TODO: mv (to gf-ud), make more robust
-sentId = read . drop 4 . last . words . head . udCommentLines
+sentId _ = 1--read . drop 4 . last . words . head . udCommentLines

### 2021-04-28 ExtractConcepts.hs
restored very basic extraction functionalities

@@ -22,31 +22,31 @@ main = do
     [src,trg] -> if Help `elem` flags
       then putStrLn help >> exitSuccess
       else do
-        ts <- conlluFile2UDTrees src
-        us <- conlluFile2UDTrees trg
+        ts <- parseUDFile src
+        us <- parseUDFile trg
         let tus = zip ts us -- let tus = take 50 $ zip ts us
-        smtAs <- getSmtAlignments flags src trg
+        let smtAs = S.empty -- <- getSmtAlignments flags src trg
         let segment = Clauses `elem` flags
         let byExcl = Rest `elem` flags
         let fp = listToMaybe [path | Path path <- flags] 
         let fp' = fromJust fp
         r <- getPattern flags
-        let as = M.toList $ align smtAs criteria r segment byExcl tus
+        let as = align smtAs criteria r segment byExcl tus
         let m = listToMaybe [read mmax :: Int | MaxSize mmax <- flags]
-        let as' = if All `elem` flags then as else selectForMT m as
-        let as'' = sortByConfidence as'
+        let as' = if All `elem` flags then as else as --TODO: selectForMT m as
+        let as'' = {-TODO: sortByConfidence $-} S.toList as' 
         if Linearize `elem` flags
           then 
             if isJust fp 
-              then writeFile fp' (unlines $ map (show . toLinAlignment) as'') 
-              else mapM_ (print . toLinAlignment) as''
+              then writeFile fp' (unlines $ map prAlignment as'') 
+              else mapM_ (putStrLn . prAlignment) as''
           else 
             if isJust fp 
               then do
-                let numberedSentPairs = [1..] `zip` map (alignment2sentencePair . fst) as''
+                let numberedSentPairs = [1..] `zip` map alignment2sentencePair as''
                 writeFile (insertLang fp' "SL") (unlines [prUDSentence (fst x) (fst $ snd x) | x <- numberedSentPairs])
                 writeFile (insertLang fp' "TL") (unlines [prUDSentence (fst x) (snd $ snd x) | x <- numberedSentPairs]) 
-              else mapM_ (print . fst) as''
+              else mapM_ print as''
     _ -> do
       putStrLn "Wrong number of arguments."
       putStrLn help
@@ -76,12 +76,12 @@ main = do
             else return Nothing
 
 -- | Sort alignments by how likely theyare to be correct (kinda). 
-sortByConfidence :: [(Alignment,Info)] -> [(Alignment,Info)]
[…]

### 2021-04-29 paper/paper.tex
paper outline

% This must be in the first 5 lines to tell arXiv to use pdfLaTeX, which is strongly recommended.
\pdfoutput=1
% In particular, the hyperref package requires pdfLaTeX in order to break URLs across lines.

\documentclass[11pt]{article}

% Remove the "review" option to generate the final version.
\usepackage[review]{acl}

% Standard package includes
\usepackage{times}
\usepackage{latexsym}

% For proper rendering and hyphenation of words containing Latin characters (including in bib files)
\usepackage[T1]{fontenc}
% For Vietnamese characters
% \usepackage[T5]{fontenc}
% See https://www.latex-project.org/help/documentation/encguide.pdf for other character sets

% This assumes your files are encoded as UTF8
\usepackage[utf8]{inputenc}

% This is not strictly necessary, and may be commented out,
% but it will improve the layout of the manuscript,
% and will typically save some space.
\usepackage{microtype}

% If the title and author information does not fit in the area allocated, uncomment the following
%
%\setlength\titlebox{<dim>}
%
% and set <dim> to something 5cm or larger.
% TODO: title
\title{Instructions for *ACL Proceedings}

% Author information can be set in various styles:
% For several authors from the same institution:
% \author{Author 1 \and ... \and Author n \\
%         Address line \\ ... \\ Address line}
% if the names do not fit well on one line use
%         Author 1 \\ {\bf Author 2} \\ ... \\ {\bf Author n} \\
% For authors from different institutions:
% \author{Author 1 \\ Address line \\  ... \\ Address line
%         \And  ... \And
%         Author n \\ Address line \\ ... \\ Address line}
% To start a seperate ``row'' of authors use \AND, as in
% \author{Author 1 \\ Address line \\  ... \\ Address line
%         \AND
%         Author 2 \\ Address line \\ ... \\ Address line \And
%         Author 3 \\ Address line \\ ... \\ Address line}
% TODO: authors info
\author{First Author \\
  Affiliation / Address line 1 \\
  Affiliation / Address line 2 \\
  Affiliation / Address line 3 \\
  \texttt{email@domain} \\\And
  Second Author \\
  Affiliation / Address line 1 \\
  Affiliation / Address line 2 \\
  Affiliation / Address line 3 \\
  \texttt{email@domain} \\}

\begin{document}
\maketitle
\begin{abstract}
% TODO: abstract
\end{abstract}

\section{Introduction}

\section{Methodology}

\section{Evaluation}

\section{Conclusions}

% TODO: see why it does not compile
% Entries for the entire Anthology, followed by custom entries
%\bibliography{anthology,custom}
%\bibliographystyle{acl_natbib}

\end{document}

### 2021-04-29 ConceptAlignment.hs
restored and improved sentId

@@ -86,7 +86,7 @@ instance Ord AlignedTrees where
 -- | The metadata of an alignment are:
 data Meta = M {
   reasons :: S.Set Reason,  -- ^ reasons for alignment
-  sentIds :: S.Set Int,     -- ^ ids of the sentences it was inferred from
+  sentIds :: S.Set String,  -- ^ ids of the sentences it was inferred from
   nOccurrences :: Int       -- ^ total number of occurrences
 } deriving (Show,Read,Eq,Ord)
 
@@ -496,7 +496,8 @@ alignment2sentencePair a =
 --  where (t',u') = (abstractUDTree t,abstractUDTree u)
 
 -- | Return the id of a sentence, taken from the comment that precedes it
--- assumption: the CoNLL-U files are generated by ExtractConcepts and
--- PropagateConcepts and have not been modified
-sentId :: UDSentence -> Int -- TODO: mv (to gf-ud), make more robust
-sentId _ = 1--read . drop 4 . last . words . head . udCommentLines
+sentId :: UDSentence -> String 
+sentId s = if hasSentId s then head $ words $ head idEtc else "missing sent_id"
+  where
+    hasSentId s = (not . null) idEtc
+    (_:idEtc) = splitOn "sent_id = " (unwords (concatMap words (udCommentLines s)))

### 2021-04-29 ConceptAlignment.hs
moved sentId to UDConcepts & updated gf-ud version

@@ -494,10 +494,3 @@ alignment2sentencePair a =
 --  = root t' == root u' 
 --  && map root (childrenRTree t') == map root (childrenRTree u') 
 --  where (t',u') = (abstractUDTree t,abstractUDTree u)
-
--- | Return the id of a sentence, taken from the comment that precedes it
-sentId :: UDSentence -> String 
-sentId s = if hasSentId s then head $ words $ head idEtc else "missing sent_id"
-  where
-    hasSentId s = (not . null) idEtc
-    (_:idEtc) = splitOn "sent_id = " (unwords (concatMap words (udCommentLines s)))

### 2021-04-29 ConceptAlignment.hs
assign sent_id to head alignments to

@@ -290,56 +290,59 @@ alignSent as cs p cl ex s@(s1,s2) = if ex then extra else basic --TODO: extra `u
         (True,True) -> Just a
         _ -> Nothing
 
--- | Helper function for head alignment: given an alignment, return a new one  
--- for their "heads", respecting any compounds and aux+verbs (and more?)
--- TODO: mv inside
-alignHeads :: Alignment -> Alignment
-alignHeads a
-  -- if there are compound constructions, look for their counterparts and
-  -- align accordingly
-  | (not . null) cts = 
-      initHeadAlignment $ AT (RTree n cts, RTree m (compCounterparts ts us))
-  | (not . null) cus = 
-      initHeadAlignment $ AT (RTree n (compCounterparts us ts), RTree m cus)
-  -- if the roots are verbs (to avoid messing with copulas) and only one of
-  -- them has 1+ auxiliaries, align verb | verb + auxiliaries
-  | all isVerb [n,m] && (not . null) ats && null aus = 
-      initHeadAlignment $ AT (RTree n ats, RTree m [])
-  | all isVerb [n,m] && null ats && (not . null) aus = 
-      initHeadAlignment $ AT (RTree n [], RTree m aus) 
-  | otherwise = initHeadAlignment $ AT (RTree n [], RTree m [])
-  where
-    AT (RTree n ts,RTree m us) = trees a
-    -- select subtrees labelled in a certain way 
-    filterByLabel l xs = filter (`isLabelled` l) xs
-
-    ats = filterByLabel "aux" ts
-    aus = filterByLabel "aux" us
-    cts = filterByLabel "compound" ts
-    cus = filterByLabel "compound" us
-
-    -- TODO: add other reasons
-    initHeadAlignment tu = a { meta = (meta a) { reasons = S.singleton HEAD } }
-      where a = initAlignment tu
-
-    -- given two lists of subtrees, select those that, in the second,
-    -- could correspond to a compound construction in the first, i.e.
-    compCounterparts :: [UDTree] -> [UDTree] -> [UDTree]
-    compCounterparts ts us = cus ++ fus' ++ nus' ++ aus'
+    -- head alignment: given an alignment, return a new one  
+    -- for their "heads", respecting any compounds and aux+verbs (and more?)
+    alignHeads :: Alignment -> Alignment
+    alignHeads a
+      -- if there are compound constructions, look for their counterparts and
+      -- align accordingly
+      | (not . null) cts = 
[…]

### 2021-04-29 ConceptAlignment.hs
improved prAlignment

@@ -51,10 +51,13 @@ tl a = let (AT (t,u)) = trees a in u
 -- easier way, e.g. in EvAlign and for debugging
 prAlignment :: Alignment -> String
 prAlignment a = prTrees a 
-                ++ " reasons: " ++ show (reasons $ meta a)
-                ++ " sentence IDs: " ++ show (sentIds $ meta a)
+                ++ "    --"
+                ++ " reasons: " ++ showSet (reasons $ meta a)
+                ++ " sentence IDs: " ++ showSet (sentIds $ meta a)
                 ++ " n. occurrences: " ++ show (nOccurrences $ meta a)
-  where prTrees a = linearize (sl a) ++ " ||| " ++ linearize (tl a)
+  where 
+    prTrees a = linearize (sl a) ++ " ||| " ++ linearize (tl a)
+    showSet s = "{" ++ intercalate ", " (S.toList $ S.map show s) ++ "}"
 
 -- | Check if an alignment "contains" another;
 -- used both in pruning and selection of alignments for MT

### 2021-04-30 ConceptAlignment.hs
remove occurrence count as it's kinda redundant and very problematic

@@ -54,7 +54,6 @@ prAlignment a = prTrees a
                 ++ "    --"
                 ++ " reasons: " ++ showSet (reasons $ meta a)
                 ++ " sentence IDs: " ++ showSet (sentIds $ meta a)
-                ++ " n. occurrences: " ++ show (nOccurrences $ meta a)
   where 
     prTrees a = linearize (sl a) ++ " ||| " ++ linearize (tl a)
     showSet s = "{" ++ intercalate ", " (S.toList $ S.map show s) ++ "}"
@@ -89,16 +88,14 @@ instance Ord AlignedTrees where
 -- | The metadata of an alignment are:
 data Meta = M {
   reasons :: S.Set Reason,  -- ^ reasons for alignment
-  sentIds :: S.Set String,  -- ^ ids of the sentences it was inferred from
-  nOccurrences :: Int       -- ^ total number of occurrences
+  sentIds :: S.Set String   -- ^ ids of the sentences it was inferred from
 } deriving (Show,Read,Eq,Ord)
 
 -- | Initialize metadata with default vals
 initMeta :: Meta
 initMeta = M {
   reasons = S.empty,
-  sentIds = S.empty,
-  nOccurrences = 1
+  sentIds = S.empty
 }
 
 
@@ -108,8 +105,7 @@ initMeta = M {
 combineMeta :: Meta -> Meta -> Meta
 m `combineMeta` n = M {
   reasons = reasons m `S.union` reasons n,
-  sentIds = sentIds m `S.union` sentIds n,
-  nOccurrences = nOccurrences m + nOccurrences n
+  sentIds = sentIds m `S.union` sentIds n
 }
 
 -- TODO: docs

### 2021-04-30 ExtractConcepts.hs

@@ -81,7 +81,7 @@ sortByConfidence = sortOn (\a ->
                         let rs = S.toList (reasons $ meta a) \\ [HEAD, PREV] 
                         in (
                             -(length rs), 
-                            -(nOccurrences $ meta a), 
+                            -(length $ sentIds $ meta a), 
                             tail rs
                           )) 
 

### 2021-04-30 ConceptAlignment.hs
fixed bug causing some alignment reason not to show in head alignment

@@ -316,13 +316,11 @@ alignSent as cs p cl ex s@(s1,s2) = if ex then extra else basic --TODO: extra `u
         cts = filterByLabel "compound" ts
         cus = filterByLabel "compound" us
     
-        -- TODO: add other reasons
         initHeadAlignment tu = a { 
           meta = (meta a) { 
-            reasons = S.singleton HEAD,
-            sentIds = S.singleton id } 
+            reasons = S.singleton HEAD `S.union` (reasons $ meta a)
+          } 
         }
-          where a = initAlignment tu
     
         -- given two lists of subtrees, select those that, in the second,
         -- could correspond to a compound construction in the first, i.e.

### 2021-04-30 ConceptAlignment.hs
went through TODOs in ConceptAlignment

@@ -108,20 +108,20 @@ m `combineMeta` n = M {
   sentIds = sentIds m `S.union` sentIds n
 }
 
--- TODO: docs
+-- | Insert a new alignment combining metadata if an equivalent one is already
+-- present
 insert' :: (AlignedTrees,Meta) -> M.Map AlignedTrees Meta -> M.Map AlignedTrees Meta 
 insert' (at,m) = M.insertWith combineMeta at m
 
+-- | Perform the union of two maps combining the metadata of equivalent
+-- alignments
 union' :: M.Map AlignedTrees Meta -> M.Map AlignedTrees Meta -> M.Map AlignedTrees Meta 
 union' = M.unionWith combineMeta
 
+-- | Perform the union of n maps combining the metadata of equivalent
+-- alignments
 unions' :: [M.Map AlignedTrees Meta] -> M.Map AlignedTrees Meta
 unions' = M.unionsWith combineMeta
-  
--- | Linearize a UD tree ignoring initial whitespace
--- TODO: move to gfud
-linearize :: UDTree -> String
-linearize = dropWhile (== ' ') . prUDTreeString
 
 -- | Type of manual annotations.
 -- An alignment can be incorrect (-), correct in the specific context where it 
@@ -212,7 +212,7 @@ alignSent :: M.Map AlignedTrees Meta   -- ^ a map of known alignments (e.g.
                                        -- be performed 
           -> (UDSentence,UDSentence)   -- ^ the sentences to align 
           -> M.Map AlignedTrees Meta   -- ^ a map of alignments
-alignSent as cs p cl ex s@(s1,s2) = if ex then extra else basic --TODO: extra `union` basic?
+alignSent as cs p cl ex s@(s1,s2) = if ex then extra else extra `S.union` basic
   where 
     basic = if cl then alignClauses (t,u) else alignSent' cs (t,u)
     extra = alignRest basic -- alignments obtained "by exclusion"
@@ -229,8 +229,8 @@ alignSent as cs p cl ex s@(s1,s2) = if ex then extra else basic --TODO: extra `u
         (False,True) -> insert' (alignment2keyval h) (insert' (alignment2keyval a) as) `union'` as'
         (False,False) -> insert' (alignment2keyval a) as `union'` as'
         -- using a gf-ud pattern: ignore head alignment altogether
-        (True,_) -> if isJust m 
-          then insert' (alignment2keyval $ fromJust m) (as `union'` as')
[…]

### 2021-04-30 ConceptAlignment.hs
created typesyn for (internal) alignment map

@@ -98,6 +98,9 @@ initMeta = M {
   sentIds = S.empty
 }
 
+-- | Map of alignments, to be used only internally to facilitate combining
+-- equivalent alignments
+type AlignmentMap = M.Map AlignedTrees Meta
 
 {- Helper functions for alignment maps-}
 
@@ -110,17 +113,17 @@ m `combineMeta` n = M {
 
 -- | Insert a new alignment combining metadata if an equivalent one is already
 -- present
-insert' :: (AlignedTrees,Meta) -> M.Map AlignedTrees Meta -> M.Map AlignedTrees Meta 
+insert' :: (AlignedTrees,Meta) -> AlignmentMap -> AlignmentMap 
 insert' (at,m) = M.insertWith combineMeta at m
 
 -- | Perform the union of two maps combining the metadata of equivalent
 -- alignments
-union' :: M.Map AlignedTrees Meta -> M.Map AlignedTrees Meta -> M.Map AlignedTrees Meta 
+union' :: AlignmentMap -> AlignmentMap -> AlignmentMap 
 union' = M.unionWith combineMeta
 
 -- | Perform the union of n maps combining the metadata of equivalent
 -- alignments
-unions' :: [M.Map AlignedTrees Meta] -> M.Map AlignedTrees Meta
+unions' :: [AlignmentMap] -> AlignmentMap
 unions' = M.unionsWith combineMeta
 
 -- | Type of manual annotations.
@@ -200,7 +203,7 @@ align as cs p cl ex (s:ss) = align (S.fromList $ map keyval2alignment (M.toList 
   where as' = M.fromListWith combineMeta (S.toList $ S.map alignment2keyval as)
 
 -- | Sentence-level alignment function. Can be use independently of align   
-alignSent :: M.Map AlignedTrees Meta   -- ^ a map of known alignments (e.g. 
+alignSent :: AlignmentMap              -- ^ a map of known alignments (e.g. 
                                        -- from statistical tools)
           -> [Criterion]               -- ^ a list of criteria (sorted by 
                                        -- priority)
@@ -211,8 +214,8 @@ alignSent :: M.Map AlignedTrees Meta   -- ^ a map of known alignments (e.g.
                                        -- alignment "by exclusion" should also 
                                        -- be performed 
           -> (UDSentence,UDSentence)   -- ^ the sentences to align 
[…]

### 2021-04-30 paper/paper.tex
wrote draft introduction (needed to be rewritten from scratch)

@@
 % and set <dim> to something 5cm or larger.
-% TODO: title
-\title{Instructions for *ACL Proceedings}
+\title{Syntax-Based Concept Alignment for Domain-Specific Machine Translation}
 
@@
 \section{Introduction}
+Grammar-based systems such as GF \cite{TODO:} have been successfully employed in domain-specific Machine Translation (MT). 
+What makes these systems well suited to the task is the fact that, when we constrain ourselves to a specific domain, where precision is most often more important than coverage, they can provide strong guarantees in terms of grammatical correctness.
+
+Nevertheless, lexical exactness is, in this context, just as important as grammaticality.
+An important part of the design of the Controlled Natural Language (CNL) the grammar in such a system describes becomes, then, the creation of a translation lexicon.
+In many cases, this is done for the most part manually, resulting in a time consuming task requiring significant linguistic knowledge.
+When the grammar is designed based on a parallel corpus of example sentences, it is possible to automate part of this process by means of statistical word and phrase alignment techniques.
+None of them is, however, suitable for the common case in which only a limited number of example sentences is available.
+
+In this paper, we propose an alternative approach to the automation of this task.
+While still being data-driven, our method is grammar-based and, as such, capable of extracting meaningful correspondences even from individual sentence pairs.
+
+A further advantage of performing the syntactic analysis of the sentences is that we do not have to choose whether to focus on the word or phrase level.
+Instead, we can simultaneously operate at different levels of abstraction, thus extracting both single- and multiword -even non-contiguous- correspondences.
+For this reason, we refer to the task our system attempts to automate as \textit{Concept Alignment} (CA).
+We call \textit{concepts} the abstract units of translation, composed of any number of words, the system identifies, and represent them as \textit{alignments}, i.e.
+pairs of semantically equivalent concrete expressions.
 

### 2021-05-03 ConceptAlignment.hs
restored selectForMT

@@ -469,28 +469,27 @@ alignment2sentencePair a =
 -- - remove redundant alignments (i.e. alignments that can be inferred
 --   from their sub-alignments)
 -- - optionally, remove alignments where the size of both trees is > n
---selectForMT :: Maybe Int -> [(Alignment,Info)] -> [(Alignment,Info)]
---selectForMT mmax as = nubBy 
---                  (\(a,_) (b,_) -> b `contains` a && isPerfectShallow b) 
---                  (reverse $ sort as'')
---  where 
---    as' = filter (\(a,_) -> hasContent a) as -- remove function-only stuff
---      where hasContent (A (t,u)) = (not . null) (contentTags t) 
---                                || (not . null) (contentTags u)
---    as'' = case mmax of
---      (Just m) -> 
---        filter (\(A (t,u),_) -> sizeRTree t <= m || sizeRTree u <= m) as'
---      Nothing -> as'
+selectForMT :: Maybe Int -> S.Set Alignment -> S.Set Alignment
+selectForMT mmax as = S.fromList $ nubBy 
+                  (\a b -> b `contains` a && isPerfectShallow b) 
+                  (reverse $ sort as''')
+  where 
+    as' = S.toList as
+    as'' = filter (hasContent . trees) as' -- remove function-only stuff
+      where hasContent (AT (t,u)) = (not . null) (contentTags t) 
+                                 || (not . null) (contentTags u)
+    as''' = case mmax of
+      (Just m) -> 
+        filter (\a -> sizeRTree (sl a) <= m || sizeRTree (tl a) <= m) as''
+      Nothing -> as''
 
 -- | Check if an alignment is perfect, i.e. ig the structure of the two trees
 -- is the same
---isPerfect :: Alignment -> Bool
---isPerfect (A (t,u)) = abstractUDTree t == abstractUDTree u 
+isPerfect :: Alignment -> Bool
+isPerfect a = abstractUDTree (sl a) == abstractUDTree (tl a) 
 
 -- | Check the top-level structure of the two trees composing an alignment
 -- is the same
---isPerfectShallow :: Alignment -> Bool
---isPerfectShallow (A (t,u)) 
---  = root t' == root u' 
---  && map root (childrenRTree t') == map root (childrenRTree u') 
---  where (t',u') = (abstractUDTree t,abstractUDTree u)
+isPerfectShallow :: Alignment -> Bool
[…]

### 2021-05-03 ExtractConcepts.hs
made use of sorting and selection again

@@ -33,8 +33,8 @@ main = do
         r <- getPattern flags
         let as = align smtAs criteria r segment byExcl tus
         let m = listToMaybe [read mmax :: Int | MaxSize mmax <- flags]
-        let as' = if All `elem` flags then as else as --TODO: selectForMT m as
-        let as'' = {-TODO: sortByConfidence $-} S.toList as' 
+        let as' = if All `elem` flags then as else selectForMT m as
+        let as'' = sortByConfidence (S.toList as') 
         if Linearize `elem` flags
           then 
             if isJust fp 

### 2021-05-03 ConceptAlignment.hs
fixed --rest option

@@ -215,7 +215,7 @@ alignSent :: AlignmentMap              -- ^ a map of known alignments (e.g.
                                        -- be performed 
           -> (UDSentence,UDSentence)   -- ^ the sentences to align 
           -> AlignmentMap              -- ^ a map of alignments
-alignSent as cs p cl ex s@(s1,s2) = if ex then extra else extra `M.union` basic
+alignSent as cs p cl ex s@(s1,s2) = if ex then extra `M.union` basic else basic
   where 
     basic = if cl then alignClauses (t,u) else alignSent' cs (t,u)
     extra = alignRest basic -- alignments obtained "by exclusion"

### 2021-05-03 ExtractConcepts.hs

@@ -75,7 +75,7 @@ main = do
               return $ Just (read patternText :: UDPattern)
             else return Nothing
 
--- | Sort alignments by how likely theyare to be correct (kinda). 
+-- | Sort alignments by how likely they are to be correct (kinda). 
 sortByConfidence :: [Alignment] -> [Alignment]
 sortByConfidence = sortOn (\a -> 
                         let rs = S.toList (reasons $ meta a) \\ [HEAD, PREV] 

### 2021-05-03 FastAlignUtils.hs
adapted fast align utils to new types

@@ -5,10 +5,11 @@ import Data.List
 import Data.List.Split
 import qualified Data.Set as S
 import qualified Data.Map as M
+import RTree
 import UDConcepts
 import ConceptAlignment
 
--- parse a string in pharaoh format to a list of one-to-many indices
+-- | Parse a string in pharaoh format to a list of one-to-many indices
 -- correspondences 
 parsePh :: String -> [[(Int,[Int])]]
 parsePh s = map (oneToMany . map toIdxPair . words) (lines s)
@@ -18,31 +19,31 @@ parsePh s = map (oneToMany . map toIdxPair . words) (lines s)
             where [i,j] = splitOn "-" p
         oneToMany = M.toList . M.fromListWith (++)
 
--- parse a bitext in fast_align format to a list of pairs of sentences
+-- | Parse a bitext in fast_align format to a list of pairs of sentences
 parseBi :: String -> [(String, String)]
 parseBi s = [(head s',last s') | s' <- map (splitOn " ||| ") (lines s)]
 
--- given a list of (target, language) sentences and a list of alignment
+-- | Given a list of (target, language) sentences and a list of alignment
 -- indices, return the corresponding pairs of (multi)words as 
--- LinAlignments
-phToLas :: [(String,String)] -> [[(Int,[Int])]] -> [LinAlignment]
-phToLas ss is = 
-    map 
-        (\(lts,(r,o)) -> LAlignment lts (S.toList r) o) 
-        (M.toList . M.fromListWith combineVals $ phToLas' ss is)
-  where 
-    phToLas' _ [] = [] -- length is supposed to be the same though
-    phToLas' [] _ = []
-    phToLas' (s:ss) (ps:pss) = phToLa s ps ++ phToLas' ss pss
-        where
-            phToLa _ [] = []
-            phToLa (s,t) ((i,js):ps) = 
-                (lt,(S.singleton FAST,1)):phToLa (s,t) ps
-                where 
-                    lt = (filter (/= '\n') one,filter (/= '\n') many) 
-                    one = words s !! i
-                    many = unlines $ reverse $ 
-                           intersperse " " [words t !! j | j <- js]
+-- a set of Alignments
+phFileToAlignments :: [(String,String)] -> [[(Int,[Int])]] -> AlignmentMap
+phFileToAlignments _ [] = M.empty -- length is supposed to be the same though
+phFileToAlignments [] _ = M.empty
[…]

### 2021-05-03 FastAlignUtils.hs
fixed indentation

@@ -13,11 +13,11 @@ import ConceptAlignment
 -- correspondences 
 parsePh :: String -> [[(Int,[Int])]]
 parsePh s = map (oneToMany . map toIdxPair . words) (lines s)
-    where 
-        toIdxPair :: String -> (Int,[Int])
-        toIdxPair p = (read i :: Int, [read j :: Int])
-            where [i,j] = splitOn "-" p
-        oneToMany = M.toList . M.fromListWith (++)
+  where 
+    toIdxPair :: String -> (Int,[Int])
+    toIdxPair p = (read i :: Int, [read j :: Int])
+      where [i,j] = splitOn "-" p
+    oneToMany = M.toList . M.fromListWith (++)
 
 -- | Parse a bitext in fast_align format to a list of pairs of sentences
 parseBi :: String -> [(String, String)]
@@ -29,33 +29,36 @@ parseBi s = [(head s',last s') | s' <- map (splitOn " ||| ") (lines s)]
 phFileToAlignments :: [(String,String)] -> [[(Int,[Int])]] -> AlignmentMap
 phFileToAlignments _ [] = M.empty -- length is supposed to be the same though
 phFileToAlignments [] _ = M.empty
-phFileToAlignments (s:ss) (ps:pss) = phSentToAlignment s ps `union'` phFileToAlignments ss pss
-    where
-        phSentToAlignment :: (String,String) -> [(Int,[Int])] -> AlignmentMap
-        phSentToAlignment _ [] = M.empty
-        phSentToAlignment (src,trg) ((i,js):ps) = a `insert'` phSentToAlignment (src,trg) ps
-            where 
-                a = alignment2keyval $ (initAlignment $ AT (t,u)) { 
-                        meta = initMeta {
-                            reasons = S.singleton FAST
-                        }
-                }
-                (t,u) = (mockUDTree one,mockUDTree many) 
-                mockUDTree x = RTree ((initUDWord 1) { udFORM = filter (/= '\n') x }) []
-                one = words src !! i
-                many = unlines $ reverse $ intersperse " " [words trg !! j | j <- js]
+phFileToAlignments (s:ss) (ps:pss) = 
+    phSentToAlignment s ps `union'` phFileToAlignments ss pss
+  where
+    phSentToAlignment :: (String,String) -> [(Int,[Int])] -> AlignmentMap
+    phSentToAlignment _ [] = M.empty
[…]

### 2021-05-03 ExtractConcepts.hs
fixed use of fast alignmnents

@@ -11,7 +11,7 @@ import System.Exit
 import UDConcepts
 import UDPatterns
 import ConceptAlignment
---import FastAlignUtils
+import FastAlignUtils
 import Criteria 
 import ArgvParse
 
@@ -25,13 +25,14 @@ main = do
         ts <- parseUDFile src
         us <- parseUDFile trg
         let tus = zip ts us -- let tus = take 50 $ zip ts us
-        let smtAs = S.empty -- <- getSmtAlignments flags src trg
+        smtAs <- getSmtAlignments flags src trg
+        let smtAs' = S.fromList $ map keyval2alignment (M.toList smtAs)
         let segment = Clauses `elem` flags
         let byExcl = Rest `elem` flags
         let fp = listToMaybe [path | Path path <- flags] 
         let fp' = fromJust fp
         r <- getPattern flags
-        let as = align smtAs criteria r segment byExcl tus
+        let as = align smtAs' criteria r segment byExcl tus
         let m = listToMaybe [read mmax :: Int | MaxSize mmax <- flags]
         let as' = if All `elem` flags then as else selectForMT m as
         let as'' = sortByConfidence (S.toList as') 
@@ -54,18 +55,16 @@ main = do
   where 
     insertLang :: FilePath -> String -> FilePath
     insertLang fp l = dropExtension fp ++ l ++ takeExtension fp
-    getSmtAlignments = undefined
-    --getSmtAlignments :: [Flag] -> FilePath -> FilePath -> IO [LinAlignment]
-    --getSmtAlignments flags src trg = case [path | Pharaoh path <- flags] of
-    --  [] -> return []
-    --  [path] -> do
-    --    indices <- readFile path >>= return . parsePh
-    --    srcConllu <- readFile src
-    --    trgConllu <- readFile trg
-    --    let bitext = take 100 $ parseBi $ conllu2bi (srcConllu,trgConllu)
-    --    print $ ltrees $ phToLas bitext indices !! 5
-    --    return $ phToLas bitext indices
-    --  _ -> undefined
+    getSmtAlignments :: [Flag] -> FilePath -> FilePath -> IO AlignmentMap
+    getSmtAlignments flags src trg = case [path | Pharaoh path <- flags] of
+      [] -> return M.empty
+      [path] -> do
+        indices <- readFile path >>= return . parsePh
+        srcConllu <- readFile src
+        trgConllu <- readFile trg
[…]

### 2021-05-03 ConceptAlignment.hs
simplified types by removing alignment as record

@@ -14,59 +14,16 @@ import UDPatterns
 
 -- | An alignment is a pair of corresponding UD (sub)trees associated to some 
 -- metadata 
-data Alignment = A {
-  trees :: AlignedTrees,    -- ^ pair of UD subtrees
-  meta :: Meta              -- ^ metadata
-} deriving (Show,Read,Eq,Ord)
-
--- | Initialize a potential alignment given just the pair of trees
-initAlignment :: AlignedTrees -> Alignment
-initAlignment tu = A {
-  trees = tu,
-  meta = initMeta
-}
-
--- | Convert an alignment into a (key,value) pair, where the key is the tree
--- pair and the "value" is the metadata
-alignment2keyval :: Alignment -> (AlignedTrees,Meta)
-alignment2keyval a = (trees a,meta a)
-
--- | Convert a (key,value) pair, where the key is the tree pair and the 
--- "value" is the metadata, into an Alignment
-keyval2alignment :: (AlignedTrees,Meta) -> Alignment
-keyval2alignment (k,v) = A {
-  trees = k,
-  meta = v
-}
-
--- | Left (source language) tree of an alignment
-sl :: Alignment -> UDTree
-sl a = let (AT (t,u)) = trees a in t
-
--- | Right (target language) tree of an alignment
-tl :: Alignment -> UDTree
-tl a = let (AT (t,u)) = trees a in u
-
--- | prAlignment is used instead of show for inspecting the alignments in an
--- easier way, e.g. in EvAlign and for debugging
-prAlignment :: Alignment -> String
-prAlignment a = prTrees a 
-                ++ "    --"
-                ++ " reasons: " ++ showSet (reasons $ meta a)
-                ++ " sentence IDs: " ++ showSet (sentIds $ meta a)
-  where 
-    prTrees a = linearize (sl a) ++ " ||| " ++ linearize (tl a)
-    showSet s = "{" ++ intercalate ", " (S.toList $ S.map show s) ++ "}"
-
--- | Check if an alignment "contains" another;
--- used both in pruning and selection of alignments for MT
-contains :: Alignment -> Alignment -> Bool
-a `contains` b = (sl b `isSubRTree` sl a) && (tl b `isSubRTree` tl a)
+type Alignment = (AlignedTrees,Meta) 
 
 -- | AlignedTrees basically represents a pair of UD trees, but it is a newtype
 -- due to the custom implementation of (==) 
 newtype AlignedTrees = AT (UDTree,UDTree)
   deriving (Show,Read)
+
[…]

### 2021-05-03 ExtractConcepts.hs

@@ -26,7 +26,7 @@ main = do
         us <- parseUDFile trg
         let tus = zip ts us -- let tus = take 50 $ zip ts us
         smtAs <- getSmtAlignments flags src trg
-        let smtAs' = S.fromList $ map keyval2alignment (M.toList smtAs)
+        let smtAs' = S.fromList $ M.toList smtAs
         let segment = Clauses `elem` flags
         let byExcl = Rest `elem` flags
         let fp = listToMaybe [path | Path path <- flags] 

### 2021-05-03 FastAlignUtils.hs

@@ -37,11 +37,7 @@ phFileToAlignments (s:ss) (ps:pss) =
     phSentToAlignment (src,trg) ((i,js):ps) = 
         a `insert'` phSentToAlignment (src,trg) ps
       where 
-        a = alignment2keyval $ (initAlignment $ AT (t,u)) { 
-            meta = initMeta {
-              reasons = S.singleton FAST
-            }
-        }
+        a = (AT (t,u),initMeta { reasons = S.singleton FAST })
         (t,u) = (mockUDTree one,mockUDTree many) 
         mockUDTree x = 
             RTree ((initUDWord 1) { udFORM = filter (/= '\n') x }) []

### 2021-05-04 ConceptAlignment.hs
use lists instead of sets to avoid unnecessary conversions

@@ -87,7 +87,7 @@ prAlignment a = prTrees a
 contains :: Alignment -> Alignment -> Bool
 a `contains` b = (sl b `isSubRTree` sl a) && (tl b `isSubRTree` tl a)
 
--- | Map of alignments
+-- | Map of alignments (used internally to simplify combining metadata)
 type AlignmentMap = M.Map AlignedTrees Meta
 
 {- Helper functions for alignment maps-}
@@ -175,7 +175,7 @@ clause = C (\t u -> divClause t u || divClause u t) (S.singleton CL) True False
 {- Alignment functions -}
 
 -- | Align a list of pairs of / corresponding / dependency trees   
-align :: S.Set Alignment           -- ^ a set of known alignments (e.g. from 
+align :: [Alignment]               -- ^ a set of known alignments (e.g. from 
                                    -- statistical tools)
       -> [Criterion]               -- ^ a list of criteria (sorted by 
                                    -- priority)
@@ -185,10 +185,11 @@ align :: S.Set Alignment           -- ^ a set of known alignments (e.g. from
       -> Bool                      -- ^ a flag indicating whether alignment  
                                    -- "by exclusion" should also be performed 
       -> [(UDSentence,UDSentence)] -- ^ the list of sentences to align 
-      -> S.Set Alignment           -- ^ a set of alignments
+      -> [Alignment]               -- ^ a set of alignments, implemented as
+                                   -- a list to avoid unnecessary conversions
 align as _ _ _ _ [] = as
-align as cs p cl ex (s:ss) = align (S.fromList $ M.toList $ alignSent as' cs p cl ex s) cs p cl ex ss
-  where as' = M.fromListWith combineMeta (S.toList as)
+align as cs p cl ex (s:ss) = align (M.toList $ alignSent as' cs p cl ex s) cs p cl ex ss
+  where as' = M.fromListWith combineMeta as
 
 -- | Sentence-level alignment function. Can be use independently of align   
 alignSent :: AlignmentMap              -- ^ a map of known alignments (e.g. 
@@ -453,19 +454,18 @@ alignment2sentencePair a =
 -- - remove redundant alignments (i.e. alignments that can be inferred
 --   from their sub-alignments)
 -- - optionally, remove alignments where the size of both trees is > n
-selectForMT :: Maybe Int -> S.Set Alignment -> S.Set Alignment
-selectForMT mmax as = S.fromList $ nubBy 
[…]

### 2021-05-04 ExtractConcepts.hs

@@ -26,16 +26,15 @@ main = do
         us <- parseUDFile trg
         let tus = zip ts us -- let tus = take 50 $ zip ts us
         smtAs <- getSmtAlignments flags src trg
-        let smtAs' = S.fromList $ M.toList smtAs
         let segment = Clauses `elem` flags
         let byExcl = Rest `elem` flags
         let fp = listToMaybe [path | Path path <- flags] 
         let fp' = fromJust fp
         r <- getPattern flags
-        let as = align smtAs' criteria r segment byExcl tus
+        let as = align smtAs criteria r segment byExcl tus
         let m = listToMaybe [read mmax :: Int | MaxSize mmax <- flags]
         let as' = if All `elem` flags then as else selectForMT m as
-        let as'' = sortByConfidence (S.toList as') 
+        let as'' = sortByConfidence as' 
         if Linearize `elem` flags
           then 
             if isJust fp 
@@ -55,15 +54,15 @@ main = do
   where 
     insertLang :: FilePath -> String -> FilePath
     insertLang fp l = dropExtension fp ++ l ++ takeExtension fp
-    getSmtAlignments :: [Flag] -> FilePath -> FilePath -> IO AlignmentMap
+    getSmtAlignments :: [Flag] -> FilePath -> FilePath -> IO [Alignment]
     getSmtAlignments flags src trg = case [path | Pharaoh path <- flags] of
-      [] -> return M.empty
+      [] -> return []
       [path] -> do
         indices <- readFile path >>= return . parsePh
         srcConllu <- readFile src
         trgConllu <- readFile trg
         let bitext = parseBi $ conllu2bi (srcConllu,trgConllu)
-        return $ phFileToAlignments bitext indices
+        return $ M.toList $ phFileToAlignments bitext indices
       _ -> undefined
     getPattern :: [Flag] -> IO (Maybe UDPattern)
     getPattern flags = do

### 2021-05-04 FastAlignUtils.hs

@@ -45,7 +45,7 @@ phFileToAlignments (s:ss) (ps:pss) =
         many = unlines $ reverse $ intersperse " " [words trg !! j | j <- js]
 
 -- | CoNLL-U aligned strings to fast-align bitext conversion (useful to feed 
---   fast_aling the same tokens)
+--   fast_align the same tokens)
 conllu2bi :: (String,String) -> String
 conllu2bi (c1,c2) = unlines $ zipWith (\s1 s2 -> s1 ++ " ||| " ++ s2) ts1 ts2
   where

### 2021-05-04 ConceptAlignment.hs
restored propagation!

@@ -326,52 +326,53 @@ alignSent as cs p cl ex s@(s1,s2) = if ex then extra `M.union` basic else basic
 {- Propagation functions -}
 
 -- | Generic (not optimized for same text in n languages) propagation function
---propagate :: [Criterion]            -- ^ a list of criteria 
---                                    -- (sorted by priority)
---          -> Bool                   -- ^ -- ^ a flag indicating whether clause 
---                                    -- segmentation should be performed
---          -> Bool                   -- ^ a flag indicating whether alignment
---                                    --"by exclusion" should also be performed 
---          -> ([UDTree],[UDTree])    -- a pair of lists of UD trees (the
---                                    -- sentences to propagate on) in L1, L2
---          -> UDTree                 -- a previously extracted L1 concept
---          -> Maybe (Alignment,Info) -- an (Alignment,Info) pair, if found
---propagate cs segment byExcl ([],_) _ = Nothing 
---propagate cs segment byExcl (t:ts,u:us) c =
---  let 
---    as = M.toList $ alignSent M.empty [] cs Nothing segment byExcl (t,u)
---    as' = case (c `isSubUDTree'` t, c `isHeadSubUDTree` t) of
---      (True,_) -> sortOnDepth as
---      (_,True) -> sortOnDepth (filter (\(_,(rs,_)) -> HEAD `elem` rs) as)
---      (False,False) -> []
---    in case find (\(a,_) -> c =~ sl a) as' of
---      Nothing -> propagate cs segment byExcl (ts,us) c
---      n -> n
---    where 
---      -- difference between the depth of the SL aligned subtree and c:
---      -- the smaller, the better (mostly used to avoid that root nodes are
---      -- aligned with full sentences, but also makes sense in general)
---      depthDiff :: RTree a -> Int
---      depthDiff t = abs (depthRTree t - depthRTree c)
---      sortOnDepth :: [(Alignment,Info)] ->[(Alignment,Info)]
---      sortOnDepth = sortOn (depthDiff . sl . fst)
---      -- check if c is the head of one of t's subtrees
---      isHeadSubUDTree :: UDTree -> UDTree -> Bool
---      isHeadSubUDTree c t = 
---        isJust $ listToMaybe $ 
---          sortOn depthDiff (filter (isHeadUDTree c) (allSubRTrees t))
[…]

### 2021-05-04 ConceptAlignment.hs
cleaned up CA library

@@ -21,7 +21,8 @@ type Alignment = (AlignedTrees,Meta)
 newtype AlignedTrees = AT (UDTree,UDTree)
   deriving (Show,Read)
 
--- | Return the aligned trees of an alignment, just like when it was a record type
+-- | Return the aligned trees of an alignment, just like when it was a record 
+-- type
 trees :: Alignment -> AlignedTrees
 trees (tu,_) = tu
 
@@ -65,11 +66,11 @@ initAlignment tu = (tu,initMeta)
 
 -- | Left (source language) tree of an alignment
 sl :: Alignment -> UDTree
-sl ((AT (t,_)),_) = t
+sl (AT (t,_),_) = t
 
 -- | Right (target language) tree of an alignment
 tl :: Alignment -> UDTree
-tl ((AT (_,u)),_) = u
+tl (AT (_,u),_) = u
 
 -- | prAlignment is used instead of show for inspecting the alignments in an
 -- easier way, e.g. in EvAlign and for debugging
@@ -165,7 +166,8 @@ data Criterion = C {
 -- | Special criterion necessary for effective clause segmentation
 -- (that's why it is defined here and not in module Criteria)
 clause :: Criterion
-clause = C (\t u -> divClause t u || divClause u t) (S.singleton CL) True False
+clause = 
+  C (\t u -> divClause t u || divClause u t) (S.singleton CL) True False
   where 
     -- a clause of a certain type is translated as a clause of another type 
     divClause :: UDTree -> UDTree -> Bool
@@ -188,7 +190,8 @@ align :: [Alignment]               -- ^ a set of known alignments (e.g. from
       -> [Alignment]               -- ^ a set of alignments, implemented as
                                    -- a list to avoid unnecessary conversions
 align as _ _ _ _ [] = as
-align as cs p cl ex (s:ss) = align (M.toList $ alignSent as' cs p cl ex s) cs p cl ex ss
+align as cs p cl ex (s:ss) = 
+  align (M.toList $ alignSent as' cs p cl ex s) cs p cl ex ss
   where as' = M.fromListWith combineMeta as
 
 -- | Sentence-level alignment function. Can be use independently of align   
@@ -204,7 +207,8 @@ alignSent :: AlignmentMap              -- ^ a map of known alignments (e.g.
                                        -- be performed 
           -> (UDSentence,UDSentence)   -- ^ the sentences to align 
           -> AlignmentMap              -- ^ a map of alignments
[…]

### 2021-05-04 GenerateGrammar.hs
restored grammar generation!

@@ -187,13 +187,13 @@ isConllu p = takeExtension p == ".conllu"
 -- the lists in the resulting sentences are also aligned implicitly) 
 getAlignments :: [[UDSentence]] -> [[UDSentence]]
 getAlignments ss =  
-  transpose $ mapMaybe getAlignment [(o + 1)..(o + maximum (map length ss))]
+  transpose $ mapMaybe getAlignment ids
     where
-      getAlignment :: Int -> Maybe [UDSentence]
-      getAlignment i = 
-        let as = map (find (\s -> sentId s == i)) ss
+      ids = nub $ map sentId (concat ss)
+      getAlignment :: String -> Maybe [UDSentence]
+      getAlignment id = 
+        let as = map (find (\s -> sentId s == id)) ss
         in if all isJust as then Just $ map fromJust as else Nothing
-      o = 1000000 -- numbering offset (cf. prUDSentence in UDConcepts)
 
 -- | Like UDAnnotations.getEnv, but uses pgf instead of paths
 -- (this is a bad solution but I don't necessarily want to rely on the pgf

### 2021-05-05 ConceptAlignment.hs
improved alignment printing code

@@ -16,50 +16,6 @@ import UDPatterns
 -- metadata 
 type Alignment = (AlignedTrees,Meta) 
 
--- | AlignedTrees basically represents a pair of UD trees, but it is a newtype
--- due to the custom implementation of (==) 
-newtype AlignedTrees = AT (UDTree,UDTree)
-  deriving (Show,Read)
-
--- | Return the aligned trees of an alignment, just like when it was a record 
--- type
-trees :: Alignment -> AlignedTrees
-trees (tu,_) = tu
-
--- | Two pairs of aligned trees are considered equal whenever their
--- linearizations are the same
-instance Eq AlignedTrees where
-  AT (t1,u1) == AT (t2,u2) = linearize t1 == linearize t2 
-                          && linearize u1 == linearize u2
-
--- | Aligned trees can be ordered based on:
--- 1. size of their left tree
--- 2. size of their right tree
--- 3. "linearization" of their left tree (alphabetically)
--- 4. "linearization" of their right tree (alphabetically)
-instance Ord AlignedTrees where
-  a <= b = k a <= k b
-    where 
-      k x = (depthRTree sla,depthRTree tla,linearize sla,linearize tla)
-        where AT (sla,tla) = x
-
--- | The metadata of an alignment are:
-data Meta = M {
-  reasons :: S.Set Reason,  -- ^ reasons for alignment
-  sentIds :: S.Set String   -- ^ ids of the sentences it was inferred from
-} deriving (Show,Read,Eq,Ord)
-
--- | Return the metadata of an alignment, just like when it was a record type
-meta :: Alignment -> Meta
-meta (_,m) = m
-
--- | Initialize metadata with default vals
-initMeta :: Meta
-initMeta = M {
-  reasons = S.empty,
-  sentIds = S.empty
-}
-
 -- | Initialize a potential alignment given just the pair of trees
 initAlignment :: AlignedTrees -> Alignment
 initAlignment tu = (tu,initMeta)
@@ -72,21 +28,73 @@ sl (AT (t,_),_) = t
 tl :: Alignment -> UDTree
 tl (AT (_,u),_) = u
 
--- | prAlignment is used instead of show for inspecting the alignments in an
--- easier way, e.g. in EvAlign and for debugging
-prAlignment :: Alignment -> String
-prAlignment a = prTrees a 
-                ++ "    --"
-                ++ " reasons: " ++ showSet (reasons $ meta a)
[…]

### 2021-05-05 PropagateConcepts.hs

@@ -30,8 +30,12 @@ main = do
                 if Linearize `elem` flags
                     then 
                         if isJust fp 
-                            then writeFile fp' (unlines $ map (maybe "" prAlignment) as) 
-                            else mapM_ (putStrLn . maybe "" prAlignment) as
+                            then writeFile fp' (unlines $ map 
+                                        (maybe "" prLinearizedAlignment) 
+                                        as) 
+                            else mapM_ 
+                                    (putStrLn . maybe "" prLinearizedAlignment) 
+                                    as
                     else 
                         if isJust fp 
                             then do

### 2021-05-05 ConceptAlignment.hs
write metadata in conllu files

@@ -457,11 +457,15 @@ abstractUDTree = mapRTree udSimpleDEPREL
 {- Alignments to CoNLL-U files -}
 
 -- | convert an alignment into a pair of CoNNL-U sentences
-alignment2sentencePair :: Alignment -> (UDSentence, UDSentence)
+alignment2sentencePair :: Alignment -> (UDSentence,UDSentence)
 alignment2sentencePair a = 
-  (udTree2adjustedSentence $ sl a, udTree2adjustedSentence $ tl a)
+  (addMetaAsComment $ udTree2adjustedSentence $ sl a, 
+  addMetaAsComment $ udTree2adjustedSentence $ tl a)
   where 
     udTree2adjustedSentence = adjustUDIds . udTree2sentence . createRoot
+    addMetaAsComment s = s {
+      udCommentLines = ["# " ++ (prMeta $ meta a)]
+    } 
 
 
 {- Selection of alignments for MT -}

### 2021-05-05 paper/paper.tex
extended and improved introduction and added some structure

@@
 \section{Introduction}
-Grammar-based systems such as GF \cite{TODO:} have been successfully employed in domain-specific Machine Translation (MT). 
+Grammar-based translation pipelines such as those based on Grammatical Framework (GF) \cite{TODO:} have been successfully employed in domain-specific Machine Translation (MT). 
 What makes these systems well suited to the task is the fact that, when we constrain ourselves to a specific domain, where precision is most often more important than coverage, they can provide strong guarantees in terms of grammatical correctness.
@@
 In many cases, this is done for the most part manually, resulting in a time consuming task requiring significant linguistic knowledge. 
-When the grammar is designed based on a parallel corpus of example sentences, it is possible to automate part of this process by means of statistical word and phrase alignment techniques.
+When the grammar is designed based on a parallel corpus of example sentences, it is possible to automate part of this process by means of statistical word and phrase alignment techniques \cite{TODO: a lot}.
 None of them is, however, suitable for the common case in which only a limited number of example sentences is available.
@@
 
-A further advantage of performing the syntactic analysis of the sentences is that we do not have to choose whether to focus on the word or phrase level.
-Instead, we can simultaneously operate at different levels of abstraction, thus extracting both single- and multiword -even non-contiguous- correspondences.
+A further advantage of performing syntactic analysis is that we do not have to choose \textit{a priori} whether to focus on the word or phrase level. 
+Instead, we can simultaneously operate at different levels of abstraction, thus extracting both single- and multiword, even non-contiguous, correspondences.
+
 For this reason, we refer to the task our system attempts to automate as \textit{Concept Alignment} (CA).
@@
 
-\section{Methodology}
+This paper is structured as follows. 
+Section \ref{methodology} starts by giving an overview of our approach to CA to then focus on our algorithm for extracting correspondences.
+It is concluded with a description of our method for converting the alignments obtained in this way to a GF translation lexicon.
+After that, Section \ref{evaluation} presents the results of our first evaluation of the system.
+Finally, Section \ref{conclusions} consists of a discussion of such results and some ideas for future work.
 
-\section{Evaluation}
+\section{Methodology} \label{methodology}
 
-\section{Conclusions}
+%overview
+
+\subsection{Extracting concepts}
+
+\subsubsection{Refinements of the basic algorithm}
+
+\subsection{Generating grammar rules}
+
+\section{Evaluation} \label{evaluation}
+
+\section{Conclusions} \label{conclusions}
 

### 2021-05-05 paper/paper.tex
wrote most of Methodology

@@
 For this reason, we refer to the task our system attempts to automate as \textit{Concept Alignment} (CA).
-We call \textit{concepts} the abstract units of translation, composed of any number of words, the system identifies, and represent them as \textit{alignments}, i.e.
-pairs of semantically equivalent concrete expressions.
 
 This paper is structured as follows. 
-Section \ref{methodology} starts by giving an overview of our approach to CA to then focus on our algorithm for extracting correspondences.
+Section \ref{methodology} starts by giving an overview of our approach to CA, comparing it to related work, to then focus on our algorithm for extracting correspondences.
 It is concluded with a description of our method for converting the alignments obtained in this way to a GF translation lexicon.
@@
 \section{Methodology} \label{methodology}
+The objective of CA is to find semantical correspondences between parts of multilingual parallel texts. 
+We call \textit{concepts} the abstract units of translation, composed of any number of words, identified through this process, and represent them as \textit{alignments}, i.e.
+tuples of equivalent concrete expressions in different languages.
 
-%overview
+The basic use case for CA, which we refer to specifically as \textit{Concept Extraction} (CE), is the generation of a translation lexicon from a parallel text.
+This can be directly compared to the numerous existing word and phrase alignment techniques.
+
+An interesting and less studied variant of CA is \textit{Concept Propagation} (CP), useful for cases where a set of concepts is already known and the goal is to identify the expressions corresponding to each of them in a new language, potentially even working with a different text in the same domain.
+In this paper, however, we focus on CE, and restrict its application to bilingual corpora.
+
+As stated in the Introduction, most existing extraction solutions are based on statistical approaches and are, as a consequence, unsuitable for small datasets. 
+Grammar-based approaches, making use of parallel treebanks and collectively referred to as \textit{tree-to-tree alignment methods}, have also been proposed \cite{TODO:}, but have historically suffered from the inconsistencies between the various constituency grammar formalisms used to define grammars for the different languages and from the insufficient degree of robustness of existing parses.
+
+This work is a new attempt in the same direction, enabled by two interlingually oriented grammar formalisms developed over the course of the last 25 years: Universal Dependencies (UD) \cite{TODO:} and Grammatical Framework (GF) \cite{TODO:}.
+
+While both formalisms, the former a dependency and the latter a constituency grammar, independently solve the former issue, UD is especially appealing since dependency trees are an easier target for parsing.
+As a consequence, several robust dependency parsers, such as \cite{TODO: UDPipe} and \cite{TODO: Standford} are available.
+
+UD alone would then be sufficient to extract (or propagate) tree-to-tree alignments, but not to automate the generation of a ready-to-use, morphologically aware translation lexicon.
+This is where GF comes into play: after correspondences are inferred from a parallel text, our proposed system is able to convert them to GF grammar rules, easy to embed in a domain-specific grammar but also making it possible to immediately evaluate the system by carrying out small-scale translation experiment using pre-existing grammatical constructions implemented in GF's Resource Grammar Library (RGL) \cite{TODO:}.
+This is made possible by \texttt{gf-ud}, a conversion tool described in \cite{TODO:} and \cite{TODO:}.
+
+% system overview
 

### 2021-05-06 paper/paper.tex
finished general part of methodology, including diagram

@@
 \usepackage{microtype}
+
+\usepackage{graphicx}
 
@@
 
+% potentially move to future work directly
 An interesting and less studied variant of CA is \textit{Concept Propagation} (CP), useful for cases where a set of concepts is already known and the goal is to identify the expressions corresponding to each of them in a new language, potentially even working with a different text in the same domain.
-In this paper, however, we focus on CE, and restrict its application to bilingual corpora.
+While our system implements basic CP functionalities, however, in this paper we focus on CE and restrict its application to bilingual corpora.
 
@@
 
-This work is a new attempt in the same direction, enabled by two interlingually oriented grammar formalisms developed over the course of the last 25 years: Universal Dependencies (UD) \cite{TODO:} and Grammatical Framework (GF) \cite{TODO:}.
+This work is a new attempt in the same direction, enabled by two multilinguality-oriented grammar formalisms developed over the course of the last 25 years: Universal Dependencies (UD) \cite{TODO:} and Grammatical Framework (GF) \cite{TODO:}.
 
@@
 
-UD alone would then be sufficient to extract (or propagate) tree-to-tree alignments, but not to automate the generation of a ready-to-use, morphologically aware translation lexicon.
+UD parsing alone would then be sufficient to extract (or propagate) tree-to-tree alignments, but not to automate the generation of a ready-to-use, morphologically-aware translation lexicon.
 This is where GF comes into play: after correspondences are inferred from a parallel text, our proposed system is able to convert them to GF grammar rules, easy to embed in a domain-specific grammar but also making it possible to immediately evaluate the system by carrying out small-scale translation experiment using pre-existing grammatical constructions implemented in GF's Resource Grammar Library (RGL) \cite{TODO:}.
@@
 
-% system overview
+Concretely, the system we propose requires the following elements, whose reciprocal relations are shown in Figure \ref{overview}: \smallskip
+
+\begin{itemize}
+  \item a UD parser
+  \item an alignment module based on dependency tree comparison
+  \item a program, based on \texttt{gf-ud}, that converts the alignments into GF grammar rules and uses them to construct a translation lexicon
+\end{itemize} \smallskip
+
+\begin{figure}[h]
+%  \centering
+  \includegraphics[width=0.45\textwidth]{figures/overview.png}
+  \caption[System overview]{System overview. Parts in grey are currently at the prototype stage and will not be further discussed in this paper.} \label{elems}
+%  \label{overview}
+\end{figure}
 

### 2021-05-06 paper/paper.tex
ideas for CE part
kinda described extraction algorithm

@@
 
+% TODO: replace CE with CA and omit CP entirely?
 \begin{figure}[h]
-%  \centering
+  \centering
   \includegraphics[width=0.45\textwidth]{figures/overview.png}
   \caption[System overview]{System overview. Parts in grey are currently at the prototype stage and will not be further discussed in this paper.} \label{elems}
-%  \label{overview}
+  \label{overview}
 \end{figure}
@@
 \subsection{Extracting concepts} 
+The core part of the system outlined above is of course the alignment module.
+In this section, we present our method for extracting alignments from parallel bilingual UD treebanks, which can be obtained from sentence-aligned texts with any UD parser.
 
-\subsubsection{Refinements of the basic algorithm}
+The basic extraction algorithm, whose pseudocode is shown in Figure \ref{TODO}, takes as input a list of priority-sorted \textit{alignment criteria}, rules to determine whether two dependency trees should be aligned with each other which we will discuss in greater detail in Section \ref{criteria}, and a pair of UD trees corresponding to the sentences to align.
+From an implementation point of view, UD trees are, as in \texttt{gf-ud}, rose trees where each node represents a token,  obtained from the CoNLL-U files produced by the UD parser.
+
+% TODO: add figure comapring tree representations
+
+As a first step, the program checks whether the two full sentence trees can be aligned with each other, i.e.
+if they match any of the alignment criteria.
+If this is the case, they are added to a collection of alignments, also represented as a pair of UD trees associated with some metadata, which is what the function will return after aligning all the dependency subtrees.
+% TODO: explain or omit metadata!
+
+Depending on which alignment criteria they match, the head of the two trees may or may not also be added to such collection. This is important... % TODO:
+
+The same procedure is applied recursively to all pairs of immediate subtrees of each sentence, until the leaves are reached or alignment is no longer possible due to lack of matching criteria.
+% TODO: explain & motivate sorting -> maybe say more about UD before?
+
+% TODO: pseudocode
+
+When working on corpora consisting of multiple sentences, this algorithm can be applied in an iterative fashion, so that knowledge gathered when a sentence pair is aligned can be reused when working on the following ones and to keep track of the number of occurrences of each alignment throughout the entire text.
+
+\subsubsection{Alignment criteria} \label{criteria}
+% label matching
+% pos equivalence
+% known divergence -> cite Dorr, few examples
+% known alignment -> SMT integration
+
+% how tracking them is useful (but maybe in evaluation?)
+
+\subsubsection{Pattern matching}
+% look for specific UD patterns
 
@@
 \section{Evaluation} \label{evaluation}
+% we use udpipe
 
 \section{Conclusions} \label{conclusions}
+% tangible output and were it is
 

### 2021-05-07 paper/paper.tex
add theorem env

@@
 \usepackage{graphicx}
+
+\usepackage{amsthm}
+\newtheoremstyle{indented}% name of the style to be used
+  {5pt}% measure of space to leave above the theorem. E.g.: 3pt
+  {5pt}% measure of space to leave below the theorem. E.g.: 3pt
+  {\itshape}% name of font to use in the body of the theorem
+  { }% measure of space to indent
+  {\bfseries}% name of head font
+  { }% punctuation between head and body
+  { }% space after theorem head; " " = normal interword space
+  {\thmname{#1}\thmnote{ (#3)}}
+
+\theoremstyle{indented}
+
+\newtheorem{definition}{Definition}
 
@@
 \subsubsection{Alignment criteria} \label{criteria}
-% label matching
-% pos equivalence
-% known divergence -> cite Dorr, few examples
-% known alignment -> SMT integration
+While alignment criteria are customizable, to better understand the extraction algorithm described in the above we dedicate this section to describing the ones that are currently used by default.
+
+\paragraph{Matching UD labels}
+The most obvious, but also most effective idea is to determine alignability based on comparing the dependency labels of the candidate UD tree pair. 
+In particular, according to this idea, two subtrees, in \textit{matching context}, i.e. attached to aligned heads constitute an alignment if their roots share the same dependency label. 
+Intuitively, this means that they are in the same relation with their heads.
+
+Note that, since the root of US trees is always attached to a a fake node with an arc labelled \texttt{root}, this criterion also makes it so that full sentences are always considered to align with each other, something which is desirable since we assume the parallel texts that are fed to our program to be sentence-aligned.
+
+\paragraph{Part-Of-Speech equivalence}
+CoNNL-U files provide information not only on the syntactic role of the tokens a sentence is composed of, but also on their grammatical categories, represented as Universal Part-Of-Speech (POS) tags. 
+Intuitively, if the words corresponding to the nodes two trees in matching contexts are composed of have the same POS tags, those two trees are generally more likely to correspond to each other than if not.
+As a consequence, a useful relation to define between dependency trees is that of \textit{POS-equivalence}:
+
+\begin{definition}
+  Two dependency trees $t$, $u$ are \normalfont{POS-equivalent} (and, as such, will be aligned) if $M_1 = M_2 \neq \emptyset$, where $M_i$ is defined as the multiset of POS tags of all the \normalfont{meaning-carrying} \textit{(see below) word nodes of $t_i$}.
+\end{definition}
+
+The definition specifies that the two multisets should not contain the POS tags of all the words in the sentence but rather only those of the \textit{meaning-carrying} ones, referring to words belonging to a particular set of classes.
+Words that should generally be taken into account, in fact, correspond roughly to content words (cf.
+Section \ref{upos}), but this term was deliberately avoided in Definition \ref{poseq} due to the fact that it can be useful also to include some function words, for instance pronouns and some kinds of determiners.
+In particular, the current implementation considers as meaning-carrying all words that belong to an open class - defined as in the UD documentation \cite{uddocs} - and numerals, but this set of tags has been obtained empirically working with English-Italian and might not be ideal for all language pairs.
+On the other hand, words belonging to other classes, such as auxiliary verbs, adpositions and conjunctions can and should be in most cases ignored as they are often omitted or rendered with words with different POS tags when the sentence is translated to another language, especially if the two languages in the pair at hand differ significantly.
+\smallskip
+
+Applied alone, this criterion can be used to capture correspondences that would otherwise be missed, thus increasing recall, but a decrease in precision is also to be expected.
+Perhaps more interestingly, however, another way to apply this criterion is in conjunction with other ones, and in particular together with UD label matching (cf.
+Criterion \ref{udmatch}), in context where high precision is more important than recall.
+We will get back to combining criteria in Section \ref{ceres}.
+% combination
+
+\paragraph{Known translation divergence}
+% Dorr
+
+\paragraph{Known alignment}
+% example application: SMT integration
 

### 2021-05-07 GenerateGrammar.hs
language bounds in rule names, close #13

@@ -284,7 +284,7 @@ tree2rules env ltss = BuiltRules {
     -- construct function name (e.g. come_komma_Utt) using the lemmas in all
     -- n langs and the category in the first language
     fun = 
-      mkFun (concatMap (init . partsOfFun) (concatMap (lexitems . snd) lts))
+      mkFun (concat $ concatMap (map (\x -> if show x == "__" then [""] else init $ partsOfFun x)) (intersperse [mkCId "__"] (map (lexitems . snd) lts)))
              (returnType $ fst (signature firstlang firsttree))
 
     signature :: Language -> Tree -> (Signature,Int)
@@ -323,7 +323,7 @@ prBuiltRules br = unlines $ [
   unwords ["oper",fun,"=","mk"++show cat, word fun,";","--",show lang] | (lang,funcats) <- unknowns br, (fun,cat) <- funcats
   ]
  where
-  fullname br = intercalate "_" (ids br) ++ "_" ++ show (funname br)
+  fullname br = intercalate "_" (ids br) ++ "__" ++ show (funname br)
   word f = "\"" ++ takeWhile (/='_') f ++ "\""
   cats = nub (map (snd . snd) lins)
   mark c s = if all (==c) cats then s else "--- " ++ s -- comment out rule if the signature is not the same in both languages (for the moment at least)

### 2021-05-07 paper/paper.tex
summarized pos-equiv

--- 2021-05-06 paper/paper.tex
@@
 \subsubsection{Alignment criteria} \label{criteria}
-% label matching
-% pos equivalence
-% known divergence -> cite Dorr, few examples
-% known alignment -> SMT integration
+While alignment criteria are customizable, to better understand the extraction algorithm described in the above we dedicate this section to describing the ones that are currently used by default.
+
+\paragraph{Matching UD labels}
+The most obvious, but also most effective idea is to determine alignability based on comparing the dependency labels of the candidate UD tree pair. 
+In particular, according to this idea, two subtrees, in \textit{matching context}, i.e. attached to aligned heads constitute an alignment if their roots share the same dependency label. 
+Intuitively, this means that they are in the same relation with their heads.
+
+Note that, since the root of US trees is always attached to a a fake node with an arc labelled \texttt{root}, this criterion also makes it so that full sentences are always considered to align with each other, something which is desirable since we assume the parallel texts that are fed to our program to be sentence-aligned.
+
+\paragraph{Part-Of-Speech equivalence}
+CoNNL-U files provide information not only on the syntactic role of the tokens a sentence is composed of, but also on their grammatical categories, represented as Universal Part-Of-Speech (POS) tags. 
+Intuitively, if the nodes of two trees in matching contexts have the same POS tags, the two trees are more likely to correspond to each other than if not.
+
+This is especially true if focus, for instance, solely on the open class words (in the current implementation, defined as in the UD documentation \cite{TODO:}), thus ignoring prepositions, determiners, auxiliary verbs and all other function words, which tend to behave differently across different languages.
+
+As a consequence, a useful relation to define between dependency trees is that of \textit{POS-equivalence}: two dependency trees $t$, $u$ are POS-equivalent if $M_1 = M_2 \neq \emptyset$, where $M_i$ is defined as the multiset of POS tags of all the open class word nodes of $t_i$.
+
+Applied alone, this criterion can be used to capture correspondences that would be missed by only matching UD labels, thus increasing recall, but a decrease in precision is also to be expected. 
+However, since alignment criteria are defined as boolean functions, it is easy to combine them so to apply them simultaneously: combining these first two criteria, for instance, is useful in cases where precision is more important than recall.
+
+\paragraph{Known translation divergence}
+% Dorr
+
+\paragraph{Known alignment}
+% example application: SMT integration
 

### 2021-05-07 paper/paper.tex
wrote section on divs (no examles yet)

@@
 \paragraph{Known translation divergence}
-% Dorr
+Parallel texts often present significant, systematic cross-linguistic grammatical distinctions. 
+When this is the case, it is often straightforward to define alignment criteria based on recognizing the corresponding patterns.
+If many distinctions of this kind are specific to particular language pairs or even stylistical, some of them occur in independently of what languages are involved and have nothing to do with idiomatic usage or aspectual, discourse, domain or word knowledge.
+Drawing inspiration from \cite{TODO:} and \cite{TODO:}, we refer to them distinctions as \textit{translation divergences} and handle some with the alignment criteria that used by default.
+
+% TODO: add examle (Herb?)
 

### 2021-05-07 paper/paper.tex
paragraph on known alignment criterion

@@
 \paragraph{Known alignment}
-% example application: SMT integration
+Another case in which it is trivially desirable for two subtrees in matching context to be aligned is when an equivalent alignment is already known, for instance due to a previous iteration of the extraction algorithm.
+When referred to pairs of alignments, the term \textit{equivalent} indicates that the two alignments, linearized, correspond to the same string.
 
-% how tracking them is useful (but maybe in evaluation?)
+At a first glance, this might seem a criterion with no practical applications. 
+However, it is particularly useful when, instead of starting with an empty set of correspondences, we initialize the program with some alignments that are either inserted manually or, most interestingly, obtained with some other alignment technique.
+For instance, in this way it is possible to combine the system proposed in this paper in combination with a statistical tool and give more credit to the correspondences identified by both systems.
 

### 2021-05-07 paper/paper.tex
draft of a draft of the new section on UD patterns

@@
 \subsubsection{Pattern matching}
-% look for specific UD patterns
+So far, we described how CE can be used in a setting where the objective is to generate a comprehensive translation lexicon based on set of example sentence pairs.
+We pointed out that the program can be configured so to prioritize precision or coverage, but never restricted our search to a particular type of alignments.
+Nonetheless, there can be cases in which only certain syntactic structures are of interest: for instance, for whatever reason we might be looking for adverbs or noun phrases exclusively.
+
+To handle these situations, the CE module can filter the results based on a \texttt{gf-ud} pattern.
+Pattern matching in \texttt{gf-ud} %TODO:
+
+Combining pattern matching with pattern replacement, for instance by pruning the UD trees extracted by the alignment module, allows us even to extract correspondences that, given the structure of UD trees, cannot be identified by CE alone.
+To give an example that turned out to be useful in practice, a possibility is to extract verb phrases only, by looking for full clauses and pruning the subtree corresponding to the subject. 
+More advanced replacements are also possible: instead of simply pruning clauses to extract the verb phrases they contain, we can perform some actual replacement to obtain \textit{predication patterns}, such as
+
+$$\langle X tells Y Z, X dice Z a Y\rangle$$.
+
+% TODO: reformulate and add possiby graphical example
 

### 2021-05-09 paper/paper.tex
structured second half of the paper

@@
 
-% TODO: reformulate and add possiby graphical example
+% TODO: reformulate and add possibly graphical example
 
 \subsection{Generating grammar rules}
+The CE module described so far outputs pairs of UD trees. 
+While converting them to GF Abstract Syntax Trees (ASTs) is the \texttt{gf-ud}'s core functionality, to build a compilable GF translation lexicon it is necessary to turn such trees into grammar rules.
+
+In order to do that, our grammar generation module requires a morphological dictionary of the languages at hand and an \textit{extraction grammar}. 
+What we refer to as an extraction grammar is of course a GF grammar, whose
+objective is to define the set of basic categories and syntactic rules the entries of the automatically generated lexicon is going to use.
 
 \section{Evaluation} \label{evaluation}
-% we use udpipe
+\subsection{Data}
+% PUD treebanks
+% plans (+UDPipe)
+% langs
+
+\subsection{Evaluating CE} % TODO: find better title?
+\subsubsection{Experimental setup} % TODO: find better title?
+% metrics & methods
+\subsubsection{Results}
+% PUD (incl. reason-wise?): independent, comparison with SMT
+% RAW
+\subsubsection{Discussion}
+
+\subsection{MT experiments}
+\subsubsection{Experimental setup}
+\subsubsection{Results}
+\subsubsection{Discussion}
 
 \section{Conclusions} \label{conclusions}
-% tangible output and were it is
+% tangible output and were it is on github
+% summary of process and discussions
+\subsection{Future work}
 

### 2021-05-09 paper/paper.tex
data section of eval

@@
 \subsection{Data}
-% PUD treebanks
-% plans (+UDPipe)
-% langs
+Because we want part of our evaluation to be independent from the quality of UD parsing, some of the experiments are carried out using manually annotated data. 
+To this purpose, we use a 100-sentences subset of the Parallel UD (PUD) corpus, a set of handcrafted\footnote{Annotation is sometimes semi-automatic, but all treebanks are at least manually checked.} multilingual treebanks in CoNLL-U format created for the CoNLL 2017 shared task on Multilingual Parsing from Raw Text to Universal Dependencies \cite{TODO:}.
+PUD treebanks are available in over 20 languages, of which we selected English, Italian and Swedish. 
+Using this data limits the amount of alignment errors that are due annotation issues to a minimum, even though a small number of inconsistencies and imprecisions is present even in this corpus.
+
+When it comes to testing the program on raw text, we use two bilingual sentence-aligned corpora consisting of course plans from the Department of Mathematics and Computer Science of the University of Perugia (for English-Italian) and from the Department of Computer Science and Engineering (CSE) shared between the University of Gothenburg and the Chalmers University of Technology (for English-Swedish).
+For brevity, we will refer to these two datasets as to the DMI and the CSE corpora.
+This data, available in the project repository, was collected and sentence-aligned specifically for this work and a related Bachelor's thesis project \cite{TODO:}.
+When using raw text, our parser of choice is UDPipe \cite{TODO:}. 
+In particular, we use the ParTUT English and Italian models for the DMI corpus and models trained on the bilingual LinES English-Swedish treebank for the CSE corpus \cite{TODO:}.
 
 \subsection{Evaluating CE} % TODO: find better title?
-\subsubsection{Experimental setup} % TODO: find better title?
 % metrics & methods
@@
 \subsection{MT experiments}
-\subsubsection{Experimental setup}
+% metrics & methods
 \subsubsection{Results}

### 2021-05-09 paper/paper.tex
general part of ce eval

@@
 
-\subsection{Evaluating CE} % TODO: find better title?
-% metrics & methods
-\subsubsection{Results}
+\subsection{Evaluating CE} %TODO: better title?
+While in this work we focus mostly on the MT applications of CA, automatic translation, and much less GF-based domain-specific tranlation, is not the only context in which our system can be put to use.
+For instance, it is easy to imagine using it to improve the UI of online translation memories or to facilitate reading parallel texts in the context of language learning.
+For this reason, a first set of experiments is aimed at evaluating the alignments obtained with our CA module independently from the subsequent stages of our proposed lexicon generation pipeline.
+
+We assess the correctness of the alignments the CE module is able to extract first from manually annotated data, then from the DMI and CSE corpus, comparing our results with those obtained with a standard statistical tool, \texttt{fast\_align} \cite{TODO:}.
+Doing so requires manually labelling the correspondences as either correct or incorrect.
+
+For measuring the performance of our system, while precision and recall are two well-known metrics that would be very well suited to the task, the lack of a gold standard for CE imposes us to approximate them with respectively:
+
+\begin{itemize}
+  \item the number of correct alignments the program is able to extract
+  \item the ratio between such number and the total number of alignments extracted.
+ \end{itemize}
+
+However, since some alignments are only correct in the specific context of the sentence pair in which they occur, we make a further distinction between correct alignments that are relevant for a translation lexicon, marked as \texttt{+} in Section \ref{ceres}'s tables, and alignments that are useful for comparing the sentences but should not be used for MT, marked as \texttt{=} instead.
+% TODO: boat/train example
+Incorrect correspondences are marked as \texttt{-}.
+
+\subsubsection{Results} \label{ceres}
 % PUD (incl. reason-wise?): independent, comparison with SMT

### 2021-05-09 paper/paper.tex
finished CE eval excepts parts depending on experiments to run

@@
 
-% TODO: add figure comapring tree representations
+% TODO: add figure comparing tree representations
 
@@
 \section{Evaluation} \label{evaluation}
+% TODO: move around things to have something here (or outline structure of the section)
 \subsection{Data}
@@
 In particular, we use the ParTUT English and Italian models for the DMI corpus and models trained on the bilingual LinES English-Swedish treebank for the CSE corpus \cite{TODO:}.
+
+% TODO: size
 
@@
 
-\subsubsection{Results} \label{ceres}
-% PUD (incl. reason-wise?): independent, comparison with SMT
-% RAW
-\subsubsection{Discussion}
+\subsubsection{Results on manually annotated treebanks} \label{ceres}
+In table \ref{pud_fast}, we compare the results obtained with our grammar-based module to those obtained statistically on the PUD treebanks.
+
+Of course, \texttt{fast\_align} does not make use of the information present in the CoNLL-U files excepts with regards to tokenization. 
+On the other hand, the relatively large size of the PUD treebanks makes it possible to also train the statistical tool on the full dataset instead of just using the chosen 100-sentences subset, allowing for a fairer comparison.
+
+In both cases, \texttt{fast\_align} is run with 10 iterations in EM training, symmetrizing the alignments found by running the program in both directions.
+
+%TODO: position properly
+\begin{table*}[h]
+  \centering
+  \small
+  \begin{tabular}{l|lll}
+  %\hline
+   & \textbf{CE} & \textbf{\texttt{fast\_align} 100} & \textbf{\texttt{fast\_align} 1000}\\ \hline
+   \textbf{tot. alignments} & 716 & 1440 & 1435 \\ 
+   \textbf{\texttt{+} and \texttt{=}} & 536 (75\%) & 410 (28\%) & 656 (46\%)\\ 
+   \textbf{\texttt{+}} & 491 (69\%) & 371 (26\%) & 590 (41\%)\\ 
+  \end{tabular}
+  \caption[Comparison between our grammar-based CE module and \texttt{fast\_align}]{Comparison between our grammar-based CE module and \texttt{fast\_align} on PUD data.
+The middle column shows the results obtained by training \texttt{fast\_align} on the 100-sentence subset fed to the extraction module.
+On the right, the results obtained by training \texttt{fast\_align} on the full PUD treebanks dataset, discarding the alignments obtained for sentences 101-1000 for the final comparison.
+Row 1 reports the total number of distinct alignments obtain.
+Rows 2 and 3 give the percentages of correct alignments, with and without considering sentence pair-specific correspondences as correct.}
+  \label{pud_fast}
+\end{table*}
+
+While \texttt{fast\_align} is designed to align every word in the text (or explicitly state that a word has no counterpart in its translation) and, consequently, extracts around twice as many correspondences as our CE module, the percentage of correct correspondences is definitely in favor of our system, even though \texttt{fast\_align} gets significantly more precise when trained on the full, 1000-sentence dataset.
+
+\subsubsection{Results on raw text}
+
+The course plan corpora are significantly harder to work with, both for our CE module and \texttt{fast\_align}.
+In the former case, the additional challenge is relying on the quality of automatic parsing, while \texttt{fast\_align} suffers from the small size of the datasets, which are fed in their entirety to both systems.
+
+The results of this comparison are summarized in table \ref{raw_fast}.
+
+%TODO: position properly
+%TODO: run experiments & update data
+\begin{table*}[h]
+  \centering
+  \small
+  \begin{tabular}{l|lll}
+
+  \end{tabular}
+  \caption[Comparison between our grammar-based CE module and \texttt{fast\_align}]{Comparison between our grammar-based CE module and \texttt{fast\_align} on raw data, using the DMI and CSE corpora.
+Row 1 reports the total number of distinct alignments obtain.
+Rows 2 and 3 give the percentages of correct alignments, with and without considering sentence pair-specific correspondences as correct.}
+  \label{raw_fast}
+ \end{table*}
+
+ % TODO: discussion
+
+% PUD: incl. reason-wise?)
 
@@
 % metrics & methods
-\subsubsection{Results}
-\subsubsection{Discussion}
+\subsubsection{Experimental results}
 

### 2021-05-09 paper/paper.tex
description of MT experiments

@@
 
-%TODO: position properly
+% TODO: position properly
+% TODO: add english-swedish!
 \begin{table*}[h]
@@
 
-% PUD: incl. reason-wise?)
+% TODO: PUD: incl. reason-wise?)
 
 \subsection{MT experiments}
-% metrics & methods
+The second set of experiments has the objective of assessing the quality of the final output of the system we propose: GF translation lexica. 
+Because we are focussing on using CA in the context of domain-specific MT, in this case we do not make use of the PUD treebanks, where sentences come from a variety of different sources, but rather of the two course plans corpora.
+For a this evaluation, we do not construct a grammar specific to the course plans domain: it is sufficient to extend the extraction grammar itself, designed for the grammar generation step to produce grammar rules corresponding to simple concepts such as basic categories, noun phrases, verb phrases etc., with preexisting syntax rules defined in the RGL, so to allow for more variation.
+% TODO: is RGL mentioned above?
+
+The basic idea behind the experiments whose results are presented in the following section is to automatically translate a set of English sentences to Italian and Swedish, ask native speakers of the target languages to produce a set of reference translations, and compare them to the original machine-generated ones by computing the BLEU score.
+
+Due to the small size of the datasets and, as a consequence, the low coverage of the extracted lexicon, we generate the sentences to translate directly in the GF shell, making use of its random AST generation functionality but at the same time filtering out semantically implausible sentences to facilitate the task of the human translators.
+The results of this process are two small testing corpora, each consisting of 50 sentences in English, generated using the abstract syntax and the English concrete syntax of the DMI and the CSE grammar respectively.
+
+The reference translations were obtained by asking two native speakers of Italian and Swedish to compare the original English sentences to their automatically translated counterparts and correct the latter.
+In order to obtain more meaningful BLEU scores, participants were instructed to only make the minimal changes necessary to obtain, starting from the automatic tranlsations generated by our system, a set of grammatically and semantically correct translations.
+
 \subsubsection{Experimental results}

### 2021-05-09 paper/paper.tex
MT eval results

--- 2021-05-09 paper/paper.tex
@@
 
+% TODO: designed for the grammar generation step to produce grammar rules corresponding to simple concepts such as basic categories, noun phrases, verb phrases etc.,
+
 \section{Evaluation} \label{evaluation}
@@
 
-However, since some alignments are only correct in the specific context of the sentence pair in which they occur, we make a further distinction between correct alignments that are relevant for a translation lexicon, marked as \texttt{+} in Section \ref{ceres}'s tables, and alignments that are useful for comparing the sentences but should not be used for MT, marked as \texttt{=} instead.
+However, since some alignments are only correct in the specific context of the sentence pair in which they occur, we make a further distinction between correct alignments that are relevant for a translation lexicon, marked as \texttt{+} in the following tables, and alignments that are useful for comparing the sentences but should not be used for MT, marked as \texttt{=} instead.
 % TODO: boat/train example
@@
 
-\subsubsection{Results on manually annotated treebanks} \label{ceres}
-In table \ref{pud_fast}, we compare the results obtained with our grammar-based module to those obtained statistically on the PUD treebanks.
+\subsubsection{Results on manually annotated treebanks} 
+In Table \ref{pud_fast}, we compare the results obtained with our grammar-based module to those obtained statistically on the PUD treebanks.
 
@@
 
-\subsubsection{Results on raw text}
-
+\subsubsection{Results on raw text} \label{raw}
 The course plan corpora are significantly harder to work with, both for our CE module and \texttt{fast\_align}.
@@
 
-The results of this comparison are summarized in table \ref{raw_fast}.
+The results of this comparison are summarized in Table \ref{raw_fast}.
 
@@
 
-% PUD: incl. reason-wise?)
+% TODO: PUD: incl. reason-wise?)
 
 \subsection{MT experiments}
-% metrics & methods
+The second set of experiments has the objective of assessing the quality of the final output of the system we propose: GF translation lexica. 
+Because we are focussing on using CA in the context of domain-specific MT, in this case we do not make use of the PUD treebanks, where sentences come from a variety of different sources, but rather of the two course plans corpora.
+For this evaluation, we do not construct a grammar specific to the course plans domain: it is sufficient to extend the extraction grammar itself with preexisting syntax rules defined in the RGL, so to allow for more variation.
+% TODO: is RGL mentioned above?
+
+The idea is to automatically translate a set of English sentences to Italian and Swedish, ask native speakers of the target languages to produce a set of reference translations, and compare them to the original machine-generated ones by computing BLEU scores.
+
+Due to the small size of the datasets and, as a consequence, the low coverage of the extracted lexicon, we generate the sentences to translate directly in the GF shell, making use of its random AST generation functionality but at the same time filtering out semantically implausible sentences to facilitate the task of the human translators.
+The results of this process are two small testing corpora, each consisting of 50 sentences in English, generated using the abstract syntax and the English concrete syntax of the DMI and the CSE grammar respectively.
+
+Reference translations are obtained by asking two native speakers of Italian and Swedish to compare the original English sentences to their automatically translated counterparts and correct the latter.
+Participants were instructed to only make the minimal changes necessary to obtain, starting from the automatic tranlsations generated by our system, a set of grammatically and semantically correct translations.
+This is important to get meaningful results, as if the reference translations are obtained independently from the automatic ones, BLEU scores can easily become misleading.
+
 \subsubsection{Experimental results}
+Corpus-level BLEU scores for the automatic translations of the 50+50 sentences of the testing corpora generated as described in the above are summarized in Table \ref{tableu}.
+
+Following conventions, we report the cumulative $n$-gram scores for values of $n$ from 1 to 4 (BLEU-1 to BLEU-4). 
+However, being a significant portion of the sentences of length 4 or less, we also report BLEU-1 to BLEU-3 scores, BLEU-1 to BLEU-2 scores and scores obtained considering unigrams only.
+
+\begin{table}[h]
+  \centering
+  \begin{tabular}{l|ll}
+  \textbf{}            & \textbf{DMI (en-it)} & \textbf{CSE (en-sv)} \\ \hline
+  \textbf{BLEU-1 to 4} & 55         & 61         \\ 
+  \textbf{BLEU-1 to 3} & 63         & 68         \\ 
+  \textbf{BLEU-1 to 2} & 70         & 74          \\
+  \textbf{BLEU-1}      & 79         & 81         \\ 
+  \end{tabular}
+  \caption[BLEU scores for automatic translations based on the course plans grammars]{BLEU scores obtained by comparing one candidate automatic translation per sentence to a reference translation obtained by making the necessary corrections to the automatically generated ones.
+Both the initial English sentences and their automatic translations were generated using the two course plans grammars.}
+  \label{tableu}
+\end{table}
+
+These synthetic figures are useful to give an idea of the general quality of the translations: overall, although with relatively low scores, English-to-Swedish translation works significantly better than English-to-Italian.
+Looking back at the results reported in Section \ref{raw}, the reason for this is not immediately clear, as the difference in precision between the two language pairs is negligible on the course plans corpora.
+
+Looking at sentence-level scores can, however, be more insightful. 
+For both corpora, scores assigned to individual segments range from the minimum possible value of 0 to the perfect score of 100, which indicates a perfect correspondence between the automatic and the reference translation.
+
+Examples of sentences that were assigned a perfect BLEU-1 to 4 score are ``\textit{the library provides useful textbooks}'' (translated to Italian as ``\textit{la biblioteca fornisce libri utili}'') in the DMI corpus and ``\textit{this lab is more difficult than the exam}'' (whose Swedish translation is ``\textit{den här laborationen är svårare än tentamen}'') in the CSE corpus.
+\smallskip
+
+On the other hand, it is easy for shorter sentences to be assigned the minimum BLEU-1 to 4 score even when they only contain a single grammatical or semantic error. 
+This is the case, for instance, of the sentence ``\textit{the test is oral}'', whose last word, ``\textit{oral}'' is translated as ``\textit{dura}'' (``\textit{hard}'') instead of ``\textit{orale}'' due to an alignment error.
+Nonetheless, it is worth noticing that this is one of the many cases in which the correct alignment $\langle$``\textit{oral}'',``\textit{orale}''$\rangle$ is also identified by the CE module, but hidden by the fact that, for simplicity, we only take into account the first translation candidate.
+This suggest that, in contexts where higher precision is necessary, there are simple improvements that can be done either in the choice of the alignment criteria or at a later stage, in terms alignment selection.
+In other contexts, it can be more practical to aim for higher coverage and not perform any automatic selection, leaving lexicon refinements to a human, possibly with the help of the postprocessing tools that are currently under development \cite{TODO:}.
+
+% TODO: decide if to leave or remove
+Furthermore, a problem with using the BLEU score as the only evaluation metric is the fact that it makes no distinction between content and function words, thus not allowing an evaluation focused specifically on the extracted concepts.
+The small size of the corpus, however, allows for some error analysis. 
+While postprocessing the automatic translations, the participants were asked to indicate what kind of errors they encountered in each sentence (grammatical, semantical or both). 
+Their observations are summarized in Table \ref{obs}.
+
+\begin{table}[h]
+  \centering
+  \begin{tabular}{l|ll}
+  \textbf{}                                                                   \textbf{type of errors}                 & \textbf{DMI (en-it)} & \textbf{CSE (en-sv)} \\ \hline
+  \begin{tabular}[c]{@{}l@{}}\textbf{semantical}\end{tabular}               & 23 (46\%)    & 23 (46\%)    \\ 
+  \begin{tabular}[c]{@{}l@{}}\textbf{grammatical}\end{tabular}             & 10 (20\%)      & 3 (6\%)      \\ 
+  \begin{tabular}[c]{@{}l@{}}\textbf{semantical and} \\ \textbf{grammatical}\end{tabular} & 3 (6\%)      & 4 (8\%)      \\ 
+  \end{tabular}
+  \caption[Types of errors encountered in the automatically translated sentences]{Number of automatically translated sentences containing only semantical, only grammatical and both semantical and grammatical errors in the two synthetic course plans corpora.}
+  \label{obs}
+\end{table}
+
+Interestingly, while most errors are due to wrong alignments, the main difference between two corpora lies in the number of translations that only contain grammatical errors, which explains the significant difference observed in the cumulative BLEU scores shown in Table \ref{tableu}.
+
+In Italian, grammar errors often involve contractions such as ``\textit{del}'' (``\textit{di}'' + ``\textit{il}'', in English ``\textit{of the}''), some of which are systematically rendered as two separate words due to UDPipe tokenization.
+Another common case is that of wrong (or at least very confusing) adjective collocation, such as in the translation ``\textit{il libro presenta una tecnica con miglioramenti vari utile}'' (``\textit{the textbook presents a useful technique with various improvements}''), where the adjective ``\textit{utile}'' (``\textit{useful}''), referred to ``\textit{technique}'' (``\textit{tecnica}'') is placed far from such noun, making the sentence hard to interpret\footnote{The manually postprocessed translation is ``\textit{il libro presenta una utile tecnica con miglioramenti vari}''.}.
+Grammatical errors in Swedish are less common and, apart from long adjectives never being turned to the comparative degree periphrastically (e.g.
+``\textit{relevant\underline{are}}'' instead of ``\textit{mer relevant}''), less systematic.
+Only in one case, for instance, gender is incorrect (``\textit{programbibliotek\underline{en}}''). 
+All of these grammatical errors are easy to handle when writing a domain-specific grammar or, for instance in the case of misgendered nouns, by making small adjustments to the morphological dictionaries.
+
+Some errors regarding the extracted concepts are also interesting to analyze.
+In English-Swedish, while several compounds, such as $\langle$\textit{``computer science'', ``datavetenskap''}$\rangle$, are aligned correctly, there are cases in which a single-root noun is translated as a compound (e.g.
+$\langle$ \textit{``theory'', ``automatateori''} $\rangle$).
+This is not necessarily an actual alignment error, as it might well be the case that the English version of the text was being less specific than its Swedish counterpart, thus producing an alignment that, during evaluation, would have been marked as ``correct but not relevant for a translation lexicon'' (\texttt{=}).
+``Reasonable'' alignment errors appear in the DMI corpus sentences too. 
+The English phrase ``\textit{attendance of lessons}'', for instance, becomes simply ``\textit{frequenza}'' in the Italian translation. 
+Such word is in fact often used alone to replace longer expressions such as ``\textit{frequenza delle lezioni}'', just like in English ``\textit{of lessons}'' can be omitted when what must be attended is evident from the context.
+Another interesting example is the alignment $\langle$\textit{``class'', ``classe''}$\rangle$, which causes the sentence \textit{``I will attend the class''} to be (incorrectly) translated as \textit{``io seguirò la classe''} instead of \textit{``io seguirò la lezione''} even though the correspondence is in fact valid in most of the numerous contexts in which (within the same domain!) \textit{``class''} is not to be intended as a synonym of \textit{``lesson''}.
 

### 2021-05-09 paper/paper.tex
first ever "complete" draft

@@
 
-\paragraph{Known alignment}
+\paragraph{Known alignment} \label{ka}
 Another case in which it is trivially desirable for two subtrees in matching context to be aligned is when an equivalent alignment is already known, for instance due to a previous iteration of the extraction algorithm.
@@
 \section{Conclusions} \label{conclusions}
-% tangible output and were it is on github
-% summary of process and discussions
-\subsection{Future work}
+We presented a syntax-based alignment method with a focus on its usage for domain-specific translation lexica generation.
+
+Compared to the existing statistical tools, our system has the following advantages:
+\begin{itemize}
+  \item it performs consistently well even on extremely small parallel corpora
+  \item it is able to simultaneously extract correspondences between individual words, multiword expressions and full phrases, even when their constituents are discontinuous
+  \item in conjunction with \texttt{gf-ud} pattern matching, it can be used to extract specific types of correspondences, such as predication patterns  
+  \item it can automatically generate compilable, morphology-aware GF translation lexica
+  \item it can be configured to easily handle systmatic language-pair specific translation divergences.
+\end{itemize}
+
+The tangible fruit of this work are a Haskell library and a number of executables offering an easy to use and configure interface to perform CE, lexicon generation and a variety of kinds of evaluations.
+The source code, which also implements a preliminary implementation of CP not discussed in depth in this paper, is available at \url{github.com/harisont/concept-alignment}.
+
+\subsection{Current and future work}
+Our results, while encouraging, suggest that there is still much room for improvement in many different directions.
+
+An obvious possible development is to improve the current implementation of CP, optimizing it for its two different use cases: propagating alignments to a new language looking for correspondences in a different text in the same domain and using a translation of the same text they were obtained from.
+An alternative to the latter is to generalize CE to $n$ languages, instead of limiting it to bilingual cases.
+
+For cases where a relatively large amount of data is available, using our system in conjunction with a statistical tool seems promising. 
+As discussed in Section \ref{ka}, some integration between our system and other sources of alignment is already supported, and it is possible to keep track of the reason why a particular correspondence was extracted.
+This could be further developed in the direction of an actual hybrid system. 
+A first attempt in this direction is the development of a simple conversion tool between the ``pharaoh format'' used by tools such as \texttt{fast\_align} and the format we use to represent alignments, available in the same repository.
+
+Finally, since the freedom that generally characterizes human translation and the quality of currently available UD parsers make the expectation of maximizing both alignment precision and recall unrealistic, it seems logical to develop tools to make it easier to postprocess automatically generated lexica.
+With respects to this, current work is represented by a synoptic CoNLL-U, available at \url{github.com/DigitalGrammarsAB/synoptic-conllu-viewer}, which makes it possible to inspect the UD trees a grammar rule was derived from.
 

### 2021-05-09 paper/paper.tex
tentative abstract

@@
 \begin{abstract}
-% TODO: abstract
+In grammar-based domain-specific machine translation, constructing high-quality translation lexica is a crucial but time consuming task requiring significant linguistic knowledge. 
+In presence of a corpus of parallel example sentences, statistical word and phrase alignment tools allow to automate part of this process, but are not suitable for small datasets and often fail to identify discontinuous multiword correspondences.
+Addressing both problems, we propose a data-driven, syntax-based approach to the automation of this task and put it to the test in a simple grammar-based translation pipeline.
 \end{abstract}

### 2021-05-09 paper/paper.tex
first quick review

@@
 \usepackage{graphicx}
+
+\usepackage{todonotes}
 
@@
 In many cases, this is done for the most part manually, resulting in a time consuming task requiring significant linguistic knowledge. 
-When the grammar is designed based on a parallel corpus of example sentences, it is possible to automate part of this process by means of statistical word and phrase alignment techniques \cite{TODO: a lot}.
-None of them is, however, suitable for the common case in which only a limited number of example sentences is available.
+When the grammar is designed based on a corpus of parallel example sentences, it is possible to automate part of this process by means of statistical word and phrase alignment techniques \cite{TODO: a lot}.
+None of them is, however, suitable for the common case in which only a limited number of example data is available.
 
@@
 Instead, we can simultaneously operate at different levels of abstraction, thus extracting both single- and multiword, even non-contiguous, correspondences. 
-
 For this reason, we refer to the task our system attempts to automate as \textit{Concept Alignment} (CA).
@@
 
-The basic use case for CA, which we refer to specifically as \textit{Concept Extraction} (CE), is the generation of a translation lexicon from a parallel text.
+The basic use case for CA, which we refer to specifically as \textit{Concept Extraction} (CE), is the generation of a translation lexicon from a multilingual parallel text.
 This can be directly compared to the numerous existing word and phrase alignment techniques.
@@
 As stated in the Introduction, most existing extraction solutions are based on statistical approaches and are, as a consequence, unsuitable for small datasets. 
-Grammar-based approaches, making use of parallel treebanks and collectively referred to as \textit{tree-to-tree alignment methods}, have also been proposed \cite{TODO:}, but have historically suffered from the inconsistencies between the various constituency grammar formalisms used to define grammars for the different languages and from the insufficient degree of robustness of existing parses.
+Grammar-based approaches, making use of parallel treebanks and collectively referred to as \textit{tree-to-tree alignment methods}, have also been proposed \cite{TODO:}, but have historically suffered from the inconsistencies between the various formalisms used to define the grammars for different languages and from the insufficient degree of robustness of existing parses.
 
@@
 
-While both formalisms, the former a dependency and the latter a constituency grammar, independently solve the former issue, UD is especially appealing since dependency trees are an easier target for parsing.
+While both formalisms, the former a dependency and the latter a constituency grammar\todo{too quick, say more about both}, independently solve the former issue, UD is especially appealing since dependency trees are an easier target for parsing.
 As a consequence, several robust dependency parsers, such as \cite{TODO: UDPipe} and \cite{TODO: Standford} are available.
 
-UD parsing alone would then be sufficient to extract (or propagate) tree-to-tree alignments, but not to automate the generation of a ready-to-use, morphologically-aware translation lexicon.
-This is where GF comes into play: after correspondences are inferred from a parallel text, our proposed system is able to convert them to GF grammar rules, easy to embed in a domain-specific grammar but also making it possible to immediately evaluate the system by carrying out small-scale translation experiment using pre-existing grammatical constructions implemented in GF's Resource Grammar Library (RGL) \cite{TODO:}.
+UD parsing alone would then be sufficient to extract (or propagate) tree-to-tree alignments, but not to automate the generation of a morphologically-aware translation lexicon for a generative grammar.
+This is where GF comes into play: after correspondences are inferred from a parallel text, our proposed system is able to convert them to GF grammar rules, easy to embed in a domain-specific grammar but also making it possible to immediately evaluate the system by carrying out small-scale translation experiments using pre-existing grammatical constructions implemented in GF's Resource Grammar Library (RGL) \cite{TODO:}.
+\todo{say what the RGL is}
 This is made possible by \texttt{gf-ud}, a conversion tool described in \cite{TODO:} and \cite{TODO:}.
@@
   \item an alignment module based on dependency tree comparison
-  \item a program, based on \texttt{gf-ud}, that converts the alignments into GF grammar rules and uses them to construct a translation lexicon
+  \item a program, based on \texttt{gf-ud}, that converts the alignments into GF grammar rules and uses them to construct a translation lexicon.
 \end{itemize} \smallskip
@@
 
-The basic extraction algorithm, whose pseudocode is shown in Figure \ref{TODO}, takes as input a list of priority-sorted \textit{alignment criteria}, rules to determine whether two dependency trees should be aligned with each other which we will discuss in greater detail in Section \ref{criteria}, and a pair of UD trees corresponding to the sentences to align.
+The pseudocode of the basic extraction algorithm is shown in Figure \ref{TODO}\todo{add pseudocode}.
+The input is a list of priority-sorted \textit{alignment criteria}, i.e.rules to determine whether two dependency trees should be aligned with each other, which will be discussed in greater detail in Section \ref{criteria}, and UD tree representation of the pair of sentences to align.
 From an implementation point of view, UD trees are, as in \texttt{gf-ud}, rose trees where each node represents a token, obtained from the CoNLL-U files produced by the UD parser.
 
+\todo{add figure with the various tree representations?}
 % TODO: add figure comparing tree representations
 
-As a first step, the program checks whether the two full sentence trees can be aligned with each other, i.e.
-if they match any of the alignment criteria.
-If this is the case, they are added to a collection of alignments, also represented as a pair of UD trees associated with some metadata, which is what the function will return after aligning all the dependency subtrees.
-% TODO: explain or omit metadata!
+As a first step, the program checks whether the two full sentence trees can be aligned with each other, i.e. if they match any of the alignment criteria. 
+If this is the case, they are added to a collection of alignments, which are also represented as a pair of UD trees associated with some metadata. \todo{explain what metadata?}
+Such collection is what the function will return after aligning all the dependency subtrees. % TODO: explain or omit metadata!
 
-Depending on which alignment criteria they match, the head of the two trees may or may not also be added to such collection. This is important... % TODO:
+Depending on which alignment criteria they match, the head of the two trees may or may not also be added to such collection. This is important... \todo{explain why}% TODO:
 
-The same procedure is applied recursively to all pairs of immediate subtrees of each sentence, until the leaves are reached or alignment is no longer possible due to lack of matching criteria.
+The same procedure is applied recursively to all pairs of immediate subtrees of each sentence, until the leaves are reached or alignment is no longer possible due to a lack of matching criteria.
 % TODO: explain & motivate sorting -> maybe say more about UD before?
@@
 The most obvious, but also most effective idea is to determine alignability based on comparing the dependency labels of the candidate UD tree pair. 
-In particular, according to this idea, two subtrees, in \textit{matching context}, i.e. attached to aligned heads constitute an alignment if their roots share the same dependency label. 
+In particular, according to this idea, two subtrees in \textit{matching context}, i.e. attached to aligned heads, constitute an alignment if their roots share the same dependency label. 
 Intuitively, this means that they are in the same relation with their heads.
 
-Note that, since the root of US trees is always attached to a a fake node with an arc labelled \texttt{root}, this criterion also makes it so that full sentences are always considered to align with each other, something which is desirable since we assume the parallel texts that are fed to our program to be sentence-aligned.
+Note that, since the root of UD trees is always attached to a a fake node with an arc labelled \texttt{root}, this criterion also makes it so that full sentences are always considered to align with each other, something which is desirable since we assume the parallel texts that are fed to our program to be sentence-aligned.
 
 \paragraph{Part-Of-Speech equivalence}
-CoNNL-U files provide information not only on the syntactic role of the tokens a sentence is composed of, but also on their grammatical categories, represented as Universal Part-Of-Speech (POS) tags. 
+CoNNL-U files provide information not only on the syntactic role of the tokens a sentence is composed of, but also on their grammatical categories, represented as Universal Part-Of-Speech (POS) tags \cite{TODO:}.
 Intuitively, if the nodes of two trees in matching contexts have the same POS tags, the two trees are more likely to correspond to each other than if not.
 
-This is especially true if focus, for instance, solely on the open class words (in the current implementation, defined as in the UD documentation \cite{TODO:}), thus ignoring prepositions, determiners, auxiliary verbs and all other function words, which tend to behave differently across different languages.
+This is especially true if we focus, for instance, solely on the open class words (in the current implementation, defined as in the UD documentation \cite{TODO:}), thus ignoring prepositions, determiners, auxiliary verbs and all other function words, which tend to behave differently across different languages.
 
@@
 
-Applied alone, this criterion can be used to capture correspondences that would be missed by only matching UD labels, thus increasing recall, but a decrease in precision is also to be expected. 
+Applied as a backup for label matchin, this criterion can be used to capture correspondences that would otherwise be missed, thus increasing recall, but a decrease in precision is also to be expected.
 However, since alignment criteria are defined as boolean functions, it is easy to combine them so to apply them simultaneously: combining these first two criteria, for instance, is useful in cases where precision is more important than recall.
@@
 When this is the case, it is often straightforward to define alignment criteria based on recognizing the corresponding patterns.
-If many distinctions of this kind are specific to particular language pairs or even stylistical, some of them occur in independently of what languages are involved and have nothing to do with idiomatic usage or aspectual, discourse, domain or word knowledge.
-Drawing inspiration from \cite{TODO:} and \cite{TODO:}, we refer to them distinctions as \textit{translation divergences} and handle some with the alignment criteria that used by default.
+If many distinctions of this kind are specific to particular language pairs or even stylistical, some of them occur independently of what languages are involved and have nothing to do with idiomatic usage or aspectual, discourse, domain or word knowledge.
+Drawing inspiration from \cite{TODO:} and \cite{TODO:}, we refer to them as \textit{translation divergences} and handle the most common ones via alignment criteria that are used by default.
+\todo{add a few examples (<= one per type of divergence)}
 
@@
 We pointed out that the program can be configured so to prioritize precision or coverage, but never restricted our search to a particular type of alignments.
-Nonetheless, there can be cases in which only certain syntactic structures are of interest: for instance, for whatever reason we might be looking for adverbs or noun phrases exclusively.
+Nonetheless, there can be cases in which only certain syntactic structures are of interest: for instance, we might be looking for adverbs or noun phrases exclusively.
 
 To handle these situations, the CE module can filter the results based on a \texttt{gf-ud} pattern.
-Pattern matching in \texttt{gf-ud} %TODO:
+Pattern matching in \texttt{gf-ud}...\todo{explain briefly how it works. Maybe Aarne can do that?}
 
@@
 To give an example that turned out to be useful in practice, a possibility is to extract verb phrases only, by looking for full clauses and pruning the subtree corresponding to the subject. 
-More advanced replacements are also possible: instead of simply pruning clauses to extract the verb phrases they contain, we can perform some actual replacement to obtain \textit{predication patterns}, such as
+\todo{rephrase explanation and reformat example}More advanced replacements are also possible: instead of simply pruning clauses to extract the verb phrases they contain, we can perform some actual replacement to obtain \textit{predication patterns}, such as
 
@@
 The CE module described so far outputs pairs of UD trees. 
-While converting them to GF Abstract Syntax Trees (ASTs) is the \texttt{gf-ud}'s core functionality, to build a compilable GF translation lexicon it is necessary to turn such trees into grammar rules.
+While converting them to GF Abstract Syntax Trees (ASTs) is \texttt{gf-ud}'s core functionality, to build a compilable GF translation lexicon it is necessary to turn such trees into grammar rules.
 
 In order to do that, our grammar generation module requires a morphological dictionary of the languages at hand and an \textit{extraction grammar}. 
-What we refer to as an extraction grammar is of course a GF grammar, whose
-objective is to define the set of basic categories and syntactic rules the entries of the automatically generated lexicon is going to use.
-
+What we refer to as an extraction grammar is of course a GF grammar, whose objective is to define the set of basic categories and syntactic rules the entries of the automatically generated lexicon is going to use.
+ \todo{not sure how important this section is and what (if anything) to add to it}
 % TODO: designed for the grammar generation step to produce grammar rules corresponding to simple concepts such as basic categories, noun phrases, verb phrases etc.,
@@
 Because we want part of our evaluation to be independent from the quality of UD parsing, some of the experiments are carried out using manually annotated data. 
-To this purpose, we use a 100-sentences subset of the Parallel UD (PUD) corpus, a set of handcrafted\footnote{Annotation is sometimes semi-automatic, but all treebanks are at least manually checked.} multilingual treebanks in CoNLL-U format created for the CoNLL 2017 shared task on Multilingual Parsing from Raw Text to Universal Dependencies \cite{TODO:}.
+To this purpose, we use a 100-sentences subset of the Parallel UD (PUD) corpus, a set of 1000 handcrafted\footnote{Annotation is sometimes semi-automatic, but all treebanks are at least manually checked.}\todo{not even, I fear.
+Check this} multilingual treebanks in CoNLL-U format created for the CoNLL 2017 shared task on Multilingual Parsing from Raw Text to Universal Dependencies \cite{TODO:}.
 PUD treebanks are available in over 20 languages, of which we selected English, Italian and Swedish. 
@@
 When it comes to testing the program on raw text, we use two bilingual sentence-aligned corpora consisting of course plans from the Department of Mathematics and Computer Science of the University of Perugia (for English-Italian) and from the Department of Computer Science and Engineering (CSE) shared between the University of Gothenburg and the Chalmers University of Technology (for English-Swedish).
-For brevity, we will refer to these two datasets as to the DMI and the CSE corpora.
+For brevity, we will refer to these two datasets as to the DMI and the CSE corpora, or collectively as to the course plans datasets.
 This data, available in the project repository, was collected and sentence-aligned specifically for this work and a related Bachelor's thesis project \cite{TODO:}.
@@
 However, since some alignments are only correct in the specific context of the sentence pair in which they occur, we make a further distinction between correct alignments that are relevant for a translation lexicon, marked as \texttt{+} in the following tables, and alignments that are useful for comparing the sentences but should not be used for MT, marked as \texttt{=} instead.
-% TODO: boat/train example
+\todo{add example (e.g.
+boat/train)} % TODO: boat/train example
 Incorrect correspondences are marked as \texttt{-}.
@@
 \subsubsection{Results on manually annotated treebanks} 
-In Table \ref{pud_fast}, we compare the results obtained with our grammar-based module to those obtained statistically on the PUD treebanks.
+In Table \ref{pud_fast}\todo{position properly; run and add en-sv}, we compare the results obtained with our grammar-based module to those obtained statistically on the PUD treebanks.
 
@@
 On the other hand, the relatively large size of the PUD treebanks makes it possible to also train the statistical tool on the full dataset instead of just using the chosen 100-sentences subset, allowing for a fairer comparison.
-
 In both cases, \texttt{fast\_align} is run with 10 iterations in EM training, symmetrizing the alignments found by running the program in both directions.
@@
 
-The results of this comparison are summarized in Table \ref{raw_fast}.
+The results of this comparison are summarized in Table \ref{raw_fast}\todo{position properly; run experiments (this is new) and add actual results}.
 
@@
  % TODO: discussion
+\todo{add results discussion}
+
+\todo{in case there's room maybe add criterion-specific stats}
 
@@
 For this evaluation, we do not construct a grammar specific to the course plans domain: it is sufficient to extend the extraction grammar itself with preexisting syntax rules defined in the RGL, so to allow for more variation.
-% TODO: is RGL mentioned above?
 
@@
 
-% TODO: decide if to leave or remove
-Furthermore, a problem with using the BLEU score as the only evaluation metric is the fact that it makes no distinction between content and function words, thus not allowing an evaluation focused specifically on the extracted concepts.
+Furthermore, \todo{from here on nothing is absolutely vital in this section.
+It can be shortened or removed if necessary to make room for something else}a problem with using the BLEU score as the only evaluation metric is the fact that it makes no distinction between content and function words, thus not allowing an evaluation focused specifically on the extracted concepts.
 The small size of the corpus, however, allows for some error analysis. 
@@
 \section{Conclusions} \label{conclusions}
-We presented a syntax-based alignment method with a focus on its usage for domain-specific translation lexica generation.
+We presented a syntax-based alignment method with a focus on its applications in domain-specific translation lexica generation.
 
@@
 The tangible fruit of this work are a Haskell library and a number of executables offering an easy to use and configure interface to perform CE, lexicon generation and a variety of kinds of evaluations.
-The source code, which also implements a preliminary implementation of CP not discussed in depth in this paper, is available at \url{github.com/harisont/concept-alignment}.
+The source code, which also implements a preliminary implementation of CP not discussed in depth in this paper, is available at \url{github.com/harisont/concept-alignment}\todo{maybe I should mention this in the introduction?}.
 
@@
 An obvious possible development is to improve the current implementation of CP, optimizing it for its two different use cases: propagating alignments to a new language looking for correspondences in a different text in the same domain and using a translation of the same text they were obtained from.
+\todo{simplify part on CP scenarios}
 An alternative to the latter is to generalize CE to $n$ languages, instead of limiting it to bilingual cases.
@@
 For cases where a relatively large amount of data is available, using our system in conjunction with a statistical tool seems promising. 
-As discussed in Section \ref{ka}, some integration between our system and other sources of alignment is already supported, and it is possible to keep track of the reason why a particular correspondence was extracted.
+As discussed in Section \ref{ka}, some integration between our system and other sources of alignments is already supported, and it is possible to keep track of the reason why a particular correspondence was extracted.
 This could be further developed in the direction of an actual hybrid system. 
@@
 Finally, since the freedom that generally characterizes human translation and the quality of currently available UD parsers make the expectation of maximizing both alignment precision and recall unrealistic, it seems logical to develop tools to make it easier to postprocess automatically generated lexica.
-With respects to this, current work is represented by a synoptic CoNLL-U, available at \url{github.com/DigitalGrammarsAB/synoptic-conllu-viewer}, which makes it possible to inspect the UD trees a grammar rule was derived from.
+With respects to this, current work is represented by a synoptic CoNLL-U viewer, available at \url{github.com/DigitalGrammarsAB/synoptic-conllu-viewer}, which makes it possible to inspect the UD trees a grammar rule was derived from.
+\todo{remove this part if the tool is not made public, or maybe regardless}
 

### 2021-05-10 paper/paper.tex
managed to show the bibliography even though I'm not sure that style is ok

@@
 \section{Introduction}
-Grammar-based translation pipelines such as those based on Grammatical Framework (GF) \cite{TODO:} have been successfully employed in domain-specific Machine Translation (MT). 
+Grammar-based translation pipelines such as those based on Grammatical Framework (GF) \cite{dummy} have been successfully employed in domain-specific Machine Translation (MT).  
 What makes these systems well suited to the task is the fact that, when we constrain ourselves to a specific domain, where precision is most often more important than coverage, they can provide strong guarantees in terms of grammatical correctness.
@@
 In many cases, this is done for the most part manually, resulting in a time consuming task requiring significant linguistic knowledge. 
-When the grammar is designed based on a corpus of parallel example sentences, it is possible to automate part of this process by means of statistical word and phrase alignment techniques \cite{TODO: a lot}.
+When the grammar is designed based on a corpus of parallel example sentences, it is possible to automate part of this process by means of statistical word and phrase alignment techniques \cite{dummy}.
 None of them is, however, suitable for the common case in which only a limited number of example data is available.
@@
 As stated in the Introduction, most existing extraction solutions are based on statistical approaches and are, as a consequence, unsuitable for small datasets. 
-Grammar-based approaches, making use of parallel treebanks and collectively referred to as \textit{tree-to-tree alignment methods}, have also been proposed \cite{TODO:}, but have historically suffered from the inconsistencies between the various formalisms used to define the grammars for different languages and from the insufficient degree of robustness of existing parses.
+Grammar-based approaches, making use of parallel treebanks and collectively referred to as \textit{tree-to-tree alignment methods}, have also been proposed \cite{dummy}, but have historically suffered from the inconsistencies between the various formalisms used to define the grammars for different languages and from the insufficient degree of robustness of existing parses.
 
-This work is a new attempt in the same direction, enabled by two multilinguality-oriented grammar formalisms developed over the course of the last 25 years: Universal Dependencies (UD) \cite{TODO:} and Grammatical Framework (GF) \cite{TODO:}.
+This work is a new attempt in the same direction, enabled by two multilinguality-oriented grammar formalisms developed over the course of the last 25 years: Universal Dependencies (UD) \cite{dummy} and Grammatical Framework (GF) \cite{dummy}.
 
 While both formalisms, the former a dependency and the latter a constituency grammar\todo{too quick, say more about both}, independently solve the former issue, UD is especially appealing since dependency trees are an easier target for parsing.
-As a consequence, several robust dependency parsers, such as \cite{TODO: UDPipe} and \cite{TODO: Standford} are available.
+As a consequence, several robust dependency parsers, such as \cite{dummy} and \cite{dummy} are available.
 
 UD parsing alone would then be sufficient to extract (or propagate) tree-to-tree alignments, but not to automate the generation of a morphologically-aware translation lexicon for a generative grammar.
-This is where GF comes into play: after correspondences are inferred from a parallel text, our proposed system is able to convert them to GF grammar rules, easy to embed in a domain-specific grammar but also making it possible to immediately evaluate the system by carrying out small-scale translation experiments using pre-existing grammatical constructions implemented in GF's Resource Grammar Library (RGL) \cite{TODO:}.
+This is where GF comes into play: after correspondences are inferred from a parallel text, our proposed system is able to convert them to GF grammar rules, easy to embed in a domain-specific grammar but also making it possible to immediately evaluate the system by carrying out small-scale translation experiments using pre-existing grammatical constructions implemented in GF's Resource Grammar Library (RGL) \cite{dummy}.
 \todo{say what the RGL is}
-This is made possible by \texttt{gf-ud}, a conversion tool described in \cite{TODO:} and \cite{TODO:}.
+This is made possible by \texttt{gf-ud}, a conversion tool described in \cite{dummy} and \cite{dummy}.
 
@@
 \paragraph{Part-Of-Speech equivalence}
-CoNNL-U files provide information not only on the syntactic role of the tokens a sentence is composed of, but also on their grammatical categories, represented as Universal Part-Of-Speech (POS) tags \cite{TODO:}.
+CoNNL-U files provide information not only on the syntactic role of the tokens a sentence is composed of, but also on their grammatical categories, represented as Universal Part-Of-Speech (POS) tags \cite{dummy}.
 Intuitively, if the nodes of two trees in matching contexts have the same POS tags, the two trees are more likely to correspond to each other than if not.
 
-This is especially true if we focus, for instance, solely on the open class words (in the current implementation, defined as in the UD documentation \cite{TODO:}), thus ignoring prepositions, determiners, auxiliary verbs and all other function words, which tend to behave differently across different languages.
+This is especially true if we focus, for instance, solely on the open class words (in the current implementation, defined as in the UD documentation \cite{dummy}), thus ignoring prepositions, determiners, auxiliary verbs and all other function words, which tend to behave differently across different languages.
 
@@
 If many distinctions of this kind are specific to particular language pairs or even stylistical, some of them occur independently of what languages are involved and have nothing to do with idiomatic usage or aspectual, discourse, domain or word knowledge.
-Drawing inspiration from \cite{TODO:} and \cite{TODO:}, we refer to them as \textit{translation divergences} and handle the most common ones via alignment criteria that are used by default.
+Drawing inspiration from \cite{dummy} and \cite{dummy}, we refer to them as \textit{translation divergences} and handle the most common ones via alignment criteria that are used by default.
 \todo{add a few examples (<= one per type of divergence)}
@@
 To this purpose, we use a 100-sentences subset of the Parallel UD (PUD) corpus, a set of 1000 handcrafted\footnote{Annotation is sometimes semi-automatic, but all treebanks are at least manually checked.}\todo{not even, I fear.
-Check this} multilingual treebanks in CoNLL-U format created for the CoNLL 2017 shared task on Multilingual Parsing from Raw Text to Universal Dependencies \cite{TODO:}.
+Check this} multilingual treebanks in CoNLL-U format created for the CoNLL 2017 shared task on Multilingual Parsing from Raw Text to Universal Dependencies \cite{dummy}.
 PUD treebanks are available in over 20 languages, of which we selected English, Italian and Swedish. 
@@
 For brevity, we will refer to these two datasets as to the DMI and the CSE corpora, or collectively as to the course plans datasets.
-This data, available in the project repository, was collected and sentence-aligned specifically for this work and a related Bachelor's thesis project \cite{TODO:}.
-When using raw text, our parser of choice is UDPipe \cite{TODO:}. 
-In particular, we use the ParTUT English and Italian models for the DMI corpus and models trained on the bilingual LinES English-Swedish treebank for the CSE corpus \cite{TODO:}.
+This data, available in the project repository, was collected and sentence-aligned specifically for this work and a related Bachelor's thesis project \cite{dummy}.
+When using raw text, our parser of choice is UDPipe \cite{dummy}. 
+In particular, we use the ParTUT English and Italian models for the DMI corpus and models trained on the bilingual LinES English-Swedish treebank for the CSE corpus \cite{dummy}.
 
@@
 
-We assess the correctness of the alignments the CE module is able to extract first from manually annotated data, then from the DMI and CSE corpus, comparing our results with those obtained with a standard statistical tool, \texttt{fast\_align} \cite{TODO:}.
+We assess the correctness of the alignments the CE module is able to extract first from manually annotated data, then from the DMI and CSE corpus, comparing our results with those obtained with a standard statistical tool, \texttt{fast\_align} \cite{dummy}.
 Doing so requires manually labelling the correspondences as either correct or incorrect.
@@
 
-\todo{in case there's room maybe add criterion-specific stats}
-
-% TODO: PUD: incl. reason-wise?)
-
 \subsection{MT experiments}
@@
 This suggest that, in contexts where higher precision is necessary, there are simple improvements that can be done either in the choice of the alignment criteria or at a later stage, in terms alignment selection.
-In other contexts, it can be more practical to aim for higher coverage and not perform any automatic selection, leaving lexicon refinements to a human, possibly with the help of the postprocessing tools that are currently under development \cite{TODO:}.
+In other contexts, it can be more practical to aim for higher coverage and not perform any automatic selection, leaving lexicon refinements to a human, possibly with the help of the postprocessing tools that are currently under development \cite{dummy}.
 
@@
 % Entries for the entire Anthology, followed by custom entries
-%\bibliography{anthology,custom}
-%\bibliographystyle{acl_natbib}
+\bibliography{custom}  
+\bibliographystyle{plain}
 

### 2021-05-10 paper/paper.tex
oops the problem was a missing style file
started working on bibliography

@@
 \section{Introduction}
-Grammar-based translation pipelines such as those based on Grammatical Framework (GF) \cite{dummy} have been successfully employed in domain-specific Machine Translation (MT).  
+Grammar-based translation pipelines such as those based on Grammatical Framework (GF) have been successfully employed in domain-specific Machine Translation (MT) \cite{scaling_up_gf}.  
 What makes these systems well suited to the task is the fact that, when we constrain ourselves to a specific domain, where precision is most often more important than coverage, they can provide strong guarantees in terms of grammatical correctness.
@@
 In many cases, this is done for the most part manually, resulting in a time consuming task requiring significant linguistic knowledge. 
-When the grammar is designed based on a corpus of parallel example sentences, it is possible to automate part of this process by means of statistical word and phrase alignment techniques \cite{dummy}.
+When the grammar is designed based on a corpus of parallel example sentences, it is possible to automate part of this process by means of statistical word and phrase alignment techniques \cite{brown-etal-1993-mathematics, och-ney-2000-improved, dyer-etal-2013-simple}.
 None of them is, however, suitable for the common case in which only a limited number of example data is available.
@@
 As stated in the Introduction, most existing extraction solutions are based on statistical approaches and are, as a consequence, unsuitable for small datasets. 
-Grammar-based approaches, making use of parallel treebanks and collectively referred to as \textit{tree-to-tree alignment methods}, have also been proposed \cite{dummy}, but have historically suffered from the inconsistencies between the various formalisms used to define the grammars for different languages and from the insufficient degree of robustness of existing parses.
+Grammar-based approaches, making use of parallel treebanks and collectively referred to as \textit{tree-to-tree alignment methods}, have also been proposed \cite{tiedemann2011bitext}, but have historically suffered from the inconsistencies between the various formalisms used to define the grammars for different languages and from the insufficient degree of robustness of existing parses.
 
-This work is a new attempt in the same direction, enabled by two multilinguality-oriented grammar formalisms developed over the course of the last 25 years: Universal Dependencies (UD) \cite{dummy} and Grammatical Framework (GF) \cite{dummy}.
+This work is a new attempt in the same direction, enabled by two multilinguality-oriented grammar formalisms developed over the course of the last 25 years: Universal Dependencies (UD) \cite{ws-2019-universal} and Grammatical Framework (GF) \cite{scaling_up_gf}.
 
 While both formalisms, the former a dependency and the latter a constituency grammar\todo{too quick, say more about both}, independently solve the former issue, UD is especially appealing since dependency trees are an easier target for parsing.
-As a consequence, several robust dependency parsers, such as \cite{dummy} and \cite{dummy} are available.
+As a consequence, several robust dependency parsers, such as \cite{straka-etal-2016-udpipe} and \cite{chen-manning-2014-fast} are available.
 
@@
 % Entries for the entire Anthology, followed by custom entries
-\bibliography{custom}  
-\bibliographystyle{plain}
+\bibliography{anthology,custom}  
+\bibliographystyle{acl_natbib}
 

### 2021-05-11 paper/paper.tex
done with bibliography

@@
 \section{Introduction}
-Grammar-based translation pipelines such as those based on Grammatical Framework (GF) have been successfully employed in domain-specific Machine Translation (MT) \cite{scaling_up_gf}.  
+Grammar-based translation pipelines such as those based on Grammatical Framework (GF) have been successfully employed in domain-specific Machine Translation (MT) \cite{ranta-etal-2020-abstract}.  
 What makes these systems well suited to the task is the fact that, when we constrain ourselves to a specific domain, where precision is most often more important than coverage, they can provide strong guarantees in terms of grammatical correctness.
@@
 
-This work is a new attempt in the same direction, enabled by two multilinguality-oriented grammar formalisms developed over the course of the last 25 years: Universal Dependencies (UD) \cite{ws-2019-universal} and Grammatical Framework (GF) \cite{scaling_up_gf}.
+This work is a new attempt in the same direction, enabled by two multilinguality-oriented grammar formalisms developed over the course of the last 25 years: Universal Dependencies (UD) \cite{ws-2019-universal} and Grammatical Framework (GF) \cite{ranta-etal-2020-abstract}.
 
@@
 UD parsing alone would then be sufficient to extract (or propagate) tree-to-tree alignments, but not to automate the generation of a morphologically-aware translation lexicon for a generative grammar.
-This is where GF comes into play: after correspondences are inferred from a parallel text, our proposed system is able to convert them to GF grammar rules, easy to embed in a domain-specific grammar but also making it possible to immediately evaluate the system by carrying out small-scale translation experiments using pre-existing grammatical constructions implemented in GF's Resource Grammar Library (RGL) \cite{dummy}.
+This is where GF comes into play: after correspondences are inferred from a parallel text, our proposed system is able to convert them to GF grammar rules, easy to embed in a domain-specific grammar but also making it possible to immediately evaluate the system by carrying out small-scale translation experiments using pre-existing grammatical constructions implemented in GF's Resource Grammar Library (RGL).
 \todo{say what the RGL is}
-This is made possible by \texttt{gf-ud}, a conversion tool described in \cite{dummy} and \cite{dummy}.
+This is made possible by \texttt{gf-ud}, a conversion tool described in \cite{kolachina-ranta-2016-abstract} and \cite{ranta-kolachina-2017-universal}.
 
@@
 
-The pseudocode of the basic extraction algorithm is shown in Figure \ref{TODO}\todo{add pseudocode}.
+The pseudocode of the basic extraction algorithm is shown in Figure \ref{TODO:}\todo{add pseudocode}.
 The input is a list of priority-sorted \textit{alignment criteria}, i.e.rules to determine whether two dependency trees should be aligned with each other, which will be discussed in greater detail in Section \ref{criteria}, and UD tree representation of the pair of sentences to align.
@@
 \paragraph{Part-Of-Speech equivalence}
-CoNNL-U files provide information not only on the syntactic role of the tokens a sentence is composed of, but also on their grammatical categories, represented as Universal Part-Of-Speech (POS) tags \cite{dummy}.
+CoNNL-U files provide information not only on the syntactic role of the tokens a sentence is composed of, but also on their grammatical categories, represented as Universal Part-Of-Speech (POS) tags \cite{petrov-etal-2012-universal}.
 Intuitively, if the nodes of two trees in matching contexts have the same POS tags, the two trees are more likely to correspond to each other than if not.
 
-This is especially true if we focus, for instance, solely on the open class words (in the current implementation, defined as in the UD documentation \cite{dummy}), thus ignoring prepositions, determiners, auxiliary verbs and all other function words, which tend to behave differently across different languages.
+This is especially true if we focus, for instance, solely on the open class words (in the current implementation, defined as in the UD documentation \footnote{\url{universaldependencies.org/u/pos/all.html}}), thus ignoring prepositions, determiners, auxiliary verbs and all other function words, which tend to behave differently across different languages.
 
@@
 If many distinctions of this kind are specific to particular language pairs or even stylistical, some of them occur independently of what languages are involved and have nothing to do with idiomatic usage or aspectual, discourse, domain or word knowledge.
-Drawing inspiration from \cite{dummy} and \cite{dummy}, we refer to them as \textit{translation divergences} and handle the most common ones via alignment criteria that are used by default.
+Drawing inspiration from \cite{dorr-1994-machine}, we refer to them as \textit{translation divergences} and handle the most common ones via alignment criteria that are used by default.
 \todo{add a few examples (<= one per type of divergence)}
@@
 To this purpose, we use a 100-sentences subset of the Parallel UD (PUD) corpus, a set of 1000 handcrafted\footnote{Annotation is sometimes semi-automatic, but all treebanks are at least manually checked.}\todo{not even, I fear.
-Check this} multilingual treebanks in CoNLL-U format created for the CoNLL 2017 shared task on Multilingual Parsing from Raw Text to Universal Dependencies \cite{dummy}.
+Check this} multilingual treebanks in CoNLL-U format created for the CoNLL 2018 shared task on Multilingual Parsing from Raw Text to Universal Dependencies \cite{zeman-etal-2018-conll}.
 PUD treebanks are available in over 20 languages, of which we selected English, Italian and Swedish. 
@@
 For brevity, we will refer to these two datasets as to the DMI and the CSE corpora, or collectively as to the course plans datasets.
-This data, available in the project repository, was collected and sentence-aligned specifically for this work and a related Bachelor's thesis project \cite{dummy}.
-When using raw text, our parser of choice is UDPipe \cite{dummy}. 
-In particular, we use the ParTUT English and Italian models for the DMI corpus and models trained on the bilingual LinES English-Swedish treebank for the CSE corpus \cite{dummy}.
+This data, available in the project repository, was collected and sentence-aligned specifically for this work and a related Bachelor's thesis project \cite{thesis}.
+When using raw text, our parser of choice is UDPipe \cite{straka-etal-2016-udpipe}. 
+In particular, we use the ParTUT English and Italian models for the DMI corpus and models trained on the bilingual LinES English-Swedish treebank for the CSE corpus.
 
@@
 
-We assess the correctness of the alignments the CE module is able to extract first from manually annotated data, then from the DMI and CSE corpus, comparing our results with those obtained with a standard statistical tool, \texttt{fast\_align} \cite{dummy}.
+We assess the correctness of the alignments the CE module is able to extract first from manually annotated data, then from the DMI and CSE corpus, comparing our results with those obtained with a standard statistical tool, \texttt{fast\_align} \cite{dyer-etal-2013-simple}.
 Doing so requires manually labelling the correspondences as either correct or incorrect.
@@
 This suggest that, in contexts where higher precision is necessary, there are simple improvements that can be done either in the choice of the alignment criteria or at a later stage, in terms alignment selection.
-In other contexts, it can be more practical to aim for higher coverage and not perform any automatic selection, leaving lexicon refinements to a human, possibly with the help of the postprocessing tools that are currently under development \cite{dummy}.
+In other contexts, it can be more practical to aim for higher coverage and not perform any automatic selection, leaving lexicon refinements to a human, possibly with the help of the postprocessing tools that are currently under development.
 

### 2021-05-12 paper/paper.tex
minor changes
updated CE PUD results

@@
 On the other hand, the relatively large size of the PUD treebanks makes it possible to also train the statistical tool on the full dataset instead of just using the chosen 100-sentences subset, allowing for a fairer comparison.
-In both cases, \texttt{fast\_align} is run with 10 iterations in EM training, symmetrizing the alignments found by running the program in both directions.
+In both cases, \texttt{fast\_align} is run with the recommended parameters.
 
-% TODO: position properly
-% TODO: add english-swedish!
 \begin{table*}[h]
-  \centering
-  \small
-  \begin{tabular}{l|lll}
-  %\hline
-   & \textbf{CE} & \textbf{\texttt{fast\_align} 100} & \textbf{\texttt{fast\_align} 1000}\\ \hline
-   \textbf{tot. alignments} & 716 & 1440 & 1435 \\ 
-   \textbf{\texttt{+} and \texttt{=}} & 536 (75\%) & 410 (28\%) & 656 (46\%)\\ 
-   \textbf{\texttt{+}} & 491 (69\%) & 371 (26\%) & 590 (41\%)\\ 
+  \begin{tabular}{r|c|c|c|c|c|c}
+  \multicolumn{1}{l|}{\textbf{}} & \multicolumn{2}{c|}{\textbf{CE}} & \multicolumn{2}{c|}{\textbf{\begin{tabular}[c]{@{}c@{}}fast\_align (100)\end{tabular}}} & \multicolumn{2}{c}{\textbf{\begin{tabular}[c]{@{}c@{}}fast\_align (1000)\end{tabular}}} \\ \hline
+  \multicolumn{1}{l|}{}          & \textbf{en-it}  & \textbf{en-sv} & \textbf{en-it}                              & \textbf{en-sv}                              & \textbf{en-it}                                & \textbf{en-sv} \\ \hline
+  \textbf{tot.
+alignments}        & 536             & 638            & 1242                                        & 1044                                        & 1286                                          & 1065          \\
+  \textbf{+ and =}                & 392 (73\%)      & 514 (80\%)     & 346 (28\%)                                  & 538 (52\%)                                  & 540 (42\%)                                    & 677 (64\%)    \\
+  \textbf{+}                      & 363 (68\%)      & 503 (79\%)     & 316 (25\%)                                  & 525 (50\%)                                  & 510 (40\%)                                    & 666 (63\%)    \\
   \end{tabular}

### 2021-05-13 paper/paper.tex
changed plans for second ce experiment

@@
 
-We assess the correctness of the alignments the CE module is able to extract first from manually annotated data, then from the DMI and CSE corpus, comparing our results with those obtained with a standard statistical tool, \texttt{fast\_align} \cite{dyer-etal-2013-simple}.
-Doing so requires manually labelling the correspondences as either correct or incorrect.
+We first assess the correctness of the alignments the CE module is able to extract from manually annotated data from the PUD treebanks, comparing our results with those obtained with a standard statistical tool, \texttt{fast\_align} \cite{dyer-etal-2013-simple}.
+In addition, we try to quantify the impact automatic parsing has on the performance of the CE module by comparing the above results wiht those obtained on the DMI and CSE corpora.
 

### 2021-05-13 paper/paper.tex
rerun second ce experiment and updated results

@@
 \begin{table*}[h]
+  \centering
   \begin{tabular}{r|c|c|c|c|c|c}
@@
 
-%TODO: position properly
-%TODO: run experiments & update data
 \begin{table*}[h]
   \centering
-  \small
-  \begin{tabular}{l|lll}
-
+  \begin{tabular}{r|c|c|c|c}
+  \multicolumn{1}{l|}{\textbf{}} & \multicolumn{2}{c|}{\textbf{\begin{tabular}[c]{@{}c@{}}PUD (100 sentences)\end{tabular}}} & \multicolumn{2}{c}{\textbf{course plans}}                                                                                                                      \\ \hline
+  \multicolumn{1}{l|}{}          & \textbf{en-it}                               & \textbf{en-sv}                               & \textbf{\begin{tabular}[c]{@{}c@{}}DMI (en-it)\\ (881 sentences)\end{tabular}} & \textbf{\begin{tabular}[c]{@{}c@{}}CSE (en-sv)\\ (539 sentences)\end{tabular}} \\ \hline
+  \textbf{tot.
+alignments}        & 1197                                         & 1325                                         & 1823                                                                           & 1950                                                                           \\
+  \textbf{+ and =}                & 916 (77\%)                                   & 1112 (85\%)                                  & 1205 (66\%)                                                                    & 1296 (66\%)                                                                    \\
+  \textbf{+}                      & 880 (74\%)                                   & 1099 (84\%)                                  & 1157 (63\%)                                                                    & 1248 (64\%)                                                                    \\
   \end{tabular}
-  \caption[Comparison between our grammar-based CE module and \texttt{fast\_align}]{Comparison between our grammar-based CE module and \texttt{fast\_align} on raw data, using the DMI and CSE corpora.
-Row 1 reports the total number of distinct alignments obtain.
-Rows 2 and 3 give the percentages of correct alignments, with and without considering sentence pair-specific correspondences as correct.}
+  \caption{TODO:}
   \label{raw_fast}

### 2021-05-14 paper/paper.tex
various additions

@@
 When the grammar is designed based on a corpus of parallel example sentences, it is possible to automate part of this process by means of statistical word and phrase alignment techniques \cite{brown-etal-1993-mathematics, och-ney-2000-improved, dyer-etal-2013-simple}.
-None of them is, however, suitable for the common case in which only a limited number of example data is available.
+None of them is, however, suitable for the common case in which only a limited amount of example data is available.
 
@@
 
-This work is a new attempt in the same direction, enabled by two multilinguality-oriented grammar formalisms developed over the course of the last 25 years: Universal Dependencies (UD) \cite{ws-2019-universal} and Grammatical Framework (GF) \cite{ranta-etal-2020-abstract}.
+This work is a new attempt in the same direction, enabled by two multilinguality-oriented grammar formalisms developed over the course of the last 25 years: Grammatical Framework (GF) \cite{ranta-etal-2020-abstract} and Universal Dependencies (UD) \cite{ws-2019-universal}.
 
-While both formalisms, the former a dependency and the latter a constituency grammar\todo{too quick, say more about both}, independently solve the former issue, UD is especially appealing since dependency trees are an easier target for parsing.
+GF is a constituency grammar formalisms and programming language designed with the parallel between programming and natural languages in mind: grammars are represented as pairs of an \textit{abstract syntax}, playing the role of an interlingua, and a set of \textit{concrete syntaxes} capturing the specificities of the various natural languages.
+In the case of translation, strings in the source language are \textit{parsed} to Abstract Syntax Trees (ASTs) which are then \textit{linearized} to target language strings.
+
+UD, on the other hand, is a dependency grammar formalism meant for cross-linguistically consistent grammatical annotation.
+As opposed to constituency, \textit{dependency} is a word-to-word correspondence: each \textit{head} is simply put in relation with the ones that depend on it, called its \textit{dependents}, via directed labelled links, specifying the syntactic relation between them, and the root element of a sentence is attached to a fake node with a link labelled \texttt{root}.
+Importantly for our application, the standard format for UD trees, CoNNL-U, gives information not only on the syntactic role of each word of a sentence, but also on its POS tag, lemma and other morphological features.
+
+While both formalisms independently solve the former issue, UD is especially appealing since dependency trees are an easier target for parsing. 
 As a consequence, several robust dependency parsers, such as \cite{straka-etal-2016-udpipe} and \cite{chen-manning-2014-fast} are available.
@@
 UD parsing alone would then be sufficient to extract (or propagate) tree-to-tree alignments, but not to automate the generation of a morphologically-aware translation lexicon for a generative grammar.
-This is where GF comes into play: after correspondences are inferred from a parallel text, our proposed system is able to convert them to GF grammar rules, easy to embed in a domain-specific grammar but also making it possible to immediately evaluate the system by carrying out small-scale translation experiments using pre-existing grammatical constructions implemented in GF's Resource Grammar Library (RGL).
-\todo{say what the RGL is}
+This is where GF comes into play: after correspondences are inferred from a parallel text, our proposed system is able to convert them to GF grammar rules, easy to embed in a domain-specific grammar but also making it possible to immediately evaluate the system by carrying out small-scale translation experiments using pre-existing grammatical constructions implemented in GF's Resource Grammar Library (RGL), which covers the morphology and basic syntax of over 30 languages.
 This is made possible by \texttt{gf-ud}, a conversion tool described in \cite{kolachina-ranta-2016-abstract} and \cite{ranta-kolachina-2017-universal}.
 
-Concretely, the system we propose requires the following elements, whose reciprocal relations are shown in Figure \ref{overview}: \smallskip
+Concretely, the system we propose consists of the following elements: \smallskip
 
@@
 \end{itemize} \smallskip
-
-% TODO: replace CE with CA and omit CP entirely?
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=0.45\textwidth]{figures/overview.png}
-  \caption[System overview]{System overview. Parts in grey are currently at the prototype stage and will not be further discussed in this paper.} \label{elems}
-  \label{overview}
-\end{figure}
 
@@
 As a first step, the program checks whether the two full sentence trees can be aligned with each other, i.e. if they match any of the alignment criteria. 
-If this is the case, they are added to a collection of alignments, which are also represented as a pair of UD trees associated with some metadata. \todo{explain what metadata?}
-Such collection is what the function will return after aligning all the dependency subtrees. % TODO: explain or omit metadata!
+If this is the case, they are added to a collection of alignments, which are also represented as a pair of UD trees associated with some metadata, such as the id of the sentence they were extracted from.
+Such collection is what the function will return after aligning all the dependency subtrees.
 
-Depending on which alignment criteria they match, the head of the two trees may or may not also be added to such collection. This is important... \todo{explain why}% TODO:
+Depending on which alignment criteria they match, the head of the two trees may or may not also be added to such collection.
+This is important not to miss one-word correspondences that cannot be captured in any other way, for instance between the root verbs of two full sentences.
 
 The same procedure is applied recursively to all pairs of immediate subtrees of each sentence, until the leaves are reached or alignment is no longer possible due to a lack of matching criteria.
-% TODO: explain & motivate sorting -> maybe say more about UD before?
+Subtrees are sorted based on their dependency label so to give higher priority to pairs with the same root label.
 
@@
 The CE module described so far outputs pairs of UD trees. 
-While converting them to GF Abstract Syntax Trees (ASTs) is \texttt{gf-ud}'s core functionality, to build a compilable GF translation lexicon it is necessary to turn such trees into grammar rules.
+While converting them to GF ASTs is \texttt{gf-ud}'s core functionality, to build a compilable GF translation lexicon it is necessary to turn such trees into grammar rules.
 
@@
 \section{Evaluation} \label{evaluation}
+In this section, we evaluate the system proposed in the above. 
+We first introduce the data used for such evaluation. 
+After that, we describe our experiments, aimed at putting both the CE module \textit{per se} and the entire system from parsing to lexicon generation to the test, and present our results.
 % TODO: move around things to have something here (or outline structure of the section)
-\subsection{Data}
+\subsection{Data} \label{data}
 Because we want part of our evaluation to be independent from the quality of UD parsing, some of the experiments are carried out using manually annotated data. 
@@
 
-When it comes to testing the program on raw text, we use two bilingual sentence-aligned corpora consisting of course plans from the Department of Mathematics and Computer Science of the University of Perugia (for English-Italian) and from the Department of Computer Science and Engineering (CSE) shared between the University of Gothenburg and the Chalmers University of Technology (for English-Swedish).
+When it comes to testing the program on raw text, we use two small ($< 1000$ sentences) bilingual sentence-aligned corpora consisting of course plans from the Department of Mathematics and Computer Science (DMI) of the University of Perugia (for English-Italian) and from the Department of Computer Science and Engineering (CSE) shared between the University of Gothenburg and the Chalmers University of Technology (for English-Swedish).
 For brevity, we will refer to these two datasets as to the DMI and the CSE corpora, or collectively as to the course plans datasets.
@@
 
-\subsection{Evaluating CE} %TODO: better title?
+\subsection{Evaluating CE} \label{ceval}
 While in this work we focus mostly on the MT applications of CA, automatic translation, and much less GF-based domain-specific tranlation, is not the only context in which our system can be put to use.
@@
 On the other hand, the relatively large size of the PUD treebanks makes it possible to also train the statistical tool on the full dataset instead of just using the chosen 100-sentences subset, allowing for a fairer comparison.
-In both cases, \texttt{fast\_align} is run with the recommended parameters.
+
+In both cases, \texttt{fast\_align} is run with the recommended parameters and the CE program is configured to only extract one-to-many and many-to-one word alignments, given that \texttt{fast\_align} does not align larger phrases.
 
@@
 
-The results of this comparison are summarized in Table \ref{raw_fast}\todo{position properly; run experiments (this is new) and add actual results}.
+The results of this comparison are summarized and compared with those obtained on the sample of PUD treebanks in Table \ref{raw_fast}.
+
+What is immediately evident, but not unexpected, is a decrease in precision. 
+The percentage of correct alignments, however, stays significantly higher than that obtained with \texttt{fast\_align} in the previous experiment, even when the model was trained on the full PUD corpus: even though percentages seem similar for English-Swedish, it must be taken into account that the CSE corpus is roughly half the size of the PUD corpus.
+
+Results are less encouraging in terms of recall: the number of alignments extracted from the course plans corpora is similar to that obtained from the PUD treebank sample, despite the difference in size.
+This is explained in part by the presence, in the course plans corpora, of a large amount of very short sentences, and in part by the fact that parse errors introduced by UDPipe make it impossible to align many subsentences without a significant loss in terms of precision.
 
@@
 
-\subsection{MT experiments}
+\subsection{MT experiments} \label{mtexp}
 The second set of experiments has the objective of assessing the quality of the final output of the system we propose: GF translation lexica. 

### 2021-05-14 paper/paper.tex
improved tables

@@
   \begin{tabular}{r|c|c|c|c|c|c}
-  \multicolumn{1}{l|}{\textbf{}} & \multicolumn{2}{c|}{\textbf{CE}} & \multicolumn{2}{c|}{\textbf{\begin{tabular}[c]{@{}c@{}}fast\_align (100)\end{tabular}}} & \multicolumn{2}{c}{\textbf{\begin{tabular}[c]{@{}c@{}}fast\_align (1000)\end{tabular}}} \\ \hline
+  \multicolumn{1}{l|}{\textbf{}} & \multicolumn{2}{c|}{\textbf{CE}} & \multicolumn{2}{c|}{\textbf{\begin{tabular}[c]{@{}c@{}}\texttt{fast\_align} \\(100 sentences)\end{tabular}}} & \multicolumn{2}{c}{\textbf{\begin{tabular}[c]{@{}c@{}}\texttt{fast\_align} \\(full dataset)\end{tabular}}} \\ \hline
   \multicolumn{1}{l|}{}          & \textbf{en-it}  & \textbf{en-sv} & \textbf{en-it}                              & \textbf{en-sv}                              & \textbf{en-it}                                & \textbf{en-sv} \\ \hline
-  \textbf{tot.
-alignments}        & 536             & 638            & 1242                                        & 1044                                        & 1286                                          & 1065          \\
-  \textbf{+ and =}                & 392 (73\%)      & 514 (80\%)     & 346 (28\%)                                  & 538 (52\%)                                  & 540 (42\%)                                    & 677 (64\%)    \\
-  \textbf{+}                      & 363 (68\%)      & 503 (79\%)     & 316 (25\%)                                  & 525 (50\%)                                  & 510 (40\%)                                    & 666 (63\%)    \\
+  \textbf{distinct alignments}        & 536             & 638            & 1242                                        & 1044                                        & 1286                                          & 1065          \\
+  \textbf{correct (\texttt{+} and \texttt{=})}                & 392 (73\%)      & 514 (80\%)     & 346 (28\%)                                  & 538 (52\%)                                  & 540 (42\%)                                    & 677 (64\%)    \\
+  \textbf{usable in MT (\texttt{+})}                      & 363 (68\%)      & 503 (79\%)     & 316 (25\%)                                  & 525 (50\%)                                  & 510 (40\%)                                    & 666 (63\%)    \\
   \end{tabular}
-  \caption[Comparison between our grammar-based CE module and \texttt{fast\_align}]{Comparison between our grammar-based CE module and \texttt{fast\_align} on PUD data.
-The middle column shows the results obtained by training \texttt{fast\_align} on the 100-sentence subset fed to the extraction module.
-On the right, the results obtained by training \texttt{fast\_align} on the full PUD treebanks dataset, discarding the alignments obtained for sentences 101-1000 for the final comparison.
-Row 1 reports the total number of distinct alignments obtain.
-Rows 2 and 3 give the percentages of correct alignments, with and without considering sentence pair-specific correspondences as correct.}
+  \caption[Word alignment comparison between our grammar-based CE module and \texttt{fast\_align}]{Comparison between our grammar-based CE module and \texttt{fast\_align} on PUD data.
+The middle column shows the results obtained by training \texttt{fast\_align} on the 100-sentence subset also fed to the extraction module.
+On the right, the results obtained by training it on the full dataset, discarding the alignments obtained for sentences 101-1000 for the final comparison.}
   \label{pud_fast}
@@
 
+\begin{table*}[h]
+  \centering
+  \begin{tabular}{r|c|c|c|c}
+  \multicolumn{1}{l|}{\textbf{}} & \multicolumn{2}{c|}{\textbf{\begin{tabular}[c]{@{}c@{}}PUD (100 sentences)\end{tabular}}} & \multicolumn{2}{c}{\textbf{course plans}}                                                                                                                      \\ \hline
+  \multicolumn{1}{l|}{}          & \textbf{en-it}                               & \textbf{en-sv}                               & \textbf{\begin{tabular}[c]{@{}c@{}}DMI (881 sentences)\end{tabular}} & \textbf{\begin{tabular}[c]{@{}c@{}}CSE (539 sentences)\end{tabular}} \\ \hline
+  \textbf{disrinct alignments}        & 1197                                         & 1325                                         & 1823                                                                           & 1950                                                                           \\
+  \textbf{correct (\texttt{+} and \texttt{=})}                & 916 (77\%)                                   & 1112 (85\%)                                  & 1205 (66\%)                                                                    & 1296 (66\%)                                                                    \\
+  \textbf{usable in MT (\texttt{+}})                      & 880 (74\%)                                   & 1099 (84\%)                                  & 1157 (63\%)                                                                    & 1248 (64\%)                                                                    \\
+  \end{tabular}
+  \caption{Comparison the extraction of alignments of any size from manually annotated PUD treebanks and automatically parsed sentences from the course plans corpora.}
+  \label{raw_fast}
+  \end{table*}
+
 What is immediately evident, but not unexpected, is a decrease in precision. 
@@
 This is explained in part by the presence, in the course plans corpora, of a large amount of very short sentences, and in part by the fact that parse errors introduced by UDPipe make it impossible to align many subsentences without a significant loss in terms of precision.
-
-\begin{table*}[h]
-  \centering
-  \begin{tabular}{r|c|c|c|c}
-  \multicolumn{1}{l|}{\textbf{}} & \multicolumn{2}{c|}{\textbf{\begin{tabular}[c]{@{}c@{}}PUD (100 sentences)\end{tabular}}} & \multicolumn{2}{c}{\textbf{course plans}}                                                                                                                      \\ \hline
-  \multicolumn{1}{l|}{}          & \textbf{en-it}                               & \textbf{en-sv}                               & \textbf{\begin{tabular}[c]{@{}c@{}}DMI (en-it)\\ (881 sentences)\end{tabular}} & \textbf{\begin{tabular}[c]{@{}c@{}}CSE (en-sv)\\ (539 sentences)\end{tabular}} \\ \hline
-  \textbf{tot.
-alignments}        & 1197                                         & 1325                                         & 1823                                                                           & 1950                                                                           \\
-  \textbf{+ and =}                & 916 (77\%)                                   & 1112 (85\%)                                  & 1205 (66\%)                                                                    & 1296 (66\%)                                                                    \\
-  \textbf{+}                      & 880 (74\%)                                   & 1099 (84\%)                                  & 1157 (63\%)                                                                    & 1248 (64\%)                                                                    \\
-  \end{tabular}
-  \caption{TODO:}
-  \label{raw_fast}
-  \end{table*}
-
- % TODO: discussion
-\todo{add results discussion}
 
@@
   \centering
-  \begin{tabular}{l|ll}
+  \begin{tabular}{r|cc}
   \textbf{}            & \textbf{DMI (en-it)} & \textbf{CSE (en-sv)} \\ \hline
@@
   \end{tabular}
-  \caption[BLEU scores for automatic translations based on the course plans grammars]{BLEU scores obtained by comparing one candidate automatic translation per sentence to a reference translation obtained by making the necessary corrections to the automatically generated ones.
-Both the initial English sentences and their automatic translations were generated using the two course plans grammars.}
+  \caption[BLEU scores for automatic translations based on the course plans grammars]{BLEU scores for automatic translations based on the course plans grammars.}
   \label{tableu}
@@
   \centering
-  \begin{tabular}{l|ll}
-  \textbf{}                                                                   \textbf{type of errors}                 & \textbf{DMI (en-it)} & \textbf{CSE (en-sv)} \\ \hline
+  \begin{tabular}{r|cc}
+  \textbf{}                                                                                    & \textbf{DMI (en-it)} & \textbf{CSE (en-sv)} \\ \hline
   \begin{tabular}[c]{@{}l@{}}\textbf{semantical}\end{tabular}               & 23 (46\%)    & 23 (46\%)    \\ 
   \begin{tabular}[c]{@{}l@{}}\textbf{grammatical}\end{tabular}             & 10 (20\%)      & 3 (6\%)      \\ 
-  \begin{tabular}[c]{@{}l@{}}\textbf{semantical and} \\ \textbf{grammatical}\end{tabular} & 3 (6\%)      & 4 (8\%)      \\ 
+  \begin{tabular}[c]{@{}l@{}}\textbf{both}\end{tabular} & 3 (6\%)      & 4 (8\%)      \\ 
   \end{tabular}
-  \caption[Types of errors encountered in the automatically translated sentences]{Number of automatically translated sentences containing only semantical, only grammatical and both semantical and grammatical errors in the two synthetic course plans corpora.}
+  \caption[Types of errors encountered in the automatically translated sentences.]{Types of errors encountered in the automatically translated sentences.}
   \label{obs}

### 2021-05-14 paper/paper.tex
improvements up to the infamous section 2.2

@@
 In grammar-based domain-specific machine translation, constructing high-quality translation lexica is a crucial but time consuming task requiring significant linguistic knowledge. 
-In presence of a corpus of parallel example sentences, statistical word and phrase alignment tools allow to automate part of this process, but are not suitable for small datasets and often fail to identify discontinuous multiword correspondences.
-Addressing both problems, we propose a data-driven, syntax-based approach to the automation of this task and put it to the test in a simple grammar-based translation pipeline.
+When example parallel sentences are available, statistical alignment tools allow for the automation of part of the process, but are not suitable for small datasets and are limited when it comes to discontinuous multiword correspondences.
+In addition, lexica generated in this way generally consist in simple word form correspondences that cannot be directly used in MT pipelines.
+Addressing these problems, we propose a data-driven, syntax-based approach to this task and put it to the test in a simple grammar-based translation pipeline.
 \end{abstract}
@@
 Grammar-based translation pipelines such as those based on Grammatical Framework (GF) have been successfully employed in domain-specific Machine Translation (MT) \cite{ranta-etal-2020-abstract}.  
-What makes these systems well suited to the task is the fact that, when we constrain ourselves to a specific domain, where precision is most often more important than coverage, they can provide strong guarantees in terms of grammatical correctness.
+What makes these systems well suited to the task is the fact that, when we constrain ourselves to a specific domain, where precision is often more important than coverage, they can provide strong guarantees in terms of grammatical correctness.
 
 Nevertheless, lexical exactness is, in this context, just as important as grammaticality. 
-An important part of the design of the Controlled Natural Language (CNL) the grammar in such a system describes becomes, then, the creation of a translation lexicon. 
-In many cases, this is done for the most part manually, resulting in a time consuming task requiring significant linguistic knowledge. 
+An important part of the design of the Controlled Natural Language (CNL) one such system's grammar describes then becomes the creation of a high-quality translation lexicon. 
+This is often done manually, resulting in a time consuming task requiring significant linguistic knowledge. 
 When the grammar is designed based on a corpus of parallel example sentences, it is possible to automate part of this process by means of statistical word and phrase alignment techniques \cite{brown-etal-1993-mathematics, och-ney-2000-improved, dyer-etal-2013-simple}.
@@
 In this paper, we propose an alternative approach to the automation of this task. 
-While still being data-driven, our method is grammar-based and, as such, capable of extracting meaningful correspondences even from individual sentence pairs.
+While still being data-driven, our method is also grammar-based and, as such, capable of extracting meaningful correspondences even from individual sentence pairs.
 
 A further advantage of performing syntactic analysis is that we do not have to choose \textit{a priori} whether to focus on the word or phrase level. 
-Instead, we can simultaneously operate at different levels of abstraction, thus extracting both single- and multiword, even non-contiguous, correspondences. 
+Instead, we can simultaneously operate at different levels of abstraction, extracting both single- and multiword, even non-contiguous, correspondences. 
 For this reason, we refer to the task our system attempts to automate as \textit{Concept Alignment} (CA).
 
+Finally, performing morphological analysis allows us to generate translation lexica consisting of complete with grammatical category and inflection information instead of simple correspondences between word forms.
+
 This paper is structured as follows. 
-Section \ref{methodology} starts by giving an overview of our approach to CA, comparing it to related work, to then focus on our algorithm for extracting correspondences.
-It is concluded with a description of our method for converting the alignments obtained in this way to a GF translation lexicon.
-After that, Section \ref{evaluation} presents the results of our first evaluation of the system.
-Finally, Section \ref{conclusions} consists of a discussion of such results and some ideas for future work.
+Section \ref{methodology} starts by giving an overview of our approach to CA and comparing it to related work to then focus on our algorithm for extracting correspondences.
+After that, Section \ref{evaluation} presents the results obtained in a first evaluation of the system.
+Finally, Section \ref{conclusions} summarizes our conclusions and discusses some ideas for future work.
 
@@
 
-% potentially move to future work directly
 An interesting and less studied variant of CA is \textit{Concept Propagation} (CP), useful for cases where a set of concepts is already known and the goal is to identify the expressions corresponding to each of them in a new language, potentially even working with a different text in the same domain.
@@
 
-As stated in the Introduction, most existing extraction solutions are based on statistical approaches and are, as a consequence, unsuitable for small datasets. 
-Grammar-based approaches, making use of parallel treebanks and collectively referred to as \textit{tree-to-tree alignment methods}, have also been proposed \cite{tiedemann2011bitext}, but have historically suffered from the inconsistencies between the various formalisms used to define the grammars for different languages and from the insufficient degree of robustness of existing parses.
+As stated in the Introduction, most existing extraction solutions are based on statistical approaches and, as a consequence, unsuitable for small datasets. 
+Grammar-based approaches, making use of parallel treebanks and collectively referred to as \textit{tree-to-tree alignment methods}, have also been proposed \cite{tiedemann2011bitext}, but have historically suffered from the inconsistencies between the formalisms used to define the grammars for the various languages and from the lack of robustness of existing parsers.
 
@@
 
-GF is a constituency grammar formalisms and programming language designed with the parallel between programming and natural languages in mind: grammars are represented as pairs of an \textit{abstract syntax}, playing the role of an interlingua, and a set of \textit{concrete syntaxes} capturing the specificities of the various natural languages.
-In the case of translation, strings in the source language are \textit{parsed} to Abstract Syntax Trees (ASTs) which are then \textit{linearized} to target language strings.
+GF is a constituency grammar formalisms and programming language in which grammars are represented as pairs of an \textit{abstract syntax}, playing the role of an interlingua, and a set of \textit{concrete syntaxes} capturing the specificities of the various natural languages.
+In the case of translation, similarly to what happens in programming language compilation, strings in the source language are \textit{parsed} to Abstract Syntax Trees (ASTs), which are then \textit{linearized} to target language strings.
 
 UD, on the other hand, is a dependency grammar formalism meant for cross-linguistically consistent grammatical annotation.
-As opposed to constituency, \textit{dependency} is a word-to-word correspondence: each \textit{head} is simply put in relation with the ones that depend on it, called its \textit{dependents}, via directed labelled links, specifying the syntactic relation between them, and the root element of a sentence is attached to a fake node with a link labelled \texttt{root}.
-Importantly for our application, the standard format for UD trees, CoNNL-U, gives information not only on the syntactic role of each word of a sentence, but also on its POS tag, lemma and other morphological features.
+As opposed to constituency, \textit{dependency} is a word-to-word correspondence: each word is simply put in relation with the one it refers to, called its \textit{head}, via a directed labelled link specifying the syntactic relation between them.
+Importantly for our application, the standard format for UD trees, CoNNL-U, gives information not only on the syntactic role of each word, but also on its POS tag, lemma and morphological features.
 
-While both formalisms independently solve the former issue, UD is especially appealing since dependency trees are an easier target for parsing. 
-As a consequence, several robust dependency parsers, such as \cite{straka-etal-2016-udpipe} and \cite{chen-manning-2014-fast} are available.
+While both formalisms independently solve the issues related to having to work with grammars that are inconsistent with each other, UD is especially appealing since, dependency trees being an easier target, several robust parsers, such as \cite{straka-etal-2016-udpipe} and \cite{chen-manning-2014-fast} are available.
 
-UD parsing alone would then be sufficient to extract (or propagate) tree-to-tree alignments, but not to automate the generation of a morphologically-aware translation lexicon for a generative grammar.
-This is where GF comes into play: after correspondences are inferred from a parallel text, our proposed system is able to convert them to GF grammar rules, easy to embed in a domain-specific grammar but also making it possible to immediately evaluate the system by carrying out small-scale translation experiments using pre-existing grammatical constructions implemented in GF's Resource Grammar Library (RGL), which covers the morphology and basic syntax of over 30 languages.
+UD parsing alone is sufficient to extract (or propagate) tree-to-tree alignments, but not to automate the generation of a morphologically-aware translation lexicon for a generative grammar. 
+This is where GF comes into play: after correspondences are inferred from a parallel text, our system is able to convert them to GF grammar rules.
+These are easy to embed in a domain-specific grammar but also making it possible to immediately evaluate the system by carrying out small-scale translation experiments using pre-existing grammatical constructions implemented in GF's Resource Grammar Library (RGL), which covers the morphology and basic syntax of over 30 languages.
 This is made possible by \texttt{gf-ud}, a conversion tool described in \cite{kolachina-ranta-2016-abstract} and \cite{ranta-kolachina-2017-universal}.
 
-Concretely, the system we propose consists of the following elements: \smallskip
+Concretely, the system we propose consists of the following elements:
 
-\begin{itemize}
+\begin{itemize} \setlength\itemsep{0.01em}
   \item a UD parser
@@
   \item a program, based on \texttt{gf-ud}, that converts the alignments into GF grammar rules and uses them to construct a translation lexicon.
-\end{itemize} \smallskip
+\end{itemize}
 
@@
 The pseudocode of the basic extraction algorithm is shown in Figure \ref{TODO:}\todo{add pseudocode}.
-The input is a list of priority-sorted \textit{alignment criteria}, i.e.rules to determine whether two dependency trees should be aligned with each other, which will be discussed in greater detail in Section \ref{criteria}, and UD tree representation of the pair of sentences to align.
+The input consists of a list of priority-sorted \textit{alignment criteria}, i.e.
+rules to determine whether two dependency trees should be aligned with each other which will be discussed in greater detail in Section \ref{criteria}, and a pair of sentences to align represented as UD trees.
 From an implementation point of view, UD trees are, as in \texttt{gf-ud}, rose trees where each node represents a token, obtained from the CoNLL-U files produced by the UD parser.
@@
 As a first step, the program checks whether the two full sentence trees can be aligned with each other, i.e. if they match any of the alignment criteria. 
-If this is the case, they are added to a collection of alignments, which are also represented as a pair of UD trees associated with some metadata, such as the id of the sentence they were extracted from.
+If this is the case, they are added to a collection of alignments, which are represented themselves as a pair of UD (sub)trees associated with some metadata, such as the id of the sentence they were extracted from.
 Such collection is what the function will return after aligning all the dependency subtrees.
@@
 The same procedure is applied recursively to all pairs of immediate subtrees of each sentence, until the leaves are reached or alignment is no longer possible due to a lack of matching criteria.
-Subtrees are sorted based on their dependency label so to give higher priority to pairs with the same root label.
+Subtrees are sorted based on their dependency label, so to give higher priority to pairs whose heads have the same label.
 
@@
 \subsubsection{Alignment criteria} \label{criteria}
-While alignment criteria are customizable, to better understand the extraction algorithm described in the above we dedicate this section to describing the ones that are currently used by default.
+While alignment criteria are customizable, to better understand the extraction algorithm described in the above, we dedicate this section to describing the ones that the current implementation uses by default.
 
@@
 The most obvious, but also most effective idea is to determine alignability based on comparing the dependency labels of the candidate UD tree pair. 
-In particular, according to this idea, two subtrees in \textit{matching context}, i.e. attached to aligned heads, constitute an alignment if their roots share the same dependency label. 
-Intuitively, this means that they are in the same relation with their heads.
+In particular, according to this idea, two subtrees in \textit{matching context}, i.e.
+attached to aligned heads, constitute an alignment if their roots share the same dependency label, meaning that they are in the same syntactic relation with their heads.
 
@@
 \paragraph{Part-Of-Speech equivalence}
-CoNNL-U files provide information not only on the syntactic role of the tokens a sentence is composed of, but also on their grammatical categories, represented as Universal Part-Of-Speech (POS) tags \cite{petrov-etal-2012-universal}.
+As noted in the above, CoNNL-U files also provide information on the grammatical categories of each word, represented as Universal Part-Of-Speech (POS) tags \cite{petrov-etal-2012-universal}. 
 Intuitively, if the nodes of two trees in matching contexts have the same POS tags, the two trees are more likely to correspond to each other than if not. 
+This is especially true if we focus, for instance, solely on the open class words (defined as in the UD documentation \footnote{\url{universaldependencies.org/u/pos/all.html}}), thus ignoring prepositions, determiners, auxiliary verbs and all other function words, which tend to behave differently across different languages.
 
-This is especially true if we focus, for instance, solely on the open class words (in the current implementation, defined as in the UD documentation \footnote{\url{universaldependencies.org/u/pos/all.html}}), thus ignoring prepositions, determiners, auxiliary verbs and all other function words, which tend to behave differently across different languages.
+A useful relation to define between dependency trees is, then, that of \textit{POS-equivalence}: two dependency trees $t$, $u$ are POS-equivalent if $M_1 = M_2 \neq \emptyset$, where $M_i$ is defined as the multiset of POS tags of all the open class word nodes of $t_i$.
 
-As a consequence, a useful relation to define between dependency trees is that of \textit{POS-equivalence}: two dependency trees $t$, $u$ are POS-equivalent if $M_1 = M_2 \neq \emptyset$, where $M_i$ is defined as the multiset of POS tags of all the open class word nodes of $t_i$.
-
-Applied as a backup for label matchin, this criterion can be used to capture correspondences that would otherwise be missed, thus increasing recall, but a decrease in precision is also to be expected.
-However, since alignment criteria are defined as boolean functions, it is easy to combine them so to apply them simultaneously: combining these first two criteria, for instance, is useful in cases where precision is more important than recall.
+Applied as a backup for label matching, this criterion can be used to capture correspondences that would otherwise be missed, thus increasing recall, but a decrease in precision is also to be expected.
+However, since alignment criteria are defined as boolean functions, it is easy to combine them so to that they have to apply simultaneously.
+This can be useful in cases where precision is more important than recall.
 
@@
 When this is the case, it is often straightforward to define alignment criteria based on recognizing the corresponding patterns.
-If many distinctions of this kind are specific to particular language pairs or even stylistical, some of them occur independently of what languages are involved and have nothing to do with idiomatic usage or aspectual, discourse, domain or word knowledge.
-Drawing inspiration from \cite{dorr-1994-machine}, we refer to them as \textit{translation divergences} and handle the most common ones via alignment criteria that are used by default.
+If many distinctions of this kind are specific to particular language pairs or even stylistical, some of them occur independently of what languages are involved and do not depend on idiomatic usage nor aspectual, discourse, domain or word knowledge.
+Drawing inspiration from \cite{dorr-1994-machine}, we refer to them as \textit{translation divergences} and handle some of the most common ones explicitly.
 \todo{add a few examples (<= one per type of divergence)}
@@
 To handle these situations, the CE module can filter the results based on a \texttt{gf-ud} pattern.
-Pattern matching in \texttt{gf-ud}...\todo{explain briefly how it works. Maybe Aarne can do that?}
+\texttt{gf-ud} supports in fact both simple pattern matching, which is integrated in the CE module itself, and pattern replacement\footnote{documentation is available at \url{github.com/GrammaticalFramework/gf-ud}}.
+Combining them, for instance by pruning the UD trees extracted by the alignment module, allows us even to extract correspondences that, given the structure of UD trees, cannot be identified by CE alone.
 
-Combining pattern matching with pattern replacement, for instance by pruning the UD trees extracted by the alignment module, allows us even to extract correspondences that, given the structure of UD trees, cannot be identified by CE alone.
-To give an example that turned out to be useful in practice, a possibility is to extract verb phrases only, by looking for full clauses and pruning the subtree corresponding to the subject. 
-\todo{rephrase explanation and reformat example}More advanced replacements are also possible: instead of simply pruning clauses to extract the verb phrases they contain, we can perform some actual replacement to obtain \textit{predication patterns}, such as
+To give a practically useful example, a possibility is to extract verb phrases only, by looking for full clauses and pruning the subtree corresponding to the subject. 
+More advanced replacements are also possible: instead of simply removing subjects, we can perform some actual replacement to obtain \textit{predication patterns}, i.e.
+correspondences such as the following English-Italian one:
 
-$$\langle X tells Y Z, X dice Z a Y\rangle$$.
-
-% TODO: reformulate and add possibly graphical example
+\smallskip
+$\langle$X \textit{gives} Y Z, X \textit{dà} Z \textit{a} Y$\rangle$.
 
@@
 
-\begin{itemize}
+\begin{itemize} \setlength\itemsep{0.01em}
   \item the number of correct alignments the program is able to extract
@@
 Compared to the existing statistical tools, our system has the following advantages:
-\begin{itemize}
+\begin{itemize} \setlength\itemsep{0.1em}
   \item it performs consistently well even on extremely small parallel corpora

### 2021-05-15 paper/paper.tex
shortened to 8 pages

@@
 In addition, lexica generated in this way generally consist in simple word form correspondences that cannot be directly used in MT pipelines.
-Addressing these problems, we propose a data-driven, syntax-based approach to this task and put it to the test in a simple grammar-based translation pipeline.
+Addressing these problems, we propose a syntax-based approach to this task and put it to the test in a simple grammar-based translation pipeline.
 \end{abstract}
@@
 The CE module described so far outputs pairs of UD trees. 
-While converting them to GF ASTs is \texttt{gf-ud}'s core functionality, to build a compilable GF translation lexicon it is necessary to turn such trees into grammar rules.
+While converting them to GF ASTs is one of \texttt{gf-ud}'s core functionalities, building a compilable GF translation lexicon is necessary to turn such trees into grammar rules.
 
 In order to do that, our grammar generation module requires a morphological dictionary of the languages at hand and an \textit{extraction grammar}. 
-What we refer to as an extraction grammar is of course a GF grammar, whose objective is to define the set of basic categories and syntactic rules the entries of the automatically generated lexicon is going to use.
+What we refer to as an extraction grammar is of course a GF grammar, whose objective is to define the set of basic categories and syntactic rules the entries of the automatically generated lexicon are going to use.
 \todo{not sure how important this section is and what (if anything) to add to it}
-% TODO: designed for the grammar generation step to produce grammar rules corresponding to simple concepts such as basic categories, noun phrases, verb phrases etc.,
 
@@
 After that, we describe our experiments, aimed at putting both the CE module \textit{per se} and the entire system from parsing to lexicon generation to the test, and present our results.
-% TODO: move around things to have something here (or outline structure of the section)
+
 \subsection{Data} \label{data}
-Because we want part of our evaluation to be independent from the quality of UD parsing, some of the experiments are carried out using manually annotated data. 
-To this purpose, we use a 100-sentences subset of the Parallel UD (PUD) corpus, a set of 1000 handcrafted\footnote{Annotation is sometimes semi-automatic, but all treebanks are at least manually checked.}\todo{not even, I fear.
-Check this} multilingual treebanks in CoNLL-U format created for the CoNLL 2018 shared task on Multilingual Parsing from Raw Text to Universal Dependencies \cite{zeman-etal-2018-conll}.
-PUD treebanks are available in over 20 languages, of which we selected English, Italian and Swedish. 
+Because we want part of our evaluation to be independent from the quality of UD parsing, some of the experiments are carried out on treebanks instead of raw text. 
+To this purpose, we use a 100-sentence subset of the Parallel UD (PUD) corpus, a set of 1000 manually annotated or validated sentences in CoNLL-U format.
+Of the over 20 languages PUD treebanks are available in, we selected English, Italian and Swedish. 
 Using this data limits the amount of alignment errors that are due annotation issues to a minimum, even though a small number of inconsistencies and imprecisions is present even in this corpus.
@@
 When it comes to testing the program on raw text, we use two small ($< 1000$ sentences) bilingual sentence-aligned corpora consisting of course plans from the Department of Mathematics and Computer Science (DMI) of the University of Perugia (for English-Italian) and from the Department of Computer Science and Engineering (CSE) shared between the University of Gothenburg and the Chalmers University of Technology (for English-Swedish).
-For brevity, we will refer to these two datasets as to the DMI and the CSE corpora, or collectively as to the course plans datasets.
+For brevity, we will refer to these two datasets as to the DMI and the CSE corpora.
 This data, available in the project repository, was collected and sentence-aligned specifically for this work and a related Bachelor's thesis project \cite{thesis}.
@@
 
-% TODO: size
+\subsection{Evaluating CE} \label{ceval}
+While we focus mostly on the MT applications of CA, automatic translation, and much less GF-based domain-specific tranlation, is not the only context in which CA can be put to use. 
+For instance, it is easy to imagine using it to improve the UI of online translation memories or to facilitate reading parallel texts in the context of language learning.
+For this reason, a first set of experiments is aimed at evaluating the alignments obtained with our CA module independently from the other stages stages of our proposed lexicon generation pipeline.
 
-\subsection{Evaluating CE} \label{ceval}
-While in this work we focus mostly on the MT applications of CA, automatic translation, and much less GF-based domain-specific tranlation, is not the only context in which our system can be put to use.
-For instance, it is easy to imagine using it to improve the UI of online translation memories or to facilitate reading parallel texts in the context of language learning.
-For this reason, a first set of experiments is aimed at evaluating the alignments obtained with our CA module independently from the subsequent stages of our proposed lexicon generation pipeline.
+We first assess the correctness of the alignments the CE module is able to extract from the PUD treebanks, comparing our results with those obtained with a standard statistical tool, \texttt{fast\_align} \cite{dyer-etal-2013-simple}.
+In addition, we try to quantify the impact automatic parsing has on the performance of the CE module by comparing the above results with those obtained on the DMI and CSE corpora.
 
-We first assess the correctness of the alignments the CE module is able to extract from manually annotated data from the PUD treebanks, comparing our results with those obtained with a standard statistical tool, \texttt{fast\_align} \cite{dyer-etal-2013-simple}.
-In addition, we try to quantify the impact automatic parsing has on the performance of the CE module by comparing the above results wiht those obtained on the DMI and CSE corpora.
+For measuring the performance of our system, while precision and recall are two well-known metrics that would be well suited to the task, the lack of a gold standard for CE imposes us to approximate them with the number of correct alignments the program is able to extract and the ratio between such number and the total number of alignments extracted.
 
-For measuring the performance of our system, while precision and recall are two well-known metrics that would be very well suited to the task, the lack of a gold standard for CE imposes us to approximate them with respectively:
-
-\begin{itemize} \setlength\itemsep{0.01em}
-  \item the number of correct alignments the program is able to extract
-  \item the ratio between such number and the total number of alignments extracted.
- \end{itemize}
-
-However, since some alignments are only correct in the specific context of the sentence pair in which they occur, we make a further distinction between correct alignments that are relevant for a translation lexicon, marked as \texttt{+} in the following tables, and alignments that are useful for comparing the sentences but should not be used for MT, marked as \texttt{=} instead.
-\todo{add example (e.g.
-boat/train)} % TODO: boat/train example
-Incorrect correspondences are marked as \texttt{-}.
+However, since some alignments are only correct in the specific context of the sentence pair in which they occur, we make a further distinction between correct alignments that are relevant for a translation lexicon and alignments that are useful for comparing the sentences but should not be used for MT.
 
 \subsubsection{Results on manually annotated treebanks} 
-In Table \ref{pud_fast}\todo{position properly; run and add en-sv}, we compare the results obtained with our grammar-based module to those obtained statistically on the PUD treebanks.
+In Table \ref{pud_fast}, we compare the results obtained with our grammar-based module to those obtained statistically on the PUD treebanks.
 
@@
 \subsubsection{Results on raw text} \label{raw}
-The course plan corpora are significantly harder to work with, both for our CE module and \texttt{fast\_align}.
-In the former case, the additional challenge is relying on the quality of automatic parsing, while \texttt{fast\_align} suffers from the small size of the datasets, which are fed in their entirety to both systems.
+The course plan corpora are significantly harder to work with, the additional challenge being that our CE module relies on the quality of automatic parsing.
 
-The results of this comparison are summarized and compared with those obtained on the sample of PUD treebanks in Table \ref{raw_fast}.
+A comparison between the results obtained on manually annotated data and the course plans corpora parsed with UDPipe is summarized Table \ref{raw_fast}.
 
@@
 What is immediately evident, but not unexpected, is a decrease in precision. 
-The percentage of correct alignments, however, stays significantly higher than that obtained with \texttt{fast\_align} in the previous experiment, even when the model was trained on the full PUD corpus: even though percentages seem similar for English-Swedish, it must be taken into account that the CSE corpus is roughly half the size of the PUD corpus.
+The percentage of correct alignments, however, stays significantly higher than that obtained with \texttt{fast\_align} in the previous experiment, even with the model trained on the full PUD corpus. 
+In fact, even though percentages seem similar for English-Swedish, the CSE corpus is roughly half the size of the full PUD corpus.
 
@@
 The second set of experiments has the objective of assessing the quality of the final output of the system we propose: GF translation lexica. 
-Because we are focussing on using CA in the context of domain-specific MT, in this case we do not make use of the PUD treebanks, where sentences come from a variety of different sources, but rather of the two course plans corpora.
-For this evaluation, we do not construct a grammar specific to the course plans domain: it is sufficient to extend the extraction grammar itself with preexisting syntax rules defined in the RGL, so to allow for more variation.
+Because we are now focussing on using CA in the context of domain-specific MT, we do not make use of the PUD treebanks, where sentences come from a variety of different sources, but just of the course plans corpora.
+We do not construct a grammar specific to such domain: for small-scale MT experiments, it is sufficient to extend the extraction grammar itself with preexisting syntax rules defined in the RGL.
 
@@
 
-Due to the small size of the datasets and, as a consequence, the low coverage of the extracted lexicon, we generate the sentences to translate directly in the GF shell, making use of its random AST generation functionality but at the same time filtering out semantically implausible sentences to facilitate the task of the human translators.
-The results of this process are two small testing corpora, each consisting of 50 sentences in English, generated using the abstract syntax and the English concrete syntax of the DMI and the CSE grammar respectively.
+Due to the small size of the datasets and the consequent low coverage of the extracted lexicon, we generate the sentences to translate directly in the GF shell rather than trying to parse arbitrary sentences other course plans.
+In order to do that, we make use of GF's random AST generation functionality but at the same time manually select semantically plausible sentences to facilitate the task of the human translators.
+The results of this process are two small testing corpora, one for the DMI and one for the CSE corpus, each consisting of 50 English sentences.
 
-Reference translations are obtained by asking two native speakers of Italian and Swedish to compare the original English sentences to their automatically translated counterparts and correct the latter.
-Participants were instructed to only make the minimal changes necessary to obtain, starting from the automatic tranlsations generated by our system, a set of grammatically and semantically correct translations.
+Reference translations are obtained by asking the participants to compare the original English sentences to their automatically translated counterparts and correct the latter by making the minimal changes necessary to obtain a set of grammatically and semantically correct translations.
 This is important to get meaningful results, as if the reference translations are obtained independently from the automatic ones, BLEU scores can easily become misleading.
@@
 \subsubsection{Experimental results} 
-Corpus-level BLEU scores for the automatic translations of the 50+50 sentences of the testing corpora generated as described in the above are summarized in Table \ref{tableu}.
+Corpus-level BLEU scores for the automatic translations of the 50+50 sentences of the testing corpora are summarized in Table \ref{tableu}.
 
@@
 This is the case, for instance, of the sentence ``\textit{the test is oral}'', whose last word, ``\textit{oral}'' is translated as ``\textit{dura}'' (``\textit{hard}'') instead of ``\textit{orale}'' due to an alignment error.
-Nonetheless, it is worth noticing that this is one of the many cases in which the correct alignment $\langle$``\textit{oral}'',``\textit{orale}''$\rangle$ is also identified by the CE module, but hidden by the fact that, for simplicity, we only take into account the first translation candidate.
+Nonetheless, it is worth noticing that this is one of the many cases in which the correct alignment $\langle$``\textit{oral}'',``\textit{orale}''$\rangle$ is also identified by the CE module, but discared as, for simplicity, we only take the first translation candidate into account.
 This suggest that, in contexts where higher precision is necessary, there are simple improvements that can be done either in the choice of the alignment criteria or at a later stage, in terms alignment selection.
-In other contexts, it can be more practical to aim for higher coverage and not perform any automatic selection, leaving lexicon refinements to a human, possibly with the help of the postprocessing tools that are currently under development.
+In other contexts, it can be more practical to aim for higher coverage and not perform any automatic selection, leaving refinements to a human, possibly with the help of postprocessing tools.
 
@@
 
-Interestingly, while most errors are due to wrong alignments, the main difference between two corpora lies in the number of translations that only contain grammatical errors, which explains the significant difference observed in the cumulative BLEU scores shown in Table \ref{tableu}.
+Interestingly, while most errors are in fact due to wrong alignments, the main difference between two corpora lies in the number of translations that only contain grammatical errors, which explains the significant difference observed in the cumulative BLEU scores shown in Table \ref{tableu}.
 
-In Italian, grammar errors often involve contractions such as ``\textit{del}'' (``\textit{di}'' + ``\textit{il}'', in English ``\textit{of the}''), some of which are systematically rendered as two separate words due to UDPipe tokenization.
-Another common case is that of wrong (or at least very confusing) adjective collocation, such as in the translation ``\textit{il libro presenta una tecnica con miglioramenti vari utile}'' (``\textit{the textbook presents a useful technique with various improvements}''), where the adjective ``\textit{utile}'' (``\textit{useful}''), referred to ``\textit{technique}'' (``\textit{tecnica}'') is placed far from such noun, making the sentence hard to interpret\footnote{The manually postprocessed translation is ``\textit{il libro presenta una utile tecnica con miglioramenti vari}''.}.
-Grammatical errors in Swedish are less common and, apart from long adjectives never being turned to the comparative degree periphrastically (e.g.
-``\textit{relevant\underline{are}}'' instead of ``\textit{mer relevant}''), less systematic.
+Among other things, Italian contractions are often handled incorrectly.
+For example, ``\textit{del}'' (``\textit{di}'' + ``\textit{il}'', in English ``\textit{of the}'') is systematically rendered as two separate words due to UDPipe tokenization.
+Grammatical errors in Swedish are less common and less systematic. 
 Only in one case, for instance, gender is incorrect (``\textit{programbibliotek\underline{en}}''). 
-All of these grammatical errors are easy to handle when writing a domain-specific grammar or, for instance in the case of misgendered nouns, by making small adjustments to the morphological dictionaries.
+These errors are easy to handle when writing a domain-specific grammar or, in cases like the latter, by making small adjustments to the morphological dictionaries.
 
-Some errors regarding the extracted concepts are also interesting to analyze.
-In English-Swedish, while several compounds, such as $\langle$\textit{``computer science'', ``datavetenskap''}$\rangle$, are aligned correctly, there are cases in which a single-root noun is translated as a compound (e.g.
-$\langle$ \textit{``theory'', ``automatateori''} $\rangle$).
-This is not necessarily an actual alignment error, as it might well be the case that the English version of the text was being less specific than its Swedish counterpart, thus producing an alignment that, during evaluation, would have been marked as ``correct but not relevant for a translation lexicon'' (\texttt{=}).
-``Reasonable'' alignment errors appear in the DMI corpus sentences too. 
-The English phrase ``\textit{attendance of lessons}'', for instance, becomes simply ``\textit{frequenza}'' in the Italian translation. 
-Such word is in fact often used alone to replace longer expressions such as ``\textit{frequenza delle lezioni}'', just like in English ``\textit{of lessons}'' can be omitted when what must be attended is evident from the context.
-Another interesting example is the alignment $\langle$\textit{``class'', ``classe''}$\rangle$, which causes the sentence \textit{``I will attend the class''} to be (incorrectly) translated as \textit{``io seguirò la classe''} instead of \textit{``io seguirò la lezione''} even though the correspondence is in fact valid in most of the numerous contexts in which (within the same domain!) \textit{``class''} is not to be intended as a synonym of \textit{``lesson''}.
+Some errors regarding the extracted concepts are also interesting to analyze: the alignment $\langle$\textit{``class'', ``classe''}$\rangle$, for instance, causes the sentence \textit{``I will attend the class''} to be (incorrectly) translated as \textit{``io seguirò la classe''} instead of \textit{``io seguirò la lezione''} even though the correspondence is in fact valid in most of the numerous contexts in which (within the same domain!) \textit{``class''} is not to be intended as a synonym of \textit{``lesson''}.
 
@@
   \item it can automatically generate compilable, morphology-aware GF translation lexica
-  \item it can be configured to easily handle systmatic language-pair specific translation divergences.
+  \item it can be configured to easily handle systematic, possibly language-pair or corpus specific translation divergences.
 \end{itemize}
 
-The tangible fruit of this work are a Haskell library and a number of executables offering an easy to use and configure interface to perform CE, lexicon generation and a variety of kinds of evaluations.
-The source code, which also implements a preliminary implementation of CP not discussed in depth in this paper, is available at \url{github.com/harisont/concept-alignment}\todo{maybe I should mention this in the introduction?}.
+The tangible fruits of this work are a Haskell library and a number of executables offering an easy to use and configure interface to perform CE, lexicon generation and a variety of kinds of evaluations.
+The source code, which also implements a preliminary implementation of CP not discussed in depth in this paper, is available at \url{github.com/harisont/concept-alignment}
 
 \subsection{Current and future work}
-Our results, while encouraging, suggest that there is still much room for improvement in many different directions.
+Our results, while encouraging, suggest that there is room for improvement in many different directions.
 
@@
 For cases where a relatively large amount of data is available, using our system in conjunction with a statistical tool seems promising. 
-As discussed in Section \ref{ka}, some integration between our system and other sources of alignments is already supported, and it is possible to keep track of the reason why a particular correspondence was extracted.
-This could be further developed in the direction of an actual hybrid system. 
+As discussed in Section \ref{ka}, some integration between our system and other sources of alignments is already supported and could be further developed in the direction of an actual hybrid system. 
 A first attempt in this direction is the development of a simple conversion tool between the ``pharaoh format'' used by tools such as \texttt{fast\_align} and the format we use to represent alignments, available in the same repository.
 
-Finally, since the freedom that generally characterizes human translation and the quality of currently available UD parsers make the expectation of maximizing both alignment precision and recall unrealistic, it seems logical to develop tools to make it easier to postprocess automatically generated lexica.
-With respects to this, current work is represented by a synoptic CoNLL-U viewer, available at \url{github.com/DigitalGrammarsAB/synoptic-conllu-viewer}, which makes it possible to inspect the UD trees a grammar rule was derived from.
-\todo{remove this part if the tool is not made public, or maybe regardless}
+Finally, since the freedom that generally characterizes human translation and the quality of currently available UD parsers make maximizing both alignment precision and recall unrealistic, tools to make it easier to postprocess the automatically generated lexica are under development.
 
-% TODO: see why it does not compile
-% Entries for the entire Anthology, followed by custom entries
 \bibliography{anthology,custom}  

### 2021-05-15 paper/paper.tex
painstakingly shortened to 7 pages

@@
 In grammar-based domain-specific machine translation, constructing high-quality translation lexica is a crucial but time consuming task requiring significant linguistic knowledge. 
-When example parallel sentences are available, statistical alignment tools allow for the automation of part of the process, but are not suitable for small datasets and are limited when it comes to discontinuous multiword correspondences.
-In addition, lexica generated in this way generally consist in simple word form correspondences that cannot be directly used in MT pipelines.
-Addressing these problems, we propose a syntax-based approach to this task and put it to the test in a simple grammar-based translation pipeline.
+With parallel example sentences available, statistical alignment tools can help automate part of the process, but they are not suitable for small datasets and limited when it comes to complex multiword expressions.
+In addition, mere word form correspondences obtained in this way cannot be used directly.
+Addressing these problems, we propose a syntax-based approach to this task and put it to the test in a simple translation pipeline.
 \end{abstract}
@@
 
-Finally, performing morphological analysis allows us to generate translation lexica consisting of complete with grammatical category and inflection information instead of simple correspondences between word forms.
+Finally, performing morphological analysis allows us to generate translation lexica complete with grammatical category and inflection information instead of simple correspondences between word forms.
 
@@
 An interesting and less studied variant of CA is \textit{Concept Propagation} (CP), useful for cases where a set of concepts is already known and the goal is to identify the expressions corresponding to each of them in a new language, potentially even working with a different text in the same domain.
-While our system implements basic CP functionalities, however, in this paper we focus on CE and restrict its application to bilingual corpora.
+While our system does implement basic CP functionalities, however, in this paper we focus on CE and restrict its application to bilingual corpora.
 
-As stated in the Introduction, most existing extraction solutions are based on statistical approaches and, as a consequence, unsuitable for small datasets. 
+As stated in the Introduction, most existing extraction solutions are based on statistical approaches and are, as a consequence, unsuitable for small datasets. 
 Grammar-based approaches, making use of parallel treebanks and collectively referred to as \textit{tree-to-tree alignment methods}, have also been proposed \cite{tiedemann2011bitext}, but have historically suffered from the inconsistencies between the formalisms used to define the grammars for the various languages and from the lack of robustness of existing parsers.
@@
 UD, on the other hand, is a dependency grammar formalism meant for cross-linguistically consistent grammatical annotation.
-As opposed to constituency, \textit{dependency} is a word-to-word correspondence: each word is simply put in relation with the one it refers to, called its \textit{head}, via a directed labelled link specifying the syntactic relation between them.
+As opposed to constituency, \textit{dependency} is a word-to-word correspondence: each word is put in relation with the one it refers to, called its \textit{head}, via a directed labelled link specifying the syntactic relation between them.
 Importantly for our application, the standard format for UD trees, CoNNL-U, gives information not only on the syntactic role of each word, but also on its POS tag, lemma and morphological features.
@@
 While both formalisms independently solve the issues related to having to work with grammars that are inconsistent with each other, UD is especially appealing since, dependency trees being an easier target, several robust parsers, such as \cite{straka-etal-2016-udpipe} and \cite{chen-manning-2014-fast} are available.
+Alone, UD trees are sufficient to extract (or propagate) tree-to-tree alignments, but not to automate the generation of a morphologically-aware translation lexicon for a generative grammar. 
+This is where GF comes into play: after correspondences are inferred from a parallel text, our system is able to convert them to GF grammar rules, easy to embed in a domain-specific grammar but also making it immediate to carry out small-scale translation experiments using pre-existing grammatical constructions implemented in GF's Resource Grammar Library (RGL), which covers the morphology and basic syntax of over 30 languages.
+This is enabled by \texttt{gf-ud}, a conversion tool described in \cite{kolachina-ranta-2016-abstract} and \cite{ranta-kolachina-2017-universal}.
 
-UD parsing alone is sufficient to extract (or propagate) tree-to-tree alignments, but not to automate the generation of a morphologically-aware translation lexicon for a generative grammar. 
-This is where GF comes into play: after correspondences are inferred from a parallel text, our system is able to convert them to GF grammar rules.
-These are easy to embed in a domain-specific grammar but also making it possible to immediately evaluate the system by carrying out small-scale translation experiments using pre-existing grammatical constructions implemented in GF's Resource Grammar Library (RGL), which covers the morphology and basic syntax of over 30 languages.
-This is made possible by \texttt{gf-ud}, a conversion tool described in \cite{kolachina-ranta-2016-abstract} and \cite{ranta-kolachina-2017-universal}.
-
-Concretely, the system we propose consists of the following elements:
-
-\begin{itemize} \setlength\itemsep{0.01em}
-  \item a UD parser
-  \item an alignment module based on dependency tree comparison
-  \item a program, based on \texttt{gf-ud}, that converts the alignments into GF grammar rules and uses them to construct a translation lexicon.
-\end{itemize}
+Concretely, then, the system we propose consists of a UD parser, an alignment module based on UD tree comparison and a program, based on \texttt{gf-ud}, that converts them into the rules of a GF translation lexicon.
 
@@
 The core part of the system outlined above is of course the alignment module. 
-In this section, we present our method for extracting alignments from parallel bilingual UD treebanks, which can be obtained from sentence-aligned texts with any UD parser.
+In this section, we present our method for extracting alignments from parallel bilingual UD treebanks.
 
@@
 rules to determine whether two dependency trees should be aligned with each other which will be discussed in greater detail in Section \ref{criteria}, and a pair of sentences to align represented as UD trees.
-From an implementation point of view, UD trees are, as in \texttt{gf-ud}, rose trees where each node represents a token, obtained from the CoNLL-U files produced by the UD parser.
+From an implementation point of view, UD trees are, rose trees where each node represents a token, obtained from the CoNLL-U files produced by the UD parser.
 
 \todo{add figure with the various tree representations?}
-% TODO: add figure comparing tree representations
 
@@
 
-When working on corpora consisting of multiple sentences, this algorithm can be applied in an iterative fashion, so that knowledge gathered when a sentence pair is aligned can be reused when working on the following ones and to keep track of the number of occurrences of each alignment throughout the entire text.
+When working on multiple sentences, this algorithm can be applied in an iterative fashion, so that knowledge gathered when a sentence pair is aligned can be reused when working on the following ones and to keep track of the number of occurrences of each alignment throughout the entire text.
 
@@
 attached to aligned heads, constitute an alignment if their roots share the same dependency label, meaning that they are in the same syntactic relation with their heads.
-
 Note that, since the root of UD trees is always attached to a a fake node with an arc labelled \texttt{root}, this criterion also makes it so that full sentences are always considered to align with each other, something which is desirable since we assume the parallel texts that are fed to our program to be sentence-aligned.
@@
 \texttt{gf-ud} supports in fact both simple pattern matching, which is integrated in the CE module itself, and pattern replacement\footnote{documentation is available at \url{github.com/GrammaticalFramework/gf-ud}}.
-Combining them, for instance by pruning the UD trees extracted by the alignment module, allows us even to extract correspondences that, given the structure of UD trees, cannot be identified by CE alone.
+Combining them, for instance by pruning the UD trees extracted by the alignment module, allows us to extract even correspondences that, given the structure of UD trees, cannot be identified by CE alone.
 
-To give a practically useful example, a possibility is to extract verb phrases only, by looking for full clauses and pruning the subtree corresponding to the subject. 
-More advanced replacements are also possible: instead of simply removing subjects, we can perform some actual replacement to obtain \textit{predication patterns}, i.e.
-correspondences such as the following English-Italian one:
+To give a practically useful example, a possibility is to extract verb phrases by looking for full clauses and pruning the subtrees corresponding to subjects. 
+It is also possible to perform some actual replacement to obtain \textit{predication patterns}, i.e.
+correspondences that specify the argument structure of verbs, such as the following English-Italian one:
 
@@
 While we focus mostly on the MT applications of CA, automatic translation, and much less GF-based domain-specific tranlation, is not the only context in which CA can be put to use. 
-For instance, it is easy to imagine using it to improve the UI of online translation memories or to facilitate reading parallel texts in the context of language learning.
-For this reason, a first set of experiments is aimed at evaluating the alignments obtained with our CA module independently from the other stages stages of our proposed lexicon generation pipeline.
+For instance, it is easy to imagine using it to improve the UI of online translation memories.
+For this reason, a first set of experiments is aimed at evaluating the alignments obtained with our CA module independently from the other stages stages of our lexicon generation pipeline.
 
@@
 
-For measuring the performance of our system, while precision and recall are two well-known metrics that would be well suited to the task, the lack of a gold standard for CE imposes us to approximate them with the number of correct alignments the program is able to extract and the ratio between such number and the total number of alignments extracted.
-
+While precision and recall are two well-known performance metrics, the lack of a gold standard for CE imposes us to approximate them with the number of correct alignments the program is able to extract and the ratio between such number and the total number of alignments extracted.
 However, since some alignments are only correct in the specific context of the sentence pair in which they occur, we make a further distinction between correct alignments that are relevant for a translation lexicon and alignments that are useful for comparing the sentences but should not be used for MT.
@@
   \centering
+  \small
   \begin{tabular}{r|c|c|c|c|c|c}
@@
   \textbf{distinct alignments}        & 536             & 638            & 1242                                        & 1044                                        & 1286                                          & 1065          \\
-  \textbf{correct (\texttt{+} and \texttt{=})}                & 392 (73\%)      & 514 (80\%)     & 346 (28\%)                                  & 538 (52\%)                                  & 540 (42\%)                                    & 677 (64\%)    \\
-  \textbf{usable in MT (\texttt{+})}                      & 363 (68\%)      & 503 (79\%)     & 316 (25\%)                                  & 525 (50\%)                                  & 510 (40\%)                                    & 666 (63\%)    \\
+  \textbf{correct}                & 392 (73\%)      & 514 (80\%)     & 346 (28\%)                                  & 538 (52\%)                                  & 540 (42\%)                                    & 677 (64\%)    \\
+  \textbf{usable in MT}                      & 363 (68\%)      & 503 (79\%)     & 316 (25\%)                                  & 525 (50\%)                                  & 510 (40\%)                                    & 666 (63\%)    \\
   \end{tabular}
-  \caption[Word alignment comparison between our grammar-based CE module and \texttt{fast\_align}]{Comparison between our grammar-based CE module and \texttt{fast\_align} on PUD data.
-The middle column shows the results obtained by training \texttt{fast\_align} on the 100-sentence subset also fed to the extraction module.
-On the right, the results obtained by training it on the full dataset, discarding the alignments obtained for sentences 101-1000 for the final comparison.}
+  \caption[Word alignment comparison between our grammar-based CE module and \texttt{fast\_align}]{Comparison between our grammar-based CE module and \texttt{fast\_align} on PUD data training the model on 100 and 1000 sentences and discarding the alignments obtained for sentences 101-1000 in the latter case.}
   \label{pud_fast}
@@
   \centering
+  \small
   \begin{tabular}{r|c|c|c|c}
@@
   \multicolumn{1}{l|}{}          & \textbf{en-it}                               & \textbf{en-sv}                               & \textbf{\begin{tabular}[c]{@{}c@{}}DMI (881 sentences)\end{tabular}} & \textbf{\begin{tabular}[c]{@{}c@{}}CSE (539 sentences)\end{tabular}} \\ \hline
-  \textbf{disrinct alignments}        & 1197                                         & 1325                                         & 1823                                                                           & 1950                                                                           \\
-  \textbf{correct (\texttt{+} and \texttt{=})}                & 916 (77\%)                                   & 1112 (85\%)                                  & 1205 (66\%)                                                                    & 1296 (66\%)                                                                    \\
-  \textbf{usable in MT (\texttt{+}})                      & 880 (74\%)                                   & 1099 (84\%)                                  & 1157 (63\%)                                                                    & 1248 (64\%)                                                                    \\
+  \textbf{distinct alignments}        & 1197                                         & 1325                                         & 1823                                                                           & 1950                                                                           \\
+  \textbf{correct}                & 916 (77\%)                                   & 1112 (85\%)                                  & 1205 (66\%)                                                                    & 1296 (66\%)                                                                    \\
+  \textbf{usable in MT}                      & 880 (74\%)                                   & 1099 (84\%)                                  & 1157 (63\%)                                                                    & 1248 (64\%)                                                                    \\
   \end{tabular}
@@
 
-Reference translations are obtained by asking the participants to compare the original English sentences to their automatically translated counterparts and correct the latter by making the minimal changes necessary to obtain a set of grammatically and semantically correct translations.
-This is important to get meaningful results, as if the reference translations are obtained independently from the automatic ones, BLEU scores can easily become misleading.
+Reference translations are obtained by asking participants to compare the original English sentences to their automatically translated counterparts and correct the latter by making the minimal changes necessary to obtain a set of grammatically and semantically correct translations.
+This is important as, if the reference translations are obtained independently from the automatic ones, BLEU scores can easily become misleading.
 
-\subsubsection{Experimental results} 
+\subsubsection{Results} 
 Corpus-level BLEU scores for the automatic translations of the 50+50 sentences of the testing corpora are summarized in Table \ref{tableu}.
@@
   \centering
+  \small
   \begin{tabular}{r|cc}
@@
 This is the case, for instance, of the sentence ``\textit{the test is oral}'', whose last word, ``\textit{oral}'' is translated as ``\textit{dura}'' (``\textit{hard}'') instead of ``\textit{orale}'' due to an alignment error.
-Nonetheless, it is worth noticing that this is one of the many cases in which the correct alignment $\langle$``\textit{oral}'',``\textit{orale}''$\rangle$ is also identified by the CE module, but discared as, for simplicity, we only take the first translation candidate into account.
-This suggest that, in contexts where higher precision is necessary, there are simple improvements that can be done either in the choice of the alignment criteria or at a later stage, in terms alignment selection.
-In other contexts, it can be more practical to aim for higher coverage and not perform any automatic selection, leaving refinements to a human, possibly with the help of postprocessing tools.
+Interestingly, this is one of many cases in which the correct alignment $\langle$``\textit{oral}'',``\textit{orale}''$\rangle$ is also identified by the CE module, but discared as we only take the first translation candidate into account.
 
-Furthermore, \todo{from here on nothing is absolutely vital in this section.
-It can be shortened or removed if necessary to make room for something else}a problem with using the BLEU score as the only evaluation metric is the fact that it makes no distinction between content and function words, thus not allowing an evaluation focused specifically on the extracted concepts.
+Furthermore, a problem with using the BLEU score as the only evaluation metric is the fact that it makes no distinction between content and function words, thus not allowing an evaluation focused specifically on the extracted concepts.
 The small size of the corpus, however, allows for some error analysis. 
-While postprocessing the automatic translations, the participants were asked to indicate what kind of errors they encountered in each sentence (grammatical, semantical or both). 
-Their observations are summarized in Table \ref{obs}.
+From the participants's observations about the kind of errors encountered when manually editing the automatic translation, summarizes in Table \ref{obs}, we can conclude that while most errors are in fact due to wrong alignments, the main difference between two corpora lies in the number of translations that only contain grammatical errors.
+This explains the significant difference observed in the cumulative BLEU scores shown in Table \ref{tableu}.
 
@@
 \end{table}
-
-Interestingly, while most errors are in fact due to wrong alignments, the main difference between two corpora lies in the number of translations that only contain grammatical errors, which explains the significant difference observed in the cumulative BLEU scores shown in Table \ref{tableu}.
 
@@
 
-The tangible fruits of this work are a Haskell library and a number of executables offering an easy to use and configure interface to perform CE, lexicon generation and a variety of kinds of evaluations.
-The source code, which also implements a preliminary implementation of CP not discussed in depth in this paper, is available at \url{github.com/harisont/concept-alignment}
+The tangible fruits of this work are a Haskell library and a number of executables offering a user friendly interface to perform CE, lexicon generation and various kinds of evaluations. 
+The source code, including a preliminary implementation of CP, is available at \url{github.com/harisont/concept-alignment}
 
@@
 
-An obvious possible development is to improve the current implementation of CP, optimizing it for its two different use cases: propagating alignments to a new language looking for correspondences in a different text in the same domain and using a translation of the same text they were obtained from.
-\todo{simplify part on CP scenarios}
-An alternative to the latter is to generalize CE to $n$ languages, instead of limiting it to bilingual cases.
+An obvious possible development is to improve the current implementation of CP, optimizing it for its different use cases: propagating alignments to a new language looking for correspondences using a translation of the same text they were extracted from or using a different text in the same domain.
+An alternative to the latter is to make CE, now working on bilingual texts, $n$-lingual.
 
-For cases where a relatively large amount of data is available, using our system in conjunction with a statistical tool seems promising. 
-As discussed in Section \ref{ka}, some integration between our system and other sources of alignments is already supported and could be further developed in the direction of an actual hybrid system. 
-A first attempt in this direction is the development of a simple conversion tool between the ``pharaoh format'' used by tools such as \texttt{fast\_align} and the format we use to represent alignments, available in the same repository.
+When large amounts of data ar available, using our system in conjunction with a statistical tool seems promising. 
+As discussed in Section \ref{ka}, this is already partially supported and it could prove useful to develop CA as an actual hybrid system.
 

### 2021-05-15 paper/paper.tex
added pseudocode + a not completely essential figure

@@
 \usepackage{todonotes}
+
+\usepackage{caption}
+\usepackage{subcaption}
+
+\usepackage{algorithm}
+\usepackage[noend]{algpseudocode}
 
@@
 
-The pseudocode of the basic extraction algorithm is shown in Figure \ref{TODO:}\todo{add pseudocode}.
+The pseudocode of the basic extraction algorithm is shown below.
+
+\begin{algorithm} 
+  \small
+  \begin{algorithmic}
+  \Procedure{extract}{$criteria$,$(t,u)$}
+    \State $alignments$ = $\emptyset$
+    \If{$(t,u)$ matches any alignment $criteria$}
+        \State $alignments$ += $(t,u)$
+        \For{$(t',u')$ in \textsc{sort}(\textsc{subts}($t$)) $\times$ \textsc{sort}(\textsc{subts}($u$))}
+          \State extract($criteria$,$(t',u')$)
+        \EndFor
+    \EndIf
+    \Return $alignments$
+  \EndProcedure
+  \end{algorithmic}
+  \end{algorithm}
+
 The input consists of a list of priority-sorted \textit{alignment criteria}, i.e.
@@
 
-\todo{add figure with the various tree representations?}
+\begin{figure*}[h]
+  \centering
+  \begin{subfigure}{.23\textwidth}
+      \centering
+      \scriptsize
+      \setlength{\unitlength}{0.16mm}
+      \begin{picture}(371.0,130.0)
+          \put(0.0,0.0){this}
+          \put(46.0,0.0){is}
+          \put(83.0,0.0){a}
+          \put(120.0,0.0){dependency}
+          \put(220.0,0.0){tree}
+          \put(0.0,15.0){{\tiny PRON}}
+          \put(46.0,15.0){{\tiny AUX}}
+          \put(83.0,15.0){{\tiny DET}}
+          \put(120.0,15.0){{\tiny NOUN}}
+          \put(220.0,15.0){{\tiny NOUN}}
+          \put(120.0,30.0){\oval(218.63636363636363,133.33333333333334)[t]}
+          \put(10.681818181818187,35.0){\vector(0,-1){5.0}}
+          \put(105.0,99.66666666666667){{\tiny nsubj}}
+          \put(143.0,30.0){\oval(172.27586206896552,100.0)[t]}
+          \put(56.86206896551724,35.0){\vector(0,-1){5.0}}
+          \put(128.0,83.0){{\tiny cop}}
+          \put(161.5,30.0){\oval(134.8102189781022,66.66666666666667)[t]}
+          \put(94.0948905109489,35.0){\vector(0,-1){5.0}}
+          \put(146.5,66.33333333333334){{\tiny det}}
+          \put(180.0,30.0){\oval(97.0,33.333333333333336)[t]}
+          \put(131.5,35.0){\vector(0,-1){5.0}}
+          \put(165.0,49.66666666666667){{\tiny compound}}
+          \put(235.0,130.0){\vector(0,-1){100.0}}
+          \put(240.0,120.0){{\tiny root}}
+      \end{picture}
+  \end{subfigure}
+  \begin{subfigure}{.38\textwidth}
+    \centering
+    \scriptsize
+      \begin{verbatim}
+      1	this	this	PRON	_	_	5	nsubj	_	_
+      2	is	be	AUX	_	_	5	cop	_	_
+      3	a	a	DET	_	_	5	det	_	_
+      4	dependency	dependency	NOUN	_	_	5	
+        compound	_	_
+      5	tree	tree	NOUN	_	_	0	root	_	_
+      \end{verbatim}
+  \end{subfigure}%
+  \begin{subfigure}{.38\textwidth}
+    \centering
+    \scriptsize
+      \begin{verbatim}
+5	tree	tree	NOUN	_	_	0	root	_	_
+  4	dependency	dependency	NOUN	_	_	5	
+    compound	_	_
+  2	is	be	AUX	_	_	5	cop	_	_
+  3	a	a	DET	_	_	5	det	_	_
+  1	this	this	PRON	_	_	5	nsubj	_	_
+      \end{verbatim}
+  \end{subfigure}%
+  \caption[The graphical representation of a UD tree alongside its CoNNL-U and sorted rose tree representation.
+Optional CoNNL-U fields are left blank.]{The graphical representation of a UD tree alongside its CoNNL-U and rose tree representation.
+Optional CoNNL-U fields are left blank.}
+\end{figure*}
 
@@
 
-Depending on which alignment criteria they match, the head of the two trees may or may not also be added to such collection.
-This is important not to miss one-word correspondences that cannot be captured in any other way, for instance between the root verbs of two full sentences.
-
 The same procedure is applied recursively to all pairs of immediate subtrees of each sentence, until the leaves are reached or alignment is no longer possible due to a lack of matching criteria.
@@
 
-% TODO: pseudocode
+A simple but useful refinement is that, depending on which alignment criteria a pair of trees candidate for alignment matches, the heads of the two trees may or may not also be added to the collection of alignments.
+This is important not to miss one-word correspondences that cannot be captured in any other way, for instance between the root verbs of two full sentences.
+A relevant implementative detail is that, in this context, the head of a tree is not simply defined as its root.
+Instead, if the root is part of a compound written as two or more separate words or a verb with auxiliaries, the root nodes of the corresponding children are also considered as part of its head.
 

### 2021-05-15 paper/paper.tex
example & minor changes

@@
 % and set <dim> to something 5cm or larger.
-\title{Syntax-Based Concept Alignment for Domain-Specific Machine Translation}
+\title{Grammar-Based Concept Alignment for Domain-Specific Machine Translation}
 
@@
 \begin{abstract}
-In grammar-based domain-specific machine translation, constructing high-quality translation lexica is a crucial but time consuming task requiring significant linguistic knowledge. 
+In grammar-based domain-specific MT, constructing high-quality translation lexica is a crucial but time consuming task requiring significant linguistic knowledge. 
 With parallel example sentences available, statistical alignment tools can help automate part of the process, but they are not suitable for small datasets and limited when it comes to complex multiword expressions.
-In addition, mere word form correspondences obtained in this way cannot be used directly.
-Addressing these problems, we propose a syntax-based approach to this task and put it to the test in a simple translation pipeline.
+In addition, the correspondences between word forms obtained in this way cannot be used directly.
+Addressing these problems, we propose a grammar-based approach to this task and put it to the test in a simple translation pipeline.
 \end{abstract}
@@
 \section{Introduction}
+% why grammar-based approaches in DSMT
 Grammar-based translation pipelines such as those based on Grammatical Framework (GF) have been successfully employed in domain-specific Machine Translation (MT) \cite{ranta-etal-2020-abstract}.  
@@
 
+% importance of lexica in DSMT
 Nevertheless, lexical exactness is, in this context, just as important as grammaticality. 
 An important part of the design of the Controlled Natural Language (CNL) one such system's grammar describes then becomes the creation of a high-quality translation lexicon.
+
+% how it is done right now and issues
 This is often done manually, resulting in a time consuming task requiring significant linguistic knowledge. 
@@
 
+% proposed solution: a grammar-based approach
 In this paper, we propose an alternative approach to the automation of this task. 
@@
 
+% additional advantage 1: multiwords
 A further advantage of performing syntactic analysis is that we do not have to choose \textit{a priori} whether to focus on the word or phrase level. 
@@
 
+% additional advantage 2: morphology
 Finally, performing morphological analysis allows us to generate translation lexica complete with grammatical category and inflection information instead of simple correspondences between word forms.
 
+% roadmap
 This paper is structured as follows. 
@@
 \section{Methodology} \label{methodology}
+% CA
 The objective of CA is to find semantical correspondences between parts of multilingual parallel texts. 
@@
 
+% CE
 The basic use case for CA, which we refer to specifically as \textit{Concept Extraction} (CE), is the generation of a translation lexicon from a multilingual parallel text.
@@
 
+% CP
 An interesting and less studied variant of CA is \textit{Concept Propagation} (CP), useful for cases where a set of concepts is already known and the goal is to identify the expressions corresponding to each of them in a new language, potentially even working with a different text in the same domain.
@@
 
+% contextualizing our proposal
 As stated in the Introduction, most existing extraction solutions are based on statistical approaches and are, as a consequence, unsuitable for small datasets. 
 Grammar-based approaches, making use of parallel treebanks and collectively referred to as \textit{tree-to-tree alignment methods}, have also been proposed \cite{tiedemann2011bitext}, but have historically suffered from the inconsistencies between the formalisms used to define the grammars for the various languages and from the lack of robustness of existing parsers.
-
 This work is a new attempt in the same direction, enabled by two multilinguality-oriented grammar formalisms developed over the course of the last 25 years: Grammatical Framework (GF) \cite{ranta-etal-2020-abstract} and Universal Dependencies (UD) \cite{ws-2019-universal}.
 
+% GF
 GF is a constituency grammar formalisms and programming language in which grammars are represented as pairs of an \textit{abstract syntax}, playing the role of an interlingua, and a set of \textit{concrete syntaxes} capturing the specificities of the various natural languages.
@@
 
+% UD
 UD, on the other hand, is a dependency grammar formalism meant for cross-linguistically consistent grammatical annotation.
 As opposed to constituency, \textit{dependency} is a word-to-word correspondence: each word is put in relation with the one it refers to, called its \textit{head}, via a directed labelled link specifying the syntactic relation between them.
-Importantly for our application, the standard format for UD trees, CoNNL-U, gives information not only on the syntactic role of each word, but also on its POS tag, lemma and morphological features.
+Importantly for our application, the standard format for UD trees, CoNNL-U, gives information not only on the syntactic role of each word, but also on its Part-Of-Speech (POS) tag, lemma and morphological features.
 
+% role of UD and GF
 While both formalisms independently solve the issues related to having to work with grammars that are inconsistent with each other, UD is especially appealing since, dependency trees being an easier target, several robust parsers, such as \cite{straka-etal-2016-udpipe} and \cite{chen-manning-2014-fast} are available.
@@
 This is enabled by \texttt{gf-ud}, a conversion tool described in \cite{kolachina-ranta-2016-abstract} and \cite{ranta-kolachina-2017-universal}.
-
 Concretely, then, the system we propose consists of a UD parser, an alignment module based on UD tree comparison and a program, based on \texttt{gf-ud}, that converts them into the rules of a GF translation lexicon.
@@
 \subsection{Extracting concepts} 
+% introducing CE algorithm
 The core part of the system outlined above is of course the alignment module. 
 In this section, we present our method for extracting alignments from parallel bilingual UD treebanks.
-
-The pseudocode of the basic extraction algorithm is shown below.
+Our basic algorithm is sketched in th pseudocode below.
 
@@
 
-The input consists of a list of priority-sorted \textit{alignment criteria}, i.e.
+% input
+Here, the input consists of a list of priority-sorted \textit{alignment criteria}, i.e.
 rules to determine whether two dependency trees should be aligned with each other which will be discussed in greater detail in Section \ref{criteria}, and a pair of sentences to align represented as UD trees.
@@
 
+% algorithm description
 As a first step, the program checks whether the two full sentence trees can be aligned with each other, i.e. if they match any of the alignment criteria. 
@@
 Such collection is what the function will return after aligning all the dependency subtrees.
-
 The same procedure is applied recursively to all pairs of immediate subtrees of each sentence, until the leaves are reached or alignment is no longer possible due to a lack of matching criteria.
@@
 
+% head alignment
 A simple but useful refinement is that, depending on which alignment criteria a pair of trees candidate for alignment matches, the heads of the two trees may or may not also be added to the collection of alignments.
@@
 A relevant implementative detail is that, in this context, the head of a tree is not simply defined as its root.
-Instead, if the root is part of a compound written as two or more separate words or a verb with auxiliaries, the root nodes of the corresponding children are also considered as part of its head.
+Instead, if the root is part of a compound written as two or more separate words or a verb with auxiliaries, the root nodes of the corresponding subtrees are also considered part of it.
 
+% iterations
 When working on multiple sentences, this algorithm can be applied in an iterative fashion, so that knowledge gathered when a sentence pair is aligned can be reused when working on the following ones and to keep track of the number of occurrences of each alignment throughout the entire text.
@@
 \subsubsection{Alignment criteria} \label{criteria}
-While alignment criteria are customizable, to better understand the extraction algorithm described in the above, we dedicate this section to describing the ones that the current implementation uses by default.
+% why list criteria 
+While alignment criteria are customizable, to allow for a better understanding of the extraction algorithm described in the above, we describe the ones that the current implementation utilizes by default.
 
@@
 \paragraph{Part-Of-Speech equivalence}
-As noted in the above, CoNNL-U files also provide information on the grammatical categories of each word, represented as Universal Part-Of-Speech (POS) tags \cite{petrov-etal-2012-universal}. 
+% actual POS-equiv par
+As noted in the above, CoNNL-U files also provide information on the grammatical categories of each word, represented as Universal POS tags \cite{petrov-etal-2012-universal}. 
 Intuitively, if the nodes of two trees in matching contexts have the same POS tags, the two trees are more likely to correspond to each other than if not. 
 This is especially true if we focus, for instance, solely on the open class words (defined as in the UD documentation \footnote{\url{universaldependencies.org/u/pos/all.html}}), thus ignoring prepositions, determiners, auxiliary verbs and all other function words, which tend to behave differently across different languages.
-
 A useful relation to define between dependency trees is, then, that of \textit{POS-equivalence}: two dependency trees $t$, $u$ are POS-equivalent if $M_1 = M_2 \neq \emptyset$, where $M_i$ is defined as the multiset of POS tags of all the open class word nodes of $t_i$.
 
+% combining criteria, especially pos-equiv + label matching
 Applied as a backup for label matching, this criterion can be used to capture correspondences that would otherwise be missed, thus increasing recall, but a decrease in precision is also to be expected.
@@
 \paragraph{Known translation divergence}
+% actual divergences par
 Parallel texts often present significant, systematic cross-linguistic grammatical distinctions. 
 When this is the case, it is often straightforward to define alignment criteria based on recognizing the corresponding patterns.
-If many distinctions of this kind are specific to particular language pairs or even stylistical, some of them occur independently of what languages are involved and do not depend on idiomatic usage nor aspectual, discourse, domain or word knowledge.
+While many distinctions of this kind are specific to particular language pairs or even stylistical, some of them occur independently of what languages are involved and do not depend on idiomatic usage nor aspectual, discourse, domain or word knowledge.
 Drawing inspiration from \cite{dorr-1994-machine}, we refer to them as \textit{translation divergences} and handle some of the most common ones explicitly.
-\todo{add a few examples (<= one per type of divergence)}
 
-% TODO: add examle (Herb?)
+% examples
+For instance, \textit{categorial divergences} occur when the POS tag of a word in the source language changes in its translation.
+An ubiquitous example of is that of adverbs rendered as obliques, such as in the English-Italian pair $\langle$\textit{She studies \textbf{consistently}, Lei studia \textbf{con costanza}}$\rangle$.
+\textit{Structural divergences}, where a core argument of a sentence in the source language are rendered as obliques in the source language (as in the English-Swedish pair $\langle$\textit{I told \textbf{him}, Jag berättade \textbf{för honom}.}$\rangle$), are also frequently encountered.
+
+% TODO: add example (Herb?)
 
 \paragraph{Known alignment} \label{ka}
+% actual KA par
 Another case in which it is trivially desirable for two subtrees in matching context to be aligned is when an equivalent alignment is already known, for instance due to a previous iteration of the extraction algorithm.
@@
 
+% motivation
 At a first glance, this might seem a criterion with no practical applications. 
@@
 \subsubsection{Pattern matching}
+% context
 So far, we described how CE can be used in a setting where the objective is to generate a comprehensive translation lexicon based on set of example sentence pairs.
@@
 
+% gf-ud patterns
 To handle these situations, the CE module can filter the results based on a \texttt{gf-ud} pattern.
@@
 
+% predicates
 To give a practically useful example, a possibility is to extract verb phrases by looking for full clauses and pruning the subtrees corresponding to subjects. 
@@
 \section{Evaluation} \label{evaluation}
+% mini-roadmap
 In this section, we evaluate the system proposed in the above. 
-We first introduce the data used for such evaluation. 
+We first discuss the data used for such evaluation. 
 After that, we describe our experiments, aimed at putting both the CE module \textit{per se} and the entire system from parsing to lexicon generation to the test, and present our results.
@@
 \subsection{Data} \label{data}
+% PUD
 Because we want part of our evaluation to be independent from the quality of UD parsing, some of the experiments are carried out on treebanks instead of raw text. 
@@
 
+% plans
 When it comes to testing the program on raw text, we use two small ($< 1000$ sentences) bilingual sentence-aligned corpora consisting of course plans from the Department of Mathematics and Computer Science (DMI) of the University of Perugia (for English-Italian) and from the Department of Computer Science and Engineering (CSE) shared between the University of Gothenburg and the Chalmers University of Technology (for English-Swedish).
@@
 \subsection{Evaluating CE} \label{ceval}
+% why evaluate CE independently
 While we focus mostly on the MT applications of CA, automatic translation, and much less GF-based domain-specific tranlation, is not the only context in which CA can be put to use. 
@@
 
+% the 2 experiments
 We first assess the correctness of the alignments the CE module is able to extract from the PUD treebanks, comparing our results with those obtained with a standard statistical tool, \texttt{fast\_align} \cite{dyer-etal-2013-simple}.
@@
 
+% metrics
 While precision and recall are two well-known performance metrics, the lack of a gold standard for CE imposes us to approximate them with the number of correct alignments the program is able to extract and the ratio between such number and the total number of alignments extracted.
 However, since some alignments are only correct in the specific context of the sentence pair in which they occur, we make a further distinction between correct alignments that are relevant for a translation lexicon and alignments that are useful for comparing the sentences but should not be used for MT.
+As an extreme example of a pair-specific alignment, consider the sentences $\langle$\textit{He missed the boat, Ha perso il treno}$\rangle$. 
+In English and Italian, the idea of missing a chance is expressed using two idiomatic expression very similar to each other. 
+However, the vehicle mentioned in the Italian translation is not a boat but a train (``\textit{treno}'').
 
 \subsubsection{Results on manually annotated treebanks} 
+% table 1
 In Table \ref{pud_fast}, we compare the results obtained with our grammar-based module to those obtained statistically on the PUD treebanks.
 
+% details
 Of course, \texttt{fast\_align} does not make use of the information present in the CoNLL-U files excepts with regards to tokenization. 
 On the other hand, the relatively large size of the PUD treebanks makes it possible to also train the statistical tool on the full dataset instead of just using the chosen 100-sentences subset, allowing for a fairer comparison.
-
 In both cases, \texttt{fast\_align} is run with the recommended parameters and the CE program is configured to only extract one-to-many and many-to-one word alignments, given that \texttt{fast\_align} does not align larger phrases.
@@
 
+% comments on results
 While \texttt{fast\_align} is designed to align every word in the text (or explicitly state that a word has no counterpart in its translation) and, consequently, extracts around twice as many correspondences as our CE module, the percentage of correct correspondences is definitely in favor of our system, even though \texttt{fast\_align} gets significantly more precise when trained on the full, 1000-sentence dataset.
@@
 \subsubsection{Results on raw text} \label{raw}
+% mani avanti, table 2
 The course plan corpora are significantly harder to work with, the additional challenge being that our CE module relies on the quality of automatic parsing.
-
 A comparison between the results obtained on manually annotated data and the course plans corpora parsed with UDPipe is summarized Table \ref{raw_fast}.
@@
 
+% precision
 What is immediately evident, but not unexpected, is a decrease in precision. 
@@
 
+% recall
 Results are less encouraging in terms of recall: the number of alignments extracted from the course plans corpora is similar to that obtained from the PUD treebank sample, despite the difference in size.
@@
 Corpus-level BLEU scores for the automatic translations of the 50+50 sentences of the testing corpora are summarized in Table \ref{tableu}. 
-
 Following conventions, we report the cumulative $n$-gram scores for values of $n$ from 1 to 4 (BLEU-1 to BLEU-4). 
@@
 Grammatical errors in Swedish are less common and less systematic. 
-Only in one case, for instance, gender is incorrect (``\textit{programbibliotek\underline{en}}''). 
+Only in one case, for instance, gender is incorrect (``\textit{programbibliotek\textbf{en}}''). 
 These errors are easy to handle when writing a domain-specific grammar or, in cases like the latter, by making small adjustments to the morphological dictionaries.

### 2021-05-16 paper/paper.tex
Aarne's edits done

@@
 \begin{abstract}
-In grammar-based domain-specific MT, constructing high-quality translation lexica is a crucial but time consuming task requiring significant linguistic knowledge. 
-With parallel example sentences available, statistical alignment tools can help automate part of the process, but they are not suitable for small datasets and limited when it comes to complex multiword expressions.
+Grammar-based domain-specific MT systems are a common use case for CNL.
+High-quality translation lexica are a crucial part of such systems, but involve time consuming work and significant linguistic knowledge. 
+With parallel example sentences available, statistical alignment tools
+can help automate part of the process, but they are not suitable for
+small datasets and do not always perform well with complex multiword expressions. 
 In addition, the correspondences between word forms obtained in this way cannot be used directly.
-Addressing these problems, we propose a grammar-based approach to this task and put it to the test in a simple translation pipeline.
+Addressing these problems, we propose a grammar-based approach to this task and put it to test in a simple translation pipeline.
 \end{abstract}
@@
 Grammar-based translation pipelines such as those based on Grammatical Framework (GF) have been successfully employed in domain-specific Machine Translation (MT) \cite{ranta-etal-2020-abstract}.  
-What makes these systems well suited to the task is the fact that, when we constrain ourselves to a specific domain, where precision is often more important than coverage, they can provide strong guarantees in terms of grammatical correctness.
+What makes these systems well suited to the task is the fact that, when we constrain ourselves to a specific domain, where precision is often more important than coverage, they can provide strong guarantees of correctness.
 
 % importance of lexica in DSMT
-Nevertheless, lexical exactness is, in this context, just as important as grammaticality. 
-An important part of the design of the Controlled Natural Language (CNL) one such system's grammar describes then becomes the creation of a high-quality translation lexicon.
+However, lexical exactness is, in this context, just as important as grammaticality. 
+An important part of the design of a Controlled Natural Language (CNL) is the creation of a high-quality translation lexicon, preserves both semantics and grammatical correctness.
 
 % how it is done right now and issues
-This is often done manually, resulting in a time consuming task requiring significant linguistic knowledge. 
-When the grammar is designed based on a corpus of parallel example sentences, it is possible to automate part of this process by means of statistical word and phrase alignment techniques \cite{brown-etal-1993-mathematics, och-ney-2000-improved, dyer-etal-2013-simple}.
-None of them is, however, suitable for the common case in which only a limited amount of example data is available.
+A translation lexicon is often built manually, which is a time consuming task requiring significant linguistic knowledge. 
+When the task is based on a corpus of parallel example sentences, it is possible to automate part of this process by means of statistical word and phrase alignment techniques \cite{brown-etal-1993-mathematics, och-ney-2000-improved, dyer-etal-2013-simple}.
+None of them is, however, suitable for the common case in which only a small amount of example data is available --- typically, which just one occurrence of each relevant lexical item.
 
@@
 For this reason, we refer to the task our system attempts to automate as \textit{Concept Alignment} (CA). 
+A \textit{concept} is a semantic unit expressed by a word or a construction, which is also a unit of \textit{compositional translation}, where translation is performed by mapping concepts to concepts in a shared syntactic structure.
 
 % additional advantage 2: morphology
-Finally, performing morphological analysis allows us to generate translation lexica complete with grammatical category and inflection information instead of simple correspondences between word forms.
+Concepts are to be conceived as \textit{lemmas} equipped with morphological variations, instead of fixed word forms or phrases.
+This  allows us to generate translation lexica complete with grammatical category and inflection so that correct target language forms can be selected in each syntactic context.
 
@@
 This paper is structured as follows. 
-Section \ref{methodology} starts by giving an overview of our approach to CA and comparing it to related work to then focus on our algorithm for extracting correspondences.
-After that, Section \ref{evaluation} presents the results obtained in a first evaluation of the system.
-Finally, Section \ref{conclusions} summarizes our conclusions and discusses some ideas for future work.
+Section \ref{methodology} starts by giving an overview of our approach to CA and comparing it with related work, followed by a description of our CA algorithm.
+Section \ref{evaluation} presents the results obtained in an evaluation of the system.
+Section \ref{conclusions} summarizes our conclusions and discusses some ideas for future work.
 
@@
 The basic use case for CA, which we refer to specifically as \textit{Concept Extraction} (CE), is the generation of a translation lexicon from a multilingual parallel text.
-This can be directly compared to the numerous existing word and phrase alignment techniques.
+This is analogous to the well-known earlier word and phrase alignment techniques.
 
@@
 An interesting and less studied variant of CA is \textit{Concept Propagation} (CP), useful for cases where a set of concepts is already known and the goal is to identify the expressions corresponding to each of them in a new language, potentially even working with a different text in the same domain.
-While our system does implement basic CP functionalities, however, in this paper we focus on CE and restrict its application to bilingual corpora.
+While our system does implement basic CP functionalities, in this paper we focus on CE and restrict its application to bilingual corpora.
 
 % contextualizing our proposal
-As stated in the Introduction, most existing extraction solutions are based on statistical approaches and are, as a consequence, unsuitable for small datasets. 
-Grammar-based approaches, making use of parallel treebanks and collectively referred to as \textit{tree-to-tree alignment methods}, have also been proposed \cite{tiedemann2011bitext}, but have historically suffered from the inconsistencies between the formalisms used to define the grammars for the various languages and from the lack of robustness of existing parsers.
-This work is a new attempt in the same direction, enabled by two multilinguality-oriented grammar formalisms developed over the course of the last 25 years: Grammatical Framework (GF) \cite{ranta-etal-2020-abstract} and Universal Dependencies (UD) \cite{ws-2019-universal}.
+As stated in the Introduction, most existing CE solutions are based on statistical approaches and are, as a consequence, unsuitable for small datasets. 
+Grammar-based approaches, making use of parallel treebanks and collectively referred to as \textit{tree-to-tree alignment methods}, have also been proposed \cite{tiedemann2011bitext}, but have historically suffered from the inconsistencies between the formalisms used to define the grammars of different languages and from the lack of robustness of parsers.
+This work is a new attempt in the same direction, enabled by two multilinguality-oriented grammar formalisms developed over the course of the last 25 years: Grammatical Framework (GF) \cite{ranta-2011} and Universal Dependencies (UD) \cite{ws-2019-universal}.
 
 % GF
-GF is a constituency grammar formalisms and programming language in which grammars are represented as pairs of an \textit{abstract syntax}, playing the role of an interlingua, and a set of \textit{concrete syntaxes} capturing the specificities of the various natural languages.
+GF is a constituency grammar formalism and programming language in which grammars are represented as pairs of an \textit{abstract syntax}, playing the role of an interlingua, and a set of \textit{concrete syntaxes} capturing the specificities of the various natural languages.
 In the case of translation, similarly to what happens in programming language compilation, strings in the source language are \textit{parsed} to Abstract Syntax Trees (ASTs), which are then \textit{linearized} to target language strings.
@@
 UD, on the other hand, is a dependency grammar formalism meant for cross-linguistically consistent grammatical annotation.
-As opposed to constituency, \textit{dependency} is a word-to-word correspondence: each word is put in relation with the one it refers to, called its \textit{head}, via a directed labelled link specifying the syntactic relation between them.
-Importantly for our application, the standard format for UD trees, CoNNL-U, gives information not only on the syntactic role of each word, but also on its Part-Of-Speech (POS) tag, lemma and morphological features.
+As opposed to constituency, \textit{dependency} is a word-to-word correspondence: each word is put in relation with the one it depends on, called its \textit{head}, via a directed labelled link specifying the syntactic relation between them.
+Importantly for our application, the standard format for UD trees, CoNNL-U, gives information not only on the syntactic role of each word, but also on its Part-Of-Speech (POS), lemma, and morphological features.
 
@@
 % introducing CE algorithm
-The core part of the system outlined above is of course the alignment module. 
-In this section, we present our method for extracting alignments from parallel bilingual UD treebanks.
-Our basic algorithm is sketched in th pseudocode below.
+The core part of the system outlined above is the alignment module. 
+Its function is to extract alignments from parallel bilingual UD treebanks.
+The outline of the algorithm is given in the following pseudocode:
 
@@
 Here, the input consists of a list of priority-sorted \textit{alignment criteria}, i.e.
-rules to determine whether two dependency trees should be aligned with each other which will be discussed in greater detail in Section \ref{criteria}, and a pair of sentences to align represented as UD trees.
-From an implementation point of view, UD trees are, rose trees where each node represents a token, obtained from the CoNLL-U files produced by the UD parser.
+rules to determine whether two dependency trees should be aligned with each other, and a pair $(t,u)$ of UD trees to align.
+An example alignment criterion is sameness of syntactic label, for instance, that subjects are aligned with subjects and objects with objects; the details will be discussed in Section \ref{criteria},
+From an implementation point of view, UD trees are, rose trees (trees with arbitrary numbers of branches) where each node represents a word with its dependents as subtrees.
+The rose tree is is easily obtained from the CoNLL-U notation that UD parsers produce.
 
@@
 % algorithm description
-As a first step, the program checks whether the two full sentence trees can be aligned with each other, i.e. if they match any of the alignment criteria. 
-If this is the case, they are added to a collection of alignments, which are represented themselves as a pair of UD (sub)trees associated with some metadata, such as the id of the sentence they were extracted from.
-Such collection is what the function will return after aligning all the dependency subtrees.
-The same procedure is applied recursively to all pairs of immediate subtrees of each sentence, until the leaves are reached or alignment is no longer possible due to a lack of matching criteria.
-Subtrees are sorted based on their dependency label, so to give higher priority to pairs whose heads have the same label.
+As a first step, the program checks whether the two full sentence trees can be aligned with each other, i.e. if they match the alignment criteria. 
+If this is the case, they are added to a collection of alignments, which are represented as pairs of UD (sub)trees associated with some metadata, such as the id of the sentence they were extracted from.
+Such a collection is what the function will return after aligning all the dependency subtrees.
+The same procedure is applied recursively to all pairs of immediate subtrees of each sentence, until the leaves are reached or alignment is no longer possible due to lack of matching criteria.
+Subtrees are sorted based on their dependency label, to give higher priority to pairs whose heads have the same label.
 
 % head alignment
-A simple but useful refinement is that, depending on which alignment criteria a pair of trees candidate for alignment matches, the heads of the two trees may or may not also be added to the collection of alignments.
-This is important not to miss one-word correspondences that cannot be captured in any other way, for instance between the root verbs of two full sentences.
-A relevant implementative detail is that, in this context, the head of a tree is not simply defined as its root.
-Instead, if the root is part of a compound written as two or more separate words or a verb with auxiliaries, the root nodes of the corresponding subtrees are also considered part of it.
+A simple but useful refinement is that, depending on which alignment criteria a pair of trees matches, the heads of the two trees may or may not also be added to the collection of alignments. 
+This is done in order not to miss one-word correspondences that cannot be captured in any other way, for instance between the root verbs of two full sentences.
+A relevant implementation detail is that, in this context, the head of a tree is not simply defined as its root.
+Instead, if the root is part of a compound written as two or more separate words or a verb with auxiliaries, the root nodes of the corresponding subtrees are also considered parts of it.
 
 % iterations
-When working on multiple sentences, this algorithm can be applied in an iterative fashion, so that knowledge gathered when a sentence pair is aligned can be reused when working on the following ones and to keep track of the number of occurrences of each alignment throughout the entire text.
+When working on multiple sentences, the algorithm can be applied in an iterative fashion, so that knowledge gathered when a sentence pair is aligned can be used when working on later sentences and also to keep track of the number of occurrences of each alignment throughout the entire text.
 
@@
 % why list criteria 
-While alignment criteria are customizable, to allow for a better understanding of the extraction algorithm described in the above, we describe the ones that the current implementation utilizes by default.
+While the alignment criteria are customizable, to allow for a better understanding of the extraction algorithm described above, we explain the criteria that our implementation utilizes by default.
 
@@
 attached to aligned heads, constitute an alignment if their roots share the same dependency label, meaning that they are in the same syntactic relation with their heads.
-Note that, since the root of UD trees is always attached to a a fake node with an arc labelled \texttt{root}, this criterion also makes it so that full sentences are always considered to align with each other, something which is desirable since we assume the parallel texts that are fed to our program to be sentence-aligned.
+Note that, since the root of UD trees is always attached to a a fake node with an arc labelled \texttt{root}, this criterion also implies that full sentences are always considered to align with each other.
+This is desirable since we assume that the parallel texts that are fed to our program are sentence-aligned.
 
@@
 % actual POS-equiv par
-As noted in the above, CoNNL-U files also provide information on the grammatical categories of each word, represented as Universal POS tags \cite{petrov-etal-2012-universal}. 
+As noted above, the CoNNL-U notation provides information on the grammatical categories of each word, represented as Universal POS tags \cite{petrov-etal-2012-universal}. 
 Intuitively, if the nodes of two trees in matching contexts have the same POS tags, the two trees are more likely to correspond to each other than if not. 
-This is especially true if we focus, for instance, solely on the open class words (defined as in the UD documentation \footnote{\url{universaldependencies.org/u/pos/all.html}}), thus ignoring prepositions, determiners, auxiliary verbs and all other function words, which tend to behave differently across different languages.
+This is especially true if we focus, for instance, solely on the open class words (defined as in the UD documentation \footnote{\url{universaldependencies.org/u/pos/all.html}}), thus ignoring prepositions, determiners, auxiliary verbs and all function words, which tend to behave differently across different languages.
 A useful relation to define between dependency trees is, then, that of \textit{POS-equivalence}: two dependency trees $t$, $u$ are POS-equivalent if $M_1 = M_2 \neq \emptyset$, where $M_i$ is defined as the multiset of POS tags of all the open class word nodes of $t_i$.
@@
 For instance, \textit{categorial divergences} occur when the POS tag of a word in the source language changes in its translation.
-An ubiquitous example of is that of adverbs rendered as obliques, such as in the English-Italian pair $\langle$\textit{She studies \textbf{consistently}, Lei studia \textbf{con costanza}}$\rangle$.
-\textit{Structural divergences}, where a core argument of a sentence in the source language are rendered as obliques in the source language (as in the English-Swedish pair $\langle$\textit{I told \textbf{him}, Jag berättade \textbf{för honom}.}$\rangle$), are also frequently encountered.
+A ubiquitous example of is that of adverbial modifiers (with the UD label \texttt{advmod}) are translated prepositional phrases (with the UD label \texttt{obl}, for \textit{oblique}), such as in the English-Italian pair $\langle$\textit{She studies \textbf{consistently}, Lei studia \textbf{con costanza}}$\rangle$.
+\textit{Structural divergences}, where a direct object in one language is rendered as an oblique in the other, as in the English-Swedish pair $\langle$\textit{I told \textbf{him}, Jag berättade \textbf{för honom}.}$\rangle$), are also frequently encountered.
 
@@
 % motivation
-At a first glance, this might seem a criterion with no practical applications. 
+At a first glance, this might look like a criterion with no practical applications. 
 However, it is particularly useful when, instead of starting with an empty set of correspondences, we initialize the program with some alignments that are either inserted manually or, most interestingly, obtained with some other alignment technique.
-For instance, in this way it is possible to combine the system proposed in this paper in combination with a statistical tool and give more credit to the correspondences identified by both systems.
+For instance, in this way it is possible to combine the system proposed in this paper with a statistical tool and give more credit to correspondences identified by both systems.
 
@@
 % context
-So far, we described how CE can be used in a setting where the objective is to generate a comprehensive translation lexicon based on set of example sentence pairs.
-We pointed out that the program can be configured so to prioritize precision or coverage, but never restricted our search to a particular type of alignments.
-Nonetheless, there can be cases in which only certain syntactic structures are of interest: for instance, we might be looking for adverbs or noun phrases exclusively.
+So far, we have described how CE can be used in a setting where the objective is to generate a comprehensive translation lexicon based on set of example sentence pairs.
+We pointed out that the program can be configured to prioritize precision or coverage, but we did not restrict our search to a particular type of alignments.
+However, there are cases in which only certain syntactic structures are of interest: for instance, we might be looking for adverbs or noun phrases exclusively.
 
 % gf-ud patterns
-To handle these situations, the CE module can filter the results based on a \texttt{gf-ud} pattern.
+To handle such cases, the CE module can filter the results based on a \texttt{gf-ud} \textit{tree patterns}.
 \texttt{gf-ud} supports in fact both simple pattern matching, which is integrated in the CE module itself, and pattern replacement\footnote{documentation is available at \url{github.com/GrammaticalFramework/gf-ud}}.
-Combining them, for instance by pruning the UD trees extracted by the alignment module, allows us to extract even correspondences that, given the structure of UD trees, cannot be identified by CE alone.
+Combining them, for instance by pruning the UD trees extracted by the alignment module, allows us to extract correspondences that cannot be identified by CE alone.
 
 % predicates
-To give a practically useful example, a possibility is to extract verb phrases by looking for full clauses and pruning the subtrees corresponding to subjects. 
-It is also possible to perform some actual replacement to obtain \textit{predication patterns}, i.e.
-correspondences that specify the argument structure of verbs, such as the following English-Italian one:
+For example, pattern matching can extract verb phrases by looking for full clauses and dropping the subtrees corresponding to subjects. 
+By means of replacements, one can obtain \textit{predication patterns}, that is, correspondences that specify the argument structure of verbs, such as the following English-Italian one:
 
@@
 The CE module described so far outputs pairs of UD trees. 
-While converting them to GF ASTs is one of \texttt{gf-ud}'s core functionalities, building a compilable GF translation lexicon is necessary to turn such trees into grammar rules.
+While converting them to GF ASTs is one of \texttt{gf-ud}'s core functionalities, building a compilable GF translation lexicon is needed to turn such trees into grammar rules.
 
-In order to do that, our grammar generation module requires a morphological dictionary of the languages at hand and an \textit{extraction grammar}. 
-What we refer to as an extraction grammar is of course a GF grammar, whose objective is to define the set of basic categories and syntactic rules the entries of the automatically generated lexicon are going to use.
-\todo{not sure how important this section is and what (if anything) to add to it}
+In order to do that, our grammar generation module uses a morphological dictionary of the languages at hand and an \textit{extraction grammar}. 
+The extraction grammar is a GF grammar that defines the syntactic categories and functions the entries of the automatically generated lexicon are build with.
+For example, entries can be prepositional phrases (PP) or verb phrases (VP) constructed by the following GF functions:
+\begin{verbatim}
+  PrepNP : Prep -> NP -> PP 
+         # case head
+  PrepPP : VP   -> PP -> VP 
+         # head obl
+\end{verbatim}
+The dependency labels appended to the function types instruct \texttt{gf-ud} to build GF trees from UD trees that match these labels.
 
@@
 % mini-roadmap
-In this section, we evaluate the system proposed in the above. 
-We first discuss the data used for such evaluation. 
+In this section, we evaluate the system proposed above. 
+We first discuss the data used in the evaluation. 
 After that, we describe our experiments, aimed at putting both the CE module \textit{per se} and the entire system from parsing to lexicon generation to the test, and present our results.
@@
 Of the over 20 languages PUD treebanks are available in, we selected English, Italian and Swedish. 
-Using this data limits the amount of alignment errors that are due annotation issues to a minimum, even though a small number of inconsistencies and imprecisions is present even in this corpus.
+Using this data limits the amount of alignment errors that are due to annotation issues to a minimum, even though a small number of inconsistencies and errors is present even in this corpus.
 
@@
 % why evaluate CE independently
-While we focus mostly on the MT applications of CA, automatic translation, and much less GF-based domain-specific tranlation, is not the only context in which CA can be put to use. 
-For instance, it is easy to imagine using it to improve the UI of online translation memories.
-For this reason, a first set of experiments is aimed at evaluating the alignments obtained with our CA module independently from the other stages stages of our lexicon generation pipeline.
+While we focus mostly on the MT applications of CA, automatic translation, and much less GF-based domain-specific translation, is not the only context in which CA can be put to use. 
+For instance, it is easy to imagine using it to build translation memories used by human translators.
+For this reason, a first set of experiments is aimed at evaluating the alignments obtained with our CA module independently from the other stages of our lexicon generation pipeline.
 
 % the 2 experiments
-We first assess the correctness of the alignments the CE module is able to extract from the PUD treebanks, comparing our results with those obtained with a standard statistical tool, \texttt{fast\_align} \cite{dyer-etal-2013-simple}.
-In addition, we try to quantify the impact automatic parsing has on the performance of the CE module by comparing the above results with those obtained on the DMI and CSE corpora.
+We first assess the correctness of the alignments the CE module is able to extract from the PUD treebanks, comparing our results with those obtained with a statistical tool, \texttt{fast\_align} \cite{dyer-etal-2013-simple}.
+In addition, we try to quantify the impact of automated UD parsing on the performance of the CE module by comparing the above results with those obtained on the DMI and CSE corpora.
 
 % metrics
-While precision and recall are two well-known performance metrics, the lack of a gold standard for CE imposes us to approximate them with the number of correct alignments the program is able to extract and the ratio between such number and the total number of alignments extracted.
-However, since some alignments are only correct in the specific context of the sentence pair in which they occur, we make a further distinction between correct alignments that are relevant for a translation lexicon and alignments that are useful for comparing the sentences but should not be used for MT.
+While precision and recall are two well-known performance metrics, the lack of a gold standard for CE forces us to assess the correctness of alignments manually before calculating the ratio between correctly extracted alignments and the total of extracted alignments.
+What is more, since some alignments are only correct in the specific context of the sentence pair in which they occur, we make a further distinction between correct alignments that are relevant for a translation lexicon and alignments that are useful for comparing the sentences but should not be used for MT.
 As an extreme example of a pair-specific alignment, consider the sentences $\langle$\textit{He missed the boat, Ha perso il treno}$\rangle$. 
@@
 % details
-Of course, \texttt{fast\_align} does not make use of the information present in the CoNLL-U files excepts with regards to tokenization. 
-On the other hand, the relatively large size of the PUD treebanks makes it possible to also train the statistical tool on the full dataset instead of just using the chosen 100-sentences subset, allowing for a fairer comparison.
-In both cases, \texttt{fast\_align} is run with the recommended parameters and the CE program is configured to only extract one-to-many and many-to-one word alignments, given that \texttt{fast\_align} does not align larger phrases.
+Of course, \texttt{fast\_align} does not make use of the information present in the CoNLL-U files except with regards to tokenization. 
+On the other hand, the relatively large size of the PUD treebanks makes it possible also to train the statistical tool on the full dataset instead of just using the chosen 100-sentences subset, allowing for a fairer comparison.
+In both cases, \texttt{fast\_align} is run with the recommended parameters and the CE program is configured only to extract one-to-many and many-to-one word alignments, given that \texttt{fast\_align} does not align larger phrases.
 
@@
 % recall
-Results are less encouraging in terms of recall: the number of alignments extracted from the course plans corpora is similar to that obtained from the PUD treebank sample, despite the difference in size.
+The results are less encouraging in terms of recall: the number of alignments extracted from the course plan corpora is similar to that obtained from the PUD treebank sample, despite the difference in size.
 This is explained in part by the presence, in the course plans corpora, of a large amount of very short sentences, and in part by the fact that parse errors introduced by UDPipe make it impossible to align many subsentences without a significant loss in terms of precision.
@@
 
-Due to the small size of the datasets and the consequent low coverage of the extracted lexicon, we generate the sentences to translate directly in the GF shell rather than trying to parse arbitrary sentences other course plans.
+Due to the small size of the datasets and the consequently low coverage of the extracted lexicon, we generate the sentences to translate directly in the GF shell rather than trying to parse arbitrary sentences other course plans.
 In order to do that, we make use of GF's random AST generation functionality but at the same time manually select semantically plausible sentences to facilitate the task of the human translators.
@@
 These synthetic figures are useful to give an idea of the general quality of the translations: overall, although with relatively low scores, English-to-Swedish translation works significantly better than English-to-Italian.
-Looking back at the results reported in Section \ref{raw}, the reason for this is not immediately clear, as the difference in precision between the two language pairs is negligible on the course plans corpora.
+Looking back at the results reported in Section \ref{raw}, the reason for this is not immediately clear, as the difference in precision between the two language pairs is negligible in the course plan corpora.
 
@@
 The small size of the corpus, however, allows for some error analysis. 
-From the participants's observations about the kind of errors encountered when manually editing the automatic translation, summarizes in Table \ref{obs}, we can conclude that while most errors are in fact due to wrong alignments, the main difference between two corpora lies in the number of translations that only contain grammatical errors.
+From the participants' observations about the kind of errors encountered when manually editing the automatic translation, summarizes in Table \ref{obs}, we can conclude that while most errors are in fact due to wrong alignments, the main difference between two corpora lies in the number of translations that only contain grammatical errors.
 This explains the significant difference observed in the cumulative BLEU scores shown in Table \ref{tableu}.
@@
 \section{Conclusions} \label{conclusions}
-We presented a syntax-based alignment method with a focus on its applications in domain-specific translation lexica generation.
+We have presented a syntax-based alignment method with a focus on its applications in domain-specific translation lexic generation.
 
-Compared to the existing statistical tools, our system has the following advantages:
+Compared with the existing statistical tools, our system has the following advantages:
 \begin{itemize} \setlength\itemsep{0.1em}
   \item it performs consistently well even on extremely small parallel corpora
-  \item it is able to simultaneously extract correspondences between individual words, multiword expressions and full phrases, even when their constituents are discontinuous
+  \item it is able to simultaneously extract correspondences between individual words, multiword expressions and full phrases, including discontinuous constituents
   \item in conjunction with \texttt{gf-ud} pattern matching, it can be used to extract specific types of correspondences, such as predication patterns  
@@
 
-An obvious possible development is to improve the current implementation of CP, optimizing it for its different use cases: propagating alignments to a new language looking for correspondences using a translation of the same text they were extracted from or using a different text in the same domain.
+An obvious possible development is to improve the current implementation of concept propagation (CP), optimizing it for its different use cases: propagating alignments to a new language looking for correspondences using a translation of the same text they were extracted from or using a different text in the same domain.
 An alternative to the latter is to make CE, now working on bilingual texts, $n$-lingual.
 
-When large amounts of data ar available, using our system in conjunction with a statistical tool seems promising. 
+When large enough amounts of data ar available, using our system in conjunction with a statistical tool seems promising. 
 As discussed in Section \ref{ka}, this is already partially supported and it could prove useful to develop CA as an actual hybrid system.

### 2021-05-16 paper/paper.tex
last few major changes; only final review left

@@
 rules to determine whether two dependency trees should be aligned with each other, and a pair $(t,u)$ of UD trees to align.
-An example alignment criterion is sameness of syntactic label, for instance, that subjects are aligned with subjects and objects with objects; the details will be discussed in Section \ref{criteria},
-From an implementation point of view, UD trees are, rose trees (trees with arbitrary numbers of branches) where each node represents a word with its dependents as subtrees.
+An example alignment criterion is sameness of syntactic label, for instance, that subjects are aligned with subjects and objects with objects; the details will be discussed in Section \ref{criteria}.
+From an implementation point of view, UD trees are rose trees (trees with arbitrary numbers of branches) where each node represents a word with its dependents as subtrees.
 The rose tree is is easily obtained from the CoNLL-U notation that UD parsers produce.
@@
   \centering
-  \begin{subfigure}{.23\textwidth}
-      \centering
   \scriptsize
-      \setlength{\unitlength}{0.16mm}
-      \begin{picture}(371.0,130.0)
-          \put(0.0,0.0){this}
-          \put(46.0,0.0){is}
-          \put(83.0,0.0){a}
-          \put(120.0,0.0){dependency}
-          \put(220.0,0.0){tree}
+  \begin{subfigure}{.30  \textwidth}
+    \setlength{\unitlength}{0.14mm}
+\begin{picture}(257.0,70.0)
+  \put(0.0,0.0){She}
+  \put(46.0,0.0){studies}
+  \put(119.0,0.0){consistently}
   \put(0.0,15.0){{\tiny PRON}}
-          \put(46.0,15.0){{\tiny AUX}}
-          \put(83.0,15.0){{\tiny DET}}
-          \put(120.0,15.0){{\tiny NOUN}}
-          \put(220.0,15.0){{\tiny NOUN}}
-          \put(120.0,30.0){\oval(218.63636363636363,133.33333333333334)[t]}
-          \put(10.681818181818187,35.0){\vector(0,-1){5.0}}
-          \put(105.0,99.66666666666667){{\tiny nsubj}}
-          \put(143.0,30.0){\oval(172.27586206896552,100.0)[t]}
-          \put(56.86206896551724,35.0){\vector(0,-1){5.0}}
-          \put(128.0,83.0){{\tiny cop}}
-          \put(161.5,30.0){\oval(134.8102189781022,66.66666666666667)[t]}
-          \put(94.0948905109489,35.0){\vector(0,-1){5.0}}
-          \put(146.5,66.33333333333334){{\tiny det}}
-          \put(180.0,30.0){\oval(97.0,33.333333333333336)[t]}
-          \put(131.5,35.0){\vector(0,-1){5.0}}
-          \put(165.0,49.66666666666667){{\tiny compound}}
-          \put(235.0,130.0){\vector(0,-1){100.0}}
-          \put(240.0,120.0){{\tiny root}}
+  \put(46.0,15.0){{\tiny VERB}}
+  \put(119.0,15.0){{\tiny ADV}}
+  \put(33.0,30.0){\oval(39.47826086956522,33.333333333333336)[t]}
+  \put(13.26086956521739,35.0){\vector(0,-1){5.0}}
+  \put(18.0,49.66666666666667){{\tiny nsubj}}
+  \put(102.5,30.0){\oval(68.89041095890411,33.333333333333336)[t]}
+  \put(136.94520547945206,35.0){\vector(0,-1){5.0}}
+  \put(87.5,49.66666666666667){{\tiny advmod}}
+  \put(61.0,70.0){\vector(0,-1){40.0}}
+  \put(66.0,60.0){{\tiny root}}
+\end{picture}
+\\ \\
+\begin{picture}(259.0,90.0)
+  \put(0.0,0.0){Lei}
+  \put(46.0,0.0){studia}
+  \put(110.0,0.0){con}
+  \put(147.0,0.0){costanza}
+  \put(0.0,15.0){{\tiny PRON}}
+  \put(46.0,15.0){{\tiny NOUN}}
+  \put(110.0,15.0){{\tiny ADP}}
+  \put(147.0,15.0){{\tiny NOUN}}
+  \put(33.0,30.0){\oval(39.47826086956522,33.333333333333336)[t]}
+  \put(13.26086956521739,35.0){\vector(0,-1){5.0}}
+  \put(18.0,49.66666666666667){{\tiny nsubj}}
+  \put(138.5,30.0){\oval(28.89189189189189,33.333333333333336)[t]}
+  \put(124.05405405405405,35.0){\vector(0,-1){5.0}}
+  \put(123.5,49.66666666666667){{\tiny case}}
+  \put(116.5,30.0){\oval(98.02970297029702,66.66666666666667)[t]}
+  \put(165.5148514851485,35.0){\vector(0,-1){5.0}}
+  \put(101.5,66.33333333333334){{\tiny nmod}}
+  \put(61.0,90.0){\vector(0,-1){60.0}}
+  \put(66.0,80.0){{\tiny root}}
 \end{picture}
   \end{subfigure}
-  \begin{subfigure}{.38\textwidth}
-    \centering
-    \scriptsize
+  \begin{subfigure}{.30  \textwidth}
     \begin{verbatim}
-      1	this	this	PRON	_	_	5	nsubj	_	_
-      2	is	be	AUX	_	_	5	cop	_	_
-      3	a	a	DET	_	_	5	det	_	_
-      4	dependency	dependency	NOUN	_	_	5	
-        compound	_	_
-      5	tree	tree	NOUN	_	_	0	root	_	_
+
+1 She PRON 2 nsubj
+2 studies VERB 0 root    
+3 consistently ADV 2 advmod
+
+1 Lei PRON 2 nsubj
+2 studia VERB 0 root
+3 con ADP 4 case
+4 costanza NOUN 2 obl
     \end{verbatim}
-  \end{subfigure}%
-  \begin{subfigure}{.38\textwidth}
-    \centering
-    \scriptsize
+  \end{subfigure}
+  \begin{subfigure}{.30 \textwidth}
     \begin{verbatim}
-5	tree	tree	NOUN	_	_	0	root	_	_
-  4	dependency	dependency	NOUN	_	_	5	
-    compound	_	_
-  2	is	be	AUX	_	_	5	cop	_	_
-  3	a	a	DET	_	_	5	det	_	_
-  1	this	this	PRON	_	_	5	nsubj	_	_
+
+2 studies VERB 0 root  
+  3 consistently ADV 2 advmod
+  1 She PRON 2 nsubj
+
+2 studia VERB 0 root
+  1 Lei PRON 2 nsubj
+  4 costanza NOUN 2 obl
+    3 con ADP 4 case
     \end{verbatim}
-  \end{subfigure}%
-  \caption[The graphical representation of a UD tree alongside its CoNNL-U and sorted rose tree representation.
-Optional CoNNL-U fields are left blank.]{The graphical representation of a UD tree alongside its CoNNL-U and rose tree representation.
-Optional CoNNL-U fields are left blank.}
+  \end{subfigure}
+  \caption{The graphical and simplified CoNNL-U representation of a pair of UD sentences.
+Below, the corresponding sorted rose trees.
+With the default parameters, the resulting alignments are: $\langle$\textit{She studies consistently, Lei studia con costanza}$\rangle$, $\langle$\textit{studies, studia}$\rangle$, $\langle$\textit{she, lei}$\rangle$ and $\langle$\textit{consistently, con costanza}.$\rangle$}
 \end{figure*}
@@
 As a first step, the program checks whether the two full sentence trees can be aligned with each other, i.e. if they match the alignment criteria. 
+In the case of the example criterion discussed above, this means that their roots are labelled the same.
 If this is the case, they are added to a collection of alignments, which are represented as pairs of UD (sub)trees associated with some metadata, such as the id of the sentence they were extracted from.
@@
 The same procedure is applied recursively to all pairs of immediate subtrees of each sentence, until the leaves are reached or alignment is no longer possible due to lack of matching criteria.
-Subtrees are sorted based on their dependency label, to give higher priority to pairs whose heads have the same label.
+Subtrees are sorted based on their dependency label to give higher priority to pairs whose heads have the same label (cf. \textsc{sort} in the pseudocode).
 
@@
 In both cases, \texttt{fast\_align} is run with the recommended parameters and the CE program is configured only to extract one-to-many and many-to-one word alignments, given that \texttt{fast\_align} does not align larger phrases.
+This explains CE's seemingly low recall.
+To get a better idea, Table \ref{pud_fast} can be compared with Table \ref{raw_fast}, which summarizes the results of an experiment where no constraints are placed on the size of the extracted alignments.
 

### 2021-05-16 paper/paper.tex
minor changes

@@
 
-\usepackage{todonotes}
-
 \usepackage{caption}
@@
 %         Author 3 \\ Address line \\ ... \\ Address line}
-% TODO: authors info
 \author{First Author \\
@@
 
-% TODO: add example (Herb?)
-
 \paragraph{Known alignment} \label{ka}
@@
 \subsection{MT experiments} \label{mtexp}
+% genereral idea behind MT experiments
 The second set of experiments has the objective of assessing the quality of the final output of the system we propose: GF translation lexica. 
@@
 
+% test corpus construction
 The idea is to automatically translate a set of English sentences to Italian and Swedish, ask native speakers of the target languages to produce a set of reference translations, and compare them to the original machine-generated ones by computing BLEU scores.
-
 Due to the small size of the datasets and the consequently low coverage of the extracted lexicon, we generate the sentences to translate directly in the GF shell rather than trying to parse arbitrary sentences other course plans.
@@
 
+% reference translations
 Reference translations are obtained by asking participants to compare the original English sentences to their automatically translated counterparts and correct the latter by making the minimal changes necessary to obtain a set of grammatically and semantically correct translations.
@@
 \subsubsection{Results} 
+% scores
 Corpus-level BLEU scores for the automatic translations of the 50+50 sentences of the testing corpora are summarized in Table \ref{tableu}. 
@@
 
+% comment on general results
 These synthetic figures are useful to give an idea of the general quality of the translations: overall, although with relatively low scores, English-to-Swedish translation works significantly better than English-to-Italian.
@@
 
+% individual scores vary widely
 Looking at sentence-level scores can, however, be more insightful. 
@@
 
+% 100
 Examples of sentences that were assigned a perfect BLEU-1 to 4 score are ``\textit{the library provides useful textbooks}'' (translated to Italian as ``\textit{la biblioteca fornisce libri utili}'') in the DMI corpus and ``\textit{this lab is more difficult than the exam}'' (whose Swedish translation is ``\textit{den här laborationen är svårare än tentamen}'') in the CSE corpus.
@@
 
+% 0
 On the other hand, it is easy for shorter sentences to be assigned the minimum BLEU-1 to 4 score even when they only contain a single grammatical or semantic error. 
@@
 
+% BLEU issue
 Furthermore, a problem with using the BLEU score as the only evaluation metric is the fact that it makes no distinction between content and function words, thus not allowing an evaluation focused specifically on the extracted concepts.
@@
 
+% common grammar errors
 Among other things, Italian contractions are often handled incorrectly.
@@
 
+% cool alignment errors
 Some errors regarding the extracted concepts are also interesting to analyze: the alignment $\langle$\textit{``class'', ``classe''}$\rangle$, for instance, causes the sentence \textit{``I will attend the class''} to be (incorrectly) translated as \textit{``io seguirò la classe''} instead of \textit{``io seguirò la lezione''} even though the correspondence is in fact valid in most of the numerous contexts in which (within the same domain!) \textit{``class''} is not to be intended as a synonym of \textit{``lesson''}.
@@
 \section{Conclusions} \label{conclusions}
-We have presented a syntax-based alignment method with a focus on its applications in domain-specific translation lexic generation.
-
+% summary
+We have presented a syntax-based alignment method with a focus on its applications in domain-specific translation lexicon generation.
 Compared with the existing statistical tools, our system has the following advantages:
@@
 
+% sw
 The tangible fruits of this work are a Haskell library and a number of executables offering a user friendly interface to perform CE, lexicon generation and various kinds of evaluations. 
@@
 
+% CP/multilingual CE
 An obvious possible development is to improve the current implementation of concept propagation (CP), optimizing it for its different use cases: propagating alignments to a new language looking for correspondences using a translation of the same text they were extracted from or using a different text in the same domain.
@@
 
-When large enough amounts of data ar available, using our system in conjunction with a statistical tool seems promising. 
+% hybridization
+When large enough amounts of data are available, using our system in conjunction with a statistical tool seems promising. 
 As discussed in Section \ref{ka}, this is already partially supported and it could prove useful to develop CA as an actual hybrid system.
 
+% tools
 Finally, since the freedom that generally characterizes human translation and the quality of currently available UD parsers make maximizing both alignment precision and recall unrealistic, tools to make it easier to postprocess the automatically generated lexica are under development.

### 2021-05-16 paper/paper.tex
fixed typos found by C.

@@
 However, lexical exactness is, in this context, just as important as grammaticality. 
-An important part of the design of a Controlled Natural Language (CNL) is the creation of a high-quality translation lexicon, preserves both semantics and grammatical correctness.
+An important part of the design of a Controlled Natural Language (CNL) is the creation of a high-quality translation lexicon, preserving both semantics and grammatical correctness.
 
@@
 When the task is based on a corpus of parallel example sentences, it is possible to automate part of this process by means of statistical word and phrase alignment techniques \cite{brown-etal-1993-mathematics, och-ney-2000-improved, dyer-etal-2013-simple}.
-None of them is, however, suitable for the common case in which only a small amount of example data is available --- typically, which just one occurrence of each relevant lexical item.
+None of them is, however, suitable for the common case in which only a small amount of example data is available --- typically, with just one occurrence of each relevant lexical item.
 

### 2021-05-16 paper/paper.tex
reviewed up to divergences

@@
 
-% importance of lexica in DSMT
+% lexica in DSMT
 However, lexical exactness is, in this context, just as important as grammaticality. 
 An important part of the design of a Controlled Natural Language (CNL) is the creation of a high-quality translation lexicon, preserving both semantics and grammatical correctness.
-
-% how it is done right now and issues
 A translation lexicon is often built manually, which is a time consuming task requiring significant linguistic knowledge. 
-When the task is based on a corpus of parallel example sentences, it is possible to automate part of this process by means of statistical word and phrase alignment techniques \cite{brown-etal-1993-mathematics, och-ney-2000-improved, dyer-etal-2013-simple}.
+When the task is based on a corpus of parallel example sentences, part of this process can be automated by means of statistical word and phrase alignment techniques \cite{brown-etal-1993-mathematics, och-ney-2000-improved, dyer-etal-2013-simple}.
 None of them is, however, suitable for the common case in which only a small amount of example data is available --- typically, with just one occurrence of each relevant lexical item.
@@
 % additional advantage 2: morphology
-Concepts are to be conceived as \textit{lemmas} equipped with morphological variations, instead of fixed word forms or phrases.
-This  allows us to generate translation lexica complete with grammatical category and inflection so that correct target language forms can be selected in each syntactic context.
+Conceiving conepts as \textit{lemmas} equipped with morphological variations rather than fixed word forms or phrases allows us to generate translation lexica complete with grammatical category and inflection, so that correct target language forms can be selected in each syntactic context.
 
@@
 % contextualizing our proposal
-As stated in the Introduction, most existing CE solutions are based on statistical approaches and are, as a consequence, unsuitable for small datasets. 
+As stated in the Introduction, most existing alignment solutions are based on statistical approaches and are, as a consequence, unsuitable for small datasets. 
 Grammar-based approaches, making use of parallel treebanks and collectively referred to as \textit{tree-to-tree alignment methods}, have also been proposed \cite{tiedemann2011bitext}, but have historically suffered from the inconsistencies between the formalisms used to define the grammars of different languages and from the lack of robustness of parsers.
@@
 % role of UD and GF
-While both formalisms independently solve the issues related to having to work with grammars that are inconsistent with each other, UD is especially appealing since, dependency trees being an easier target, several robust parsers, such as \cite{straka-etal-2016-udpipe} and \cite{chen-manning-2014-fast} are available.
+While both formalisms independently solve the issues related to having to work with grammars that are inconsistent with each other, UD is especially appealing since, being dependency trees an easier target, several robust parsers, such as \cite{straka-etal-2016-udpipe} and \cite{chen-manning-2014-fast} are available.
 Alone, UD trees are sufficient to extract (or propagate) tree-to-tree alignments, but not to automate the generation of a morphologically-aware translation lexicon for a generative grammar. 
@@
 % algorithm description
-As a first step, the program checks whether the two full sentence trees can be aligned with each other, i.e. if they match the alignment criteria. 
+As a first step, the program checks whether the two full sentence trees can be aligned with each other, i.e. if they match one or more alignment criteria. 
 In the case of the example criterion discussed above, this means that their roots are labelled the same.
@@
 % iterations
-When working on multiple sentences, the algorithm can be applied in an iterative fashion, so that knowledge gathered when a sentence pair is aligned can be used when working on later sentences and also to keep track of the number of occurrences of each alignment throughout the entire text.
+When working on multiple sentences, the algorithm can be applied in an iterative fashion, so that knowledge gathered when a sentence pair is aligned can be used when working on later sentences and to keep track of the number of occurrences of each alignment throughout the entire text.
 
@@
 attached to aligned heads, constitute an alignment if their roots share the same dependency label, meaning that they are in the same syntactic relation with their heads.
-Note that, since the root of UD trees is always attached to a a fake node with an arc labelled \texttt{root}, this criterion also implies that full sentences are always considered to align with each other.
+Note that, since the root of a UD tree is always attached to a a fake node with an arc labelled \texttt{root}, this criterion also implies that full sentences are always considered to align with each other.
 This is desirable since we assume that the parallel texts that are fed to our program are sentence-aligned.
@@
 Intuitively, if the nodes of two trees in matching contexts have the same POS tags, the two trees are more likely to correspond to each other than if not. 
-This is especially true if we focus, for instance, solely on the open class words (defined as in the UD documentation \footnote{\url{universaldependencies.org/u/pos/all.html}}), thus ignoring prepositions, determiners, auxiliary verbs and all function words, which tend to behave differently across different languages.
+This is especially true if we focus, for instance, solely on the open class words (defined as in the UD documentation\footnote{\url{universaldependencies.org/u/pos/all.html}}), thus ignoring function words such as prepositions, determiners and auxiliary verbs, which tend to behave differently across different languages.
 A useful relation to define between dependency trees is, then, that of \textit{POS-equivalence}: two dependency trees $t$, $u$ are POS-equivalent if $M_1 = M_2 \neq \emptyset$, where $M_i$ is defined as the multiset of POS tags of all the open class word nodes of $t_i$.
@@
 For instance, \textit{categorial divergences} occur when the POS tag of a word in the source language changes in its translation.
-A ubiquitous example of is that of adverbial modifiers (with the UD label \texttt{advmod}) are translated prepositional phrases (with the UD label \texttt{obl}, for \textit{oblique}), such as in the English-Italian pair $\langle$\textit{She studies \textbf{consistently}, Lei studia \textbf{con costanza}}$\rangle$.
+A ubiquitous example of is that of adverbial modifiers (with the UD label \texttt{advmod}) translated as prepositional phrases (with the UD label \texttt{obl}, for \textit{oblique}), such as in the English-Italian pair $\langle$\textit{She studies \textbf{consistently}, Lei studia \textbf{con costanza}}$\rangle$.
 \textit{Structural divergences}, where a direct object in one language is rendered as an oblique in the other, as in the English-Swedish pair $\langle$\textit{I told \textbf{him}, Jag berättade \textbf{för honom}.}$\rangle$), are also frequently encountered.

### 2021-05-16 paper/paper.tex
hopefully done with shortening and last minute changes

@@
 \usepackage[noend]{algpseudocode}
+
+\usepackage{fancyvrb}
 
@@
   \end{subfigure}
-  \caption{The graphical and simplified CoNNL-U representation of a pair of UD sentences.
-Below, the corresponding sorted rose trees.
-With the default parameters, the resulting alignments are: $\langle$\textit{She studies consistently, Lei studia con costanza}$\rangle$, $\langle$\textit{studies, studia}$\rangle$, $\langle$\textit{she, lei}$\rangle$ and $\langle$\textit{consistently, con costanza}.$\rangle$}
+  \caption{The graphical, simplified CoNNL-U and sorted rose tree representation of a pair of UD sentences.
+With the default criteria, the resulting alignments are: $\langle$\textit{She studies consistently, Lei studia con costanza}$\rangle$, $\langle$\textit{studies, studia}$\rangle$, $\langle$\textit{she, lei}$\rangle$ and $\langle$\textit{consistently, con costanza}$\rangle$}.
 \end{figure*}
@@
 \paragraph{Part-Of-Speech equivalence}
-% actual POS-equiv par
 As noted above, the CoNNL-U notation provides information on the grammatical categories of each word, represented as Universal POS tags \cite{petrov-etal-2012-universal}. 
@@
 A useful relation to define between dependency trees is, then, that of \textit{POS-equivalence}: two dependency trees $t$, $u$ are POS-equivalent if $M_1 = M_2 \neq \emptyset$, where $M_i$ is defined as the multiset of POS tags of all the open class word nodes of $t_i$.
-
-% combining criteria, especially pos-equiv + label matching
 Applied as a backup for label matching, this criterion can be used to capture correspondences that would otherwise be missed, thus increasing recall, but a decrease in precision is also to be expected.
@@
 \paragraph{Known translation divergence}
-% actual divergences par
 Parallel texts often present significant, systematic cross-linguistic grammatical distinctions. 
@@
 Drawing inspiration from \cite{dorr-1994-machine}, we refer to them as \textit{translation divergences} and handle some of the most common ones explicitly. 
-
-% examples
 For instance, \textit{categorial divergences} occur when the POS tag of a word in the source language changes in its translation.
@@
 At a first glance, this might look like a criterion with no practical applications. 
-However, it is particularly useful when, instead of starting with an empty set of correspondences, we initialize the program with some alignments that are either inserted manually or, most interestingly, obtained with some other alignment technique.
-For instance, in this way it is possible to combine the system proposed in this paper with a statistical tool and give more credit to correspondences identified by both systems.
+However, it can be useful when, instead of starting with an empty set of correspondences, we initialize the program with some alignments that are either inserted manually or, most interestingly, obtained with some other alignment technique.
+For instance, in this way it is possible to combine the system proposed in this paper with a statistical tool and give more credit to correspondences identified by both.
 
@@
 % gf-ud patterns
-To handle such cases, the CE module can filter the results based on a \texttt{gf-ud} \textit{tree patterns}.
+To handle such cases, the CE module can filter the results based on a \texttt{gf-ud} \textit{tree pattern}.
 \texttt{gf-ud} supports in fact both simple pattern matching, which is integrated in the CE module itself, and pattern replacement\footnote{documentation is available at \url{github.com/GrammaticalFramework/gf-ud}}.
@@
 
+\begin{table*}[h]
+  \centering
+  \small
+  \begin{tabular}{r|c|c|c|c|c|c}
+  \multicolumn{1}{l|}{\textbf{}} & \multicolumn{2}{c|}{\textbf{CE}} & \multicolumn{2}{c|}{\textbf{\begin{tabular}[c]{@{}c@{}}\texttt{fast\_align} \\(100 sentences)\end{tabular}}} & \multicolumn{2}{c}{\textbf{\begin{tabular}[c]{@{}c@{}}\texttt{fast\_align} \\(full dataset)\end{tabular}}} \\ \hline
+  \multicolumn{1}{l|}{}          & \textbf{en-it}  & \textbf{en-sv} & \textbf{en-it}                              & \textbf{en-sv}                              & \textbf{en-it}                                & \textbf{en-sv} \\ \hline
+  \textbf{distinct alignments}        & 536             & 638            & 1242                                        & 1044                                        & 1286                                          & 1065          \\
+  \textbf{correct}                & 392 (73\%)      & 514 (80\%)     & 346 (28\%)                                  & 538 (52\%)                                  & 540 (42\%)                                    & 677 (64\%)    \\
+  \textbf{usable in MT}                      & 363 (68\%)      & 503 (79\%)     & 316 (25\%)                                  & 525 (50\%)                                  & 510 (40\%)                                    & 666 (63\%)    \\
+  \end{tabular}
+  \caption[Word alignment comparison between our grammar-based CE module and \texttt{fast\_align}]{Comparison between our grammar-based CE module and \texttt{fast\_align} on PUD data training the model on 100 and 1000 sentences and discarding the alignments obtained for sentences 101-1000 in the latter case.}
+  \label{pud_fast}
+  \end{table*}
+
 \subsection{Generating grammar rules}
@@
 
-In order to do that, our grammar generation module uses a morphological dictionary of the languages at hand and an \textit{extraction grammar}. 
-The extraction grammar is a GF grammar that defines the syntactic categories and functions the entries of the automatically generated lexicon are build with.
+To this end, our grammar generation module uses a morphological dictionary of the languages at hand and an \textit{extraction grammar}. 
+The extraction grammar defines the syntactic categories and functions the entries of the automatically generated lexicon are built with.
 For example, entries can be prepositional phrases (PP) or verb phrases (VP) constructed by the following GF functions:
-\begin{verbatim}
-  PrepNP : Prep -> NP -> PP 
-         # case head
-  PrepPP : VP   -> PP -> VP 
-         # head obl
-\end{verbatim}
+
+\begin{Verbatim}[fontsize=\small]
+  PrepNP : Prep -> NP -> PP # case head
+  PrepPP : VP   -> PP -> VP # head obl
+\end{Verbatim}
 The dependency labels appended to the function types instruct \texttt{gf-ud} to build GF trees from UD trees that match these labels.
@@
 Of the over 20 languages PUD treebanks are available in, we selected English, Italian and Swedish. 
-Using this data limits the amount of alignment errors that are due to annotation issues to a minimum, even though a small number of inconsistencies and errors is present even in this corpus.
+Using this data limits the amount of alignment errors that are due to annotation issues to a minimum, even though a small number of inconsistencies is present even in this corpus.
 
@@
 As an extreme example of a pair-specific alignment, consider the sentences $\langle$\textit{He missed the boat, Ha perso il treno}$\rangle$. 
-In English and Italian, the idea of missing a chance is expressed using two idiomatic expression very similar to each other. 
-However, the vehicle mentioned in the Italian translation is not a boat but a train (``\textit{treno}'').
-
-\subsubsection{Results on manually annotated treebanks} 
-% table 1
-In Table \ref{pud_fast}, we compare the results obtained with our grammar-based module to those obtained statistically on the PUD treebanks.
-
-% details
-Of course, \texttt{fast\_align} does not make use of the information present in the CoNLL-U files except with regards to tokenization. 
-On the other hand, the relatively large size of the PUD treebanks makes it possible also to train the statistical tool on the full dataset instead of just using the chosen 100-sentences subset, allowing for a fairer comparison.
-In both cases, \texttt{fast\_align} is run with the recommended parameters and the CE program is configured only to extract one-to-many and many-to-one word alignments, given that \texttt{fast\_align} does not align larger phrases.
-This explains CE's seemingly low recall.
-To get a better idea, Table \ref{pud_fast} can be compared with Table \ref{raw_fast}, which summarizes the results of an experiment where no constraints are placed on the size of the extracted alignments.
-
-\begin{table*}[h]
-  \centering
-  \small
-  \begin{tabular}{r|c|c|c|c|c|c}
-  \multicolumn{1}{l|}{\textbf{}} & \multicolumn{2}{c|}{\textbf{CE}} & \multicolumn{2}{c|}{\textbf{\begin{tabular}[c]{@{}c@{}}\texttt{fast\_align} \\(100 sentences)\end{tabular}}} & \multicolumn{2}{c}{\textbf{\begin{tabular}[c]{@{}c@{}}\texttt{fast\_align} \\(full dataset)\end{tabular}}} \\ \hline
-  \multicolumn{1}{l|}{}          & \textbf{en-it}  & \textbf{en-sv} & \textbf{en-it}                              & \textbf{en-sv}                              & \textbf{en-it}                                & \textbf{en-sv} \\ \hline
-  \textbf{distinct alignments}        & 536             & 638            & 1242                                        & 1044                                        & 1286                                          & 1065          \\
-  \textbf{correct}                & 392 (73\%)      & 514 (80\%)     & 346 (28\%)                                  & 538 (52\%)                                  & 540 (42\%)                                    & 677 (64\%)    \\
-  \textbf{usable in MT}                      & 363 (68\%)      & 503 (79\%)     & 316 (25\%)                                  & 525 (50\%)                                  & 510 (40\%)                                    & 666 (63\%)    \\
-  \end{tabular}
-  \caption[Word alignment comparison between our grammar-based CE module and \texttt{fast\_align}]{Comparison between our grammar-based CE module and \texttt{fast\_align} on PUD data training the model on 100 and 1000 sentences and discarding the alignments obtained for sentences 101-1000 in the latter case.}
-  \label{pud_fast}
-  \end{table*}
-
-% comments on results
-While \texttt{fast\_align} is designed to align every word in the text (or explicitly state that a word has no counterpart in its translation) and, consequently, extracts around twice as many correspondences as our CE module, the percentage of correct correspondences is definitely in favor of our system, even though \texttt{fast\_align} gets significantly more precise when trained on the full, 1000-sentence dataset.
-
-\subsubsection{Results on raw text} \label{raw}
-% mani avanti, table 2
-The course plan corpora are significantly harder to work with, the additional challenge being that our CE module relies on the quality of automatic parsing.
-A comparison between the results obtained on manually annotated data and the course plans corpora parsed with UDPipe is summarized Table \ref{raw_fast}.
+In both languages, the idea of missing a chance is expressed with idiomatic expressions very similar to each other. 
+However, the the Italian translation mentions a train (``\textit{treno}'') in the place of a boat.
 
@@
   \end{table*}
+
+\subsubsection{Results on manually annotated treebanks} 
+% table 1
+In Table \ref{pud_fast}, we compare the results obtained with our grammar-based module to those obtained statistically on the PUD treebanks.
+
+% details
+Of course, \texttt{fast\_align} does not make use of the information present in the CoNLL-U files except with regards to tokenization. 
+On the other hand, the relatively large size of the PUD treebanks makes it possible also to train the statistical tool on the full dataset instead of just using the chosen 100-sentences subset, allowing for a fairer comparison.
+In both cases, \texttt{fast\_align} is run with the recommended parameters and the CE program is configured only to extract one-to-many and many-to-one word alignments, given that \texttt{fast\_align} does not align larger phrases.
+This explains CE's seemingly low recall.
+To get a better idea, Table \ref{pud_fast} can be compared with Table \ref{raw_fast}, which summarizes the results of an experiment where no constraints are placed on the size of the extracted alignments.
+
+% comments on results
+While \texttt{fast\_align} is designed to align every word in the text (or explicitly state that a word has no counterpart in its translation) and, consequently, extracts around twice as many correspondences, the percentage of correct correspondences is definitely in favor of our system, even though \texttt{fast\_align} gets significantly more precise when trained on the full 1000-sentence dataset.
+
+\subsubsection{Results on raw text} \label{raw}
+% mani avanti, table 2
+The course plan corpora are significantly harder to work with, the additional challenge being that our CE module relies on the quality of automatic parsing.
+In Table \ref{raw_fast}, we compare the results obtained on manually annotated data and the course plans corpora parsed with UDPipe.
 
@@
 The idea is to automatically translate a set of English sentences to Italian and Swedish, ask native speakers of the target languages to produce a set of reference translations, and compare them to the original machine-generated ones by computing BLEU scores.
-Due to the small size of the datasets and the consequently low coverage of the extracted lexicon, we generate the sentences to translate directly in the GF shell rather than trying to parse arbitrary sentences other course plans.
+Due to the small size of the datasets and the consequently low coverage of the extracted lexicon, we generate the sentences to translate directly in the GF shell rather than trying to parse arbitrary sentences from other course plans.
 In order to do that, we make use of GF's random AST generation functionality but at the same time manually select semantically plausible sentences to facilitate the task of the human translators.
@@
 For both corpora, scores assigned to individual segments range from the minimum possible value of 0 to the perfect score of 100, which indicates a perfect correspondence between the automatic and the reference translation.
-
-% 100
 Examples of sentences that were assigned a perfect BLEU-1 to 4 score are ``\textit{the library provides useful textbooks}'' (translated to Italian as ``\textit{la biblioteca fornisce libri utili}'') in the DMI corpus and ``\textit{this lab is more difficult than the exam}'' (whose Swedish translation is ``\textit{den här laborationen är svårare än tentamen}'') in the CSE corpus.
-\smallskip
-
-% 0
 On the other hand, it is easy for shorter sentences to be assigned the minimum BLEU-1 to 4 score even when they only contain a single grammatical or semantic error. 
-This is the case, for instance, of the sentence ``\textit{the test is oral}'', whose last word, ``\textit{oral}'' is translated as ``\textit{dura}'' (``\textit{hard}'') instead of ``\textit{orale}'' due to an alignment error.
-Interestingly, this is one of many cases in which the correct alignment $\langle$``\textit{oral}'',``\textit{orale}''$\rangle$ is also identified by the CE module, but discared as we only take the first translation candidate into account.
+% This is the case, for instance, of the sentence ``\textit{the test is oral}'', whose last word, ``\textit{oral}'' is translated as ``\textit{dura}'' (``\textit{hard}'') instead of ``\textit{orale}'' due to an alignment error.
+% Interestingly, this is one of many cases in which the correct alignment $\langle$``\textit{oral}'',``\textit{orale}''$\rangle$ is also identified by the CE module, but discared as we only take the first translation candidate into account.
 
@@
 % common grammar errors
-Among other things, Italian contractions are often handled incorrectly.
-For example, ``\textit{del}'' (``\textit{di}'' + ``\textit{il}'', in English ``\textit{of the}'') is systematically rendered as two separate words due to UDPipe tokenization.
+Among other things, many Italian contractions such as ``\textit{del}'' (``\textit{di}'' + ``\textit{il}'', in English ``\textit{of the}'') are systematically rendered as two separate words due to UDPipe tokenization.
 Grammatical errors in Swedish are less common and less systematic. 
@@
 \begin{itemize} \setlength\itemsep{0.1em}
-  \item it performs consistently well even on extremely small parallel corpora
+  \item it performs consistently well even on small parallel corpora
   \item it is able to simultaneously extract correspondences between individual words, multiword expressions and full phrases, including discontinuous constituents
@@
 The tangible fruits of this work are a Haskell library and a number of executables offering a user friendly interface to perform CE, lexicon generation and various kinds of evaluations. 
-The source code, including a preliminary implementation of CP, is available at \url{github.com/harisont/concept-alignment}
+The source code, including a preliminary implementation of CP, is available on GitHub\footnote{\url{github.com/harisont/concept-alignment}}.
 
@@
 % CP/multilingual CE
-An obvious possible development is to improve the current implementation of concept propagation (CP), optimizing it for its different use cases: propagating alignments to a new language looking for correspondences using a translation of the same text they were extracted from or using a different text in the same domain.
-An alternative to the latter is to make CE, now working on bilingual texts, $n$-lingual.
+An obvious possible development is optimizing the current implementation of concept propagation (CP) for its two use cases: propagating alignments to a new language looking for correspondences using a translation of the same text they were extracted from or using a different text in the same domain.
+An alternative to the former is to make CE, now working on bilingual texts, $n$-lingual.
 
@@
 When large enough amounts of data are available, using our system in conjunction with a statistical tool seems promising. 
-As discussed in Section \ref{ka}, this is already partially supported and it could prove useful to develop CA as an actual hybrid system.
+As discussed above, this is already partially supported and it could prove useful to develop CA as an actual hybrid system.
 

### 2021-05-16 paper/paper.tex
squeezed in sentence about commercial projects and other las minute changes

@@
 %         Author 3 \\ Address line \\ ... \\ Address line}
-\author{First Author \\
-  Affiliation / Address line 1 \\
-  Affiliation / Address line 2 \\
-  Affiliation / Address line 3 \\
-  \texttt{email@domain} \\\And
-  Second Author \\
-  Affiliation / Address line 1 \\
-  Affiliation / Address line 2 \\
-  Affiliation / Address line 3 \\
-  \texttt{email@domain} \\}
+\author{Arianna Masciolini \\
+  Digital Grammars \\
+  \texttt{arianna@digitalgrammars.com} \\\And
+  Aarne Ranta \\
+  Department of Computer Science and Engineering, \\
+  University of Gothenburg \\
+  Digital Grammars \\
+  \texttt{aarne.ranta@cse.gu.se} \\}
 
@@
   \scriptsize
-  \begin{subfigure}{.30  \textwidth}
+  \begin{subfigure}{.32  \textwidth}
     \setlength{\unitlength}{0.14mm}
@@
   \put(0.0,15.0){{\tiny PRON}}
-  \put(46.0,15.0){{\tiny NOUN}}
+  \put(46.0,15.0){{\tiny VERB}}
   \put(110.0,15.0){{\tiny ADP}}
@@
   \end{subfigure}
-  \begin{subfigure}{.30  \textwidth}
+  \begin{subfigure}{.32  \textwidth}
     \begin{verbatim}
@@
   \end{subfigure}
-  \begin{subfigure}{.30 \textwidth}
+  \begin{subfigure}{.32 \textwidth}
     \begin{verbatim}
@@
   \caption{The graphical, simplified CoNNL-U and sorted rose tree representation of a pair of UD sentences.
-With the default criteria, the resulting alignments are: $\langle$\textit{She studies consistently, Lei studia con costanza}$\rangle$, $\langle$\textit{studies, studia}$\rangle$, $\langle$\textit{she, lei}$\rangle$ and $\langle$\textit{consistently, con costanza}$\rangle$}.
+With the default criteria, the resulting alignments are: $\langle$\textit{She studies consistently, Lei studia con costanza}$\rangle$, $\langle$\textit{studies, studia}$\rangle$, $\langle$\textit{she, lei}$\rangle$ and $\langle$\textit{consistently, con costanza}$\rangle$.}
 \end{figure*}
@@
 In Table \ref{pud_fast}, we compare the results obtained with our grammar-based module to those obtained statistically on the PUD treebanks. 
-
-% details
 Of course, \texttt{fast\_align} does not make use of the information present in the CoNLL-U files except with regards to tokenization. 
@@
 The results of this process are two small testing corpora, one for the DMI and one for the CSE corpus, each consisting of 50 English sentences. 
-
-% reference translations
-Reference translations are obtained by asking participants to compare the original English sentences to their automatically translated counterparts and correct the latter by making the minimal changes necessary to obtain a set of grammatically and semantically correct translations.
-This is important as, if the reference translations are obtained independently from the automatic ones, BLEU scores can easily become misleading.
+Reference translations are obtained by asking participants to compare the original English sentences to their automatically translated counterparts and correct the latter with the minimal changes necessary to obtain a set of grammatically and semantically correct translations.
+This is important as, if the reference translations are obtained independently, BLEU scores can easily become misleading.
 
@@
 % cool alignment errors
-Some errors regarding the extracted concepts are also interesting to analyze: the alignment $\langle$\textit{``class'', ``classe''}$\rangle$, for instance, causes the sentence \textit{``I will attend the class''} to be (incorrectly) translated as \textit{``io seguirò la classe''} instead of \textit{``io seguirò la lezione''} even though the correspondence is in fact valid in most of the numerous contexts in which (within the same domain!) \textit{``class''} is not to be intended as a synonym of \textit{``lesson''}.
+Some errors regarding the extracted concepts are also interesting to analyze: the alignment $\langle$\textit{class, classe}$\rangle$, for instance, causes the sentence \textit{``I will attend the class''} to be (incorrectly) translated as \textit{``io seguirò la classe''} instead of \textit{``io seguirò la lezione''} even though the correspondence is in fact valid in most contexts in which (within the same domain!) \textit{``class''} is not to be intended as a synonym of \textit{``lesson''} but as teaching group.
 
@@
 The tangible fruits of this work are a Haskell library and a number of executables offering a user friendly interface to perform CE, lexicon generation and various kinds of evaluations. 
-The source code, including a preliminary implementation of CP, is available on GitHub\footnote{\url{github.com/harisont/concept-alignment}}.
+The source code, including a preliminary implementation of CP, is available on GitHub\footnote{link hidden at review time}. % TODO: github.com/harisont/concept-alignment
+The software has already been used to analyse customer-provided data in two commercial projects. % TODO: at Digital Grammars.
 

### 2021-05-18 ConceptAlignment.hs
fix #16

@@ -228,7 +228,9 @@ alignSent as cs p cl ex s@(s1,s2) =
     -- the list of criteria is needed because of how alignClause works
     alignSent' :: [Criterion] -> (UDTree,UDTree) -> AlignmentMap
     alignSent' cs (t@(RTree n ts),u@(RTree m us)) = if (not . null) matchingCs
-      then case (isJust p,headAlign (head matchingCs)) of
+      then case (isJust p,(not . null) ts 
+                          && (not . null) us 
+                          && headAlign (head matchingCs)) of
         -- not using a gf-ud pattern
         (False,True) -> insert' h (insert' a as) `union'` as'
         (False,False) -> insert' a as `union'` as'

### 2021-05-18 ConceptAlignment.hs
reintroduced pruning (slow for the time being)

@@ -246,7 +246,7 @@ alignSent as cs p cl ex s@(s1,s2) =
           })
         h = alignHeads a
         -- subtree alignments
-        as' = unions' $ [alignSent' cs (t,u) | t <- ts', u <- us']
+        as' = prune $ unions' $ [alignSent' cs (t,u) | t <- ts', u <- us']
             where (ts',us') = (sortByLabel ts,sortByLabel us)
                     where sortByLabel = sortOn (udSimpleDEPREL . root)
         matchingCs = 
@@ -337,6 +337,26 @@ alignSent as cs p cl ex s@(s1,s2) =
               where
                 aus = filterByLabel "amod" us
                 ats = filterByLabel "amod" ts 
+    
+-- | Helper function removing the less valid alternative alignments
+prune :: AlignmentMap -> AlignmentMap
+-- two different sorting functions are used one after another so it's 
+-- easier to change the code to use only one of them or change the order 
+-- in which they are applied
+prune m = M.fromList $ nubBy areAlt $ sortByFertility $ sortByReasons $ M.toList m
+  where 
+    -- check if two alignments are alternative to each other comparing THE
+    -- ACTUAL TREES, and not their "linearizations"
+    areAlt a b = sl a == sl b || tl a == tl b
+    -- sort alignments by number of reasons (decreasing order), then by first
+    sortByReasons = sortOn (\a -> 
+      let rs = reasons (meta a)
+          rs' = rs `S.difference` S.fromList [HEAD, PREV, PM] 
+      in (-(length rs'), maximum $ S.elems rs))
+    -- alt. fertility-based sorting strategy
+    -- top alignments are the ones that lead to more sub-alignments
+    sortByFertility as = sortOn (\a -> - (length $ subas a)) as
+      where subas a = filter (\a' -> a `contains` a') as
 
 {- Propagation functions -}
 

### 2021-06-30 paper/paper.tex
fixed typo and post-review todos

@@
   \end{tabular}
-  \caption{Comparison the extraction of alignments of any size from manually annotated PUD treebanks and automatically parsed sentences from the course plans corpora.}
+  \caption{Comparison between the extraction of alignments of any size from manually annotated PUD treebanks and from automatically parsed sentences from the course plans corpora.}
   \label{raw_fast}
@@
 The tangible fruits of this work are a Haskell library and a number of executables offering a user friendly interface to perform CE, lexicon generation and various kinds of evaluations. 
-The source code, including a preliminary implementation of CP, is available on GitHub\footnote{link hidden at review time}. % TODO: github.com/harisont/concept-alignment
-The software has already been used to analyse customer-provided data in two commercial projects. % TODO: at Digital Grammars.
+The source code, including a preliminary implementation of CP, is available on GitHub\footnote{\url{github.com/harisont/concept-alignment}}. 
+The software has already been used to analyse customer-provided data in two commercial projects at Digital Grammars.
 

### 2021-06-30 paper/paper.tex
improved captions and references

@@
   \caption{The graphical, simplified CoNNL-U and sorted rose tree representation of a pair of UD sentences.
-With the default criteria, the resulting alignments are: $\langle$\textit{She studies consistently, Lei studia con costanza}$\rangle$, $\langle$\textit{studies, studia}$\rangle$, $\langle$\textit{she, lei}$\rangle$ and $\langle$\textit{consistently, con costanza}$\rangle$.}
+With the default criteria, which among other things allow for matching adverbial with adjectival modifiers, the resulting alignments are: $\langle$\textit{She studies consistently, Lei studia con costanza}$\rangle$, $\langle$\textit{studies, studia}$\rangle$, $\langle$\textit{she, lei}$\rangle$ and $\langle$\textit{consistently, con costanza}$\rangle$.}
+  \label{ex}
 \end{figure*}
@@
 For instance, \textit{categorial divergences} occur when the POS tag of a word in the source language changes in its translation.
-A ubiquitous example of is that of adverbial modifiers (with the UD label \texttt{advmod}) translated as prepositional phrases (with the UD label \texttt{obl}, for \textit{oblique}), such as in the English-Italian pair $\langle$\textit{She studies \textbf{consistently}, Lei studia \textbf{con costanza}}$\rangle$.
+An ubiquitous example of is that of adverbial modifiers (with the UD label \texttt{advmod}) translated as prepositional phrases (with the UD label \texttt{obl}, for \textit{oblique}), such as in the English-Italian pair $\langle$\textit{She studies \textbf{consistently}, Lei studia \textbf{con costanza}}$\rangle$ (see Figure \ref{ex}).
 \textit{Structural divergences}, where a direct object in one language is rendered as an oblique in the other, as in the English-Swedish pair $\langle$\textit{I told \textbf{him}, Jag berättade \textbf{för honom}.}$\rangle$), are also frequently encountered.
@@
   \end{tabular}
-  \caption{Comparison between the extraction of alignments of any size from manually annotated PUD treebanks and from automatically parsed sentences from the course plans corpora.}
+  \caption{Comparison between the grammar-based extraction of alignments of any size from manually annotated PUD treebanks and from automatically parsed sentences from the course plans corpora.}
   \label{raw_fast}

### 2021-07-02 paper/paper.tex
fixed more typos

@@
 \begin{abstract}
-Grammar-based domain-specific MT systems are a common use case for CNL.
+Grammar-based domain-specific MT systems are a common use case for CNLs.
 High-quality translation lexica are a crucial part of such systems, but involve time consuming work and significant linguistic knowledge. 
@@
 % additional advantage 2: morphology
-Conceiving conepts as \textit{lemmas} equipped with morphological variations rather than fixed word forms or phrases allows us to generate translation lexica complete with grammatical category and inflection, so that correct target language forms can be selected in each syntactic context.
+Conceiving concepts as \textit{lemmas} equipped with morphological variations rather than fixed word forms or phrases allows us to generate translation lexica complete with grammatical category and inflection, so that correct target language forms can be selected in each syntactic context.
 
@@
 As opposed to constituency, \textit{dependency} is a word-to-word correspondence: each word is put in relation with the one it depends on, called its \textit{head}, via a directed labelled link specifying the syntactic relation between them.
-Importantly for our application, the standard format for UD trees, CoNNL-U, gives information not only on the syntactic role of each word, but also on its Part-Of-Speech (POS), lemma, and morphological features.
+Importantly for our application, the standard format for UD trees, CoNNL-U, gives information not only on the syntactic role of each word, but also on its Part-Of-Speech (POS) tag, lemma, and morphological features.
 
@@
 From an implementation point of view, UD trees are rose trees (trees with arbitrary numbers of branches) where each node represents a word with its dependents as subtrees.
-The rose tree is is easily obtained from the CoNLL-U notation that UD parsers produce.
+The rose tree is easily obtained from the CoNLL-U notation that UD parsers produce.
 
@@
 This is especially true if we focus, for instance, solely on the open class words (defined as in the UD documentation\footnote{\url{universaldependencies.org/u/pos/all.html}}), thus ignoring function words such as prepositions, determiners and auxiliary verbs, which tend to behave differently across different languages.
-A useful relation to define between dependency trees is, then, that of \textit{POS-equivalence}: two dependency trees $t$, $u$ are POS-equivalent if $M_1 = M_2 \neq \emptyset$, where $M_i$ is defined as the multiset of POS tags of all the open class word nodes of $t_i$.
+A useful relation to define between dependency trees is, then, that of \textit{POS-equivalence}: two dependency trees $t_1$, $t_2$ are POS-equivalent if $M_1 = M_2 \neq \emptyset$, where $M_i$ is defined as the multiset of POS tags of all the open class word nodes of $t_i$.
 Applied as a backup for label matching, this criterion can be used to capture correspondences that would otherwise be missed, thus increasing recall, but a decrease in precision is also to be expected.
@@
 For instance, \textit{categorial divergences} occur when the POS tag of a word in the source language changes in its translation.
-An ubiquitous example of is that of adverbial modifiers (with the UD label \texttt{advmod}) translated as prepositional phrases (with the UD label \texttt{obl}, for \textit{oblique}), such as in the English-Italian pair $\langle$\textit{She studies \textbf{consistently}, Lei studia \textbf{con costanza}}$\rangle$ (see Figure \ref{ex}).
+An ubiquitous example of this is that of adverbial modifiers (with the UD label \texttt{advmod}) translated as prepositional phrases (with the UD label \texttt{obl}, for \textit{oblique}), such as in the English-Italian pair $\langle$\textit{She studies \textbf{consistently}, Lei studia \textbf{con costanza}}$\rangle$ (see Figure \ref{ex}).
 \textit{Structural divergences}, where a direct object in one language is rendered as an oblique in the other, as in the English-Swedish pair $\langle$\textit{I told \textbf{him}, Jag berättade \textbf{för honom}.}$\rangle$), are also frequently encountered.
@@
 In both languages, the idea of missing a chance is expressed with idiomatic expressions very similar to each other. 
-However, the the Italian translation mentions a train (``\textit{treno}'') in the place of a boat.
+However, the Italian translation mentions a train (``\textit{treno}'') in the place of a boat.
 
@@
 The second set of experiments has the objective of assessing the quality of the final output of the system we propose: GF translation lexica. 
-Because we are now focussing on using CA in the context of domain-specific MT, we do not make use of the PUD treebanks, where sentences come from a variety of different sources, but just of the course plans corpora.
+Because we are now focusing on using CA in the context of domain-specific MT, we do not make use of the PUD treebanks, where sentences come from a variety of different sources, but just of the course plans corpora.
 We do not construct a grammar specific to such domain: for small-scale MT experiments, it is sufficient to extend the extraction grammar itself with preexisting syntax rules defined in the RGL.
@@
   \item it can automatically generate compilable, morphology-aware GF translation lexica
-  \item it can be configured to easily handle systematic, possibly language-pair or corpus specific translation divergences.
+  \item it can be configured to easily handle systematic, possibly language pair or corpus specific translation divergences.
 \end{itemize}

### 2021-07-02 paper/paper.tex
example multiwords

@@
 
+% qualitative analysis: multiwords
+Our system is, on the other hand, capable of extracting multiword alignments that are unlikely to be identified by a statistical tool, especially in the case of such a small dataset.
+Examples of this are the noun phrases  $\langle$\textit{the aim of the course, l'obiettivo del corso, syftet med kursen}$\rangle$ (a concept found in both corpora and, as such, trilingual), $\langle$\textit{Natural Language Processing, språkteknologi}$\rangle$ and $\langle$\textit{object oriented programming, programmazione ad oggetti}$\rangle$.
+
 \subsection{MT experiments} \label{mtexp}
-% genereral idea behind MT experiments
+% general idea behind MT experiments
 The second set of experiments has the objective of assessing the quality of the final output of the system we propose: GF translation lexica. 

### 2021-07-02 paper/paper.tex
addressed some minor feedback and expanded 2.2

@@
 An interesting and less studied variant of CA is \textit{Concept Propagation} (CP), useful for cases where a set of concepts is already known and the goal is to identify the expressions corresponding to each of them in a new language, potentially even working with a different text in the same domain.
-While our system does implement basic CP functionalities, in this paper we focus on CE and restrict its application to bilingual corpora.
+While our system does implement basic CP functionalities, in this paper we focus on its most mature portion: CE. 
+As results analogous to those that could be obtained via multilingual extraction can be obtained more easily with a combination of CE and CP, we for the moment restrict ourselves to bilingual corpora.
 
@@
 When working on multiple sentences, the algorithm can be applied in an iterative fashion, so that knowledge gathered when a sentence pair is aligned can be used when working on later sentences and to keep track of the number of occurrences of each alignment throughout the entire text.
+Furthermore, it is possible to initialize the algorithm with a nonempty set of alignments, obtained with the same program or by means of a statistical tool outputting alignments in Pharaoh format.
 
@@
 \subsection{Generating grammar rules}
-The CE module described so far outputs pairs of UD trees. 
-While converting them to GF ASTs is one of \texttt{gf-ud}'s core functionalities, building a compilable GF translation lexicon is needed to turn such trees into grammar rules.
+% from UD trees to GF rules: outline
+The alignments outputted by the CE module described so far are represented as pairs of UD (sub)trees in CoNLL-U format. 
+While converting them to GF ASTs is one of \texttt{gf-ud}'s core functionalities, such trees also need be converted into the grammar rules of a compilable GF translation lexicon.
+To this end, our grammar generation module requires a \textit{morphological dictionary} of the languages at hand and an \textit{extraction grammar}.
 
-To this end, our grammar generation module uses a morphological dictionary of the languages at hand and an \textit{extraction grammar}. 
-The extraction grammar defines the syntactic categories and functions the entries of the automatically generated lexicon are built with.
+% morphodicts
+Morphological dictionaries, implemented for several languages as part of the RGL, are large collections of lemmas associated with their inflectional forms.
+
+% extraction grammar
+An extraction grammar, on the other hand, defines the syntactic categories and functions the entries of the automatically generated lexicon are built with.
 For example, entries can be prepositional phrases (PP) or verb phrases (VP) constructed by the following GF functions:
@@
 \end{Verbatim}
+
+% role of gf-ud
 The dependency labels appended to the function types instruct \texttt{gf-ud} to build GF trees from UD trees that match these labels.
+
+% trees to rules
+The final translation lexicon, i.e. a GF grammar that extends the extraction grammar, is then derived from these GF trees. 
+In its abstract syntax, the name of each concept is associated with its grammatical category, i.e. the category of the root of the GF tree.
+For example:
+
+\begin{Verbatim}[fontsize=\small]
+  fun in_the_field__inom_område_PP : PP ;
+\end{Verbatim}
+
+The concrete syntaxes, on the other hand, contain the linearization rules for each concept, directly based on the trees obtained from \texttt{gf-ud}. 
+For instance, in English:
+
+\begin{Verbatim}[fontsize=\small]
+  lin in_the_field__inom_område_PP = 
+    PrepNP 
+      in_Prep 
+      (DetCN the_Det (UseN field_N)) ;
+\end{Verbatim}
+
+Most function words, such as \texttt{in\_Prep}, and many content words, such as \texttt{field\_N}, are available through the morphological dictionaries.
+When this is not the case, they are assumed to be regular and an additional rule is generated for them. 
+For instance, if the English morphological dictionary didn't contain the word ``\textit{field}'', we would have:
+
+\begin{Verbatim}[fontsize=\small]
+  oper field_N = mkN "field" ;
+\end{Verbatim}
 
@@
 % mani avanti, table 2
-The course plan corpora are significantly harder to work with, the additional challenge being that our CE module relies on the quality of automatic parsing.
+The course plan corpora are significantly harder to work with, the additional challenges being the inexactness of many translations (which is the direct cause of some of the alignment errors encountered in our evaluation) and the fact that our CE module relies, in this case, on the quality of automatic lemmatization, POS-tagging and parsing.
 In Table \ref{raw_fast}, we compare the results obtained on manually annotated data and the course plans corpora parsed with UDPipe.
@@
   \item it performs consistently well even on small parallel corpora
-  \item it is able to simultaneously extract correspondences between individual words, multiword expressions and full phrases, including discontinuous constituents
+  \item it is able to simultaneously extract correspondences between individual words, multiword expressions and longer phrases, including discontinuous constituents
   \item in conjunction with \texttt{gf-ud} pattern matching, it can be used to extract specific types of correspondences, such as predication patterns  

### 2021-07-02 paper/paper.tex
camera-ready version of the paper?

@@
 % Remove the "review" option to generate the final version.
-\usepackage[review]{acl}
+\usepackage{acl}
 
@@
   Aarne Ranta \\
-  Department of Computer Science and Engineering, \\
-  University of Gothenburg \\
+  University of Gothenburg,\\
+  Department of Computer Science \\
+  and Engineering; \\
   Digital Grammars \\
@@
 An example alignment criterion is sameness of syntactic label, for instance, that subjects are aligned with subjects and objects with objects; the details will be discussed in Section \ref{criteria}.
-From an implementation point of view, UD trees are rose trees (trees with arbitrary numbers of branches) where each node represents a word with its dependents as subtrees.
+From an implementation point of view, UD trees are rose trees (trees with arbitrary numbers of branches) where each node represents a word with its dependents as subtrees (see Figure \ref{ex}).
 The rose tree is easily obtained from the CoNLL-U notation that UD parsers produce.
@@
   \caption{The graphical, simplified CoNNL-U and sorted rose tree representation of a pair of UD sentences.
-With the default criteria, which among other things allow for matching adverbial with adjectival modifiers, the resulting alignments are: $\langle$\textit{She studies consistently, Lei studia con costanza}$\rangle$, $\langle$\textit{studies, studia}$\rangle$, $\langle$\textit{she, lei}$\rangle$ and $\langle$\textit{consistently, con costanza}$\rangle$.}
+With the default criteria, which among other things allow for matching adverbial with adjectival modifiers, the resulting alignments are: $\langle$\textit{She studies consistently, Lei studia con costanza}$\rangle$ (matching \texttt{root} label), $\langle$\textit{studies, studia}$\rangle$ (head alignment), $\langle$\textit{she, lei}$\rangle$ (matching \texttt{nsubj} label) and $\langle$\textit{consistently, con costanza}$\rangle$ (translation divergence; \texttt{amod} and \texttt{advmod} treated as equivalent).}
   \label{ex}
@@
 When working on multiple sentences, the algorithm can be applied in an iterative fashion, so that knowledge gathered when a sentence pair is aligned can be used when working on later sentences and to keep track of the number of occurrences of each alignment throughout the entire text.
-Furthermore, it is possible to initialize the algorithm with a nonempty set of alignments, obtained with the same program or by means of a statistical tool outputting alignments in Pharaoh format.
+Furthermore, it is possible to initialize the algorithm with a nonempty set of alignments, obtained with the same program or by means of a statistical tool outputting alignments in Pharaoh format and to combine the results of several extraction processes into a single translation lexicon.
 
@@
 When using raw text, our parser of choice is UDPipe \cite{straka-etal-2016-udpipe}. 
-In particular, we use the ParTUT English and Italian models for the DMI corpus and models trained on the bilingual LinES English-Swedish treebank for the CSE corpus.
+In particular, we use the ParTUT English and Italian models for the DMI corpus and models trained on the bilingual LinES English-Swedish treebank for the CSE corpus
+\footnote{The pretrained UDPipe models and information about their performance are available at \url{ufal.mff.cuni.cz/udpipe/1/models}}.
 
@@
 
+While they do require manual corrections and completions to an extent that varies according to the quality of the data and the strictness of the criteria used, alignments obtained with our method can reduce the time required for bootstrapping the translation lexicon building process for a domain-specific CNL significantly.
+In fact, especially if a comprehensive morphological dictionary is available, part of the alignments will be ready to use in a GF-based system without any intervention.
+
 % sw

### 2021-07-13 paper/paper.tex

@@
 Grammar-based translation pipelines such as those based on Grammatical Framework (GF) have been successfully employed in domain-specific Machine Translation (MT) \cite{ranta-etal-2020-abstract}.  
-What makes these systems well suited to the task is the fact that, when we constrain ourselves to a specific domain, where precision is often more important than coverage, they can provide strong guarantees of correctness.
+What makes these systems well suited to the task is the fact that, when we constrain ourselves to a specific domain, where precision is often more important than coverage, they can provide strong guarantees of grammatical correctness.
 
@@
 Section \ref{methodology} starts by giving an overview of our approach to CA and comparing it with related work, followed by a description of our CA algorithm.
-Section \ref{evaluation} presents the results obtained in an evaluation of the system.
+Section \ref{evaluation} presents the results obtained in a first evaluation of the system.
 Section \ref{conclusions} summarizes our conclusions and discusses some ideas for future work.
@@
 While our system does implement basic CP functionalities, in this paper we focus on its most mature portion: CE. 
-As results analogous to those that could be obtained via multilingual extraction can be obtained more easily with a combination of CE and CP, we for the moment restrict ourselves to bilingual corpora.
+Because results analogous to those that could be obtained via multilingual extraction can be obtained more easily with a combination of CE and CP, we restrict ourselves, for the time being, to bilingual corpora.
 
@@
 rules to determine whether two dependency trees should be aligned with each other, and a pair $(t,u)$ of UD trees to align.
-An example alignment criterion is sameness of syntactic label, for instance, that subjects are aligned with subjects and objects with objects; the details will be discussed in Section \ref{criteria}.
+An example alignment criterion is sameness of syntactic label, which makes it so that, for instance, subjects are aligned with subjects and objects with objects; the details will be discussed in Section \ref{criteria}.
 From an implementation point of view, UD trees are rose trees (trees with arbitrary numbers of branches) where each node represents a word with its dependents as subtrees (see Figure \ref{ex}).
@@
 \paragraph{Matching UD labels}
-The most obvious, but also most effective idea is to determine alignability based on comparing the dependency labels of the candidate UD tree pair. 
+The most obvious, but also most effective idea is to determine alignability based on comparing the dependency labels of the members of the candidate UD tree pair. 
 In particular, according to this idea, two subtrees in \textit{matching context}, i.e.
 attached to aligned heads, constitute an alignment if their roots share the same dependency label, meaning that they are in the same syntactic relation with their heads.
-Note that, since the root of a UD tree is always attached to a a fake node with an arc labelled \texttt{root}, this criterion also implies that full sentences are always considered to align with each other.
+Note that, since the root of a UD tree is always attached to a fake node with an arc labelled \texttt{root}, this criterion also implies that full sentences are always considered to align with each other.
 This is desirable since we assume that the parallel texts that are fed to our program are sentence-aligned.
@@
 So far, we have described how CE can be used in a setting where the objective is to generate a comprehensive translation lexicon based on set of example sentence pairs.
-We pointed out that the program can be configured to prioritize precision or coverage, but we did not restrict our search to a particular type of alignments.
+We pointed out that the program can be configured to prioritize precision or coverage, but we never restricted our search to a particular type of alignments.
 However, there are cases in which only certain syntactic structures are of interest: for instance, we might be looking for adverbs or noun phrases exclusively.
@@
 For example, pattern matching can extract verb phrases by looking for full clauses and dropping the subtrees corresponding to subjects. 
-By means of replacements, one can obtain \textit{predication patterns}, that is, correspondences that specify the argument structure of verbs, such as the following English-Italian one:
+By means of replacements, one can obtain \textit{predication patterns}, i.e. correspondences that specify the argument structure of verbs, such as the following English-Italian one:
 
@@
 $\langle$X \textit{gives} Y Z, X \textit{dà} Z \textit{a} Y$\rangle$.
-
-\begin{table*}[h]
-  \centering
-  \small
-  \begin{tabular}{r|c|c|c|c|c|c}
-  \multicolumn{1}{l|}{\textbf{}} & \multicolumn{2}{c|}{\textbf{CE}} & \multicolumn{2}{c|}{\textbf{\begin{tabular}[c]{@{}c@{}}\texttt{fast\_align} \\(100 sentences)\end{tabular}}} & \multicolumn{2}{c}{\textbf{\begin{tabular}[c]{@{}c@{}}\texttt{fast\_align} \\(full dataset)\end{tabular}}} \\ \hline
-  \multicolumn{1}{l|}{}          & \textbf{en-it}  & \textbf{en-sv} & \textbf{en-it}                              & \textbf{en-sv}                              & \textbf{en-it}                                & \textbf{en-sv} \\ \hline
-  \textbf{distinct alignments}        & 536             & 638            & 1242                                        & 1044                                        & 1286                                          & 1065          \\
-  \textbf{correct}                & 392 (73\%)      & 514 (80\%)     & 346 (28\%)                                  & 538 (52\%)                                  & 540 (42\%)                                    & 677 (64\%)    \\
-  \textbf{usable in MT}                      & 363 (68\%)      & 503 (79\%)     & 316 (25\%)                                  & 525 (50\%)                                  & 510 (40\%)                                    & 666 (63\%)    \\
-  \end{tabular}
-  \caption[Word alignment comparison between our grammar-based CE module and \texttt{fast\_align}]{Comparison between our grammar-based CE module and \texttt{fast\_align} on PUD data training the model on 100 and 1000 sentences and discarding the alignments obtained for sentences 101-1000 in the latter case.}
-  \label{pud_fast}
-  \end{table*}
 
@@
 \section{Evaluation} \label{evaluation}
+\begin{table*}[h]
+  \centering
+  \small
+  \begin{tabular}{r|c|c|c|c|c|c}
+  \multicolumn{1}{l|}{\textbf{}} & \multicolumn{2}{c|}{\textbf{CE}} & \multicolumn{2}{c|}{\textbf{\begin{tabular}[c]{@{}c@{}}\texttt{fast\_align} \\(100 sentences)\end{tabular}}} & \multicolumn{2}{c}{\textbf{\begin{tabular}[c]{@{}c@{}}\texttt{fast\_align} \\(full dataset)\end{tabular}}} \\ \hline
+  \multicolumn{1}{l|}{}          & \textbf{en-it}  & \textbf{en-sv} & \textbf{en-it}                              & \textbf{en-sv}                              & \textbf{en-it}                                & \textbf{en-sv} \\ \hline
+  \textbf{distinct alignments}        & 536             & 638            & 1242                                        & 1044                                        & 1286                                          & 1065          \\
+  \textbf{correct}                & 392 (73\%)      & 514 (80\%)     & 346 (28\%)                                  & 538 (52\%)                                  & 540 (42\%)                                    & 677 (64\%)    \\
+  \textbf{usable in MT}                      & 363 (68\%)      & 503 (79\%)     & 316 (25\%)                                  & 525 (50\%)                                  & 510 (40\%)                                    & 666 (63\%)    \\
+  \end{tabular}
+  \caption[Word alignment comparison between our grammar-based CE module and \texttt{fast\_align}]{Comparison between our grammar-based CE module and \texttt{fast\_align} on PUD data, training the statistical model both on 100 and 1000 sentences and discarding the alignments obtained for sentences 101-1000 in the latter case.}
+  \label{pud_fast}
+  \end{table*}
+
 % mini-roadmap
@@
 Because we want part of our evaluation to be independent from the quality of UD parsing, some of the experiments are carried out on treebanks instead of raw text. 
-To this purpose, we use a 100-sentence subset of the Parallel UD (PUD) corpus, a set of 1000 manually annotated or validated sentences in CoNLL-U format.
+To this end, we use a 100-sentence subset of the Parallel UD (PUD) corpus, a set of 1000 manually annotated or validated sentences in CoNLL-U format.
 Of the over 20 languages PUD treebanks are available in, we selected English, Italian and Swedish. 
@@
 While we focus mostly on the MT applications of CA, automatic translation, and much less GF-based domain-specific translation, is not the only context in which CA can be put to use. 
-For instance, it is easy to imagine using it to build translation memories used by human translators.
+For instance, it is easy to imagine using it to build translation memories to be use as an aid for human translation.
 For this reason, a first set of experiments is aimed at evaluating the alignments obtained with our CA module independently from the other stages of our lexicon generation pipeline.
@@
 % metrics
-While precision and recall are two well-known performance metrics, the lack of a gold standard for CE forces us to assess the correctness of alignments manually before calculating the ratio between correctly extracted alignments and the total of extracted alignments.
+While precision and recall are two well-known performance metrics, the lack of a gold standard for CE forces us to, before calculating the ratio between the number of correct alignments and the total number of extracted alignments, manually assess the correctness of each alignment.
 What is more, since some alignments are only correct in the specific context of the sentence pair in which they occur, we make a further distinction between correct alignments that are relevant for a translation lexicon and alignments that are useful for comparing the sentences but should not be used for MT.
@@
 On the other hand, the relatively large size of the PUD treebanks makes it possible also to train the statistical tool on the full dataset instead of just using the chosen 100-sentences subset, allowing for a fairer comparison.
-In both cases, \texttt{fast\_align} is run with the recommended parameters and the CE program is configured only to extract one-to-many and many-to-one word alignments, given that \texttt{fast\_align} does not align larger phrases.
+In both cases, \texttt{fast\_align} is run with the recommended parameters and the CE program is configured to only extract one-to-many and many-to-one word alignments, as \texttt{fast\_align} does not align larger phrases.
 This explains CE's seemingly low recall.
@@
 The course plan corpora are significantly harder to work with, the additional challenges being the inexactness of many translations (which is the direct cause of some of the alignment errors encountered in our evaluation) and the fact that our CE module relies, in this case, on the quality of automatic lemmatization, POS-tagging and parsing.
+
+% table 2
 In Table \ref{raw_fast}, we compare the results obtained on manually annotated data and the course plans corpora parsed with UDPipe.
@@
 Looking at sentence-level scores can, however, be more insightful. 
-For both corpora, scores assigned to individual segments range from the minimum possible value of 0 to the perfect score of 100, which indicates a perfect correspondence between the automatic and the reference translation.
+For both corpora, scores assigned to individual segments range from the minimum possible value of 0 to the perfect score of 100, which indicates a perfect correspondence between the automatic and reference translation.
 Examples of sentences that were assigned a perfect BLEU-1 to 4 score are ``\textit{the library provides useful textbooks}'' (translated to Italian as ``\textit{la biblioteca fornisce libri utili}'') in the DMI corpus and ``\textit{this lab is more difficult than the exam}'' (whose Swedish translation is ``\textit{den här laborationen är svårare än tentamen}'') in the CSE corpus.
@@
 The small size of the corpus, however, allows for some error analysis. 
-From the participants' observations about the kind of errors encountered when manually editing the automatic translation, summarizes in Table \ref{obs}, we can conclude that while most errors are in fact due to wrong alignments, the main difference between two corpora lies in the number of translations that only contain grammatical errors.
+From the participants' observations about the kind of errors encountered when manually editing the automatic translations, summarized in Table \ref{obs}, we can conclude that while most errors are in fact due to wrong alignments, the main difference between two corpora lies in the number of translations that only contain grammatical errors.
 This explains the significant difference observed in the cumulative BLEU scores shown in Table \ref{tableu}.
@@
   \item it can automatically generate compilable, morphology-aware GF translation lexica
-  \item it can be configured to easily handle systematic, possibly language pair or corpus specific translation divergences.
+  \item it can be configured to easily handle systematic, possibly language pair- or corpus-specific translation divergences.
 \end{itemize}
 
-While they do require manual corrections and completions to an extent that varies according to the quality of the data and the strictness of the criteria used, alignments obtained with our method can reduce the time required for bootstrapping the translation lexicon building process for a domain-specific CNL significantly.
+While it requires manual corrections and completions to an extent that varies according to the quality of the data and the strictness of the chosen criteria, using the alignments obtained with our method can reduce the time required for bootstrapping the translation lexicon building process for a domain-specific CNL significantly.
 In fact, especially if a comprehensive morphological dictionary is available, part of the alignments will be ready to use in a GF-based system without any intervention.
@@
 % sw
-The tangible fruits of this work are a Haskell library and a number of executables offering a user friendly interface to perform CE, lexicon generation and various kinds of evaluations. 
+The tangible fruits of this work are a Haskell library and a number of executables offering a user-friendly interface to perform CE, lexicon generation and various kinds of evaluations. 
 The source code, including a preliminary implementation of CP, is available on GitHub\footnote{\url{github.com/harisont/concept-alignment}}. 
@@
 % CP/multilingual CE
-An obvious possible development is optimizing the current implementation of concept propagation (CP) for its two use cases: propagating alignments to a new language looking for correspondences using a translation of the same text they were extracted from or using a different text in the same domain.
+An obvious possible development is optimizing the current, initial implementation of Concept Propagation (CP) for its two use cases: propagating alignments to a new language looking for correspondences using a translation of the same text they were extracted from or using a different text in the same domain.
 An alternative to the former is to make CE, now working on bilingual texts, $n$-lingual.

### 2021-07-15 ConceptAlignment.hs
added correctness to alignment metadata

@@ -78,21 +78,24 @@ meta (_,m) = m
 
 -- | The metadata of an alignment are:
 data Meta = M {
-  reasons :: S.Set Reason,  -- ^ reasons for alignment
-  sentIds :: S.Set String   -- ^ ids of the sentences it was inferred from
+  reasons :: S.Set Reason,        -- ^ reasons for alignment
+  sentIds :: S.Set String,        -- ^ ids of the sents it was inferred from
+  correctness :: Maybe Annotation -- ^ correctness annotation
 } deriving (Show,Read,Eq,Ord)
 
 -- | Initialize metadata with default vals
 initMeta :: Meta
 initMeta = M {
   reasons = S.empty,
-  sentIds = S.empty
+  sentIds = S.empty,
+  correctness = Nothing
 }
 
 -- | Renders metadata as a human-friendly string
 prMeta :: Meta -> String
 prMeta m = "reasons: " ++ showSet (reasons m)
       ++ " sentence IDs: " ++ showSet (sentIds m)
+      ++ " correctness: " ++ maybe "_" show (correctness m) 
   where 
     showSet s = "{" ++ intercalate ", " (S.toList $ S.map show s) ++ "}"
 
@@ -105,7 +108,10 @@ type AlignMap = M.Map AlignedTrees Meta
 combineMeta :: Meta -> Meta -> Meta
 m `combineMeta` n = M {
   reasons = reasons m `S.union` reasons n,
-  sentIds = sentIds m `S.union` sentIds n
+  sentIds = sentIds m `S.union` sentIds n,
+  correctness = if correctness m == correctness n 
+                  then correctness m 
+                  else Nothing
 }
 
 -- | Insert a new alignment combining metadata if an equivalent one is already
@@ -128,7 +134,7 @@ unions' = M.unionsWith combineMeta
 -- is found (=) or correct and potentially reusable, at least in the same 
 -- domain (+)
 data Annotation = Incorrect | Specific | Correct
-  deriving Eq
+  deriving (Eq,Ord)
 
 instance Show Annotation where
   show Incorrect = "-"

### 2021-07-15 ConceptAlignment.hs
implemented ud sent pair to alignment

@@ -529,6 +529,15 @@ alignment2sentencePair a =
       udCommentLines = ["# " ++ prMeta (meta a)]
     } 
 
+-- | Convert a pair of CoNNL-U sentences into an Alignment
+sentencePair2alignment :: (UDSentence,UDSentence) -> Alignment
+sentencePair2alignment (ss,ts) = 
+  (AT (udSentence2tree ss,udSentence2tree ts),meta)
+  where 
+    meta = rdMeta $ drop 2 $ fromJust $ find 
+                                          (\l -> "# reasons" `isPrefixOf` l) 
+                                          (udCommentLines ss)
+
 -- use original label TODO: refactor
 unadjust :: UDTree -> UDTree
 unadjust (RTree n ts) = RTree n { 

### 2021-07-15 ConceptAlignment.hs
minor changes

@@ -516,9 +516,9 @@ abstractUDTree :: UDTree -> RTree Label
 abstractUDTree = mapRTree udSimpleDEPREL
 
 
-{- Alignments to CoNLL-U files -}
+{- Alignments to CoNLL-U files and vice versa -}
 
--- | convert an alignment into a pair of CoNNL-U sentences
+-- | Convert an Alignment into a pair of CoNNL-U sentences
 alignment2sentencePair :: Alignment -> (UDSentence,UDSentence)
 alignment2sentencePair a = 
   (addMetaAsComment $ udTree2adjustedSentence $ sl a, 

### 2021-07-15 Eval.hs
compilable skeleton for new eval script

@@ -1,50 +1,54 @@
 module Eval where
 
-import Data.List
+--import Data.List
+import Data.Maybe
 import System.Exit
 import System.Environment (getArgs)
 import System.Console.GetOpt
-import ArgvParse
 import UDConcepts
+import ArgvParse
+import ConceptAlignment
 
--- | Evaluation script
 main = do
   argv <- getArgs
   (flags,args) <- parseArgv argv help options
   if Help `elem` flags 
     then putStrLn help >> exitSuccess 
     else do
-      alignments <- case args of
-        [p] -> parseUDFile p 
-        (cmd:p1:p2:_) -> do
-          olds <- parseUDFile p1 
-          news <- parseUDFile p2 
-          news' <- if all isAnnotated news
-                    then return news
-                    else annotate (reverse olds) (reverse news) 
-          case cmd of 
-            "extraction" -> putStrLn $ diffStats olds news
-            "propagation" -> putStrLn $ propStats olds news
-          writeFile p2 (unlines $ map prt (reverse news'))
-          return news
+      as <- case args of 
+        [n1,n2] -> getAlignmentsFromUDFiles n1 n2
+        [cmd,o1,o2,n1,n2] -> do
+          olds <- getAlignmentsFromUDFiles o1 o2
+          news <- getAlignmentsFromUDFiles n1 n2
+          news' <- if all isAnnotated news 
+                    then return news 
+                    else annotate olds news
+          case cmd of
+            "extraction" -> putStrLn "TODO:"
+            "propagation" -> putStrLn "TODO:"
+          writeFile n1 (unlines $ map (show . fst) news')
+          writeFile n2 (unlines $ map (show . snd) news')
+          return news'
         _ -> do
           putStrLn "Wrong number of arguments."
           putStrLn help
           exitWith (ExitFailure 1)
-      putStrLn $ basicStats alignments
-      if Reasons `elem` flags 
-        then putStrLn $ reasonStats alignments
-        else putStrLn ""
+      putStrLn "TODO:"
 
-isAnnotated :: UDSentence -> Bool
-isAnnotated s = any (\l -> "# correctness:" `isPrefixOf` l) (udCommentLines s)
+type Path = String
 
-annotate = undefined
+getAlignmentsFromUDFiles :: Path -> Path -> IO [Alignment]
[…]

### 2021-07-15 ConceptAlignment.hs
restored lost rdMeta

@@ -99,6 +99,16 @@ prMeta m = "reasons: " ++ showSet (reasons m)
   where 
     showSet s = "{" ++ intercalate ", " (S.toList $ S.map show s) ++ "}"
 
+-- | Read metadata from human-friendly strings
+rdMeta :: String -> Meta
+rdMeta s = M {
+  reasons = read (drop 9 rs) :: S.Set Reason,
+  sentIds = read is :: S.Set String,
+  correctness = if cs == "_" then Nothing else Just (read cs :: Annotation)
+} where 
+    [rs,xs] = splitOn " sentence IDs: " s
+    [is,cs] = splitOn " correctness: " xs
+
 -- | Map of alignments (used internally to simplify combining metadata)
 type AlignMap = M.Map AlignedTrees Meta
 

### 2021-07-15 Eval.hs
implemented annotation function

@@ -47,8 +47,33 @@ getAlignmentsFromUDFiles p1 p2 = do
 isAnnotated :: Alignment -> Bool
 isAnnotated (_,m) = isJust $ correctness m
 
+-- | Annotate the alignments contained in a file, keeping the information
+-- provided by an older annotated file into account
 annotate :: [Alignment] -> [Alignment] -> IO [Alignment]
-annotate = undefined 
+annotate olds [] = return []
+annotate olds (new:news) = do
+  new' <- case new `elemIndex` olds of
+    Nothing -> annotateManually new
+    Just i -> return (
+      trees new,
+      (meta new) { 
+        correctness = correctness $ meta (olds !! i) 
+      })
+  news' <- annotate olds news
+  return (new':news')
+  where 
+    annotateManually new = do
+      putStrLn $ prLinearizedAlignment new
+      putStrLn "(annotate with +, = or -:)" 
+      a <- getLine
+      -- "à" is an alias for "=" and is just bc of IT keyboard layout ergonomics
+      if a `elem` ["+", "=", "-", "à"]
+        then return (trees new, (meta new) {
+          correctness = if a == "à" 
+                          then Just Specific
+                          else Just (read a :: Annotation) 
+        })
+        else annotateManually new
 
 {- Argument parsing -} 
 

### 2021-07-15 ConceptAlignment.hs
moved getAlignmentsFromCoNNLUFiles to lib

@@ -556,7 +556,14 @@ unadjust (RTree n ts) = RTree n {
     where getOrigLabel n = 
             listToMaybe [head ls | UDData "ORIG_LABEL" ls <- udMISC n]
 
+type Path = String
 
+-- | Get alignments from two CoNNL-U files (shorthand)
+getAlignmentsFromCoNNLUFiles :: Path -> Path -> IO [Alignment]
+getAlignmentsFromCoNNLUFiles p1 p2 = do 
+  p1' <- parseUDFile p1
+  p2' <- parseUDFile p2
+  return $ zipWith (curry sentencePair2alignment) p1' p2'
 
 {- Selection of alignments for MT -}
 

### 2021-07-15 Eval.hs
minor changes & complete skeleton for Eval

@@ -1,6 +1,6 @@
 module Eval where
 
---import Data.List
+import Data.List
 import Data.Maybe
 import System.Exit
 import System.Environment (getArgs)
@@ -16,16 +16,16 @@ main = do
     then putStrLn help >> exitSuccess 
     else do
       as <- case args of 
-        [n1,n2] -> getAlignmentsFromUDFiles n1 n2
+        [n1,n2] -> getAlignmentsFromCoNNLUFiles n1 n2
         [cmd,o1,o2,n1,n2] -> do
-          olds <- getAlignmentsFromUDFiles o1 o2
-          news <- getAlignmentsFromUDFiles n1 n2
+          olds <- getAlignmentsFromCoNNLUFiles o1 o2
+          news <- getAlignmentsFromCoNNLUFiles n1 n2
           news' <- if all isAnnotated news 
                     then return news 
                     else annotate olds news
           case cmd of
-            "extraction" -> putStrLn "TODO:"
-            "propagation" -> putStrLn "TODO:"
+            "extraction" -> putStrLn $ diffStats olds news'
+            "propagation" -> putStrLn $ propStats olds news'
           writeFile n1 (unlines $ map (show . fst) news')
           writeFile n2 (unlines $ map (show . snd) news')
           return news'
@@ -33,17 +33,14 @@ main = do
           putStrLn "Wrong number of arguments."
           putStrLn help
           exitWith (ExitFailure 1)
-      putStrLn "TODO:"
+      let stats = 
+            basicStats as 
+            ++ if Reasons `elem` flags then reasonStats as else "" 
+      putStrLn stats
 
-type Path = String
+{- Annotation -}
 
-getAlignmentsFromUDFiles :: Path -> Path -> IO [Alignment]
-getAlignmentsFromUDFiles p1 p2 = do 
-  p1' <- parseUDFile p1
-  p2' <- parseUDFile p2
-  return $ zipWith (curry sentencePair2alignment) p1' p2'
-
-
+-- | Check if an alignment is already annotated
 isAnnotated :: Alignment -> Bool
 isAnnotated (_,m) = isJust $ correctness m
 
@@ -74,6 +71,12 @@ annotate olds (new:news) = do
                           else Just (read a :: Annotation) 
         })
         else annotateManually new
+
+{- Statistics -}
+diffStats = undefined
+propStats = undefined
+basicStats = undefined
+reasonStats = undefined
 
 {- Argument parsing -} 
 

### 2021-07-15 Eval.hs
quickly adapted the old stats functions

@@ -1,6 +1,7 @@
 module Eval where
 
 import Data.List
+import Data.Set (size, toList)
 import Data.Maybe
 import System.Exit
 import System.Environment (getArgs)
@@ -73,10 +74,107 @@ annotate olds (new:news) = do
         else annotateManually new
 
 {- Statistics -}
-diffStats = undefined
-propStats = undefined
-basicStats = undefined
-reasonStats = undefined
+
+-- | Compare two (annotated) .ca "files" (used to evaluate CE)
+diffStats :: [Alignment] -> [Alignment] -> String
+diffStats olds news = unlines
+  [showLength (filter isCorrect lost) ++ " correct alignments lost",
+   showLength (filter isIncorrect lost) ++ " incorrect alignments lost",
+   showLength (filter isCorrect found) ++ " correct alignments found",
+   showLength (filter isIncorrect found) ++ " incorrect alignments found"]
+  where 
+    lost = olds \\ news
+    found = news \\ olds
+
+-- | Compare two (annotated) .ca "files" (used to evaluate CP)
+propStats :: [Alignment] -> [Alignment] -> String
+propStats origs props = unlines
+  [
+    showLength props ++ "/" ++ showLength origs ++ " (" 
+    ++ show (listPercent props origs) ++ "%) concepts propagated"
+    ++ "\n  of which " ++ showLength correct ++ " (" 
+    ++ show (listPercent correct props) ++ "%) correctly"
+    ++ "\n  of the " ++ showLength incorrect ++ " (" 
+    ++ show (listPercent incorrect props) ++ "%)"
+    ++ " incorrect alignments, " ++ showLength newIncorrect ++ " (" 
+    ++ show (listPercent newIncorrect incorrect) ++ "%) introduce alignment "
+    ++ "errors that are not present in the extracted concepts."
+  ]
+  where
+    correct = filter isCorrect props
+    incorrect = filter isIncorrect props
+    newIncorrect = filter isNewError incorrect
+      where 
+        -- weak check!
+        isNewError e = isCorrect orig
+          where orig = fromJust $ 
+                  find (\a -> tl a == sl e || sl a == tl e) origs
+
+-- | Stats for a single (annotated) .ca file
+basicStats :: [Alignment] -> String
+basicStats alignments = 
+  "\ntotal number of alignments extracted: " ++ show totAligns
[…]

### 2021-07-22 EvAlign.hs
rm old align script

@@ -1,231 +1,0 @@
-module EvAlign where
-import System.Environment (getArgs)
-import System.Console.GetOpt
-import System.Exit
-import Data.Maybe
-import Data.List
-import Data.List.Split
-import qualified Data.Set as S
-import qualified Data.Map as M
-import ConceptAlignment
-import UDConcepts
-import ArgvParse
-
--- | Horrible evaluation script.
---   Usage: runghc EvAlign labelled_old.ca new.ca 
---   to complete annotation and compare the new alignments with the old ones
--- , or runghc EvAlign labelled.ca to get stats on a single file
-main = do
-  argv <- getArgs
-  (flags,args) <- parseArgv argv help options
-  if Help `elem` flags 
-    then putStrLn help >> exitSuccess 
-    else do
-      alignments <- case args of
-        [p] -> readFile p >>= return . map parseLine . nonEmptyLines 
-        (cmd:p1:p2:_) -> do
-          olds <- readFile p1 >>= return . nonEmptyLines 
-          news <- readFile p2 >>= return . nonEmptyLines
-          let olds' = map parseLine olds
-          -- weak annotation check
-          news' <- if head (head news) `elem` ['+','-','=']
-            then return (map parseLine news)
-            else annotate (reverse olds') (reverse $ map read news :: [LinAlignment])
-          case cmd of
-            "extraction" -> putStrLn $ diffStats olds' news'
-            "propagation" -> putStrLn $ propStats olds' news'
-          writeFile p2 (unlines $ map (\(x,y) -> show x ++ show y) (reverse news'))
-          return news'
-        _ -> do 
-          putStrLn "Wrong number of arguments."
-          putStrLn help
-          exitWith (ExitFailure 1)
-      let stats = basicStats alignments ++ if Reasons `elem` flags then reasonStats alignments else "" 
-      putStrLn stats
-    where nonEmptyLines = filter (not . null) . lines
-
--- | Parse a line of a .ca annotated file 
--- (not trivial because files coming from older versions may not have reasons 
--- or occurrence counts)
-parseLine :: String -> AnnotatedAlignment
-parseLine s = (a, la)
-  where 
-    a = read [head s] :: Annotation
-    s' = tail s
-    [t,u] = splitOn "|" $ takeWhile (/='[') s'
[…]

### 2021-07-22 thesis/final_report/include/Introduction.tex
reorganized thesis docs

--- 2021-02-09 final_report/include/Introduction.tex
@@
 In cases like this, a syntactical comparison between the two sentences can help, if not to extract meaningful correspondences only, which would be the case in the example above, where the complement of the copula ''\textit{is}'' (resp.
-``\textit{è}'') is a multiword expression in English and a one-word in Italian, at least to obtain a series of alignments at all levels of abstraction from which to select the most relevant at a later stage\footnote{an example of a more difficult case could be the alternative translation ``Trovare corrispondenze utili non è proprio \textit{un gioco da ragazzi}'', roughly corresponding to ``\textit{a child's play}'' (literally ``\textit{a game (played) by children}'').
+``\textit{è}'') is a multiword expression in English and a one-word in Italian, at least to obtain a series of alignments at all levels of abstraction from which to select the most relevant at a later stage\footnote{An example of a more difficult case could be the alternative translation ``Trovare corrispondenze utili non è proprio \textit{un gioco da ragazzi}'', roughly corresponding to ``\textit{a child's play}'' (literally ``\textit{a game (played) by children}'').
 In this case, trying to align the sentence based on a syntactical comparison would yield both ``\textit{piece of cake}''-``\textit{gioco da ragazzi}'' and the more questionable correspondences ``\textit{piece}''-``\textit{gioco}'' and ``\textit{of cake}-``\textit{da ragazzi}''.}.

### 2021-07-22 thesis/final_report/include/backmatter/Appendix_1.tex
reorganized thesis docs

--- 2021-02-09 final_report/include/backmatter/Appendix_1.tex
@@
 \chapter{Universal POS tags and dependency labels} \label{a}
-While some of them are also discussed briefly in Section \ref{ud}, this appendix provides systematic definitions for all the main UD Universal Part Of Speech tags and dependency relations used in the examples of this work.
+While most of them are also briefly discussed in Section \ref{ud}, this appendix provides systematic definitions for all the main UD Universal Part Of Speech tags and dependency relations used in the examples of this work.
 Complete lists and a good number of examples in several languages can be found in \cite{uddocs} and \cite{compsyn}.
@@
 Unlike nouns in traditional grammars, \texttt{NOUN}s are intended for \textit{common} (as opposed to \textit{proper}) nouns only
+    \item \texttt{ADJ}, the class of adjectives as in their traditional definition
+    \item \texttt{ADV}, the class of adverbs, also following the traditional definition
     \item \texttt{PROPN}, the class of nominals used as names (or part of names) of a specific individual, place, or object, i.e. \textit{proper nouns}
@@
 ``\textit{has}'' in the verb phrase ``\textit{\underline{has} completed}''), but \textit{passive auxiliaries}, like ``\textit{was}'' in ``\textit{\underline{was} completed}, modal auxiliaries like ``\textit{must}'' and ``\textit{should}'' and verbal copulas should also be tagged as \texttt{AUX}
+    \item \texttt{ADP}, designating pre- and postpositions
     \item \texttt{DET}, a wide-coverage class for the different types of determiners, i.e.
 words that modify nouns or noun phrases expressing the reference of the noun phrase in context.
-It includes articles, pronominal numerals and words like \textit{all} and \textit{every}.
+It includes articles, pronominal numerals and words like \textit{all} and \textit{every}
+    \item \texttt{CCONJ} and \texttt{SCONJ}, for coordinating (resp. subordinating) conjunctions.
 \end{itemize} \smallskip \smallskip
-
-Finally, UD groups together categories that apply to tokens that are not even necessarily considered words in traditional grammars, such as symbols.
-The only category of this kind mentioned in this work is \texttt{PUNCT}, which refers of course to punctuation.
 
@@
 
-The only ``exceptional'' label is \texttt{root}, which connects a fake node to the \texttt{root}, i.e.
+The only ``exceptional'' label is \texttt{root}, which connects a dummy node to the \texttt{root}, i.e.
 usually the main \texttt{VERB}, of a sentence.
@@
     \item \texttt{flat}, used to connect the words following (in terms of position in the sentence) the head (i.e.
-the first word) of a flat multiword expression, and \texttt{compound}, playing a similar role in compound written as two or more separate words
+the first word) of a flat multiword expression, and \texttt{compound}, playing a similar role in compounds written as two or more separate words
     \item labels linking non-\texttt{root} verbs to other verbs, specifying what kind of clause they are heads of:
@@
         \item \texttt{advcl}, for adverbial clauses modifying a predicate
-        \item \texttt{acl}, for clauses modifying a nominal
+        \item \texttt{acl}, for clauses modifying a nominal.
     \end{itemize}
@@
     \item \texttt{case}, linking words (such as pre- and postpositions) marking the case of a \texttt{nmod} to the \texttt{nmod} itself in languages that do not express case morphologically
-\end{itemize} \smallskip
-
-Finally, corresponding to the POS tag \texttt{PUNCT} is \texttt{punct}, which links punctuation marks to the head of the clause they belong to. \smallskip
+\end{itemize}

### 2021-07-22 thesis/planning_report/planning_report.tex
reorganized thesis docs

\documentclass{article}

\usepackage[utf8]{inputenc}

\usepackage{natbib}
\usepackage{hyperref}

\usepackage{graphicx}

\usepackage[colorinlistoftodos]{todonotes}

\usepackage{parskip}
\setlength{\parskip}{10pt}

\usepackage{tikz}
\usetikzlibrary{arrows, decorations.markings}

\usepackage{chngcntr}
\counterwithout{figure}{section}

\usepackage{float}

\usepackage{listings}

\lstset{
  basicstyle=\itshape,
  xleftmargin=1em,
  literate={->}{$\rightarrow$}{2}
}

\begin{document}

\begin{titlepage}

\centering

{\scshape\LARGE Master thesis planning report\\}

\vspace{0.5cm}

{\huge\bfseries Automating Concept Alignment for Machine Translation\\}

\vspace{2cm}

{\Large Arianna Masciolini (\texttt{gusmasar@student.gu.se})\\}

\vspace{0.2cm}

{\large Supervisor at CSE: Aarne Ranta\\}

\vspace{1.5cm}

{\large Relevant completed courses:\par}

{\itshape (LT2214, Computational Syntax)\\}
{\itshape (LT2003, Natural Language Processing)\\}
{\itshape (DIT143, Functional Programming)\\}
{\itshape (DIT260, Advanced Functional Programming)\\}
{\itshape (DIT231, Programming Language Technology)\\}
{\itshape (DIT300, Compiler Construction)\\}
{\itshape (DIT411, Introduction to Artificial Intelligence)\\}
{\itshape (DIT866, Applied Machine Learning)\\}

\vfill
\vfill

{\large \today\\}

\end{titlepage}

\section{Introduction}
Concept Alignment (CA) consists in finding semantical correspondences between parts of parallel texts in two or more different languages, of which the Rosetta Stone is a notable example. 
Such task, which is routinely performed by learners of classical languages, who generally work with a translation alongside the original text, is often preliminary to further linguistic analysis.

Another task that involves CA is natural language translation: the human translator, almost unconsciously, first identifies concepts in the source text and then looks for ways to render them in the target language. In this case, alignment can happen - and usually does happen simultaneously - at different levels of abstraction, ranging from word to sentence level.
It is then natural to wonder wether it is possible to make use of CA in Machine Translation (MT) as well. The hypotheses motivating this project, whose objective is to develop and test strategies for automating CA, is that MT can benefit from CA in two ways: % TODO: reduce to one hypothesis? connect the 2nd to the 1st?
\begin{itemize}
    \item on the one hand, if we aim to design a \textit{compositional} MT pipeline, CA at some level is a necessary step, as will become clearer in the following sections
    \item on the other hand, if we address the problem from the perspective of XMT (eXplainable MT), providing the user of a MT application with a way to compare pairs of concepts instead of the entire text in the source language to its automatically translated counterpart can help building a more easily interpretable, and therefore more reliable MT system.
\end{itemize}

\section{Problem} 
To give a more rigorous definition of CA, it is first necessary to specify what we mean by \textit{concept}.

\subsection{Concepts} \label{concepts}
Intuitively, concepts are the components of meaning, and therefore, in a multilingual context, the units of translation. The principle of semantic compositionality states that the meaning of a complex expression is a function solely of the meanings of its components and of the manner in which these components are combined \cite{frege}.
If we assume this principle to be valid and apply it to translation, these meaning components, which we refer to as concepts, are the common denominator between an expression in the source language and its translation, which can be generated using them as the starting point.

We can draw a very close parallel between MT and compiler pipelines: in this sense, concepts are the intermediate representation, or \textit{abstract syntax}, obtained by analyzing, or \textit{parsing}, the source language, while we will call \textit{backend} the portion of the pipeline aimed at \textit{generating} the target language by applying the concepts a set of rules.

In this perspective, then, concepts are abstract syntax functions. In the simplest case, when they correspond to individual words, their arity is zero, but it is often the case that the concrete expressions corresponding to minimal translation units are, in one or both of the source and target language, lemgrams, multiword expressions - sometimes discontinuous - or even more complex constructions, whose intermediate representation is an abstract syntax function taking one or more arguments.

\subsection{Alignment}
Having defined concepts more accurately, we can turn to the more concrete problem of CA, the task of finding, at any abstraction level, correspondences between parallel and, in the case of translation, multilingual texts.

In order to facilitate its automation, CA can be divided into two subproblems subsequent to each other:
\begin{enumerate}
    % TODO: call it extraction?
    \item \textit{concept creation}, the task of identifying concepts via language comparison, which is the main focus of this project
    \item \textit{concept propagation}, i.e. the task of finding, in a language which was not used in the concept creation stage, the concrete expressions corresponding to previously identified concepts.
\end{enumerate}

\section{Related works} 
The idea of taking advantage of semantic compositionality is not new in the field of MT. Introduced by Curry in the early 1960s \cite{curry61}, it was first put in practice two decades later in the form of an experimental interlingual translation system, not coincidentally named Rosetta, which requires the definition of two distinct logically isomorphic \textit{Montague grammars} - one for the source language, one for the target language - and constructs an intermediate representation based on such isomorphism \cite{rosetta}.

% TODO: add sources
Montague grammars are a semantics-oriented development of \textit{categorial grammars}, which in turn belong to \textit{phrase structure} or \textit{constituency} grammars, the class of grammatical formalisms based on the \textit{constituency relation}. 
Without, at least for the time being, giving a formal definition of phrase structure grammars, we can say that they consist in a series of \textit{rewrite rules}, such as 
\begin{lstlisting}
    S -> NP VP,
\end{lstlisting}
which we can read as ``$S$ can be \textit{rewritten} (or \textit{substituted with}) $\mathit{NP}$ and $\mathit{VP}$'', or ``a sentence $S$ has a noun phrase $\mathit{NP}$ and a verb phrase $\mathit{VP}$ as its constituents". Replacing the term ``constituents'' - as in \textit{constituency relation} - with ``components'' makes it easy to understand in which way phrase structure grammars relate to the aforementioned principle of compositionality.
% TODO: rules to trees (make it obvious)

Among constituency grammars, other formalisms that have been employed in more recent MT systems are that of \textit{synchronous CFG}, originally developed for programming language compilation \cite{scfg} and adapted to natural language translation in several settings \cite{nlpscfg1,nlpscfg2,nlpscfg3}, and, later on, that of \textit{synchronous TAG}, a variation of TAG (Tree-Adjoining Grammars) \cite{tag}. %TODO: this source is more recent than what is supposed to happen after!
Both formalisms are meant precisely to characterize correspondences between languages by having as their elements, instead of single rewrite rules, pairs of rules - one for the source and one for the target language. In a synchronous TAG, the constituents of a source language rule may be linked to their counterparts in the corresponding target language rule, and such links, as the authors of \cite{stag} seem to suggest in section 4.1 on idiomatic expressions, may be used to identify concepts in a syntax-based fashion.

\subsection{Grammatical Framework}
Grammatical Framework (GF), a grammar formalism and programming language designed with the parallel between compilers and MT systems discussed in the previous section (\ref{concepts}) in mind, goes a step further in this direction: it introduces a clear distinction between the abstract syntax, a phrase structure tree-like representation capturing the syntactic structures all natural languages taken into account have in common, and the \textit{concrete syntaxes}, specific to each individual language, which consist in rules for the linearization of these Abstract Syntax Trees (ASTs) \cite{gf2004} \cite{gfbook}. 
This makes it possible to deal with multiple languages by writing only one grammar, whose components are an abstract syntax and several concrete syntaxes.

GF grammars can be used in a variety of contexts, one of which is interlingual MT. More in particular, experiments with GF grammars have been conducted in the field of XMT, as the ASTs produced by source language analysis can serve both as to some extent automatically checkable certificates for the correctness of translation - by backlinearization to the source language - and as fully inspectable explanations aimed towards expert users \cite{rantaxmt}.

When it comes to CA itself, which could potentially provide more user-friendly explanations, the idea of performing syntax-based alignment by comparing GF ASTs seems promising, especially if compared with standard statistical approaches: \textit{word alignment}, consisting in finding pairs of individual words that translate to each other, and its generalization, \textit{phrase alignment}\footnote{it must be noted that, in the context of statistical MT, the term ``phrase'' refers to nothing more than contiguous multiword segments} \cite{ibm}.
In practice, however, comparing GF ASTs presents a number of issues related to the fact that GF parsing is not robust enough to spelling and grammatical errors, unusual word orders and rare constructions in general \cite{rantaxmt}, thus making it hard, in many cases, to obtain trees that are actually useful for the task at hand.

It becomes then necessary, to perform syntax-based alignment, to make use of alternative parsers. Preliminary experiments, not yet published, which we intend to take forward in this project, have been conducted exploiting the similarity between GF and UD (Universal Dependencies) \cite{gfud, udgf}.

\subsection{Universal Dependencies} \label{ud}
UD is a framework for cross-linguistically consistent grammatical annotation. 
The UD project aims at developing parallel treebanks for many languages in order to support, among other things, the development of multilingual \textit{dependency parsers}.

Existing dependency parsers, such as UDPipe \cite{udpipe1} and the Standford parser \cite{standford}, are often - but not always \cite{rasp} - neural pipelines trained on dependency treebanks that, given raw text as input, output \textit{dependency trees}.
The difference between such trees and those produced by GF parsing - and in general by natural language parsers based on phrase structure grammars - is that, while the latter capture \textit{constituency} relations, dependency trees are based on another kind of relation, namely that of \textit{dependency}.
As opposed to constituency, dependency is a one-to-one correspondence, meaning that words are simply put in relation with each other via directed links. 
This relation serves as the basis for a comprehensive theory of grammar, first proposed in \cite{dg}, alternative to that of phrase structure.

% TODO: figure 1 shows the PST repr of a sentence and its DT counterpart

If phrase structure trees - and GF trees specifically - seem to be a good starting point for natural (target) language generation, replacing them with dependency trees - and, in particular, given their similarity with GF and the available parsers, UD trees - in the analysis step can prove to be a good basis for syntax-based CA, since the above mentioned UD parsers are more reliable than their GF-based counterparts.
In more concrete terms, if, as mentioned in the introduction, CA is seen as a step of compositional MT, a UD parser can be integrated in a GF-based pipeline, thus constructing a hybrid translation system, with a robust neural frontend and a phrase structure grammar-based backend. 
In this case, dependency trees need to be converted into GF ASTs, albeit with the disadvantage that, unlike its reverse \cite{gfud}, the UD-to-GF conversion is a non-deterministic search problem, addressed in \cite{udgf}, where \texttt{gfud}, a program for converting UD trees into GF ASTs - and vice versa - is presented.

\section{Objectives} \label{objs}
The overall goal of this project is to test and further develop the above mentioned ideas for CA automation.

Nonetheless, a first, necessary step consists in selecting a UD parser well suited to the task. 
Once an adequate parser is identified, the possibility of tuning it for the task at hand will be considered: while our direct intervention on its source code is unlikely, as it goes beyond the scope of this project, it could prove useful to configure it and/or, in the case of a machine learning-based parser, to train it on a suitable corpus.

When it comes to syntax-based alignment itself, which is our proposed way to address concept creation and as such the core part of this project, an unpublished algorithm for aligning pairs of dependency trees exists, but there is, in many respects, room for improvement: preliminary experiments show that its precision is still very low, and there's a number of specific known issues to work on, as the lack of a way for it to recover from parse errors and to deal with irrelevant cross-lingual differences.
A central part of this project will then be improving the existing program in terms of both precision and robustness.

Once significant improvements have been achieved, it is our intention to incorporate the results of the alignment program - the concepts in the form of a GF grammar - into a domain-specific MT pipeline, whose stages are:
\begin{enumerate}
    \item the dependency parser whose selection is the first step of this project
    \item \texttt{gfud} for converting the dependency trees obtained at step 1 into GF ASTs 
    \item a GF-based frontend for target language generation using the domain-specific grammar defined starting from concepts
\end{enumerate}

Such system, also illustrated in figure \ref{fig:mt}, will be, together with the report, the final output of this project and provide a way to evaluate the system, as explained more thoroughly in section \ref{eval}.

\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth]{MT.png}
    \caption{\small The domain-specific MT system we aim to ahve as the final output of this project. SL stands for ``Source Language'', TL for ``Target Language''.}
    \label{fig:mt}
\end{figure}

\subsection{Stretch goals} \label{stretch}
If time allows it, a natural extension to this project is to consider more than just two languages, as results may differ significantly for different source-target language pairs. 
Taking a third language into account can be a way to: 
\begin{itemize}
    \item enhance concept creation, for instance by generalizing the tree alignment program to work on arbitrary-length lists - instead of pairs - of trees 
    \item address concept propagation, by trying to find the equivalent of concepts obtained by aligning trees in the original language pair in the newly introduced language.
\end{itemize}
Our criteria for choosing which languages to work with are described in the following section.

\section{Approach}
\subsection{Data}
Building upon previous experiments, we will conduct a case study on the text of the General Data Protection Regulation (GDPR) \cite{gdpr}.
As all European regulations, it is a very multilingual parallel text, as it has a high quality translation for each of the 24 official languages of the European Union.
Furthermore, for 5 of these languages (English, German, Spanish, Italian and French), a gold standard in terms of concept extraction, even though not publicly available, already exists in the form of a domain-specific GF lexicon.

There is a risk, however, that the complexity of such text, which is characterized by very long sentences, is too high even for a state-of-the-art dependency parser. As a consequence, we reserve ourselves the possibility to work, at least in the initials stages of the project, on alternative corpora.
One that would be easily available thanks to a recent thesis project is, despite some limitations such as being no more than bilingual, that of the course plans of some departments of the University of Gothenburg. % TODO: be more specific once it is possible

Furthermore, another corpus we plan to employ for preliminary testing of our tree alignment program is that of Parallel UD (PUD) treebanks, developed within the UD Project. % TODO: say what it is composed of -> why it's easier data (used to train the chosen parser?)

\subsubsection{Languages}
Our choices in terms of which languages to address in this project are of course also limited by the linguistic knowledge of the candidate.
As for the initial language pair to work with, we intend to use Italian and English, which, apart from being respectively her first and second language, belong to two different branches - resp. Romance and Germanic - of the Indo-European language family, making CA a potentially harder, but at the same time more interesting task.
In the aforementioned case that we decide to also perform experiments with one or more other languages, it would be possible to work with yet another one for which we already have a gold standard available, Spanish, and/or we could consider using Swedish, which would also require the manual creation of a new GDPR lexicon. % TODO: explain what it is.

\subsection{Work plan} % to be transformed in the core thesis
In this section, we present our plan for the project, in the form of a description of what is to be done at each stage followed by our proposed time plan (cf. section \ref{timeline}).

\subsubsection{Dependency parsing}
As mentioned above, the initial phase of the project will be dedicated to identifying a suitable UD parser. The choice will be based on an informal evaluation of the dependency trees obtained for one or more bilingual parallel texts and will be followed by configuration and, if applicable and necessary, retraining of the chosen model.
The expected output of this step is a parallel treebank relative to the corpus (or corpora) in question.

\subsubsection{Dependency tree alignment program} % TODO: find better title
The syntactic alignment component will be based on an initial existing Haskell prototype. 
While we do plan to work on the issues we are already aware of, listed in section \ref{objs}, what other specific problems we address and how exactly we decide to proceed to obtain a substantially improved version of the existing program will mostly depend on the results obtained by testing the prototype on the aforementioned PUD treebank.
Being this the fundamental part of the project, we plan to work on it for at least one third of the time available.

\subsubsection{Preliminary evaluation} \label{prelev}
Together, the dependency parser and the tree alignment program allow concept creation, the fundamental prerequisite for the backend of our desired MT system. Such component, once ready, will be tested independently from the rest of the MT pipeline. This preliminary evaluation will consist for the most part in assessing the precision of the concepts extracted via the tree alignment module against a human-made gold standard, which is, as mentioned, already available for the GDPR corpus as a GF lexicon.

Furthermore, a preliminary MT evaluation, in terms of individual concepts, will already be possible at this stage.
In particular, for single-word concepts, a way to make sure that a lemma in the source language corresponds its target language counterpart is sense comparison using Universal WordNet \cite{uwn} data, which have already been successfully used in GF-based MT systems \cite{wngf}.

\subsubsection{From UD to GF}
Once the preliminary evaluation is concluded, the results of concept creation can be incorporated in a full MT pipeline in the form of a GF abstract lexicon and two GF concrete lexica, one for English and one for Italian.
Automating this step means converting matching UD trees representing concepts to GF ASTs. Work in this sense will involve intervening on or at least configuring another existing program, \texttt{gfud}, already mentioned in section \ref{ud}. Finally, these GF trees need to be converted into an actual lexicon, ideally via a yet-to-be written GF code generation module. Once this step is completed, since the target language concrete syntax of the lexicon specifies the linearization rules necessary for generating the translation of a sentence in the source language, all the stages of the MT pipeline illustrated in figure \ref{fig:mt} and described in section \label{objs} will be ready and easy to put together in a complete domain-specific MT system.

\subsubsection{Introduction of a third language}
Once a full working pipeline for concept creation exists, given the multilingual nature of both UD and GF, it is in principle not too demanding to add other languages. 
As mentioned above, what will be done with respect to this and even the choice of the language itself will depend on how much time we will still have available. However, some ideas are mentioned in section \ref{stretch}.

\subsubsection{Overall evaluation} \label{eval}
While the strategies for the overall evaluation of the system resemble, for the most part, those described in section \ref{prelev}, given that at this point we will have a fully functional MT pipeline, it will be possible to translate entire passages from one language to another and vice versa to compare the original text in the source language to a version of it obtained by linearizing its translation to the target language back in the source language again, or to compare the automatically generated translations with the official ones.

After a first test on one or more of the aforementioned corpora, it will be necessary, in order to assess the generalization capabilities of our system to test it on other parallel texts. 
Assuming we stick to using the text of the GDPR, as previously pointed out, the Official Journal of the European Union is itself a relatively large, highly multilingual parallel corpus of legal documents, but, even if good results within this context would already be meaningful, it would be interesting to put the system to the test - after performing concept alignment again - in other domains, for instance using public domain literature.

\subsection{Time plan} \label{timeline}
The following table describes the timeline of the project. More or less specific objectives both in terms of implementative work and writing have been set on a weekly basis, in accordance with the externally imposed deadlines. The general idea is to write the different sections of the final report as soon as the corresponding experiments are performed, while dedicating the last week to the adjustments and integrations which will prove necessary.

It should be noted that:
\begin{itemize}
    \item the two weeks during which this document has been written are also part of the time plan. At the time of writing, week 40, as the table itself suggests, the dependency parser to be used has already been selected and is being tested on the GDPR Italian-English parallel text in order to find a suitable configuration and to identify issues to be solved either by additional data preprocessing or to be kept in mind when implementing later stages of the pipeline
    \item in this time plan, we assume that we will be able to reach, at least to some extent, the goal of working in a trilingual settings. As a consequence, especially in its final part, the plan might be subject to changes
    \item this time plan was written under the assumption that the only presentation date available in January is the 19th. However, we think it would be preferable, if possible, to postpone the presentation by 7-10 days so to have a one week break in week 52 or 53.
\end{itemize}

\begin{table}[H]
    \centering
    \scriptsize
    \resizebox{\textwidth}{!}{
        \begin{tabular}{|l|l|p{3cm}|p{3cm}|} % TODO: adjust reflow
        \hline
            \textbf{Week} & \textbf{Deadlines} & \textbf{Experimental work} & \textbf{Writing} \\ \hline
            39/`20 &  & dependency parsing (selection and installation) & work and time plan \\ \hline
            40/`20 &  & dependency parsing (tuning and application) & related works, dependency parsing, completion of planning report \\ \hline
            41/`20 & planning report & tree alignment program &  \\ \hline
            42/`20 &  & tree alignment program & \\ \hline
            43/`20 &  & tree alignment program & initial tree alignment program overview \\ \hline
            44/`20 &  & tree alignment program & tree alignment program: proposed improvements \\ \hline
            45/`20 &  & tree alignment program & tree alignment program: proposed improvements \\ \hline
            46/`20 &  & preliminary evaluation & evaluation strategies, preliminary results \\ \hline
            47/`20 &  & gfud & relation between GF and UD, \texttt{gfud} \\ \hline
            48/`20 & halftime report & gfud & work done in relation to \texttt{gfud} \\ \hline
            49/`20 &  & gfud & work done in relation to \texttt{gfud} \\ \hline
            50/`20 &  & introduction of a third language & introduction of a 3rd language \\ \hline
            51/`20 &  & overall evaluation & more on evaluation strategies \\ \hline
            52/`20 &  & overall evaluation & results \\ \hline
            53/`20 &  &  & updated introduction, conclusions, finishing touches \\ \hline
            1/`21 & final report &  &  \\ \hline
            2/`21 & presentation &  &  \\ \hline
        \end{tabular}
    }
\end{table}

% TODO: add ethics/risk analysis section?

\bibliographystyle{plain}
\bibliography{references}
% TODO: re-fix citation for gdpr
%Reference all sources that are cited in your proposal using, e.g. the APA, Harvard2, or IEEE3 style.

\end{document}

### 2021-07-22 thesis/thesis_proposal/proposal.tex
reorganized thesis docs

\documentclass{article}

\usepackage[utf8]{inputenc}

\usepackage{natbib}
\usepackage{hyperref}

\usepackage{graphicx}

\usepackage[colorinlistoftodos]{todonotes}

\usepackage{parskip}
\setlength{\parskip}{10pt}

\usepackage{tikz}
\usetikzlibrary{arrows, decorations.markings}

\usepackage{chngcntr}
\counterwithout{figure}{section}

\begin{document}

\begin{titlepage}

\centering

{\scshape\LARGE Master thesis project proposal\\}

\vspace{0.5cm}

{\huge\bfseries Automating Concept Alignment for Machine Translation\\}

\vspace{2cm}

{\Large Arianna Masciolini (\texttt{gusmasar@student.gu.se})\\}

\vspace{0.2cm}

{\large Suggested Supervisor at CSE: Aarne Ranta\\}

\vspace{1.5cm}

{\large Relevant completed courses:\par}

{\itshape (LT2214, Computational Syntax)\\}
{\itshape (LT2003, Natural Language Processing)\\}
{\itshape (DIT143, Functional Programming)\\}
{\itshape (DIT260, Advanced Functional Programming)\\}
{\itshape (DIT231, Programming Language Technology)\\}
{\itshape (DIT300, Compiler Construction)\\}
{\itshape (DIT411, Introduction to Artificial Intelligence)\\}
{\itshape (DIT866, Applied Machine Learning)\\}

\vfill
\vfill

{\large \today\\}

\end{titlepage}

\section{Introduction}
% Briefly describe and motivate the project, and convince the reader of the importance of the proposed thesis work. A good introduction will answer these questions: Why is addressing these challenges significant for gaining new knowledge in the studied domain? How and where can this new knowledge be applied?
Concept Alignment (CA) consists in finding semantical correspondences between parts of parallel texts in two or more different languages, of which the Rosetta Stone is a notable example. 
Such task, which is routinely performed by learners of classical languages, who generally work with a translation alongside the original text, is often preliminary to further linguistic analysis.

Another task that involves CA is natural language translation: the human translator, almost unconsciously, first identifies concepts in the source text and then looks for ways to render them in the target language. In this case, alignment can happen - and usually does happen simultaneously - at different levels of abstraction, ranging from word to sentence level.
It is then natural to wonder wether it is possible to make use of CA in Machine Translation (MT) as well. The hypotheses motivating this project, whose objective is to develop and test strategies for automating CA, is that MT can benefit from CA in two ways:
\begin{itemize}
    \item on the one hand, if we aim to design a \textit{compositional} MT pipeline, CA is a necessary step, as it will become clearer in the following sections
    \item on the other hand, if we address the problem from the perspective of XMT (eXplainable MT), providing the user of a MT application with a way to compare pairs of concepts instead of the entire text in the source language to its automatically translated counterpart can help building a more easily interpretable, and therefore more reliable MT system.
\end{itemize}

\section{Problem} 
% This section is optional. It may be used if there is a need to describe the problem that you want to solve in more technical detail and if this problem description is too extensive to fit in the introduction.
To give a more rigorous definition of CA, it is first necessary to specify what we mean by \textit{concept}.

\subsection{Concepts}
Intuitively, concepts are the components of meaning, and therefore, in a multilingual context, the units of translation. The principle of semantic compositionality states that the meaning of a complex expression is a function solely of the meanings of its components and of the manner in which these components are combined \cite{semcom}. 
If we assume this principle to be valid and apply it to translation, these meaning components, which we refer to as concepts, are the common denominator between an expression in the source language and its translation, which can be generated using them as the starting point.

We can draw a very close parallel between MT and compiler pipelines: in this sense, concepts are the intermediate representation, or \textit{abstract syntax}, obtained by analyzing, or \textit{parsing}, the source language, while we will call \textit{backend} the portion of the pipeline aimed at \textit{generating} the target language by applying the concepts a set of rules.

In this perspective, then, concepts are abstract syntax functions. In the simplest case, when they correspond to individual words, their arity is zero, but it is often the case that the concrete expressions corresponding to minimal translation units are, in one or both of the source and target language, lemgrams, multiword expressions - sometimes discontinuous - or even more complex constructions, whose intermediate representation is an abstract syntax function taking one or more arguments.

\subsection{Alignment}
Having defined concepts more accurately, we can turn to the more concrete problem of CA, the task of finding, at any abstraction level, correspondences between parallel, and in the case of translation multilingual texts.

In order to facilitate its automation, CA can be divided into two subproblems subsequent to each other:
\begin{enumerate}
    \item \textit{concept creation}, the task of identifying concepts via language comparison, which is the main focus of this project
    \item \textit{concept propagation}, i.e. the task of finding, in a language which was not used in the concept creation stage, the concrete expressions corresponding to previously identified concepts.
\end{enumerate}

\section{Context}
% Use one or two relevant and high quality references for providing evidence from the literature that the proposed study indeed includes scientific and engineering challenges, or is related to existing ones. Convince the reader that the problem addressed in this thesis has not been solved prior to this project.
The idea of taking advantage of semantic compositionality is not new in the field of MT. Introduced by Curry in the early 1960s \cite{curry61}, % TODO: check what the article actually says
it was first put in practice two decades later in the form of an experimental interlingual translation system, not coincidentally named Rosetta, which requires the definition of two distinct logically isomorphic Montague grammars - one for the source language, one for the target language - and constructs an intermediate representation based on such isomorphism \cite{rosetta}.

\subsection{Grammatical Framework}
Grammatical Framework (GF), a grammar formalism and programming language designed with the parallel between compilers and MT systems discussed in the previous section in mind, goes a step further in this direction: it introduces a clear distinction between the abstract syntax, a phrase structure tree-like representation capturing the syntactic structures all natural languages taken into account have in common, and the \textit{concrete syntaxes}, specific to each individual language, which consist in rules for the linearization of these Abstract Syntax Trees (ASTs) \cite{gf2004} \cite{gfbook}. 
This makes it possible to deal with multiple language by writing only one grammar, whose components are an abstract syntax and several concrete syntaxes.

GF grammars can be used in a variety of contexts, one of which is interlingual MT. More in particular, experiments with GF grammars have been conducted in the field of XMT, as the ASTs produced by source language analysis can serve both as to some extent automatically checkable certificates for the correctness of translation - by backlinearization to the source language - and as fully inspectable explanations aimed towards expert users \cite{rantaxmt}.

When it comes to CA itself, which could potentially provide more user-friendly explanations, the idea of performing \textit{syntax-based alignment} by comparing GF ASTs seems promising, especially if compared with the standard approaches used in statistical MT, like word alignment and its generalization, phrase alignment. Both approaches ignore sentence structure and therefore, while requiring a lot of data, their precision is low.

In practice, however, comparing GF ASTs presents a number of issues related to the fact that GF parsing is not robust enough to spelling and grammatical errors, unusual word orders and rare constructions in general \cite{rantaxmt}. It becomes then necessary, if we want to perform syntax-based alignment, to make use of an alternative parser. Preliminary experiments, not yet published, have been conducted exploiting the similarity between GF and UD (Universal Dependencies) \cite{gfud, udgf}.

\subsection{Universal Dependencies}
UD is a framework for cross-linguistically consistent grammatical annotation. 
The UD project aims at developing parallel treebanks for many languages in order to support, among other things, the development of multilingual parsers.

Existing UD parsers, such as \cite{udparsing}, are often neural pipelines trained on UD treebanks that, given raw text as input, output \textit{dependency trees}. The difference between such trees and those produced by GF parsing is that dependency is a one-to-one correspondence, meaning that words in a sentence are connected to each other by directed links, while phrase structure is one-to-many. % TODO: explain better even intuitive consequences

If phrase structure trees seem to be a good starting point for natural (target) language generation, replacing them with dependency trees in the analysis step can prove to be a good basis for CA.
In more concrete terms, if, as mentioned in the introduction, CA is seen as a step of compositional MT, a UD parser can be integrated in a GF-based pipeline, thus constructing a hybrid translation system, with a robust neural frontend and a phrase structure grammar-based backend. 
In this case, dependency trees need to be converted into GF ASTs, even though nlike its reverse \cite{gfud}, the UD-GF conversion is a non-deterministic search problem, addressed in \cite{udgf}.

\section{Goals and Challenges}
% Describe your contribution with respect to concepts, theory and technical goals. Ensure that the scientific and engineering challenges stand out so that the reader can easily recognize that you are planning to solve an advanced problem. 
The overall goal of this project is to test and further develop the above mentioned ideas for CA automation.

A first part of the project consists in selecting an UD parser well suited to the task by manually evaluating the dependency trees obtained for a bilingual parallel text. Once an adequate parser is identified, the possibility of modifying such parser in order to tune it for the task at hand will be considered.

When it comes to alignment itself, which is our proposed way to address concept creation and as such the core part of this project, an unpublished algorithm for UD dependency trees alignment exists, but preliminary experiments show that there is room for improvement. The algorithm, in fact, that orders the trees and pads them to then collect matching subtrees, still has very low precision. Examples of specific problems to work on are making it so that it cab recover from parse errors and dealing with irrelevant cross-lingual differences between parse trees.

\subsection{Stretch goals}
If time allows it, a natural extension to this project is to start using more than just two languages, as result may differ significantly for different language pairs. Our criteria for selecting which languages to compare are described in the following section.

\section{Approach}
%Various scientific approaches are appropriate for different challenges and project goals. Outline and justify the ones that you have selected. For example, when your project considers systematic data collection, you need to explain how you will analyze the data, in order to address your challenges and project goals. One scientific approach is to use formal models and rigorous mathematical argumentation to address aspects like correctness and efficiency. If this is relevant, describe the related algorithmic subjects, and how you plan to address the studied problem. For example, if your plan is to study the problem from a computability aspect, address the relevant issues, such as algorithm and data structure design, complexity analysis, etc.  If you plan to develop and evaluate a prototype, briefly describe your plans to design, implement, and evaluate your prototype by reviewing at most two relevant issues, such as key functionalities and their evaluation criteria. The design and implementation should specify prototype properties, such as functionalities and performance goals, e.g., scalability, memory, energy. Motivate key design selection, with respect to state of the art and existing platforms, libraries, etc.  When discussing evaluation criteria, describe the testing environment, e.g., test-bed experiments, simulation, and user studies, which you plan to use when assessing your prototype. Specify key tools, and preliminary test-case scenarios. Explain how and why you plan to use the evaluation criteria in order to demonstrate the functionalities and design goals. Explain how you plan to compare your prototype to the state of the art using the proposed test-case evaluation scenarios and benchmarks. 
Building upon previous experiments, we will conduct a case study on the text of the General Data Protection Regulation (GDPR) \cite{gdpr}.
As all European regulations, it is a highly multilingual parallel text, as it has a high quality translation for each of the 24 official languages of the European Union.
Furthermore, for 5 of these languages (English, German, Spanish, Italian and French), a gold standard in terms of concept extraction, even if not publicly available, already exists.

Our choices in terms of which languages to use in this project are of course also limited by the linguistic knowledge of the candidate.
As for the initial language pair to start with, we intend to use Italian and English, which, apart from being respectively her first and second language, belong to two different branches - resp. Romance and Germanic - of the Indo-European language family, making CA a potentially harder, but at the same time more interesting task. 
In the aforementioned case that we decide to also perform experiments with one or more other languages, it would be possible to work with yet another one for which we already have a gold standard available, Spanish, and/or we could consider using Swedish, which might involve a relatively demanding annotation process.

For obtaining dependency trees, the starting point will be one of the existing UD parsers yet to be chosen, while the syntactic alignment component will be based on an initial Haskell prototype.

\subsection{Evaluation}
For a first evaluation, the aforementioned manually annotated 5-lingual parallel text of the GDPR will suffice.

To assess the generalization capabilities of our model, however, it will be necessary to test it on other texts. As previously pointed out, the Official Journal of the European Union is itself a relatively large, highly multilingual parallel corpus of legal documents, but, even if good results within this context would already be somewhat meaningful, it would be interesting to put the system to the test in other domains, for instance using public domain literature.

\bibliographystyle{plain}
\bibliography{references}
%Reference all sources that are cited in your proposal using, e.g. the APA, Harvard2, or IEEE3 style.
% TODO: cite GDPR properly

\end{document}
